---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# causatr

<!-- badges: start -->
[![R-CMD-check](https://github.com/etverse/causatr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/etverse/causatr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/etverse/causatr/graph/badge.svg)](https://app.codecov.io/gh/etverse/causatr)
<!-- badges: end -->

**causatr** provides a unified interface for causal effect estimation via three
complementary methods: g-computation (parametric g-formula), inverse probability
weighting (IPW), and propensity score matching. When all three methods agree,
you can be more confident in your findings — this is called
**methodological triangulation**.

The package implements the methods described in Hernán & Robins (2025)
*Causal Inference: What If* with a simple two-step API:

1. **Fit** the causal model with `causat()`
2. **Contrast** interventions with `contrast()`

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("etverse/causatr")
```

## Quick example

Estimate the average causal effect of quitting smoking on weight gain
using the NHEFS dataset from Hernán & Robins (2025):

```{r example}
library(causatr)
data("nhefs")

# Step 1: Fit the outcome model via g-computation
fit <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity,
  censoring = "censored"
)

# Step 2: Contrast interventions
result <- contrast(
  fit,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue"
)
result
```

## Methodological triangulation

Compare g-computation, IPW, and matching on the same data:

```{r triangulation}
# G-computation (outcome model)
fit_gc <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  censoring = "censored")

# IPW (treatment model)
fit_ipw <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  estimator = "ipw")

# Matching (propensity score)
fit_m <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  estimator = "matching", estimand = "ATT")

# All three estimates
rbind(
  data.frame(estimator = "gcomp", contrast(fit_gc,
    list(quit = static(1), cont = static(0)), reference = "cont")$contrasts),
  data.frame(estimator = "ipw", contrast(fit_ipw,
    list(quit = static(1), cont = static(0)), reference = "cont")$contrasts),
  data.frame(estimator = "matching", contrast(fit_m,
    list(quit = static(1), cont = static(0)), reference = "cont")$contrasts)
)
```

## Intervention types

Beyond static interventions, g-computation supports modified treatment policies:

```{r interventions}
fit_cont <- causat(nhefs, outcome = "wt82_71",
  treatment = "smokeintensity",
  confounders = ~ sex + age + race + wt71,
  censoring = "censored")

contrast(fit_cont,
  interventions = list(
    reduce10 = shift(-10),
    halved = scale_by(0.5),
    cap20 = threshold(0, 20),
    observed = NULL
  ),
  reference = "observed"
)
```

## Diagnostics

Check covariate balance and positivity after fitting:

```{r diagnostics, eval = FALSE}
diag <- diagnose(fit_ipw)
diag          # positivity + balance summary
plot(diag)    # Love plot (requires cobalt)
```

## Features

- **Three estimation methods**: g-computation, IPW (via
  [WeightIt](https://ngreifer.github.io/WeightIt/)), matching (via
  [MatchIt](https://kosukeimai.github.io/MatchIt/)). Matching is
  binary-only; continuous / categorical treatments use gcomp or IPW.
- **Longitudinal support**: ICE g-computation (Zivich et al. 2024) for
  time-varying treatments with sandwich variance via stacked estimating
  equations, plus parallel bootstrap via `boot::boot()`.
- **Flexible interventions**: `static()`, `shift()`, `scale_by()`,
  `threshold()`, `dynamic()` for modified treatment policies. IPSI is
  scaffolded for Phase 4.
- **Any outcome family**: gaussian, binomial (logit / probit /
  cloglog), Poisson, quasibinomial (fractional responses), Gamma with
  log link, plus any family you pass through `model_fn`.
- **Pluggable models**: `stats::glm`, `mgcv::gam`, splines via `ns()`
  / `bs()`, or any fit function with signature `(formula, data, family,
  weights, ...)`. A two-tier numeric-variance fallback handles model
  classes without a `sandwich::estfun` method.
- **Robust inference**: analytic sandwich SE (default, via a unified
  influence-function engine) or nonparametric bootstrap with
  percentile CIs. `confint()` respects the `level` argument on both
  paths.
- **External weights**: survey / IPCW weights pass through every
  method and propagate to the sandwich variance.
- **Contrast types**: risk difference, risk ratio, odds ratio — ratio
  and OR use log-scale CIs.
- **Estimands**: ATE, ATT, ATC, or custom subgroups via `subset=` /
  `by=`.
- **Built-in diagnostics**: positivity checks, covariate balance via
  [cobalt](https://ngreifer.github.io/cobalt/), weight summaries, Love
  plots.
- **Tidy integration**: `tidy()` / `glance()` / `confint()` / `coef()`
  / `vcov()` / `plot()` (forest plot via `forrest`) / broom-compatible
  output.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.

## Acknowledgements

This package was built with the contribution of [Claude](https://claude.ai),
Anthropic's AI assistant.
