---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# causatr

<!-- badges: start -->
[![R-CMD-check](https://github.com/etverse/causatr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/etverse/causatr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/etverse/causatr/graph/badge.svg)](https://app.codecov.io/gh/etverse/causatr)
<!-- badges: end -->

**causatr** provides a unified interface for causal effect estimation via three
complementary methods: g-computation (parametric g-formula), inverse probability
weighting (IPW), and propensity score matching. When all three methods agree,
you can be more confident in your findings — this is called
**methodological triangulation**.

The package implements the methods described in Hernán & Robins (2025)
*Causal Inference: What If* with a simple two-step API:

1. **Fit** the causal model with `causat()`
2. **Contrast** interventions with `contrast()`

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("etverse/causatr")
```

## Quick example

Estimate the average causal effect of quitting smoking on weight gain
using the NHEFS dataset from Hernán & Robins (2025):

```{r example}
library(causatr)
data("nhefs")

# Step 1: Fit the outcome model via g-computation
fit <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity,
  censoring = "censored"
)

# Step 2: Contrast interventions
result <- contrast(
  fit,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue"
)
result
```

## Methodological triangulation

Compare g-computation, IPW, and matching on the same data:

```{r triangulation}
# G-computation (outcome model)
fit_gc <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  censoring = "censored")

# IPW (treatment model)
fit_ipw <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  method = "ipw")

# Matching (propensity score)
fit_m <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  method = "matching", estimand = "ATT")

# All three estimates
rbind(
  data.frame(method = "gcomp", contrast(fit_gc,
    list(quit = static(1), cont = static(0)), reference = "cont")$contrasts),
  data.frame(method = "ipw", contrast(fit_ipw,
    list(quit = static(1), cont = static(0)), reference = "cont")$contrasts),
  data.frame(method = "matching", contrast(fit_m,
    list(quit = static(1), cont = static(0)), reference = "cont")$contrasts)
)
```

## Intervention types

Beyond static interventions, g-computation supports modified treatment policies:

```{r interventions}
fit_cont <- causat(nhefs, outcome = "wt82_71",
  treatment = "smokeintensity",
  confounders = ~ sex + age + race + wt71,
  censoring = "censored")

contrast(fit_cont,
  interventions = list(
    reduce10 = shift(-10),
    halved = scale(0.5),
    cap20 = threshold(0, 20),
    observed = NULL
  ),
  reference = "observed"
)
```

## Diagnostics

Check covariate balance and positivity after fitting:

```{r diagnostics, eval = FALSE}
diag <- diagnose(fit_ipw)
diag          # positivity + balance summary
plot(diag)    # Love plot (requires cobalt)
```

## Features

- **Three estimation methods**: g-computation, IPW (via
  [WeightIt](https://ngreifer.github.io/WeightIt/)), matching (via
  [MatchIt](https://kosukeimai.github.io/MatchIt/))
- **Built-in diagnostics**: positivity checks, covariate balance via
  [cobalt](https://ngreifer.github.io/cobalt/), weight summaries, Love plots
- **Flexible interventions**: `static()`, `shift()`, `scale()`, `threshold()`,
  `dynamic()` for modified treatment policies
- **Robust inference**: sandwich SE (default) or nonparametric bootstrap
- **Any outcome type**: continuous, binary, count (via `family` parameter)
- **Pluggable models**: pass `stats::glm`, `mgcv::gam`, or any compatible
  fitting function via `model_fn`
- **Contrast types**: risk difference, risk ratio, odds ratio
- **Estimands**: ATE, ATT, ATC, or custom subgroups via `subset`

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
