---
title: "Augmented IPW (doubly-robust estimation) with causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Augmented IPW (doubly-robust estimation) with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| message: false
library(causatr)
library(tinytable)
```

Augmented inverse probability weighting (AIPW) combines the two single-robust
estimators — g-computation (an **outcome** model) and IPW (a **treatment**
model) — into one **doubly-robust** estimator. For a binary treatment and
intervention value $a$,

$$
\hat\psi_{\mathrm{AIPW}}(a) = \underbrace{\frac{1}{n}\sum_i \hat m(a, L_i)}_{\text{g-computation}}
 + \underbrace{\frac{1}{n}\sum_i \frac{\mathbb 1\{A_i = a\}}{\hat g(a \mid L_i)}\,\bigl(Y_i - \hat m(A_i, L_i)\bigr)}_{\text{IPW augmentation}},
$$

where $\hat m$ is the fitted outcome regression and $\hat g$ the fitted
propensity. The augmentation term has mean zero when *either* model is correct,
so $\hat\psi_{\mathrm{AIPW}}$ is consistent if **either** the outcome model
**or** the propensity model is correctly specified — not necessarily both. That
is the **double-robustness** property.

causatr fits both nuisances with the user's `model_fn` (outcome) and
`propensity_model_fn` (treatment), assembles the doubly-robust functional per
intervention, and computes the variance with the stacked influence-function
sandwich (outcome block + propensity block + plug-in). This is the *classical
analytical* AIPW — distinct from `lmtp`'s TMLE/SDR with cross-fitting and machine
learning.

## Double robustness, demonstrated

The cleanest way to see double robustness is a simulated data generating process
with a **known** average treatment effect, where a confounder $L$ enters both the
propensity and the outcome **nonlinearly**. A model that omits the $L^2$ term is
misspecified.

```{r}
set.seed(2)
n <- 4000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(-0.5 + 0.7 * L + 0.6 * L^2))
Y <- 1 + 2 * A + 1.5 * L + 1.6 * L^2 + rnorm(n)
d <- data.frame(Y, A, L)

correct <- ~ L + I(L^2) # captures the L^2 confounding
wrong <- ~ L # misspecified: omits L^2
```

The true ATE is **2**. causatr lets the outcome and treatment models carry
*separate* confounder formulas (`confounders_outcome`, `confounders_treatment`),
so we can misspecify one while keeping the other correct:

```{r}
ate <- function(estimator, co, ct) {
  fit <- causat(
    d, outcome = "Y", treatment = "A",
    confounders_outcome = co, confounders_treatment = ct,
    estimator = estimator, model_fn = stats::glm,
    propensity_model_fn = stats::glm
  )
  contrast(fit, list(a1 = static(1), a0 = static(0)), reference = "a0")$contrasts$estimate[1]
}

results <- data.frame(
  estimator = c(
    "gcomp", "ipw",
    "aipw", "aipw", "aipw", "aipw"
  ),
  outcome_model = c(
    "wrong", "correct",
    "wrong", "correct", "correct", "wrong"
  ),
  propensity_model = c(
    "wrong", "wrong",
    "correct", "wrong", "correct", "wrong"
  ),
  ATE_hat = c(
    ate("gcomp", wrong, wrong),
    ate("ipw", correct, wrong),
    ate("aipw", wrong, correct),
    ate("aipw", correct, wrong),
    ate("aipw", correct, correct),
    ate("aipw", wrong, wrong)
  )
)
tt(results, digits = 3)
```

Reading the table: g-computation with a misspecified **outcome** model is biased,
and IPW with a misspecified **propensity** is biased. But AIPW recovers the true
ATE of 2 whenever **at least one** nuisance is correct — even though the other is
wrong. Only when **both** are misspecified does AIPW lose consistency (last row):
double robustness buys one free misspecification, not two.

## Real-data example: NHEFS

On observational data we never know the truth, but AIPW is the natural default
when you are unsure which nuisance you trust. Using the NHEFS quit-smoking
question (effect of `qsmk` on weight change `wt82_71`):

```{r}
data("nhefs")
nhefs_complete <- nhefs[!is.na(nhefs$wt82_71) & !is.na(nhefs$education), ]

conf <- ~ sex + age + I(age^2) + race + factor(education) +
  smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
  factor(exercise) + factor(active) + wt71 + I(wt71^2)

fit_aipw <- causat(
  nhefs_complete,
  outcome = "wt82_71", treatment = "qsmk",
  confounders = conf,
  estimator = "aipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_aipw <- contrast(
  fit_aipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
tt(tidy(res_aipw), digits = 3)
```

The point estimate sits between the pure g-computation and pure IPW estimates and
shares their interpretation: the average weight change had everyone quit smoking
versus had no one quit.

## Variance

The `"sandwich"` CI above is the stacked influence-function variance: the AIPW
influence function is the sum of the outcome-model correction, the propensity
correction, and the doubly-robust plug-in residual, aggregated by
`variance_if()`. It is asymptotically exact under correct specification of the
models being used. `ci_method = "bootstrap"` refits both nuisances on each
resample and is available as a non-parametric alternative.

```{r}
res_boot <- contrast(
  fit_aipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  ci_method = "bootstrap", n_boot = 200L
)
tt(res_boot$contrasts[, c("comparison", "estimate", "se", "ci_lower", "ci_upper")], digits = 3)
```

## Where AIPW fits

AIPW is a member of the methodological **triangle** alongside g-computation and
IPW: same estimand, different reliance on the nuisance models. Reach for it when

- you want a single estimate that is robust to misspecifying *one* of the two
  nuisance models, or
- you are triangulating: agreement between gcomp, IPW, and AIPW is reassuring,
  and AIPW disagreeing with both flags a likely misspecification in one of them.

causatr's AIPW supports the same surface as the other estimators — binary,
continuous, categorical, and multivariate treatments; `static` / `shift` /
`scale_by` / `dynamic` / `stochastic` interventions; difference / ratio / OR
contrasts; `by`-stratified estimands; stabilized weights; and transportability.
For the **longitudinal** doubly-robust estimator (ICE-AIPW, Bang & Robins 2005),
see `vignette("longitudinal")`. For the triangulation workflow across all three
estimators, see `vignette("triangulation")`.

## References

Robins JM, Rotnitzky A, Zhao LP (1994). Estimation of regression coefficients
when some regressors are not always observed. *Journal of the American
Statistical Association* 89:846–866.

Bang H, Robins JM (2005). Doubly robust estimation in missing data and causal
inference models. *Biometrics* 61:962–973.

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
