---
title: "G-computation with causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{G-computation with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

G-computation (the parametric g-formula) estimates causal effects by fitting a
model for the outcome conditional on treatment and confounders, then
standardising predictions over the target population under each hypothetical
intervention. This vignette demonstrates g-computation in causatr using the
NHEFS dataset from Hernán & Robins (2025), covering every supported combination
of treatment type, outcome type, contrast scale, and inference method for
time-fixed treatments.

## Data: NHEFS

The examples in this vignette use the **NHEFS** (National Health and
Nutrition Examination Survey — Epidemiologic Follow-up Study) dataset,
the running example in Hernán & Robins (2025). The assumed causal
structure is:

- **Treatment ($A$):** `qsmk` — quit smoking between 1971 and 1982 (binary: 0/1).
- **Outcome ($Y$):** `wt82_71` — weight change in kg over the same period
  (continuous). A derived binary version, `gained_weight` ($\mathbf{1}\{Y > 0\}$),
  is used in the binary outcome sections.
- **Confounders ($L$):** sex, age, race, education, smoking intensity
  (cigarettes/day), years smoked, exercise, physical activity, and baseline
  weight. These jointly affect both the probability of quitting and the
  subsequent weight change.
- **Censoring:** the `censored` column flags individuals lost to follow-up.
  Censored rows are excluded from model fitting; predictions are
  standardised over the full target population.

Structural causal model (assumed):

$$
\begin{aligned}
  L &\sim F_L \\
  A &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(\alpha_0 + \alpha_L^\top L)\bigr) \\
  Y &= \beta_0 + \beta_A A + \beta_L^\top L + \beta_{AL}^\top (A \cdot L) + \varepsilon,
      \quad \varepsilon \sim N(0, \sigma^2)
\end{aligned}
$$

The confounders $L$ are common causes of $A$ and $Y$; adjusting for $L$
blocks all backdoor paths, identifying $\mathbb{E}[Y^a]$ via the g-formula.

```{r}
#| message: false
library(causatr)
library(tinytable)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71), ]

nhefs_complete$gained_weight <- as.integer(nhefs_complete$wt82_71 > 0)

nhefs$sex <- factor(nhefs$sex, levels = 0:1, labels = c("Male", "Female"))
nhefs_complete$sex <- factor(nhefs_complete$sex, levels = 0:1, labels = c("Male", "Female"))
```

We create a binary outcome `gained_weight` (1 if weight increased, 0 otherwise)
for the binary outcome examples below.

## Binary treatment, continuous outcome

This is the core example from Chapter 13 of Hernán & Robins: the average
causal effect of quitting smoking (`qsmk`) on weight change (`wt82_71`).

### ATE with sandwich SE

```{r}
fit_gc <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity,
  censoring = "censored"
)

res_ate_sw <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_ate_sw
```

The book reports ATE $\approx$ 3.5 kg (95% CI: 2.6, 4.5).

### ATE with bootstrap SE

```{r}
res_ate_bs <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_ate_bs
```

Sandwich and bootstrap SEs should be in close agreement for correctly specified
GLMs.

### ATT estimand

The average treatment effect on the treated (ATT) averages only over individuals
who actually quit smoking. With g-computation, the estimand can be changed in
`contrast()` without refitting.

```{r}
res_att <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  estimand = "ATT",
  ci_method = "sandwich"
)
res_att
```

### ATC estimand

The average treatment effect on the controls (ATC) averages over those who
continued smoking.

```{r}
res_atc <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  estimand = "ATC",
  ci_method = "sandwich"
)
res_atc
```

### Subset estimand

A custom subgroup: effect among individuals aged 50 or older.

```{r}
res_sub <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  subset = quote(age >= 50),
  ci_method = "sandwich"
)
res_sub
```

### Extracting results programmatically

```{r}
coef(res_ate_sw)
confint(res_ate_sw)
```

## Comparing estimands

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for ATE, ATT, ATC, and subset estimands from g-computation."
est_df <- data.frame(
  estimand = c("ATE", "ATT", "ATC", "Subset (age >= 50)"),
  estimate = c(
    res_ate_sw$contrasts$estimate[1],
    res_att$contrasts$estimate[1],
    res_atc$contrasts$estimate[1],
    res_sub$contrasts$estimate[1]
  ),
  ci_lower = c(
    res_ate_sw$contrasts$ci_lower[1],
    res_att$contrasts$ci_lower[1],
    res_atc$contrasts$ci_lower[1],
    res_sub$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    res_ate_sw$contrasts$ci_upper[1],
    res_att$contrasts$ci_upper[1],
    res_atc$contrasts$ci_upper[1],
    res_sub$contrasts$ci_upper[1]
  )
)

tinyplot(
  estimate ~ estimand,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  xlab = "Estimand",
  ylab = "Effect on weight change (kg)",
  main = "G-computation estimates by estimand"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Binary treatment, binary outcome

Using the `gained_weight` indicator as a binary outcome, we estimate the risk
difference, risk ratio, and odds ratio of quitting smoking on gaining weight.

### Risk difference (sandwich)

```{r}
fit_bin <- causat(
  nhefs_complete,
  outcome = "gained_weight",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity,
  family = "binomial"
)

res_rd <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_rd
```

### Risk difference (bootstrap)

```{r}
res_rd_bs <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rd_bs
```

### Risk ratio

```{r}
res_rr <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

### Odds ratio

```{r}
res_or <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "or",
  ci_method = "sandwich"
)
res_or
```

## Continuous treatment

G-computation supports continuous treatments with all intervention types:
`static()` (set to a fixed value), `shift()`, `scale_by()`, `threshold()`,
`dynamic()`, and `NULL` (natural course). Here we use `smokeintensity`
(cigarettes per day) as the treatment and `wt82_71` as the outcome.

```{r}
fit_cont <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "smokeintensity",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeyrs + I(smokeyrs^2) + factor(exercise) + factor(active) +
    wt71 + I(wt71^2),
  censoring = "censored"
)
```

### Static intervention

Set smoking intensity to fixed values and compare. This answers the question:
what would happen if everyone smoked 5 vs 40 cigarettes per day?

```{r}
res_static <- contrast(
  fit_cont,
  interventions = list(light = static(5), heavy = static(40)),
  reference = "light",
  type = "difference",
  ci_method = "sandwich"
)
res_static
```

### Shift intervention

Reduce smoking intensity by 10 cigarettes per day for everyone.

```{r}
res_shift <- contrast(
  fit_cont,
  interventions = list(reduce10 = shift(-10), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_shift
```

### Scale intervention

Halve each individual's smoking intensity.

```{r}
res_scale <- contrast(
  fit_cont,
  interventions = list(halved = scale_by(0.5), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_scale
```

### Threshold intervention

Cap smoking intensity at 20 cigarettes per day for everyone.

```{r}
res_thresh <- contrast(
  fit_cont,
  interventions = list(cap20 = threshold(0, 20), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_thresh
```

### Comparing multiple interventions

```{r}
res_multi <- contrast(
  fit_cont,
  interventions = list(
    light    = static(5),
    reduce10 = shift(-10),
    halved   = scale_by(0.5),
    cap20    = threshold(0, 20),
    observed = NULL
  ),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_multi
```

### Visualising intervention effects

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for mean weight change under different smoking intensity interventions."
est <- res_multi$contrasts
tinyplot(
  estimate ~ comparison,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  xlab = "Comparison",
  ylab = "Difference in weight change (kg)",
  main = "Continuous treatment interventions"
)
abline(h = 0, lty = 2, col = "grey40")
```

### Dynamic intervention on continuous treatment

A dynamic rule that depends on individual characteristics. Here, heavy smokers
(> 20 cigarettes/day) are reduced to 20, while others keep their observed level.

```{r}
res_dyn_cont <- contrast(
  fit_cont,
  interventions = list(
    capped = dynamic(\(data, trt) pmin(trt, 20)),
    observed = NULL
  ),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_dyn_cont
```

### Continuous treatment with bootstrap

```{r}
res_shift_bs <- contrast(
  fit_cont,
  interventions = list(reduce10 = shift(-10), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_shift_bs
```

## Dynamic intervention

A dynamic intervention assigns treatment based on individual characteristics.
Here, we assign quitting smoking only to individuals who smoked more than 20
cigarettes per day at baseline.

```{r}
res_dyn <- contrast(
  fit_gc,
  interventions = list(
    rule = dynamic(\(data, trt) ifelse(data$smokeintensity > 20, 1, 0)),
    all_quit = static(1)
  ),
  reference = "all_quit",
  type = "difference",
  ci_method = "sandwich"
)
res_dyn
```

## Effect modification with `by`

The `by` argument stratifies causal effect estimates by levels of a
variable. It averages the fitted model's counterfactual predictions over
each subset — so it only produces *genuine* effect-modification estimates
when the outcome model itself contains an interaction between treatment
and the modifier. Here we refit the model with a `qsmk:sex` term and then
examine how the effect of quitting smoking differs by sex.

```{r}
fit_gc_hte <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity + qsmk:sex,
  censoring = "censored"
)

res_by_sex <- contrast(
  fit_gc_hte,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich",
  by = "sex"
)
res_by_sex
```

::: {.callout-note}
Without the `qsmk:sex` term in the outcome model, `by = "sex"` would still
run but would return near-identical estimates for both strata (any
remaining variation coming only from indirect interactions like
`qsmk:smokeintensity` combined with differing covariate distributions).
All four estimation methods support `A:modifier` terms: g-comp feeds them
to the outcome model directly, IPW expands the MSM to `Y ~ 1 + modifier`,
matching expands to `Y ~ A + modifier + A:modifier`, and ICE auto-expands
the interaction across treatment lags (`lag1_A:modifier`, etc.).
:::

## Tidy and glance

causatr results work with the broom ecosystem via `tidy()` and `glance()`:

```{r}
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Forest plot

The `plot()` method produces a forest plot using the `forrest` package. Forest
plots are most useful when displaying multiple estimates, such as effect
modification results:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot of the effect of quitting smoking on weight change, stratified by sex."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_by_sex)
```

## Categorical treatment (k > 2 levels)

G-computation handles categorical treatments natively by fitting a single
outcome model with the treatment as a factor covariate. `contrast()` then
standardises predictions under each named `static()` level and pairwise
differences are read off against the reference category.

**DGM.** A three-arm trial with one confounder. The treatment is assigned
with fixed probabilities (no confounding of treatment assignment in this
simple example; $L$ is a pure precision variable).

$$
\begin{aligned}
  L &\sim N(0,1) \\
  A &\sim \text{Categorical}(0.40,\; 0.35,\; 0.25) \quad\text{levels } \{A, B, C\} \\
  Y &= 2 + 1.5 \cdot \mathbf{1}(A{=}B) - 0.8 \cdot \mathbf{1}(A{=}C) + 0.5\,L + \varepsilon,
      \quad \varepsilon \sim N(0,1)
\end{aligned}
$$

True contrasts: $\mathbb{E}[Y^B] - \mathbb{E}[Y^A] = 1.5$;
$\mathbb{E}[Y^C] - \mathbb{E}[Y^A] = -0.8$.

```{r}
set.seed(42)
n <- 2000
L <- rnorm(n)
A <- sample(c("A", "B", "C"), n, replace = TRUE, prob = c(0.4, 0.35, 0.25))
Y <- 2 + (A == "B") * 1.5 + (A == "C") * -0.8 + 0.5 * L + rnorm(n)
cat_df <- data.frame(Y = Y, A = factor(A, levels = c("A", "B", "C")), L = L)

fit_cat <- causat(cat_df, outcome = "Y", treatment = "A", confounders = ~ L)

res_cat <- contrast(
  fit_cat,
  interventions = list(
    arm_a = static("A"),
    arm_b = static("B"),
    arm_c = static("C")
  ),
  reference = "arm_a",
  type = "difference",
  ci_method = "sandwich"
)
res_cat
```

Structural truth: `B − A = 1.5`, `C − A = −0.8`. The sandwich CIs should cover both.

## Multivariate (joint) treatment

Pass `treatment = c("A1", "A2")` for a joint intervention on two or more
treatment variables. Interventions are supplied as **named sub-lists**, one
entry per treatment column, and each sub-intervention is independent (the
rules can be different types — e.g. `static()` for one, `shift()` for
another).

**DGM.** Two treatments with separate confounders. $A_1$ is binary
(confounded by $L_1$); $A_2$ is continuous (confounded by $L_2$). The
outcome depends additively on both treatments.

$$
\begin{aligned}
  L_1,\; L_2 &\sim N(0,1) \\
  A_1 &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(0.3\,L_1)\bigr) \\
  A_2 &= 1 + 0.4\,L_2 + \eta, \quad \eta \sim N(0,1) \\
  Y &= 1 + 2\,A_1 + 0.5\,A_2 + L_1 + L_2 + \varepsilon,
      \quad \varepsilon \sim N(0,1)
\end{aligned}
$$

True contrast: $\mathbb{E}[Y^{A_1=1,\,A_2=2}] - \mathbb{E}[Y^{A_1=0,\,A_2=0}] = 2 \times 1 + 0.5 \times 2 = 3.0$.

```{r}
set.seed(1)
n <- 2000
L1 <- rnorm(n)
L2 <- rnorm(n)
A1 <- rbinom(n, 1, plogis(0.3 * L1))
A2 <- 1 + 0.4 * L2 + rnorm(n)
Y <- 1 + 2 * A1 + 0.5 * A2 + L1 + L2 + rnorm(n)
mv_df <- data.frame(Y = Y, A1 = A1, A2 = A2, L1 = L1, L2 = L2)

fit_mv <- causat(
  mv_df,
  outcome = "Y",
  treatment = c("A1", "A2"),
  confounders = ~ L1 + L2
)

res_mv <- contrast(
  fit_mv,
  interventions = list(
    both_on  = list(A1 = static(1), A2 = static(2)),
    both_off = list(A1 = static(0), A2 = static(0))
  ),
  reference = "both_off",
  type = "difference"
)
res_mv
```

Structural truth: `both_on − both_off = 2*1 + 0.5*2 = 3.0`.

## External (survey) weights

Pre-computed weights (survey weights, calibration weights, externally-fit
IPCW) are passed via `weights =` and validated upfront by `check_weights()`
(NA / Inf / negative / mis-sized vectors are rejected with a specific
error). The sandwich engine propagates the weights through both Channel 1
(target-population weighting) and Channel 2 (IWLS weighted score).

**DGM.** A simple confounded binary treatment with $\text{ATE} = 3$. The
survey weights $w_i \sim \text{Uniform}(0.5, 2)$ are deliberately
uninformative — they carry no information about the outcome — to verify
that the IF correctly propagates weights without introducing bias.

$$
\begin{aligned}
  L &\sim N(0,1) \\
  A &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(0.5\,L)\bigr) \\
  Y &= 2 + 3\,A + 1.5\,L + \varepsilon, \quad \varepsilon \sim N(0,1) \\
  w &\sim \text{Uniform}(0.5,\; 2.0)
\end{aligned}
$$

```{r}
set.seed(2)
n <- 2000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- 2 + 3 * A + 1.5 * L + rnorm(n)
# A random survey weight unrelated to outcome: verifies the IF correctly
# propagates weights without introducing bias when the weights carry no
# real information.
w <- runif(n, 0.5, 2.0)
sv_df <- data.frame(Y = Y, A = A, L = L)

fit_sv <- causat(
  sv_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  weights = w
)

res_sv <- contrast(
  fit_sv,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference"
)
res_sv
```

The weighted ATE estimate is still $\approx$ 3 (the structural truth), and the
sandwich SE reflects both the outcome-model and weighted-target sampling
uncertainty.

## Poisson count outcome

For rate / count outcomes, pass `family = "poisson"`. The outcome model
becomes a log-link GLM and `contrast(..., type = "ratio")` returns the
marginal rate ratio.

**DGM.** A binary treatment with a Poisson count outcome. The outcome's
conditional mean is log-linear in $A$ and $L$, so the conditional rate
ratio is $\exp(0.7) \approx 2.01$.

$$
\begin{aligned}
  L &\sim N(0,1) \\
  A &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(0.5\,L)\bigr) \\
  Y &\sim \text{Poisson}\!\bigl(\exp(0.5 + 0.7\,A + 0.3\,L)\bigr)
\end{aligned}
$$

True marginal rate ratio: $\approx \exp(0.7) \approx 2.01$.

```{r}
set.seed(3)
n <- 2000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- rpois(n, lambda = exp(0.5 + 0.7 * A + 0.3 * L))
pois_df <- data.frame(Y = Y, A = A, L = L)

fit_pois <- causat(
  pois_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  family = "poisson"
)

res_rr_pois <- contrast(
  fit_pois,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr_pois
```

Structural truth: marginal RR $\approx$ `exp(0.7)` $\approx$ 2.01.

## Gamma outcome

For strictly positive continuous outcomes (e.g. costs, durations), the Gamma
family with log link avoids predicting negative values. The marginal rate
ratio is a natural summary.

**DGM.** A binary treatment with a Gamma-distributed positive outcome.
The conditional mean is log-linear: $E[Y \mid A, L] = \exp(1 + 0.5\,A + 0.3\,L)$,
so the conditional rate ratio is $\exp(0.5) \approx 1.65$.

$$
\begin{aligned}
  L &\sim N(0,1) \\
  A &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(0.5\,L)\bigr) \\
  Y &\sim \text{Gamma}\!\bigl(\text{shape}=4,\; \text{rate}=4\,/\,\exp(1 + 0.5\,A + 0.3\,L)\bigr)
\end{aligned}
$$

True marginal rate ratio: $\approx \exp(0.5) \approx 1.65$.

```{r}
set.seed(4)
n <- 2000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- rgamma(n, shape = 4, rate = 4 / exp(1 + 0.5 * A + 0.3 * L))
gamma_df <- data.frame(Y = Y, A = A, L = L)

fit_gamma <- causat(
  gamma_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  family = "Gamma"
)

res_gamma <- contrast(
  fit_gamma,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "ratio",
  ci_method = "sandwich"
)
res_gamma
```

Structural truth: marginal RR $\approx$ `exp(0.5)` $\approx$ 1.65.

```{r}
tt(tidy(res_gamma), digits = 3)
```

## Quasibinomial outcome

When the outcome is a proportion in (0, 1) --- for example a bounded
continuous score --- or an over-dispersed binary response,
`family = "quasibinomial"` uses a logit link without assuming binomial
variance. Sandwich SEs handle the quasi-likelihood correctly.

**DGM.** A confounded binary treatment with a binary outcome generated
from a logistic model. Using `quasibinomial` instead of `binomial` relaxes
the strict variance assumption.

$$
\begin{aligned}
  L &\sim N(0,1) \\
  A &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(0.3\,L)\bigr) \\
  Y &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(-0.5 + 0.6\,A + 0.4\,L)\bigr)
\end{aligned}
$$

```{r}
set.seed(5)
n <- 2000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.3 * L))
Y <- rbinom(n, 1, plogis(-0.5 + 0.6 * A + 0.4 * L))
quasi_df <- data.frame(Y = Y, A = A, L = L)

fit_quasi <- causat(
  quasi_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  family = "quasibinomial"
)

res_quasi <- contrast(
  fit_quasi,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_quasi
```

```{r}
tt(tidy(res_quasi), digits = 3)
```

## GLM with splines

For flexible confounder adjustment without the overhead of a full GAM, use
`splines::ns()` (natural cubic splines) directly in the formula. This keeps
the model as a plain `glm`, so the analytic sandwich path fires --- unlike
GAM, no `mgcv` dependency is needed.

```{r}
fit_spline <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + splines::ns(age, 3) + race + factor(education) +
    splines::ns(smokeintensity, 3) + splines::ns(smokeyrs, 3) +
    factor(exercise) + factor(active) + splines::ns(wt71, 3),
  censoring = "censored"
)

res_spline <- contrast(
  fit_spline,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  ci_method = "sandwich"
)
res_spline
```

## GAM model via model_fn

Pass `mgcv::gam` instead of `stats::glm` for flexible nonlinear confounder
adjustment using splines.

```{r}
fit_gam <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + s(age) + race + factor(education) +
    s(smokeintensity) + s(smokeyrs) + factor(exercise) +
    factor(active) + s(wt71),
  censoring = "censored",
  model_fn = mgcv::gam
)

res_gam <- contrast(
  fit_gam,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  ci_method = "sandwich"
)
res_gam
```

::: {.callout-note}
## GAM sandwich SEs can be larger than GLM SEs

The sandwich variance for GAMs uses the penalised Bayesian covariance
(`model$Vp`) as the bread inverse, which accounts for the smoothing
penalty. This can produce noticeably larger SEs than GLM-based models
(polynomial or spline), especially when the effective degrees of freedom
are high relative to the sample size. If the GAM SE looks disproportionately
large, consider using bootstrap inference (`ci_method = "bootstrap"`) or
switching to a GLM with `splines::ns()` terms, which stays on the standard
(unpenalised) analytic sandwich path.
:::

### Comparing model specifications

All three approaches --- standard GLM with polynomial terms, GLM with natural
splines, and GAM --- target the same causal parameter. Differences in point
estimates reflect model flexibility; all three should agree closely when the
outcome model is correctly specified.

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot comparing ATE estimates from GLM (polynomial), GLM (splines), and GAM model specifications."
model_comp <- data.frame(
  Model = c("GLM (polynomial)", "GLM (splines)", "GAM"),
  Estimate = c(
    res_ate_sw$contrasts$estimate[1],
    res_spline$contrasts$estimate[1],
    res_gam$contrasts$estimate[1]
  ),
  SE = c(
    res_ate_sw$contrasts$se[1],
    res_spline$contrasts$se[1],
    res_gam$contrasts$se[1]
  ),
  CI_lower = c(
    res_ate_sw$contrasts$ci_lower[1],
    res_spline$contrasts$ci_lower[1],
    res_gam$contrasts$ci_lower[1]
  ),
  CI_upper = c(
    res_ate_sw$contrasts$ci_upper[1],
    res_spline$contrasts$ci_upper[1],
    res_gam$contrasts$ci_upper[1]
  )
)

tt(model_comp, digits = 3)
```

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot comparing ATE estimates from three model specifications."
tinyplot(
  Estimate ~ Model,
  data = model_comp,
  type = "pointrange",
  ymin = model_comp$CI_lower,
  ymax = model_comp$CI_upper,
  ylab = "ATE (kg)",
  main = "Model specification comparison (NHEFS)"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Summary of covered combinations

**Legend.** ✅ covered and truth-pinned in tests · 🟡 smoke test only ·
⛔ rejected with an informative error.

| Treatment | Outcome | Intervention | Estimand | Contrast | Inference | Weights | Status |
|---|---|---|---|---|---|---|---|
| Binary | Continuous (gaussian, GLM) | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Static | ATE | Difference | Bootstrap | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Static | ATT | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Static | ATC | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Static | Subset | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Static | ATE (by sex, `A:sex` in model) | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Dynamic | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM) | Static | ATE | Difference | Sandwich | survey | ✅ |
| Binary | Continuous (gaussian, GAM) | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Binary (binomial, GLM) | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Binary (binomial, GLM) | Static | ATE | Difference | Bootstrap | none | ✅ |
| Binary | Binary (binomial, GLM) | Static | ATE | Ratio | Sandwich | none | ✅ |
| Binary | Binary (binomial, GLM) | Static | ATE | OR | Sandwich | none | ✅ |
| Binary | Count (poisson, GLM) | Static | ATE | Ratio | Sandwich | none | ✅ |
| Binary | Positive continuous (Gamma, GLM) | Static | ATE | Ratio | Sandwich | none | ✅ |
| Binary | Binary (quasibinomial, GLM) | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous (gaussian, GLM+splines) | Static | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous (gaussian, GLM) | Static | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous (gaussian, GLM) | Shift | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous (gaussian, GLM) | Scale | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous (gaussian, GLM) | Threshold | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous (gaussian, GLM) | Dynamic | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous (gaussian, GLM) | Shift | ATE | Difference | Bootstrap | none | ✅ |
| Categorical (k>2) | Continuous (gaussian, GLM) | Static | ATE | Difference | Sandwich | none | ✅ |
| Multivariate | Continuous (gaussian, GLM) | Joint static | ATE | Difference | Sandwich | none | ✅ |

See [`FEATURE_COVERAGE_MATRIX.md`](https://github.com/etverse/causatr/blob/main/FEATURE_COVERAGE_MATRIX.md)
for the authoritative coverage status of every method × treatment × outcome
× intervention × variance combination, including censoring and the numeric
two-tier variance fallback for custom `model_fn` choices.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 13: Standardization and the parametric g-formula.
