---
title: "G-computation with causatr"
vignette: >
  %\VignetteIndexEntry{G-computation with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

G-computation (the parametric g-formula) estimates causal effects by fitting a
model for the outcome conditional on treatment and confounders, then
standardising predictions over the target population under each hypothetical
intervention. This vignette demonstrates g-computation in causatr using the
NHEFS dataset from Hernán & Robins (2025), covering every supported combination
of treatment type, outcome type, contrast scale, and inference method for
time-fixed treatments.

## Setup

```{r}
#| message: false
library(causatr)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71), ]

nhefs_complete$gained_weight <- as.integer(nhefs_complete$wt82_71 > 0)
```

We create a binary outcome `gained_weight` (1 if weight increased, 0 otherwise)
for the binary outcome examples below.

## Binary treatment, continuous outcome

This is the core example from Chapter 13 of Hernán & Robins: the average
causal effect of quitting smoking (`qsmk`) on weight change (`wt82_71`).

### ATE with sandwich SE

```{r}
fit_gc <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity,
  censoring = "censored"
)

res_ate_sw <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_ate_sw
```

The book reports ATE $\approx$ 3.5 kg (95% CI: 2.6, 4.5).

### ATE with bootstrap SE

```{r}
res_ate_bs <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 200L
)
res_ate_bs
```

Sandwich and bootstrap SEs should be in close agreement for correctly specified
GLMs.

### ATT estimand

The average treatment effect on the treated (ATT) averages only over individuals
who actually quit smoking. With g-computation, the estimand can be changed in
`contrast()` without refitting.

```{r}
res_att <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  estimand = "ATT",
  ci_method = "sandwich"
)
res_att
```

### ATC estimand

The average treatment effect on the controls (ATC) averages over those who
continued smoking.

```{r}
res_atc <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  estimand = "ATC",
  ci_method = "sandwich"
)
res_atc
```

### Subset estimand

A custom subgroup: effect among individuals aged 50 or older.

```{r}
res_sub <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  subset = quote(age >= 50),
  ci_method = "sandwich"
)
res_sub
```

### Extracting results programmatically

```{r}
coef(res_ate_sw)
confint(res_ate_sw)
```

## Binary treatment, binary outcome

Using the `gained_weight` indicator as a binary outcome, we estimate the risk
difference, risk ratio, and odds ratio of quitting smoking on gaining weight.

### Risk difference (sandwich)

```{r}
fit_bin <- causat(
  nhefs_complete,
  outcome = "gained_weight",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2) +
    qsmk:smokeintensity,
  family = "binomial"
)

res_rd <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_rd
```

### Risk difference (bootstrap)

```{r}
res_rd_bs <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 200L
)
res_rd_bs
```

### Risk ratio

```{r}
res_rr <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

### Odds ratio

```{r}
res_or <- contrast(
  fit_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "or",
  ci_method = "sandwich"
)
res_or
```

## Continuous treatment

G-computation also supports continuous treatments with modified treatment policy
interventions. Here we use `smokeintensity` (cigarettes per day) as the
treatment and `wt82_71` as the outcome.

```{r}
fit_cont <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "smokeintensity",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeyrs + I(smokeyrs^2) + factor(exercise) + factor(active) +
    wt71 + I(wt71^2),
  censoring = "censored"
)
```

### Shift intervention

Reduce smoking intensity by 10 cigarettes per day for everyone.

```{r}
res_shift <- contrast(
  fit_cont,
  interventions = list(reduce10 = shift(-10), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_shift
```

### Scale intervention

Halve each individual's smoking intensity.

```{r}
res_scale <- contrast(
  fit_cont,
  interventions = list(halved = scale(0.5), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_scale
```

### Threshold intervention

Cap smoking intensity at 20 cigarettes per day for everyone.

```{r}
res_thresh <- contrast(
  fit_cont,
  interventions = list(cap20 = threshold(0, 20), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_thresh
```

### Comparing multiple interventions

```{r}
res_multi <- contrast(
  fit_cont,
  interventions = list(
    reduce10 = shift(-10),
    halved   = scale(0.5),
    cap20    = threshold(0, 20),
    observed = NULL
  ),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_multi
```

### Visualising intervention effects

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot of mean weight change under different smoking intensity interventions."
est <- res_multi$contrasts
tinyplot(
  estimate ~ comparison,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  xlab = "Comparison",
  ylab = "Difference in weight change (kg)",
  main = "Continuous treatment: modified treatment policies"
)
abline(h = 0, lty = 2, col = "grey40")
```

### Continuous treatment with bootstrap

```{r}
res_shift_bs <- contrast(
  fit_cont,
  interventions = list(reduce10 = shift(-10), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 200L
)
res_shift_bs
```

## Dynamic intervention

A dynamic intervention assigns treatment based on individual characteristics.
Here, we assign quitting smoking only to individuals who smoked more than 20
cigarettes per day at baseline.

```{r}
res_dyn <- contrast(
  fit_gc,
  interventions = list(
    rule = dynamic(\(data, trt) ifelse(data$smokeintensity > 20, 1, 0)),
    all_quit = static(1)
  ),
  reference = "all_quit",
  type = "difference",
  ci_method = "sandwich"
)
res_dyn
```

## Effect modification with `by`

The `by` argument stratifies causal effect estimates by levels of a variable.
Here we examine whether the effect of quitting smoking differs by sex.

```{r}
res_by_sex <- contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich",
  by = "sex"
)
res_by_sex
```

## Tidy and glance

causatr results work with the broom ecosystem via `tidy()` and `glance()`:

```{r}
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Forest plot

The `plot()` method produces a forest plot using the `forrest` package:

```{r}
#| fig-width: 7
#| fig-height: 3
#| fig-alt: "Forest plot of the ATE from g-computation."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_ate_sw)
```

## GAM model via model_fn

Pass `mgcv::gam` instead of `stats::glm` for flexible nonlinear confounder
adjustment using splines.

```{r}
fit_gam <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + s(age) + race + factor(education) +
    s(smokeintensity) + s(smokeyrs) + factor(exercise) +
    factor(active) + s(wt71),
  censoring = "censored",
  model_fn = mgcv::gam
)

res_gam <- contrast(
  fit_gam,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  ci_method = "sandwich"
)
res_gam
```

## Summary of covered combinations

| Treatment | Outcome | Contrast | Inference | Estimand | Intervention |
|---|---|---|---|---|---|
| Binary | Continuous | Difference | Sandwich | ATE | Static |
| Binary | Continuous | Difference | Bootstrap | ATE | Static |
| Binary | Continuous | Difference | Sandwich | ATT | Static |
| Binary | Continuous | Difference | Sandwich | ATC | Static |
| Binary | Continuous | Difference | Sandwich | Subset | Static |
| Binary | Continuous | Difference | Sandwich | ATE (by sex) | Static |
| Binary | Binary | Difference | Sandwich | ATE | Static |
| Binary | Binary | Difference | Bootstrap | ATE | Static |
| Binary | Binary | Ratio | Sandwich | ATE | Static |
| Binary | Binary | OR | Sandwich | ATE | Static |
| Continuous | Continuous | Difference | Sandwich | ATE | Shift |
| Continuous | Continuous | Difference | Sandwich | ATE | Scale |
| Continuous | Continuous | Difference | Sandwich | ATE | Threshold |
| Continuous | Continuous | Difference | Bootstrap | ATE | Shift |
| Binary | Continuous | Difference | Sandwich | ATE | Dynamic |
| Binary | Continuous | Difference | Sandwich | ATE | GAM model |

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 13: Standardization and the parametric g-formula.
