---
title: "Intervention types in causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Intervention types in causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
nocite: |
  @diaz2012population, @diaz2023lmtp
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

causatr separates model fitting (`causat()`) from causal contrasts
(`contrast()`). The intervention --- the hypothetical change in treatment ---
is specified at contrast time, not fit time. This design lets you fit once and
contrast under many different interventions without re-estimating nuisance
models.

This vignette tours every intervention constructor, shows which estimators
support it, and demonstrates the calling pattern on real and simulated data.

## Intervention constructors at a glance

| Constructor | What it does | G-comp | IPW | AIPW | Matching |
|---|---|---|---|---|---|
| `static(v)` | Set treatment to `v` for everyone | All types | Binary, categorical | All types | Binary |
| `shift(d)` | Add `d` to observed treatment | Continuous | Continuous, count (integer `d`) | Continuous | --- |
| `scale_by(c)` | Multiply observed treatment by `c` | Continuous | Continuous, count (integer-preserving) | Continuous | --- |
| `threshold(lo, hi)` | Clamp treatment to `[lo, hi]` | Continuous | --- | --- | --- |
| `dynamic(rule)` | Apply a user-defined deterministic rule | All types | Binary, categorical | All types | --- |
| `ipsi(d)` | Multiply odds of treatment by `d` | --- | Binary | --- | --- |
| `stochastic(f)` | Draw treatment from user-supplied distribution | All types | --- | --- | --- |
| `grace_period(w)` | Delay treatment initiation by `w` periods (natural-history MTP) | Longitudinal, discrete | --- | --- | --- |
| `carry_forward()` | Carry the baseline treatment forward (LOCF; natural-history MTP) | Longitudinal, discrete | --- | --- | --- |
| `cap_escalation(d)` | Cap the natural per-period dose increase at `d` (natural-history MTP) | Longitudinal, ordered/count | --- | --- | --- |
| `NULL` | Natural course (observed values) | All types | All types | All types | --- |

`grace_period()`, `carry_forward()`, and `cap_escalation()` are
**natural-history** modified treatment policies (G-LMTPs): the intervened
treatment at time $t$ depends on the *counterfactual natural-value history* of
treatment, so they cannot be written as `dynamic()` rules (which see only the
observed lag). They are estimated by an augmented-data sequential regression and
require a longitudinal g-computation fit with a discrete treatment.
`cap_escalation()` targets an ordered / count dose and is consistent when the
dose enters the per-step model flexibly (`treatment_form = ~ factor(dose)`) —
see the [longitudinal vignette](longitudinal.html) and the
[natural-history MTPs article](natural-history-mtps.html) for runnable examples.

## Data: NHEFS

All examples use the **NHEFS** dataset [@hernan_whatif]:

- **Binary treatment ($A$):** `qsmk` — quit smoking (0/1). Used for
  `static()`, `dynamic()`, and `ipsi()` examples.
- **Continuous treatment ($A$):** `smokeintensity` — cigarettes smoked per
  day. Used for `shift()`, `scale_by()`, and `threshold()` examples.
- **Outcome ($Y$):** `wt82_71` — weight change in kg.
- **Confounders ($L$):** sex, age, race, education, smoking-related
  variables, exercise, activity, and baseline weight — common causes of
  both $A$ and $Y$.

```{r}
#| message: false
library(causatr)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71) & !is.na(nhefs$education), ]
```

## `static()` --- set treatment to a fixed value

The simplest and most common intervention. `static(1)` means "everyone is
treated"; `static(0)` means "no one is treated". For categorical treatments,
pass the level name as a string (e.g. `static("B")`).

### G-computation

```{r}
fit_gc <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  model_fn = stats::glm
)

contrast(
  fit_gc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference"
)
```

### IPW

```{r}
fit_ipw <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference"
)
```

### AIPW (doubly robust)

```{r}
fit_aipw <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "aipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

contrast(
  fit_aipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference"
)
```

### Matching

```{r}
fit_match <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "matching",
  estimand = "ATT"
)

contrast(
  fit_match,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference"
)
```

All four methods should give broadly similar estimates (methodological
triangulation). Discrepancies signal model misspecification in one or more
approaches.

## `shift()` --- add a constant to observed treatment

A modified treatment policy (MTP) that shifts each individual's observed
treatment by a fixed amount. For continuous treatments, this answers "what
would happen if everyone's exposure increased (or decreased) by `delta`
units?"

### G-computation

```{r}
fit_gc_cont <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "smokeintensity",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeyrs + I(smokeyrs^2) + factor(exercise) + factor(active) +
    wt71 + I(wt71^2),
  model_fn = stats::glm
)

res_shift_gc <- contrast(
  fit_gc_cont,
  interventions = list(reduce10 = shift(-10), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_shift_gc
```

### IPW

Under IPW, `shift(delta)` uses a density-ratio weight: the ratio of the
treatment density evaluated at the inverse-shifted value to the density at
the observed value. The pushforward sign matters --- `shift(delta)` evaluates
at `A - delta`, not `A + delta` --- because the weight uses the inverse map.

```{r}
fit_ipw_cont <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "smokeintensity",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeyrs + I(smokeyrs^2) + factor(exercise) + factor(active) +
    wt71 + I(wt71^2),
  estimator = "ipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_shift_ipw <- contrast(
  fit_ipw_cont,
  interventions = list(reduce10 = shift(-10), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_shift_ipw
```

### Comparing shifts across methods

```{r}
#| fig-width: 7
#| fig-height: 3
#| fig-alt: "Point estimates and confidence intervals for a shift(-10) intervention under g-computation and IPW."
est <- data.frame(
  method = c("G-comp", "IPW"),
  estimate = c(
    res_shift_gc$contrasts$estimate[1],
    res_shift_ipw$contrasts$estimate[1]
  ),
  ci_lower = c(
    res_shift_gc$contrasts$ci_lower[1],
    res_shift_ipw$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    res_shift_gc$contrasts$ci_upper[1],
    res_shift_ipw$contrasts$ci_upper[1]
  )
)

tinyplot(
  estimate ~ method,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  xlab = "Method",
  ylab = "Difference in weight change (kg)",
  main = "shift(-10): reduce smoking by 10 cigs/day"
)
abline(h = 0, lty = 2, col = "grey40")
```

::: {.callout-note}
`threshold()` is **not supported** under IPW because the pushforward of a
continuous density under a boundary clamp is a mixed measure (point masses at
the bounds plus continuous density in the interior). Use
`estimator = "gcomp"` for threshold interventions.
:::

### Count treatment (Poisson / negative binomial)

For integer-valued count treatments, pass `propensity_family = "poisson"` or
`"negbin"` to use a count density model instead of the default Gaussian.

**DGM.** A Poisson-distributed count treatment (e.g. number of clinic
visits) confounded by $L$, with a linear outcome model.

$$
\begin{aligned}
  L &\sim N(0,1) \\
  A &\sim \text{Poisson}\!\bigl(\exp(0.5\,L)\bigr) \\
  Y &= 2 + 1.5\,A + L + \varepsilon, \quad \varepsilon \sim N(0,1)
\end{aligned}
$$

True contrast: $\mathbb{E}[Y(A+1)] - \mathbb{E}[Y(A)] = 1.5$.

```{r}
set.seed(6)
n <- 3000
L <- rnorm(n)
A <- rpois(n, exp(0.5 * L))
Y <- 2 + 1.5 * A + L + rnorm(n)
count_df <- data.frame(Y = Y, A = A, L = L)

fit_pois <- causat(
  count_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  propensity_family = "poisson",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_pois <- contrast(
  fit_pois,
  interventions = list(plus1 = shift(1), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_pois
```

The shift delta must be integer --- `shift(0.5)` on a count treatment aborts
because `dpois()` returns 0 at non-integer arguments.

## `scale_by()` --- multiply observed treatment by a factor

A proportional MTP. `scale_by(0.5)` halves each individual's exposure;
`scale_by(2)` doubles it.

### G-computation

```{r}
res_scale_gc <- contrast(
  fit_gc_cont,
  interventions = list(halved = scale_by(0.5), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_scale_gc
```

### IPW

Under IPW, `scale_by(c)` evaluates the density at `A / c` and includes the
Jacobian `|1/c|` in the density-ratio weight.

```{r}
res_scale_ipw <- contrast(
  fit_ipw_cont,
  interventions = list(halved = scale_by(0.5), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_scale_ipw
```

## `threshold()` --- clamp treatment within bounds (g-comp only)

Caps treatment at a ceiling and/or floor. `threshold(0, 20)` means "nobody
smokes more than 20 cigarettes per day; those below 0 are set to 0".

```{r}
res_thresh <- contrast(
  fit_gc_cont,
  interventions = list(cap20 = threshold(0, 20), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_thresh
```

This intervention is only available under g-computation. Under IPW, the
density-ratio weight is not well-defined at the boundary clamp points
(point-mass discontinuity in the pushforward density). causatr rejects
`threshold()` under `estimator = "ipw"` with an informative error pointing
at g-comp.

## `dynamic()` --- deterministic treatment rule

A dynamic intervention assigns treatment based on individual characteristics.
The `rule` function receives two arguments --- the full data and the observed
treatment vector --- and returns a vector of counterfactual treatment values.

### G-computation (binary treatment)

Assign quitting only to heavy smokers (> 20 cigarettes/day at baseline):

```{r}
res_dyn_gc <- contrast(
  fit_gc,
  interventions = list(
    rule = dynamic(\(data, trt) ifelse(data$smokeintensity > 20, 1, 0)),
    all_quit = static(1)
  ),
  reference = "all_quit",
  type = "difference"
)
res_dyn_gc
```

### IPW (binary treatment)

Under IPW, `dynamic()` on a binary treatment uses Horwitz--Thompson indicator
weights: `w_i = I(A_i = rule_i) / f(A_i | L_i)`. This is well-defined
because the treatment takes only two values and each individual's
counterfactual is deterministic.

```{r}
res_dyn_ipw <- contrast(
  fit_ipw,
  interventions = list(
    rule = dynamic(\(data, trt) ifelse(data$smokeintensity > 20, 1, 0)),
    all_quit = static(1)
  ),
  reference = "all_quit",
  type = "difference"
)
res_dyn_ipw
```

### G-computation (continuous treatment)

Dynamic rules on continuous treatments work under g-comp (predict-then-average
on the outcome model). Here, heavy smokers are reduced to 20 while others keep
their observed level:

```{r}
res_dyn_cont <- contrast(
  fit_gc_cont,
  interventions = list(
    capped = dynamic(\(data, trt) pmin(trt, 20)),
    observed = NULL
  ),
  reference = "observed",
  type = "difference"
)
res_dyn_cont
```

::: {.callout-note}
`dynamic()` on continuous treatment is **not supported** under IPW. A
deterministic rule on a continuous treatment is a Dirac per individual,
which has no Lebesgue density --- the density ratio is not defined. Rewrite
the rule as a smooth `shift()` or `scale_by()`, or use
`estimator = "gcomp"`.
:::

## `ipsi()` --- incremental propensity score intervention (IPW only)

Kennedy [-@kennedy2019ipsi]'s incremental propensity score intervention multiplies each
individual's *odds of treatment* by `delta`. It is a stochastic MTP that
never materialises a specific counterfactual treatment value --- instead,
it shifts the entire propensity score distribution.

The closed-form weight is:
$$w_i = \frac{\delta \cdot A_i + (1 - A_i)}{\delta \cdot p_i + (1 - p_i)}$$

where $p_i = P(A = 1 \mid L_i)$ is the fitted propensity.

IPSI is **binary-treatment, IPW-only** --- g-comp and matching cannot evaluate
it because there is no counterfactual treatment column to predict at.

### Doubling the odds of quitting

```{r}
res_ipsi_2 <- contrast(
  fit_ipw,
  interventions = list(double_odds = ipsi(2), natural = NULL),
  reference = "natural",
  type = "difference"
)
res_ipsi_2
```

### Halving the odds of quitting

```{r}
res_ipsi_half <- contrast(
  fit_ipw,
  interventions = list(half_odds = ipsi(0.5), natural = NULL),
  reference = "natural",
  type = "difference"
)
res_ipsi_half
```

### IPSI dose-response

Varying `delta` traces out a dose-response curve on the odds scale:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Dose-response curve showing the estimated effect of incrementally shifting the odds of quitting smoking."
deltas <- c(0.25, 0.5, 1, 2, 4)
ipsi_results <- lapply(deltas, function(d) {
  res <- contrast(
    fit_ipw,
    interventions = list(shifted = ipsi(d), natural = NULL),
    reference = "natural",
    type = "difference"
  )
  data.frame(
    delta = d,
    estimate = res$contrasts$estimate[1],
    ci_lower = res$contrasts$ci_lower[1],
    ci_upper = res$contrasts$ci_upper[1]
  )
})
ipsi_df <- do.call(rbind, ipsi_results)

tinyplot(
  estimate ~ delta,
  data = ipsi_df,
  type = "pointrange",
  ymin = ipsi_df$ci_lower,
  ymax = ipsi_df$ci_upper,
  xlab = expression(delta ~ "(odds multiplier)"),
  ylab = "E[Y(IPSI)] - E[Y(natural)] (kg)",
  main = "IPSI dose-response: effect on weight change"
)
abline(h = 0, lty = 2, col = "grey40")
```

At `delta = 1` the intervention is the natural course, so the contrast is
zero. Larger `delta` makes quitting more likely and the effect on weight
gain grows.

## `stochastic()` --- user-defined randomised intervention (g-comp only)

A stochastic intervention draws each individual's counterfactual treatment from
a user-supplied distribution that may depend on covariates. Unlike `dynamic()`,
which assigns a single deterministic value, `stochastic()` integrates over the
randomness by Monte Carlo: the g-formula is evaluated `n_mc` times, each with
an independent draw from the sampler, and the predictions are averaged.

The `sampler` function has the same signature as a `dynamic()` rule ---
`function(data, treatment)` --- but should return an independent random draw
each time it is called.

### Binary treatment: covariate-dependent randomisation

Assign quitting with probability that depends on smoking intensity:

```{r}
set.seed(42)
res_stoch <- contrast(
  fit_gc,
  interventions = list(
    random_quit = stochastic(
      \(data, trt) rbinom(nrow(data), 1,
                          plogis(-1 + 0.05 * data$smokeintensity)),
      n_mc = 200L
    ),
    all_quit = static(1)
  ),
  reference = "all_quit",
  type = "difference"
)
res_stoch
```

The estimate is a population average over both individuals *and* the stochastic
policy.

### Continuous treatment: random shift

Add a random perturbation (rather than a fixed shift) to each individual's
smoking intensity:

```{r}
set.seed(42)
res_stoch_cont <- contrast(
  fit_gc_cont,
  interventions = list(
    noisy_reduce = stochastic(
      \(data, trt) trt + rnorm(length(trt), mean = -5, sd = 2),
      n_mc = 200L
    ),
    observed = NULL
  ),
  reference = "observed",
  type = "difference"
)
res_stoch_cont
```

::: {.callout-tip}
**Choosing `n_mc`**: 100--500 is typical. The sandwich SE accounts for the MC
draws used at estimation time, so there is no need to inflate `n_mc` just for
inference. If point estimates visibly change when you re-run with a different
seed, increase `n_mc`.
:::

::: {.callout-note}
`stochastic()` is only supported under `estimator = "gcomp"` (point and
longitudinal). IPW and matching cannot evaluate it because the density ratio of
a user-defined stochastic policy is generally intractable.
:::

## `NULL` --- the natural course

`NULL` means "do not intervene; use the observed treatment values". It is the
natural reference for MTPs:

```{r}
res_multi <- contrast(
  fit_gc_cont,
  interventions = list(
    reduce10 = shift(-10),
    halved   = scale_by(0.5),
    cap20    = threshold(0, 20),
    observed = NULL
  ),
  reference = "observed",
  type = "difference"
)
res_multi
```

## Combining interventions in one contrast

Both g-comp and IPW support multiple interventions in a single `contrast()`
call. All pairwise differences against the reference are computed, and the
variance-covariance matrix covers all interventions jointly.

### G-computation

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for multiple continuous-treatment interventions under g-computation."
est <- res_multi$contrasts
tinyplot(
  estimate ~ comparison,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  xlab = "Comparison",
  ylab = "Difference in weight change (kg)",
  main = "Multiple interventions (g-comp)"
)
abline(h = 0, lty = 2, col = "grey40")
```

### IPW

```{r}
res_multi_ipw <- contrast(
  fit_ipw_cont,
  interventions = list(
    reduce10 = shift(-10),
    halved   = scale_by(0.5),
    observed = NULL
  ),
  reference = "observed",
  type = "difference"
)
res_multi_ipw
```

## Intervention support by estimator

The table below summarises which combinations are supported and which are
rejected with an informative error. The restrictions reflect genuine
architectural limits of the density-ratio framework, not implementation
gaps.

| Intervention | G-comp (point) | G-comp (longitudinal) | IPW | AIPW | Matching |
|---|---|---|---|---|---|
| `static()` | All treatment types | All treatment types | Binary, categorical | All types | Binary only |
| `shift()` | Continuous | Continuous | Continuous, count (integer delta) | Continuous | --- |
| `scale_by()` | Continuous | Continuous | Continuous, count (integer-preserving) | Continuous | --- |
| `threshold()` | Continuous | Continuous | --- (mixed pushforward) | --- | --- |
| `dynamic()` | All treatment types | All treatment types | Binary, categorical | All types | --- |
| `ipsi()` | --- | --- | Binary only | --- | --- |
| `stochastic()` | All treatment types | All treatment types | --- (intractable density ratio) | --- | --- |
| `NULL` | All | All | All | All | --- |

**Why some combinations are rejected:**

- **`threshold()` under IPW:** The pushforward of a continuous density under
  a boundary clamp is a mixed measure --- point masses at `lo`/`hi` plus
  continuous density in the interior. The density ratio is not well-defined.
- **`dynamic()` on continuous treatment under IPW:** A deterministic rule on
  a continuous treatment is a Dirac per individual with no Lebesgue density.
- **`static(v)` on continuous treatment under IPW:** Nobody is observed
  exactly at `v`, so the HT indicator `I(A = v)` is zero almost surely.
- **`ipsi()` under g-comp/matching:** IPSI shifts the propensity, not the
  treatment value. There is no counterfactual treatment column to predict at.
- **`stochastic()` under IPW/matching:** The density ratio of an arbitrary
  user-defined stochastic policy is generally intractable. Only g-computation
  (Monte Carlo integration) supports it.
- **`static()` / `dynamic()` / `threshold()` / `ipsi()` on count treatments:**
  Count treatments (enabled via `propensity_family = "poisson"` or `"negbin"`)
  use `dpois()` / `dnbinom()` for the density ratio. `static()` is degenerate
  (HT indicator is zero almost surely on a count distribution). `dynamic()` has
  the same Dirac problem as on continuous. `threshold()` has the same mixed-
  measure problem. `ipsi()` is Bernoulli-specific. Only `shift()` (integer
  delta) and `scale_by()` (where `A / factor` is integer for every observation)
  are well-defined.

## References

::: {#refs}
:::
