---
title: "Introduction to causatr"
vignette: >
  %\VignetteIndexEntry{Introduction to causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

causatr provides a unified interface for causal effect estimation via three
complementary methods:

- **G-computation** (parametric g-formula) — fits an outcome model
  $E[Y | A, L]$ and standardises predictions over the target population.
- **Inverse probability weighting (IPW)** — reweights the data using
  propensity scores so that confounders are balanced across treatment groups.
- **Propensity score matching** — pairs treated and control units with similar
  propensity scores.

When all three methods agree, you can be more confident in your findings.
This is called **methodological triangulation**.

## Two-step API

All analyses follow the same two-step pattern:

```r
# Step 1: Fit
fit <- causat(data, outcome = "Y", treatment = "A", confounders = ~ L1 + L2,
              method = "gcomp")

# Step 2: Contrast
result <- contrast(fit,
  interventions = list(treated = static(1), control = static(0)),
  reference = "control")
```

## Quick example

```{r}
#| message: false
library(causatr)
data("nhefs")

fit <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  censoring = "censored"
)

result <- contrast(
  fit,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue"
)
result
```

## Intervention types

causatr supports several intervention types for g-computation:

| Function | Description | Example |
|---|---|---|
| `static(value)` | Set treatment to a fixed value | `static(1)` |
| `shift(delta)` | Add a constant to observed treatment | `shift(-10)` |
| `scale(factor)` | Multiply observed treatment | `scale(0.5)` |
| `threshold(lower, upper)` | Clamp treatment within bounds | `threshold(0, 20)` |
| `dynamic(rule)` | Apply a user-defined rule | `dynamic(\(d, a) ifelse(d$L > 0, 1, 0))` |
| `NULL` | Natural course (observed values) | — |

IPW and matching currently support only `static()` interventions.

## Contrast types

| Type | Meaning |
|---|---|
| `"difference"` | Risk/mean difference (default) |
| `"ratio"` | Risk ratio |
| `"or"` | Odds ratio (binary outcomes) |

## Inference methods

| Method | Description |
|---|---|
| `"sandwich"` | Robust sandwich SE (default, fast) |
| `"bootstrap"` | Nonparametric bootstrap (resamples full pipeline) |

## Diagnostics

Use `diagnose()` to check positivity and covariate balance after fitting:

```{r}
#| eval: false
diag <- diagnose(fit)
diag          # print positivity + balance summary
plot(diag)    # Love plot (requires cobalt)
```

For IPW and matching, balance is computed via
[cobalt](https://ngreifer.github.io/cobalt/) and includes standardised mean
differences before and after adjustment.

## Learning more

- `vignette("gcomp")` — G-computation in depth
- `vignette("ipw")` — Inverse probability weighting in depth
- `vignette("matching")` — Propensity score matching in depth
- `vignette("triangulation")` — Comparing all three methods
