---
title: "Introduction to causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Introduction to causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

causatr provides a unified interface for causal effect estimation via four
complementary methods:

- **G-computation** (parametric g-formula) — fits an outcome model
  $E[Y | A, L]$ and standardises predictions over the target population.
- **Inverse probability weighting (IPW)** — reweights the data using
  propensity scores so that confounders are balanced across treatment groups.
- **Augmented IPW (AIPW)** — combines the outcome model and propensity score
  into a doubly-robust estimator: consistent if *either* model is correctly
  specified, and semiparametrically efficient when both are correct.
- **Propensity score matching** — pairs treated and control units with similar
  propensity scores.

When multiple methods agree, you can be more confident in your findings.
This is called **methodological triangulation**.

## Two-step API

All analyses follow the same two-step pattern:

```r
# Step 1: Fit
fit <- causat(data, outcome = "Y", treatment = "A", confounders = ~ L1 + L2,
              estimator = "gcomp")

# Step 2: Contrast
result <- contrast(fit,
  interventions = list(treated = static(1), control = static(0)),
  reference = "control")
```

## Quick example

The example below uses the **NHEFS** (National Health and Nutrition
Examination Survey — Epidemiologic Follow-up Study) dataset from Hernán &
Robins (2025). The assumed causal structure is:

- **Treatment ($A$):** `qsmk` — whether the individual quit smoking between
  baseline (1971) and follow-up (1982). Binary (0 = continued, 1 = quit).
- **Outcome ($Y$):** `wt82_71` — change in body weight (kg) from 1971 to
  1982. Continuous.
- **Confounders ($L$):** a set of baseline covariates — sex, age, race,
  smoking intensity (cigarettes/day), years of smoking, exercise, physical
  activity level, and baseline weight — that jointly affect both the
  decision to quit smoking and the change in weight.
- **Censoring ($C$):** `censored` — whether the individual was lost to
  follow-up. Rows with $C = 1$ are excluded from model fitting; predictions
  are standardised over the full target population.

Under the causal DAG, $L$ is a common cause of both $A$ and $Y$. Adjusting
for $L$ blocks all backdoor paths from $A$ to $Y$, so the g-formula
identifies the average causal effect $\mathbb{E}[Y^{a=1}] - \mathbb{E}[Y^{a=0}]$.

```{r}
#| message: false
library(causatr)
data("nhefs")

fit <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  censoring = "censored"
)

result <- contrast(
  fit,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue"
)
result
```

## Intervention types

causatr supports several intervention types for g-computation:

| Function | Description | Example |
|---|---|---|
| `static(value)` | Set treatment to a fixed value | `static(1)` |
| `shift(delta)` | Add a constant to observed treatment | `shift(-10)` |
| `scale_by(factor)` | Multiply observed treatment | `scale_by(0.5)` |
| `threshold(lower, upper)` | Clamp treatment within bounds | `threshold(0, 20)` |
| `dynamic(rule)` | Apply a user-defined rule | `dynamic(\(d, a) ifelse(d$L > 0, 1, 0))` |
| `NULL` | Natural course (observed values) | — |

IPW supports `static()`, `shift()`, `scale_by()`, `dynamic()`, and `ipsi()`.
AIPW supports the same set as IPW (all except `threshold()` and `stochastic()`).
Matching supports only `static()` interventions.

## Contrast types

| Type | Meaning |
|---|---|
| `"difference"` | Risk/mean difference (default) |
| `"ratio"` | Risk ratio |
| `"or"` | Odds ratio (binary outcomes) |

## Inference methods

| Method | Description |
|---|---|
| `"sandwich"` | Robust sandwich SE (default, fast) |
| `"bootstrap"` | Nonparametric bootstrap (resamples full pipeline) |

## Estimands

| Estimand | Population | Applicability |
|---|---|---|
| `"ATE"` | All individuals | Always (default) |
| `"ATT"` | Observed treated (A = 1) | Binary point treatment only |
| `"ATC"` | Observed controls (A = 0) | Binary point treatment only |
| `subset = quote(...)` | Custom subgroup | Always |

For g-computation, the estimand can be changed in `contrast()` without
refitting. For IPW and matching, refit with `causat(estimand = ...)`.

## Treatment types

causatr supports binary, continuous, categorical, count, and multivariate treatments:

| Treatment type | G-comp | IPW | AIPW | Matching |
|---|---|---|---|---|
| Binary (0/1) | Yes | Yes | Yes | Yes |
| Continuous | Yes | Yes (GPS) | Yes (shift, scale\_by) | No (MatchIt binary-only) |
| Categorical (k > 2) | Yes | Yes | Yes | No (MatchIt binary-only) |
| Count (Poisson / NB) | Yes | Yes (`propensity_family =`) | Yes | No |
| Multivariate (joint) | Yes | Yes (sequential MTP) | No (Phase 8 pending) | No |

For multivariate treatments (e.g. `treatment = c("A1", "A2")`), supply
interventions as named lists with one element per treatment variable.

Effect modification via `A:modifier` interaction terms in `confounders` is
supported across all estimation methods. G-comp and AIPW feed the interaction
to the outcome model directly; IPW expands the per-intervention MSM to
`Y ~ 1 + modifier`; matching expands to `Y ~ A + modifier + A:modifier`;
and ICE auto-expands the interaction across treatment lags. Use
`by = "modifier"` in `contrast()` to obtain stratum-specific estimates.

## Effect modification

The `by` argument in `contrast()` stratifies estimates by levels of a
variable, producing subgroup-specific effects:

```{r}
#| eval: false
contrast(fit,
  interventions = list(quit = static(1), continue = static(0)),
  by = "sex"
)
```

## Survey designs and cluster-robust inference

`causat()` and `contrast()` support two inference-infrastructure
conveniences for design-aware studies.

**Sampling weights via `survey::svydesign`.** Pass a design object
directly to `weights =` and causatr unpacks both the sampling weights
(via `stats::weights()`) and the first-stage cluster (PSU):

```{r}
#| eval: false
library(survey)
des <- svydesign(ids = ~psu, weights = ~pw, data = nhefs)
fit <- causat(nhefs, outcome = "wt82_71", treatment = "qsmk",
              confounders = ~ sex + age + wt71, weights = des)
```

**Cluster-robust sandwich.** Use `cluster = "col"` on `causat()` (or
`contrast()`) to aggregate per-individual influence functions
within-cluster before squaring (Liang & Zeger 1986). Numerically
equivalent to `sandwich::vcovCL(type = "HC0")` applied to the final
predict-then-average step. Supported for gcomp / IPW / ICE; rejected
for matching (subclass clustering is already structural).

**`future` backend for bootstrap.** `contrast(parallel = "future")`
dispatches replicates through `future.apply::future_lapply()`, so any
`future::plan()` (local multisession, remote cluster,
`future.batchtools` for HPC) is honoured.

## Diagnostics

Use `diagnose()` to check positivity and covariate balance after fitting:

```{r}
#| eval: false
diag <- diagnose(fit)
diag          # print positivity + balance summary
plot(diag)    # Love plot (requires cobalt)
```

For IPW and matching, balance is computed via
[cobalt](https://ngreifer.github.io/cobalt/) and includes standardised mean
differences before and after adjustment.

## Extracting results

causatr result objects support standard R generics:

```r
coef(result)       # named vector of E[Y^a]
confint(result)    # CI matrix (respects custom `level`)
vcov(result)       # vcov of marginal means
tidy(result)       # broom-style tidy data frame
glance(result)     # broom-style one-row summary
plot(result)       # forest plot (requires forrest package)
```

## Limitations and known rejections

causatr rejects unsupported combinations with informative errors rather than
silently producing wrong answers. Key restrictions:

| What | Why |
|---|---|
| **Matching requires binary treatment** | MatchIt does not support continuous/categorical treatments |
| **ATT/ATC only for binary 0/1 treatment** | The estimand is not defined for other treatment types |
| **`threshold()` under IPW** | The pushforward density has point masses at the bounds (mixed measure) |
| **`dynamic()` on continuous treatment under IPW** | A deterministic rule on a continuous treatment has no Lebesgue density |
| **`ipsi()` only under IPW, binary only** | The IPSI weight formula requires binary treatment and a propensity model |
| **`stochastic()` only under g-comp** | The density ratio of a user-defined stochastic policy is intractable |
| **Missing treatment values** | Cannot define a counterfactual without observed treatment |
| **`na.action = na.exclude` rejected** | Causes silent IF corruption via residual padding mismatch |
| **Effect modifier must be baseline under IPW/matching** | Time-varying modifiers in MSMs bias via collider paths (not enforced at runtime) |
| **Multivariate longitudinal IPW** | Not yet supported; use ICE g-computation |

For the full matrix of supported estimator × treatment × intervention × outcome
combinations, see `vignette("interventions")` and
[`FEATURE_COVERAGE_MATRIX.md`](../FEATURE_COVERAGE_MATRIX.md).

::: {.callout-warning}
## Confounder selection: use a DAG, not data-driven methods

The `confounders =` formula in `causat()` should reflect a pre-specified
**causal DAG** (directed acyclic graph) encoding your assumptions about the
data-generating process. Do **not** select confounders via stepwise regression,
LASSO, or other data-driven variable selection. These methods optimise
predictive fit, not causal identification, and can:
- **Include colliders** — conditioning on a common effect of treatment and outcome
  opens a non-causal path, biasing the estimate toward zero or even flipping
  its sign.
- **Include instruments** — variables that affect treatment but not outcome
  inflate variance without reducing confounding bias (a pure efficiency loss
  under correct specification, but can amplify bias from unmeasured confounders).
- **Include mediators** — variables on the causal path from treatment to outcome
  block part of the effect you are trying to estimate, attenuating it toward the
  null.

Use [`dagitty`](https://cran.r-project.org/package=dagitty) to encode your DAG
and query it for the minimal sufficient adjustment set
(`dagitty::adjustmentSets()`). See Hernán & Robins (2025), Ch. 18.
:::

## Learning more

causatr includes detailed vignettes for each estimation method: g-computation,
IPW, matching, longitudinal ICE, and methodological triangulation. Use
`vignette(package = "causatr")` to see all available vignettes.
