---
title: "Introduction to causatr"
vignette: >
  %\VignetteIndexEntry{Introduction to causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

causatr provides a unified interface for causal effect estimation via three
complementary methods:

- **G-computation** (parametric g-formula) — fits an outcome model
  $E[Y | A, L]$ and standardises predictions over the target population.
- **Inverse probability weighting (IPW)** — reweights the data using
  propensity scores so that confounders are balanced across treatment groups.
- **Propensity score matching** — pairs treated and control units with similar
  propensity scores.

When all three methods agree, you can be more confident in your findings.
This is called **methodological triangulation**.

## Two-step API

All analyses follow the same two-step pattern:

```r
# Step 1: Fit
fit <- causat(data, outcome = "Y", treatment = "A", confounders = ~ L1 + L2,
              estimator = "gcomp")

# Step 2: Contrast
result <- contrast(fit,
  interventions = list(treated = static(1), control = static(0)),
  reference = "control")
```

## Quick example

```{r}
#| message: false
library(causatr)
data("nhefs")

fit <- causat(
  nhefs,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + race + smokeintensity + smokeyrs +
    factor(exercise) + factor(active) + wt71,
  censoring = "censored"
)

result <- contrast(
  fit,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue"
)
result
```

## Intervention types

causatr supports several intervention types for g-computation:

| Function | Description | Example |
|---|---|---|
| `static(value)` | Set treatment to a fixed value | `static(1)` |
| `shift(delta)` | Add a constant to observed treatment | `shift(-10)` |
| `scale_by(factor)` | Multiply observed treatment | `scale_by(0.5)` |
| `threshold(lower, upper)` | Clamp treatment within bounds | `threshold(0, 20)` |
| `dynamic(rule)` | Apply a user-defined rule | `dynamic(\(d, a) ifelse(d$L > 0, 1, 0))` |
| `NULL` | Natural course (observed values) | — |

IPW and matching currently support only `static()` interventions.

## Contrast types

| Type | Meaning |
|---|---|
| `"difference"` | Risk/mean difference (default) |
| `"ratio"` | Risk ratio |
| `"or"` | Odds ratio (binary outcomes) |

## Inference methods

| Method | Description |
|---|---|
| `"sandwich"` | Robust sandwich SE (default, fast) |
| `"bootstrap"` | Nonparametric bootstrap (resamples full pipeline) |

## Estimands

| Estimand | Population | Applicability |
|---|---|---|
| `"ATE"` | All individuals | Always (default) |
| `"ATT"` | Observed treated (A = 1) | Binary point treatment only |
| `"ATC"` | Observed controls (A = 0) | Binary point treatment only |
| `subset = quote(...)` | Custom subgroup | Always |

For g-computation, the estimand can be changed in `contrast()` without
refitting. For IPW and matching, refit with `causat(estimand = ...)`.

## Treatment types

causatr supports binary, continuous, categorical, and multivariate treatments:

| Treatment type | G-comp | IPW | Matching |
|---|---|---|---|
| Binary (0/1) | Yes | Yes | Yes |
| Continuous | Yes | Yes (GPS) | No (MatchIt binary-only) |
| Categorical (k > 2) | Yes | Yes | No (MatchIt binary-only) |
| Multivariate (joint) | Yes | Not yet (Phase 4) | Not yet (Phase 7) |

For multivariate treatments (e.g. `treatment = c("A1", "A2")`), supply
interventions as named lists with one element per treatment variable.

Effect modification in the outcome / MSM model itself (e.g. `~ L + sex + A:sex`)
is currently supported only for g-computation. IPW and matching wrap a
saturated `Y ~ A` MSM around the propensity / matched design, and reject an
`A:modifier` term up front — use `estimator = "gcomp"` for heterogeneous effects,
or `by = "modifier"` below for stratum-specific summaries of a homogeneous
effect. See `PHASE_8_INTERACTIONS.md` for the planned unified API.

## Effect modification

The `by` argument in `contrast()` stratifies estimates by levels of a
variable, producing subgroup-specific effects:

```{r}
#| eval: false
contrast(fit,
  interventions = list(quit = static(1), continue = static(0)),
  by = "sex"
)
```

## Diagnostics

Use `diagnose()` to check positivity and covariate balance after fitting:

```{r}
#| eval: false
diag <- diagnose(fit)
diag          # print positivity + balance summary
plot(diag)    # Love plot (requires cobalt)
```

For IPW and matching, balance is computed via
[cobalt](https://ngreifer.github.io/cobalt/) and includes standardised mean
differences before and after adjustment.

## Extracting results

causatr result objects support standard R generics:

```r
coef(result)       # named vector of E[Y^a]
confint(result)    # CI matrix (respects custom `level`)
vcov(result)       # vcov of marginal means
tidy(result)       # broom-style tidy data frame
glance(result)     # broom-style one-row summary
plot(result)       # forest plot (requires forrest package)
```

## Learning more

causatr includes detailed vignettes for each estimation method: g-computation,
IPW, matching, longitudinal ICE, and methodological triangulation. Use
`vignette(package = "causatr")` to see all available vignettes.
