---
title: "IPW density-ratio weights and variance theory"
toc-depth: 4
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{IPW density-ratio weights and variance theory}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
mjx-container[jax="CHTML"][display="true"] {
  overflow-x: auto;
  overflow-y: visible;
  padding: 0.5em 0;
}
</style>
```

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette develops the density-ratio weight formulas behind causatr's
self-contained IPW engine and shows how they feed into the sandwich variance
estimator. It is a companion to `vignette("variance-theory")`, which
derives the general stacked M-estimation framework for IPW in Section 4.2.
Here we zoom into the pieces that Section 4.2 treats as black boxes: the
weight formulas themselves, the `make_weight_fn()` closure that the variance
engine differentiates through, and the numerical cross-derivative
$A_{\beta\alpha}$ that propagates propensity-score uncertainty into the
final standard error.

For practical usage examples see `vignette("ipw")` and
`vignette("interventions")`.

## 1. Density-ratio weights: the general idea

IPW reweights the observed sample so that, under the reweighted
distribution, confounders are balanced and the weighted mean of $Y$
estimates the counterfactual mean $E[Y^d]$ under intervention $d$.
The key object is the **density-ratio weight**:

$$
w_i = \frac{f_d(A_i^{\text{obs}} \mid L_i)}{f(A_i^{\text{obs}} \mid L_i)},
$$

where $f(\cdot \mid L)$ is the conditional treatment density under the
observed data and $f_d(\cdot \mid L)$ is the density of the treatment
**under intervention $d$**. The nature of $f_d$ depends on the
intervention type:

- For **point-mass interventions** (static, dynamic) the intervened
  distribution is a Dirac, and the weight collapses to a
  Horvitz--Thompson indicator.
- For **smooth MTPs** (shift, scale) the intervened distribution is a
  pushforward of $f$ under a diffeomorphism, and the weight is a smooth
  density ratio.
- For **IPSI** (incremental propensity-score interventions) the weight
  has a closed-form expression that bypasses density evaluation entirely.

The rest of this section works through each case.

### 1.1 Horvitz--Thompson indicator form (static, dynamic)

When the intervention sets each individual's treatment to a deterministic
value $a^* = d(A_i, L_i)$, the intervened "density" is a point mass
$\delta_{a^*}$. The Radon--Nikodym derivative of a point mass w.r.t. a
density $f$ is:

$$
\frac{d\delta_{a^*}}{df}(A_i^{\text{obs}})
= \frac{\mathbb{1}\{A_i = a^*\}}{f(a^* \mid L_i)}.
$$

For binary treatment with `static(1)`, this becomes the familiar
$w_i = A_i / e(L_i)$ where $e(L_i) = P(A_i=1 \mid L_i)$ is the
propensity score.

#### Estimand-specific numerator (ATE / ATT / ATC)

The ATE targets the whole population; ATT and ATC target subpopulations.
To recover $E[Y^a \mid A = A^*]$ from observed data, the standard
Bayes-rule rewrite (Imbens 2004; Hernán & Robins Ch. 12) gives a
per-individual numerator $f^*_i$:

$$
w_i = \frac{\mathbb{1}\{A_i = a\} \cdot f^*_i}{f(A_i \mid L_i)},
$$

where:

| Estimand | Target subpopulation | $f^*_i$ |
|---|---|---|
| ATE | Whole population | $1$ |
| ATT | The treated ($A^* = 1$) | $p(L_i)$ |
| ATC | The controls ($A^* = 0$) | $1 - p(L_i)$ |

This single formula reproduces all three per-arm weight schemes in one
branch. For the degenerate arm (e.g. ATT with `static(1)` on the $A=1$
rows), $f^* = p$ and $f(1 \mid L) = p$ cancel, giving $w_i = A_i$ ---
the direct sample functional $E[Y \mid A=1]$ with no propensity
uncertainty, as expected.

causatr enforces ATT/ATC only for `static()` on binary 0/1 treatment.
All other interventions are ATE-only; non-ATE requests are rejected
by `check_estimand_intervention_compat()`.

::: {.callout-note}
### Categorical treatment

For a categorical treatment with $K$ levels, the same HT form applies:
$w_i = \mathbb{1}\{A_i = a^*\} / P(A_i = a^* \mid L_i)$, where the
multinomial probabilities come from a `nnet::multinom` model. Only ATE
is supported (ATT/ATC are ambiguous when there are more than two treatment
levels).
:::

### 1.2 Smooth pushforward (shift, scale_by) {#sec-pushforward}

When the intervention is a smooth, invertible map $d: a \mapsto d(a, l)$
on a continuous treatment, the intervened density is the **pushforward**
of $f$ under $d$. By the change-of-variables formula:

$$
f_d(y \mid l) = f\bigl(d^{-1}(y, l) \mid l\bigr) \cdot \lvert\det Dd^{-1}(y, l)\rvert.
$$

Evaluating at $y = A_i^{\text{obs}}$ gives the weight:

$$
w_i = \frac{f\bigl(d^{-1}(A_i, L_i) \mid L_i\bigr) \cdot |\text{Jac}|}{f(A_i \mid L_i)}.
$$ {#eq-pushforward}

For the two MTPs causatr supports:

| Intervention | Map $d(a)$ | Inverse $d^{-1}(y)$ | Jacobian $\lvert\det Dd^{-1}\rvert$ |
|---|---|---|---|
| `shift(delta)` | $a + \delta$ | $y - \delta$ | $1$ |
| `scale_by(c)` | $c \cdot a$ | $y / c$ | $1 / \lvert c \rvert$ |

::: {.callout-warning}
### The pushforward sign trap

A common mistake is to compute $f(d(A_i) \mid L_i) / f(A_i \mid L_i)$
--- i.e. evaluating the fitted density at the **intervened** treatment
value. This is wrong: the correct formula evaluates at
$d^{-1}(A_i^{\text{obs}})$, not $d(A_i^{\text{obs}})$.

For `shift(delta)`, the difference is a sign flip: evaluating at
$A_i + \delta$ gives $E[Y^{\text{shift}(-\delta)}]$ instead of
$E[Y^{\text{shift}(\delta)}]$. The mistake is easy to write and hard
to catch without a truth-based simulation test.

In `compute_density_ratio_weights()`:

```r
# shift: d(a) = a + delta, d^{-1}(y) = y - delta, |Jac| = 1.
# f_d(A_obs | l) = f(A_obs - delta | l).
a_eval <- a_obs - delta                       # <-- d^{-1}, not d
f_int  <- evaluate_density(tm, a_eval, data)
w      <- f_int / f_obs
```

For `scale_by(c)`:

```r
# d(a) = c*a, d^{-1}(y) = y/c, |Jac| = 1/|c|.
a_eval <- a_obs / fct                         # <-- d^{-1}, not d
f_int  <- evaluate_density(tm, a_eval, data)
w      <- (f_int / f_obs) / abs(fct)
```
:::

#### Why `threshold()` and `dynamic()` are rejected on continuous treatment

`threshold(lo, hi)` clamps $A$ to $[\text{lo}, \text{hi}]$. The
pushforward of a continuous density under a boundary clamp is a **mixed
measure**: continuous on $(\text{lo}, \text{hi})$ plus point masses at the
boundaries (from all the probability mass outside the interval that gets
piled up at the clamp points). A mixed measure has no Lebesgue density,
so the density ratio in @eq-pushforward is not defined. Users should
switch to `estimator = "gcomp"`, where the clamped-treatment
counterfactual is handled cleanly via predict-then-average on the outcome
model.

Similarly, `dynamic(rule)` on a continuous treatment defines a Dirac
$\delta_{d(L_i)}$ per individual --- a degenerate "density" with zero
Lebesgue density. The weight is zero almost surely. Users should rewrite
the rule as a smooth `shift()` or `scale_by()`, or switch to `gcomp`.

### 1.3 IPSI closed form (Kennedy 2019) {#sec-ipsi}

Incremental propensity-score interventions (Kennedy 2019) act on the
**propensity score** rather than on the treatment value. The intervention
multiplies the odds of treatment by a factor $\delta > 0$:

$$
\text{odds}_d(L_i) = \delta \cdot \text{odds}(L_i)
\quad\Rightarrow\quad
p_d(L_i) = \frac{\delta \, p(L_i)}{1 - p(L_i) + \delta \, p(L_i)}.
$$

Because the intervention is defined in terms of the propensity score
rather than a counterfactual treatment value, there is no pushforward
density to evaluate. The density-ratio weight simplifies to a closed
form:

$$
w_i = \frac{\delta \, A_i + (1 - A_i)}{\delta \, p_i + (1 - p_i)},
$$ {#eq-ipsi-weight}

where $p_i = P(A_i = 1 \mid L_i)$ is the fitted propensity score.

This formula has several appealing properties:

- **No density evaluation.** Unlike `shift()` / `scale_by()`, IPSI never
  evaluates $f(a \mid L)$ at a counterfactual treatment value. The weight
  is a direct function of the propensity.
- **Finite weights.** For finite $\delta > 0$, the denominator
  $\delta \, p + (1-p)$ is bounded between $\min(1, \delta)$ and
  $\max(1, \delta)$, so the weight is always finite. No positivity
  guard needed.
- **Smooth in $\delta$.** The dose--response curve
  $\delta \mapsto E[Y^{d_\delta}]$ is smooth, making it a natural
  candidate for a marginal structural model across a grid of $\delta$
  values.

In causatr:

```r
ipsi_weight_formula <- function(a_obs, p, delta) {
  (delta * a_obs + (1 - a_obs)) / (delta * p + (1 - p))
}
```

## 2. The `make_weight_fn()` closure {#sec-closure}

The sandwich variance estimator needs the cross-derivative
$A_{\beta\alpha} = -\partial\bar\psi_\beta/\partial\alpha$, which
captures how the MSM score changes when the propensity parameters shift.
causatr computes this numerically via `numDeriv::jacobian()`. The
numerical Jacobian needs a function $\alpha \mapsto w_i(\alpha)$ that
recomputes the weight vector under a candidate propensity parameter.

`make_weight_fn()` builds this closure. It captures everything it needs
**by value** at creation time:

| Captured quantity | Description | Varies under $\alpha$ perturbation? |
|---|---|---|
| `X_prop` | Propensity design matrix ($n \times q$) | No --- matrix is fixed |
| `a_obs` | Observed treatment values | No |
| `a_int` / `a_eval` | Intervened or inverse-intervened values | No --- intervention is fixed |
| `sigma` | Residual SD (continuous treatments) | No --- fixed at fit time |
| `delta` | Odds multiplier (IPSI only) | No |
| `family` tag | Dispatches to the right formula | No |

The only quantity that varies is $\alpha$ itself. Inside the closure,
predictions are recomputed from $\alpha$ via a single matrix multiply
`X_prop %*% alpha`, avoiding formula parsing on every `numDeriv` step.

### Closure branches

**IPSI:**

```r
function(alpha) {
  eta <- as.numeric(X_prop %*% alpha)
  p   <- plogis(eta)
  ipsi_weight_formula(a_obs, p, delta)
}
```

**HT indicator (Bernoulli):**

```r
function(alpha) {
  eta   <- as.numeric(X_prop %*% alpha)
  p     <- plogis(eta)
  f_obs <- ifelse(a_obs == 1, p, 1 - p)
  ind * f_star_fn(p) / f_obs
}
```

Here `ind` (the HT indicator $\mathbb{1}\{A_i = a^*\}$) and
`f_star_fn` (the estimand-specific Bayes numerator) are captured at
creation time. For ATE, `f_star_fn(p) = 1`; for ATT, `f_star_fn(p) = p`;
for ATC, `f_star_fn(p) = 1 - p`. Baking the ATT/ATC numerator into the
closure ensures that `numDeriv::jacobian()` picks up the
propensity-dependence of the numerator correctly.

**HT indicator (categorical/multinomial):**

```r
function(alpha) {
  alpha_mat <- matrix(alpha, nrow = K-1, ncol = p, byrow = TRUE)
  eta       <- X_prop %*% t(alpha_mat)
  exp_eta   <- exp(eta)
  denom     <- 1 + rowSums(exp_eta)
  prob_mat  <- cbind(1/denom, exp_eta/denom)     # softmax
  f_obs     <- prob_mat[cbind(1:n, col_idx)]
  ind / f_obs                                    # ATE only
}
```

The flattened $\alpha$ vector encodes $(K-1) \times p$ coefficients
(row-major). Each `numDeriv` step perturbs one entry; the closure
reconstructs the coefficient matrix, applies the softmax, and extracts
the per-individual probability at the observed treatment level.

**Smooth pushforward (Gaussian):**

```r
function(alpha) {
  mu    <- as.numeric(X_prop %*% alpha)
  f_obs <- dnorm(a_obs, mean = mu, sd = sigma)
  f_eval <- dnorm(a_eval, mean = mu, sd = sigma)
  (f_eval / f_obs) * jac_abs
}
```

Here `a_eval` ($= d^{-1}(A_i^{\text{obs}})$) and `jac_abs` are
precomputed once. The residual SD `sigma` is held fixed under $\alpha$
perturbation. Treating $\sigma$ as an additional nuisance parameter would
require a joint $(\mu, \sigma)$ M-estimation setup --- deferred.

::: {.callout-note}
### Why `sigma` is fixed

The conditional treatment density is $A_i \mid L_i \sim N(\mu_i, \sigma^2)$
where $\mu_i = L_i^T\alpha$. Under maximum likelihood the score for
$\alpha$ and the score for $\sigma$ are orthogonal (information matrix
is block-diagonal in $(\alpha, \sigma)$). Fixing $\sigma$ at its MLE and
treating only $\alpha$ as the propensity nuisance is equivalent to
profiling out $\sigma$ --- the resulting sandwich is asymptotically
correct for $\hat\mu_a$ up to an $O(n^{-1})$ term from the profiled
$\sigma$ uncertainty (Liang & Zeger 1986). In practice the effect on the
SE is negligible because $\sigma$ enters the weight ratio only through the
Normal pdf, where it cancels to leading order.
:::

## 3. The cross-derivative $A_{\beta\alpha}$ {#sec-cross}

`vignette("variance-theory")` Section 4.2 derives the full IF for
$\hat\mu_a$ under IPW as a three-term sum (Channel 1 + MSM correction +
propensity correction). The propensity correction chain is:

$$
\text{propensity correction}_i
= J \, A_{\beta\beta}^{-1} \, A_{\beta\alpha} \, A_{\alpha\alpha}^{-1}
  \, \psi_{\alpha,i},
$$

where $A_{\beta\alpha}$ is the cross-derivative of the weighted MSM
score w.r.t. the propensity parameters:

$$
A_{\beta\alpha}
= -\frac{1}{n} \sum_{i=1}^{n}
  \frac{\partial \psi_{\beta,i}}{\partial \alpha^T}
= -\frac{1}{n} \sum_{i=1}^{n}
  X_i \, \frac{\partial w_i}{\partial \alpha^T} \,
  \frac{(Y_i - \mu_i)\,\mu'(\eta_i)}{V(\mu_i)}.
$$

The weight derivative $\partial w_i / \partial\alpha^T$ depends on
the intervention type (HT indicator, smooth pushforward, IPSI --- each
is a different function of $\alpha$). Rather than deriving a closed-form
formula for each case, causatr evaluates $A_{\beta\alpha}$ numerically
via `numDeriv::jacobian()` on a function that computes the average
weighted MSM score as a function of $\alpha$:

$$
\bar\psi_\beta(\alpha)
= \frac{1}{n} \sum_{i=1}^{n}
  X_i \, w_i(\alpha) \,
  \frac{(Y_i - \hat\mu_i)\,\mu'(\hat\eta_i)}{V(\hat\mu_i)}.
$$

Everything except $w_i(\alpha)$ is held fixed at $\hat\beta$. The
Jacobian of $\bar\psi_\beta$ w.r.t. $\alpha$ is
$+\partial\bar\psi_\beta/\partial\alpha$, so the negative-Hessian
convention gives:

$$
A_{\beta\alpha} = -\texttt{numDeriv::jacobian}(\bar\psi_\beta, \hat\alpha).
$$

```r
phi_bar <- function(alpha) {
  w_alpha <- weight_fn(alpha)
  s_per_i <- w_alpha * mu_eta * r_fit / var_mu
  as.numeric(crossprod(X_msm, s_per_i)) / n_fit
}
A_beta_alpha <- -numDeriv::jacobian(phi_bar, x = alpha_hat)
```

The `weight_fn` here is the closure from @sec-closure, which encodes
the intervention-specific weight formula. By building the closure once
and handing it to `numDeriv::jacobian()`, we get a single code path
for the cross-derivative across all intervention types.

### Numerical verification {#sec-verify-cross}

causatr's test suite (`test-ipw-cross-derivative.R`) verifies the
numerical $A_{\beta\alpha}$ against hand-derived analytic formulas for
the static binary case, where the closed form is available:

$$
(A_{\beta\alpha})_{j,k}
= -\frac{1}{n}\sum_{i=1}^{n}
  X_{i,j}^{\text{msm}} \cdot
  \frac{(Y_i - \hat\mu_i)\,\mu'(\hat\eta_i)}{V(\hat\mu_i)} \cdot
  \frac{\partial w_i}{\partial \alpha_k}.
$$

For `static(1)` with Bernoulli treatment ($w_i = A_i / p_i$):

$$
\frac{\partial w_i}{\partial \alpha_k}
= -\frac{A_i}{p_i^2} \cdot p_i(1-p_i) \cdot X_{i,k}^{\text{prop}}
= -\frac{A_i(1-p_i)}{p_i} \cdot X_{i,k}^{\text{prop}}.
$$

For `shift(delta)` with Gaussian treatment ($w_i = \phi(a_i - \delta; \mu_i, \sigma) / \phi(a_i; \mu_i, \sigma)$):

$$
\frac{\partial w_i}{\partial \alpha_k}
= w_i \cdot \frac{\delta}{\sigma^2} \cdot X_{i,k}^{\text{prop}}.
$$

For IPSI with Bernoulli treatment:

$$
\frac{\partial w_i}{\partial \alpha_k}
= \frac{(\delta A_i + 1 - A_i) \cdot (-1)(\delta - 1) \cdot p_i(1-p_i)}
       {(\delta p_i + 1 - p_i)^2}
  \cdot X_{i,k}^{\text{prop}}.
$$

The test asserts agreement at $\sim 10^{-6}$ tolerance across all five
weight branches (static(1), static(0), shift, IPSI, natural course).

::: {.panel-tabset}

## Code

```{r}
#| label: verify-cross-deriv
library(causatr)

set.seed(42)
n <- 200
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- 2 + 3 * A + L + rnorm(n)
dat <- data.table::data.table(Y = Y, A = A, L = L)

fit <- causat(dat, outcome = "Y", treatment = "A", confounders = ~ L,
              estimator = "ipw")
res <- contrast(fit, interventions = list(a1 = static(1), a0 = static(0)),
                reference = "a0", ci_method = "sandwich")

tm <- fit$details$treatment_model
p <- as.numeric(stats::predict(tm$model, type = "response"))

# Hand-derived dw/dalpha for static(1): -A*(1-p)/p * X_prop
X_prop <- tm$X_prop
dw_dalpha_hand <- sweep(X_prop, 1, -A * (1 - p) / p, "*")

# Numerical dw/dalpha via the make_weight_fn closure
wfn <- causatr:::make_weight_fn(tm, dat, static(1))
dw_dalpha_num <- numDeriv::jacobian(wfn, x = tm$alpha_hat)

max_diff <- max(abs(dw_dalpha_hand - dw_dalpha_num))
```

## Result

```{r}
#| label: verify-cross-deriv-result
#| echo: false
cat(sprintf(
  "Max |hand - numerical| for dw/dalpha under static(1): %.2e\n",
  max_diff
))
```

:::

## 4. Assembly: the full IF for $\hat\mu_a$ under IPW {#sec-assembly}

`compute_ipw_if_self_contained_one()` assembles the per-individual
influence function for one intervention. The three channels, following
the derivation in `vignette("variance-theory")` @eq-if-ipw, are:

$$
\text{IF}_i(\hat\mu_a) =
\underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_{\text{Channel 1}}
+ \underbrace{J \, A_{\beta\beta}^{-1} \, \psi_{\beta,i}}_{\text{MSM correction}}
- \underbrace{J \, A_{\beta\beta}^{-1} \, A_{\beta\alpha} \,
  A_{\alpha\alpha}^{-1} \, \psi_{\alpha,i}}_{\text{propensity correction}}.
$$

Note the **minus** sign on the propensity correction. This comes from
block inversion of the lower-triangular bread:

$$
A^{-1} =
\begin{pmatrix}
A_{\alpha\alpha}^{-1} & 0 \\
-A_{\beta\beta}^{-1} A_{\beta\alpha} A_{\alpha\alpha}^{-1} & A_{\beta\beta}^{-1}
\end{pmatrix}.
$$

::: {.callout-note}
### Sign convention

`vignette("variance-theory")` writes the IF with a **plus** sign on the
propensity correction because it uses Wooldridge's convention
$A_{\beta\alpha}^W = +(1/n)\sum \partial\psi_\beta/\partial\alpha$,
whereas the code uses the negative-Hessian convention
$A_{\beta\alpha} = -(1/n)\sum \partial\psi_\beta/\partial\alpha$. The two
differ by a sign on $A_{\beta\alpha}$, which flips the composition sign.
The numerical result is the same.
:::

### Implementation walkthrough

1. **MSM correction.** `prepare_model_if(msm_model)` + `apply_model_correction(msm_prep, J)` gives
   $A_{\beta\beta}^{-1}\psi_{\beta,i}$ projected onto the gradient $J$. This is the same primitive used by
   g-comp and matching --- one code path for all estimators.

2. **Cross-derivative.** `numDeriv::jacobian(phi_bar, alpha_hat)` gives
   $A_{\beta\alpha}$ numerically, using the `weight_fn` closure from @sec-closure.
   The per-step cost is $O(n \cdot q)$ for a $q$-parameter propensity model ---
   one matrix multiply per perturbation.

3. **Propensity correction.** Form the gradient
   $g_{\text{prop}} = A_{\beta\alpha}^T \cdot h_{\text{msm}}$ where
   $h_{\text{msm}} = n \cdot A_{\beta\beta}^{-1} J$ (the $n$-scaling
   accounts for `bread_inv()` returning the raw $(X^TWX)^{-1}$). Then
   `apply_model_correction(prop_prep, g_prop)` gives
   $g_{\text{prop}}^T A_{\alpha\alpha}^{-1} \psi_{\alpha,i}$ per
   individual --- the propensity channel.

4. **Assembly.** `Ch1_i + msm_res$correction - prop_res$correction`.

### When does the propensity correction matter?

For static binary ATE, the propensity correction **reduces** the SE
(Lunceford & Davidian 2004; Hirano et al. 2003). Using estimated rather
than known propensity scores is more efficient, so the naive sandwich
(which ignores propensity uncertainty) is conservative.

For non-static interventions the picture is richer:

- **Shift:** the per-intervention SE typically drops 5--8% from the
  propensity correction (the negative cross-term from the stacked
  M-estimation outweighs the added propensity-uncertainty term).
- **IPSI:** the per-intervention effect on the SE is small, but the
  **contrast** SE can drop dramatically ($\sim$90%) because the shared
  propensity model induces large positive covariance between the IPSI
  and natural-course marginal means. When you subtract two marginal means
  that share the same propensity noise, the noise largely cancels.

Both effects are verified in `test-ipw-variance-regression.R`, which
asserts that the full IF-based SE materially differs from the
$J V_\beta J^T$ shortcut (which omits the propensity correction).

## 5. Intervention × family compatibility matrix

Not every combination of intervention type and treatment family is
valid under density-ratio IPW. The compatibility matrix, enforced by
`check_intervention_family_compat()`, is:

| Intervention | Bernoulli | Gaussian | Categorical |
|---|---|---|---|
| `static()` | $\checkmark$ HT | $\times$ | $\checkmark$ HT |
| `shift()` | $\times$ | $\checkmark$ pushforward | $\times$ |
| `scale_by()` | $\times$ | $\checkmark$ pushforward | $\times$ |
| `threshold()` | $\times$ | $\times$ | $\times$ |
| `dynamic()` | $\checkmark$ HT | $\times$ | $\checkmark$ HT |
| `ipsi()` | $\checkmark$ Kennedy | $\times$ | $\times$ |

Each $\times$ has a specific reason rooted in the density-ratio
framework:

- **`static(v)` on continuous**: the HT indicator $\mathbb{1}\{A_i = v\}$
  is zero almost surely for a continuous treatment.
- **`shift()` / `scale_by()` on discrete**: factors have no additive or
  multiplicative structure.
- **`threshold()` on continuous**: pushforward is a mixed measure
  (point masses at the clamp boundaries), so there is no Lebesgue
  density for the ratio.
- **`dynamic()` on continuous**: a deterministic per-individual rule
  is a Dirac per individual. Same pushforward degeneracy as `threshold()`.
- **`ipsi()` on non-binary**: Kennedy's (2019) closed-form formula
  (@eq-ipsi-weight) is Bernoulli-specific.

All rejections point the user to the right alternative (typically
`estimator = "gcomp"` or a smooth MTP constructor).

## References

- Hernán MA, Robins JM (2025). *Causal Inference: What If.* Chapman &
  Hall/CRC. Ch. 12 (IP weighting).
- Hirano K, Imbens GW, Ridder G (2003). Efficient estimation of average
  treatment effects using the estimated propensity score. *Econometrica*
  71:1161--1189.
- Imbens GW (2004). Nonparametric estimation of average treatment
  effects under exogeneity: a review. *Review of Economics and
  Statistics* 86:4--29.
- Kennedy EH (2019). Nonparametric causal effects based on incremental
  propensity score interventions. *JASA* 114:645--656.
- Lunceford JK, Davidian M (2004). Stratification and weighting via the
  propensity score in estimation of causal treatment effects: a
  comparative study. *Statistics in Medicine* 23:2937--2960.
- Díaz I, Williams N, Hoffman KL, Schenck EJ (2023). Non-parametric
  causal effects based on longitudinal modified treatment policies.
  *JASA* 118:846--857.
- Haneuse S, Rotnitzky A (2013). Estimation of the effect of
  interventions that modify the received treatment. *Statistics in
  Medicine* 32:5260--5277.
