---
title: "Inverse probability weighting with causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Inverse probability weighting with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Inverse probability weighting (IPW) estimates causal effects by reweighting
the observed data so that the distribution of confounders is balanced across
treatment groups. causatr delegates weight estimation to
[WeightIt](https://ngreifer.github.io/WeightIt/) and fits a weighted marginal
structural model (MSM) via `WeightIt::glm_weightit()`, which provides
M-estimation sandwich SEs that account for weight estimation uncertainty.

This vignette demonstrates IPW in causatr using the NHEFS dataset from Hernán
& Robins (2025), covering every supported combination of treatment type,
outcome type, contrast scale, and inference method for time-fixed treatments.

**Note:** IPW currently supports only `static()` interventions because the
weights are estimated once under the observed treatment distribution.
For modified treatment policies (shift, scale, threshold, dynamic, MTP), use
g-computation (see `vignette("gcomp")`); self-contained IPW for non-static
interventions is planned for Phase 4.

**Note:** IPW and matching reject an `A:modifier` term inside `confounders`
up front — the saturated `Y ~ A` MSM has no slot for the interaction.
Use `estimator = "gcomp"` for heterogeneous treatment effects, or `by =
"modifier"` below for stratum-specific summaries of a homogeneous effect.

## Setup

```{r}
#| message: false
library(causatr)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71), ]

nhefs_complete$gained_weight <- as.integer(nhefs_complete$wt82_71 > 0)

nhefs_complete$sex <- factor(nhefs_complete$sex, levels = 0:1, labels = c("Male", "Female"))
```

## Binary treatment, continuous outcome

This replicates the IPW analysis from Chapter 12 of Hernán & Robins: the
average causal effect of quitting smoking (`qsmk`) on weight change
(`wt82_71`).

### ATE with sandwich SE

```{r}
fit_ipw <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATE"
)

res_ate_sw <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_ate_sw
```

The book reports an IPW estimate of ATE $\approx$ 3.4 kg (95% CI: 2.4, 4.5).

### ATE with bootstrap SE

```{r}
res_ate_bs <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_ate_bs
```

### ATT estimand

For IPW, the estimand is fixed at fitting time because it determines the
weights. To estimate the ATT, refit with `estimand = "ATT"`.

```{r}
fit_ipw_att <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATT"
)

res_att <- contrast(
  fit_ipw_att,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_att
```

### ATC estimand

```{r}
fit_ipw_atc <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATC"
)

res_atc <- contrast(
  fit_ipw_atc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_atc
```

### ATT with bootstrap SE

```{r}
res_att_bs <- contrast(
  fit_ipw_att,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_att_bs
```

### Extracting results programmatically

```{r}
coef(res_ate_sw)
confint(res_ate_sw)
vcov(res_ate_sw)
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Binary treatment, binary outcome

Using the `gained_weight` indicator as a binary outcome. With IPW, the MSM
is `Y ~ A` (a linear probability / logistic model), so `contrast()` computes
marginal risks and derives contrasts via the unified influence-function
variance engine (`variance_if()`), which for IPW delegates Channel 2 to
`prepare_propensity_if()` / `apply_propensity_correction()`. Branch A (the
default) uses WeightIt's own M-estimation shortcut
(`sandwich::estfun(asympt = TRUE)` + `sandwich::bread()`) so weight-estimation
uncertainty propagates into the reported SEs automatically. Non-Mparts
WeightIt methods (`gbm`, `super`, `bart`, `optweight`, `energy`, `npcbps`)
fall back to treating weights as fixed and emit a warning at fit time — use
`ci_method = "bootstrap"` for valid inference on those.

### Risk difference (sandwich)

```{r}
fit_ipw_bin <- causat(
  nhefs_complete,
  outcome = "gained_weight",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw"
)

res_rd <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_rd
```

### Risk difference (bootstrap)

```{r}
res_rd_bs <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rd_bs
```

### Risk ratio

```{r}
res_rr <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

### Odds ratio

```{r}
res_or <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "or",
  ci_method = "sandwich"
)
res_or
```

### Risk ratio (bootstrap)

```{r}
res_rr_bs <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rr_bs
```

## Parallel bootstrap

Bootstrap replicates can be parallelised by passing `parallel` and `ncpus`
straight through to `boot::boot()`:

```{r}
#| eval: false
res_par <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  ci_method = "bootstrap",
  n_boot = 200L,
  parallel = "multicore",  # or "snow" on Windows
  ncpus = 4L
)
```

These default to `getOption("boot.parallel", "no")` and
`getOption("boot.ncpus", 1L)`, so a session-wide
`options(boot.parallel = "multicore", boot.ncpus = 4L)` flips parallelism
on for every `contrast()` call without per-call plumbing.

## Categorical treatment (k > 2 levels)

WeightIt estimates a multinomial propensity score for categorical
treatments and causatr passes them through to `glm_weightit()` as-is.
The MSM is still `Y ~ A` — a per-level marginal mean — so `contrast()`
can compute any pairwise contrast against a user-chosen reference level.

```{r}
set.seed(42)
n <- 2000
L <- rnorm(n)
probs <- exp(cbind(0, 0.5 * L, -0.3 * L))
probs <- probs / rowSums(probs)
A <- factor(
  apply(probs, 1, function(p) sample(c("A", "B", "C"), 1, prob = p)),
  levels = c("A", "B", "C")
)
Y <- 2 + (A == "B") * 1.5 + (A == "C") * -0.8 + 0.5 * L + rnorm(n)
cat_df <- data.frame(Y = Y, A = A, L = L)

fit_ipw_cat <- causat(
  cat_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw"
)

res_ipw_cat <- contrast(
  fit_ipw_cat,
  interventions = list(
    arm_a = static("A"),
    arm_b = static("B"),
    arm_c = static("C")
  ),
  reference = "arm_a",
  type = "difference"
)
res_ipw_cat
```

## Continuous treatment (generalised propensity score)

For continuous treatments WeightIt fits a generalised propensity score
(the default is a density-ratio model). IPW still only supports `static()`
interventions — the weights are valid for the observed treatment
distribution, not for a shifted one. Use `static(value)` at specific
levels of the continuous variable and read off pairwise differences.

```{r}
set.seed(5)
n <- 2000
L <- rnorm(n)
A <- 1 + 0.5 * L + rnorm(n)
Y <- 1 + 2 * A + L + rnorm(n)
cont_df <- data.frame(Y = Y, A = A, L = L)

fit_ipw_cont <- causat(
  cont_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw"
)

res_ipw_cont <- contrast(
  fit_ipw_cont,
  interventions = list(a_low = static(0), a_mid = static(1), a_high = static(2)),
  reference = "a_low",
  type = "difference"
)
res_ipw_cont
```

Structural truth: `E[Y(a)] = 1 + 2a`, so `a_mid − a_low = 2` and
`a_high − a_low = 4`.

## External (survey) weights

Pre-computed survey weights are passed via `weights =`; causatr validates
them via `check_weights()` and **multiplies** them element-wise with the
WeightIt-estimated IPW weights. The combined weight then drives both the
`glm_weightit()` MSM fit and the influence-function variance via
`sandwich::estfun(asympt = TRUE)`.

```{r}
set.seed(6)
n <- 2000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- 2 + 3 * A + 1.5 * L + rnorm(n)
w <- runif(n, 0.5, 2.0)
sv_df <- data.frame(Y = Y, A = A, L = L)

fit_ipw_sv <- causat(
  sv_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  weights = w
)

res_ipw_sv <- contrast(
  fit_ipw_sv,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference"
)
res_ipw_sv
```

## Non-Mparts WeightIt methods and the bootstrap fallback

WeightIt's M-estimation shortcut (`glm_weightit`'s `Mparts`) is only wired
up for a subset of methods (`glm`, `cbps`, `ipt`, `ebal`). For anything
else (`gbm`, `super`, `bart`, `optweight`, `energy`, `npcbps`, `cfd`)
causatr emits a warning at fit time so you know the sandwich SE will treat the
weights as fixed:

```
Warning: WeightIt method 'gbm' does not implement the M-estimation
correction (no `Mparts` attribute). Sandwich SEs from `ci_method =
'sandwich'` will treat the weights as fixed and may underestimate the
variance. Use `ci_method = 'bootstrap'` for valid inference, or switch
to a method that supports Mparts (`glm`, `cbps`, `ipt`, `ebal`).
```

When you see this warning, switch to `ci_method = "bootstrap"` — the
bootstrap resamples the full pipeline (including the weight-estimation
step) and is always valid, regardless of whether Mparts is wired up.
See `FEATURE_COVERAGE_MATRIX.md` entry for "non-Mparts WeightIt method"
for the end-to-end smoke test.

## Forwarding WeightIt arguments (CBPS)

Any extra arguments passed to `causat()` flow through `...` straight
into `WeightIt::weightit()`. Because the causatr argument is now
`estimator` (not `method`), WeightIt's own `method = "..."` is free
and you can pick a non-default propensity score estimator. The example
below uses covariate-balancing propensity scores (Imai & Ratkovic, 2014),
which support Mparts, so sandwich SEs remain valid with no warning.

```{r}
fit_cbps <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATE",
  method = "cbps"  # forwarded to WeightIt::weightit()
)

res_cbps <- contrast(
  fit_cbps,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_cbps
```

## Comparing estimands

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for ATE, ATT, and ATC estimated by IPW."
est_df <- data.frame(
  estimand = c("ATE", "ATT", "ATC"),
  estimate = c(
    res_ate_sw$contrasts$estimate[1],
    res_att$contrasts$estimate[1],
    res_atc$contrasts$estimate[1]
  ),
  ci_lower = c(
    res_ate_sw$contrasts$ci_lower[1],
    res_att$contrasts$ci_lower[1],
    res_atc$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    res_ate_sw$contrasts$ci_upper[1],
    res_att$contrasts$ci_upper[1],
    res_atc$contrasts$ci_upper[1]
  )
)

tinyplot(
  estimate ~ estimand,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  xlab = "Estimand",
  ylab = "Effect on weight change (kg)",
  main = "IPW estimates by estimand"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Effect modification with `by`

::: {.callout-important}
## IPW effect modification is a Phase 8 feature

The `by` argument stratifies the causal effect estimates that fall out of
the fitted marginal structural model (MSM). causatr currently hardcodes a
*saturated* MSM, `Y ~ A`, for IPW — so predictions are flat per
intervention, and averaging them over any subset returns the same pooled
ATE. In other words, `by = "sex"` is a no-op for IPW until the MSM can
carry an interaction term (e.g. `Y ~ A + A:sex` or `Y ~ A * time`).

This is a causatr implementation limitation, not a methodological one.
Unified effect-modification support across IPW, matching, and ICE is
planned for Phase 8 (`PHASE_8_INTERACTIONS.md`). In the meantime, use
`estimator = "gcomp"` with an `A:modifier` term in `confounders` if you
need heterogeneous treatment effects.
:::

## Tidy and glance

causatr results work with the broom ecosystem via `tidy()` and `glance()`:

```{r}
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Forest plot

The `plot()` method produces a forest plot using the `forrest` package.

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot of the IPW-estimated ATE of quitting smoking on weight change."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_ate_sw)
```

## Diagnostics

After fitting an IPW model, use `diagnose()` to assess positivity and
covariate balance. The balance diagnostics use
[cobalt](https://ngreifer.github.io/cobalt/) to compute standardised mean
differences (SMD) before and after weighting.

```{r}
diag <- diagnose(fit_ipw)
diag
```

### Love plot

A Love plot visualises covariate balance. Covariates with absolute SMD below
the threshold (dashed line at 0.1) are considered well-balanced.

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-alt: "Love plot showing covariate balance before and after IPW weighting."
plot(diag)
```

### Weight distribution

Extreme weights can inflate variance. The weight summary shows the effective
sample size (ESS) — a large drop from the nominal sample size suggests
influential weights.

```{r}
diag$weights
```

## Summary of covered combinations

**Legend.** ✅ covered and truth-pinned in tests · 🟡 smoke test only ·
⛔ rejected with an informative error.

| Treatment | Outcome | Intervention | Estimand | Contrast | Inference | Weights | Status |
|---|---|---|---|---|---|---|---|
| Binary | Continuous | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Bootstrap | none | ✅ |
| Binary | Continuous | Static | ATT | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATC | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATT | Difference | Bootstrap | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich | survey | ✅ |
| Binary | Binary | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Binary | Static | ATE | Difference | Bootstrap | none | ✅ |
| Binary | Binary | Static | ATE | Ratio | Sandwich | none | ✅ |
| Binary | Binary | Static | ATE | OR | Sandwich | none | ✅ |
| Binary | Binary | Static | ATE | Ratio | Bootstrap | none | ✅ |
| Categorical (k>2) | Continuous | Static | ATE | Difference | Sandwich | none | ✅ |
| Continuous (static levels) | Continuous | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary (non-Mparts WeightIt) | Continuous | Static | ATE | Difference | Bootstrap | none | 🟡 |
| Any | Any | Any | Any (with `A:modifier`) | — | — | — | ⛔ Phase 8 |
| Multivariate / Longitudinal | — | — | — | — | — | — | ⛔ Phase 4 |
| Any | — | Dynamic / Shift / Scale / Threshold / IPSI | — | — | — | — | ⛔ Phase 4 |

See [`FEATURE_COVERAGE_MATRIX.md`](https://github.com/etverse/causatr/blob/main/FEATURE_COVERAGE_MATRIX.md)
for the authoritative coverage status of every method × treatment × outcome
× intervention × variance combination.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 12: IP weighting and marginal structural models.

Greifer N (2024). WeightIt: Weighting Methods for Covariate Balancing.
<https://ngreifer.github.io/WeightIt/>
