---
title: "Inverse probability weighting with causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Inverse probability weighting with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Inverse probability weighting (IPW) estimates causal effects by reweighting
the observed data so that the distribution of confounders is balanced across
treatment groups. causatr ships a self-contained density-ratio engine that
fits the conditional treatment density via the user's `propensity_model_fn`
(default `stats::glm`) and builds per-intervention Hájek weights for an
intercept-only marginal structural model (MSM) refit inside `contrast()`.

This vignette demonstrates IPW in causatr using the NHEFS dataset from Hernán
& Robins (2025), covering every supported combination of treatment type,
outcome type, contrast scale, and inference method for time-fixed treatments.

For non-static interventions (shift, scale, dynamic, IPSI), see
`vignette("interventions")`.

**Note:** IPW and matching reject an `A:modifier` term inside `confounders`
up front --- the intercept-only MSM has no slot for the interaction.
Use `estimator = "gcomp"` for heterogeneous treatment effects, or `by =
"modifier"` below for stratum-specific summaries of a homogeneous effect.

## Setup

The examples in this vignette use the **NHEFS** (National Health and
Nutrition Examination Survey --- Epidemiologic Follow-up Study) dataset,
the running example in Hernán & Robins (2025). The assumed causal
structure is:

- **Treatment ($A$):** `qsmk` --- quit smoking between 1971 and 1982 (binary: 0/1).
- **Outcome ($Y$):** `wt82_71` --- weight change in kg over the same period
  (continuous). A derived binary version, `gained_weight` ($\mathbf{1}\{Y > 0\}$),
  is used in the binary outcome sections.
- **Confounders ($L$):** sex, age, race, education, smoking intensity
  (cigarettes/day), years smoked, exercise, physical activity, and baseline
  weight. These jointly affect both the probability of quitting and the
  subsequent weight change.

Structural causal model (assumed):

$$
\begin{aligned}
  L &\sim F_L \\
  A &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(\alpha_0 + \alpha_L^\top L)\bigr) \\
  Y &= \beta_0 + \beta_A A + \beta_L^\top L + \beta_{AL}^\top (A \cdot L) + \varepsilon,
      \quad \varepsilon \sim N(0, \sigma^2)
\end{aligned}
$$

The confounders $L$ are common causes of $A$ and $Y$; adjusting for $L$
blocks all backdoor paths, identifying $\mathbb{E}[Y^a]$ via the
Horvitz--Thompson reweighting estimator. IPW models the treatment
mechanism $P(A \mid L)$ directly and reweights outcomes to remove
confounding, rather than modelling $E[Y \mid A, L]$ as g-computation does.

```{r}
#| message: false
library(causatr)
library(tinytable)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71) & !is.na(nhefs$education), ]

nhefs_complete$gained_weight <- as.integer(nhefs_complete$wt82_71 > 0)

nhefs_complete$sex <- factor(nhefs_complete$sex, levels = 0:1, labels = c("Male", "Female"))
```

## Binary treatment, continuous outcome

This replicates the IPW analysis from Chapter 12 of Hernán & Robins: the
average causal effect of quitting smoking (`qsmk`) on weight change
(`wt82_71`).

### ATE with sandwich SE

```{r}
fit_ipw <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATE"
)

res_ate_sw <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_ate_sw
```

The book reports an IPW estimate of ATE $\approx$ 3.4 kg (95% CI: 2.4, 4.5).

### ATE with bootstrap SE

```{r}
res_ate_bs <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_ate_bs
```

### ATT estimand

For IPW, the estimand is fixed at fitting time because it determines the
weights. To estimate the ATT, refit with `estimand = "ATT"`.

```{r}
fit_ipw_att <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATT"
)

res_att <- contrast(
  fit_ipw_att,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_att
```

### ATC estimand

```{r}
fit_ipw_atc <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw",
  estimand = "ATC"
)

res_atc <- contrast(
  fit_ipw_atc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_atc
```

### ATT with bootstrap SE

```{r}
res_att_bs <- contrast(
  fit_ipw_att,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_att_bs
```

### Extracting results programmatically

```{r}
coef(res_ate_sw)
confint(res_ate_sw)
vcov(res_ate_sw)
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Binary treatment, binary outcome

Using the `gained_weight` indicator as a binary outcome. The IPW engine
refits an intercept-only weighted MSM per intervention; `contrast()` computes
marginal risks and derives contrasts via the unified influence-function
variance engine (`variance_if()`), which for IPW uses
`compute_ipw_if_self_contained_one()` to build the per-individual IF on the
stacked propensity + MSM M-estimation system.

### Risk difference (sandwich)

```{r}
fit_ipw_bin <- causat(
  nhefs_complete,
  outcome = "gained_weight",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  estimator = "ipw"
)

res_rd <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_rd
```

### Risk difference (bootstrap)

```{r}
res_rd_bs <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rd_bs
```

### Risk ratio

```{r}
res_rr <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

### Odds ratio

```{r}
res_or <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "or",
  ci_method = "sandwich"
)
res_or
```

### Risk ratio (bootstrap)

```{r}
res_rr_bs <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rr_bs
```

## Parallel bootstrap

Bootstrap replicates can be parallelised by passing `parallel` and `ncpus`
straight through to `boot::boot()`:

```{r}
#| eval: false
res_par <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  ci_method = "bootstrap",
  n_boot = 200L,
  parallel = "multicore",  # or "snow" on Windows
  ncpus = 4L
)
```

These default to `getOption("boot.parallel", "no")` and
`getOption("boot.ncpus", 1L)`, so a session-wide
`options(boot.parallel = "multicore", boot.ncpus = 4L)` flips parallelism
on for every `contrast()` call without per-call plumbing.

## Dynamic intervention (binary treatment)

A `dynamic()` intervention assigns each individual to a treatment value
based on their covariates. Under IPW, dynamic rules on binary treatment
use Horwitz--Thompson indicator weights: $w_i = \mathbb{1}\{A_i = d(L_i)\} / f(d(L_i) \mid L_i)$.

```{r}
rule <- function(data, trt) {
  ifelse(data$smokeintensity > 20, 1L, 0L)
}

res_dyn <- contrast(
  fit_ipw,
  interventions = list(
    targeted = dynamic(rule),
    all_quit = static(1)
  ),
  reference = "all_quit",
  type = "difference"
)
res_dyn
```

Heavy smokers (> 20 cigs/day) are assigned to quit, others are left
untreated. The contrast with `static(1)` (all quit) tells us how
much is lost by targeting only heavy smokers.

```{r}
tt(tidy(res_dyn), digits = 3)
```

## IPSI --- incremental propensity score intervention

`ipsi(delta)` multiplies each individual's odds of treatment by `delta`
(Kennedy 2019). It is a smooth stochastic intervention that avoids
positivity violations --- no one is forced to a treatment value they
would never take.

```{r}
res_ipsi <- contrast(
  fit_ipw,
  interventions = list(
    double_odds = ipsi(2),
    observed = NULL
  ),
  reference = "observed",
  type = "difference"
)
res_ipsi
```

Varying `delta` traces out a dose--response curve:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "IPSI dose-response curve showing the effect of varying the odds multiplier delta."
deltas <- c(0.25, 0.5, 1, 2, 4)
ipsi_results <- lapply(deltas, function(d) {
  r <- contrast(
    fit_ipw,
    interventions = list(ipsi = ipsi(d), obs = NULL),
    reference = "obs",
    type = "difference"
  )
  data.frame(
    delta = d,
    estimate = r$contrasts$estimate[1],
    ci_lower = r$contrasts$ci_lower[1],
    ci_upper = r$contrasts$ci_upper[1]
  )
})
ipsi_df <- do.call(rbind, ipsi_results)

tinyplot(
  estimate ~ delta,
  data = ipsi_df,
  type = "pointrange",
  ymin = ipsi_df$ci_lower,
  ymax = ipsi_df$ci_upper,
  xlab = expression(delta ~ "(odds multiplier)"),
  ylab = "E[Y(delta)] - E[Y(obs)]",
  main = "IPSI dose-response"
)
abline(h = 0, lty = 2, col = "grey40")
```

At `delta = 1` the intervention equals the natural course, so the contrast
is zero. Values below 1 reduce the odds of treatment; values above 1
increase them.

## Categorical treatment (k > 2 levels)

The density-ratio engine auto-selects a multinomial propensity model (via
`nnet::multinom`) when the treatment is a factor or character column. The
per-intervention MSM is intercept-only, so `contrast()` computes any pairwise
difference against a user-chosen reference level.

**DGM.** A single confounder $L$ influences both the three-level categorical
treatment $A \in \{A, B, C\}$ and the outcome $Y$. Treatment assignment
follows a softmax (multinomial logistic) model with category A as the
reference level. The outcome is linear in the treatment indicators and $L$.

$$
\begin{aligned}
  L &\sim N(0, 1) \\
  P(A = k \mid L) &= \frac{\exp(\gamma_k L)}{\sum_{j} \exp(\gamma_j L)},
    \quad \gamma_A = 0,\; \gamma_B = 0.5,\; \gamma_C = -0.3 \\
  Y &= 2 + 1.5 \cdot \mathbf{1}\{A = B\} - 0.8 \cdot \mathbf{1}\{A = C\}
       + 0.5\, L + \varepsilon,
    \quad \varepsilon \sim N(0, 1)
\end{aligned}
$$

True contrasts (vs. reference A): $\mathbb{E}[Y^B] - \mathbb{E}[Y^A] = 1.5$
and $\mathbb{E}[Y^C] - \mathbb{E}[Y^A] = -0.8$.

```{r}
set.seed(42)
n <- 2000
L <- rnorm(n)
probs <- exp(cbind(0, 0.5 * L, -0.3 * L))
probs <- probs / rowSums(probs)
A <- factor(
  apply(probs, 1, function(p) sample(c("A", "B", "C"), 1, prob = p)),
  levels = c("A", "B", "C")
)
Y <- 2 + (A == "B") * 1.5 + (A == "C") * -0.8 + 0.5 * L + rnorm(n)
cat_df <- data.frame(Y = Y, A = A, L = L)

fit_ipw_cat <- causat(
  cat_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw"
)

res_ipw_cat <- contrast(
  fit_ipw_cat,
  interventions = list(
    arm_a = static("A"),
    arm_b = static("B"),
    arm_c = static("C")
  ),
  reference = "arm_a",
  type = "difference"
)
res_ipw_cat
```

## Continuous treatment (shift / scale MTPs)

For continuous treatments the engine fits a Gaussian treatment density model
and uses density-ratio weights for modified treatment policies. `static()`
interventions on continuous treatments are rejected (nobody is observed
exactly at the target value, so the HT indicator is zero almost surely).
Use `shift()` or `scale_by()` instead:

**DGM.** A single confounder $L$ drives both a continuous (Gaussian)
treatment $A$ and a continuous outcome $Y$. The treatment density
is $A \mid L \sim N(1 + 0.5\, L,\; 1)$ and the outcome is linear in
$A$ and $L$ with additive noise.

$$
\begin{aligned}
  L &\sim N(0, 1) \\
  A \mid L &\sim N(1 + 0.5\, L,\; 1) \\
  Y &= 1 + 2\, A + L + \varepsilon, \quad \varepsilon \sim N(0, 1)
\end{aligned}
$$

Because $\partial Y / \partial A = 2$, a unit additive shift
$A \mapsto A + 1$ has true causal effect
$\mathbb{E}[Y(A + 1)] - \mathbb{E}[Y(A)] = 2$.

```{r}
set.seed(5)
n <- 2000
L <- rnorm(n)
A <- 1 + 0.5 * L + rnorm(n)
Y <- 1 + 2 * A + L + rnorm(n)
cont_df <- data.frame(Y = Y, A = A, L = L)

fit_ipw_cont <- causat(
  cont_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw"
)

res_shift <- contrast(
  fit_ipw_cont,
  interventions = list(up1 = shift(1), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_shift

res_scale <- contrast(
  fit_ipw_cont,
  interventions = list(doubled = scale_by(2), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_scale
```

Structural truth: `E[Y(A+1)] - E[Y(A)] = 2` (the coefficient on `A`).

See `vignette("interventions")` for a full tour of shift, scale, dynamic,
and IPSI interventions with side-by-side g-comp and IPW examples.

## Count treatment (Poisson / negative binomial)

Integer-valued treatments --- dose levels, event counts, visit counts --- default
to a Gaussian treatment density, which works when the support is dense enough
that the Normal approximation is harmless (age in years, cigarettes/day). For
sparse count data (0--5 doses), the Gaussian pdf evaluates between integers that
carry zero probability, producing badly calibrated density ratios.

Pass `propensity_family = "poisson"` or `propensity_family = "negbin"` to opt
into a Poisson or negative binomial treatment density model:

**DGM.** A confounder $L$ affects both a Poisson-distributed count
treatment $A$ and a continuous outcome $Y$. The treatment is generated
from a log-linear Poisson model and the outcome is linear in $A$ and $L$.

$$
\begin{aligned}
  L &\sim N(0, 1) \\
  A \mid L &\sim \text{Poisson}\!\bigl(\exp(0.5\, L)\bigr) \\
  Y &= 2 + 1.5\, A + L + \varepsilon, \quad \varepsilon \sim N(0, 1)
\end{aligned}
$$

Because $\partial Y / \partial A = 1.5$, a unit shift $A \mapsto A + 1$
has true causal effect $\mathbb{E}[Y(A + 1)] - \mathbb{E}[Y(A)] = 1.5$.

```{r}
set.seed(6)
n <- 3000
L <- rnorm(n)
A <- rpois(n, exp(0.5 * L))
Y <- 2 + 1.5 * A + L + rnorm(n)
count_df <- data.frame(Y = Y, A = A, L = L)

fit_pois <- causat(
  count_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  propensity_family = "poisson"
)

res_pois <- contrast(
  fit_pois,
  interventions = list(plus1 = shift(1), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_pois
```

Structural truth: `E[Y(A+1)] - E[Y(A)] = 1.5`.

For negative binomial (which nests Poisson when theta is large), pass
`propensity_family = "negbin"`. `MASS::glm.nb` is auto-selected as the
propensity fitter:

```{r}
#| eval: !expr requireNamespace("MASS", quietly = TRUE)
fit_nb <- causat(
  count_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  propensity_family = "negbin"
)

res_nb <- contrast(
  fit_nb,
  interventions = list(plus1 = shift(1), observed = NULL),
  reference = "observed",
  type = "difference"
)
res_nb
```

`scale_by()` also works on count treatments, but with a stricter constraint:
the density ratio evaluates at `A / factor`, so `A / factor` must be a
non-negative integer for *every* observed value. In practice, this means
`scale_by()` on counts is only feasible when all observed counts are multiples
of the factor.

::: {.callout-important}
## Integer constraints on count interventions

Only `shift()` with an integer delta and `scale_by()` where `A / factor` is
integer for every observation are allowed. `static()`, `dynamic()`,
`threshold()`, and `ipsi()` are rejected because the count pmf is zero at
non-integer arguments and the HT indicator / IPSI closed form require
binary structure. Use `estimator = "gcomp"` for those intervention types on
count data.
:::

## GAM propensity model

Pass `propensity_model_fn = mgcv::gam` to use a GAM for the treatment
density model while keeping the default `stats::glm` for the outcome MSM:

```{r}
#| eval: !expr requireNamespace("mgcv", quietly = TRUE)
fit_gam_ps <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + s(age) + race + factor(education) +
    s(smokeintensity) + s(smokeyrs) + factor(exercise) +
    factor(active) + s(wt71),
  estimator = "ipw",
  propensity_model_fn = mgcv::gam
)

res_gam_ps <- contrast(
  fit_gam_ps,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference"
)
res_gam_ps
```

## External (survey) weights

Pre-computed survey weights are passed via `weights =`; causatr validates
them via `check_weights()` and **multiplies** them element-wise with the
estimated density-ratio weights. The combined weight drives both the
per-intervention MSM fit and the influence-function sandwich variance.

**DGM.** A simple confounded binary treatment with a continuous outcome.
Random survey weights (independent of the data-generating process) are
attached to illustrate external weight handling.

$$
\begin{aligned}
  L &\sim N(0, 1) \\
  A \mid L &\sim \text{Bernoulli}\!\bigl(\text{logit}^{-1}(0.5\, L)\bigr) \\
  Y &= 2 + 3\, A + 1.5\, L + \varepsilon, \quad \varepsilon \sim N(0, 1) \\
  w_i &\stackrel{\text{iid}}{\sim} \text{Uniform}(0.5,\; 2.0)
\end{aligned}
$$

True ATE: $\mathbb{E}[Y^1] - \mathbb{E}[Y^0] = 3$.

```{r}
set.seed(6)
n <- 2000
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- 2 + 3 * A + 1.5 * L + rnorm(n)
w <- runif(n, 0.5, 2.0)
sv_df <- data.frame(Y = Y, A = A, L = L)

fit_ipw_sv <- causat(
  sv_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  weights = w
)

res_ipw_sv <- contrast(
  fit_ipw_sv,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference"
)
res_ipw_sv
```

## Comparing estimands

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for ATE, ATT, and ATC estimated by IPW."
est_df <- data.frame(
  estimand = c("ATE", "ATT", "ATC"),
  estimate = c(
    res_ate_sw$contrasts$estimate[1],
    res_att$contrasts$estimate[1],
    res_atc$contrasts$estimate[1]
  ),
  ci_lower = c(
    res_ate_sw$contrasts$ci_lower[1],
    res_att$contrasts$ci_lower[1],
    res_atc$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    res_ate_sw$contrasts$ci_upper[1],
    res_att$contrasts$ci_upper[1],
    res_atc$contrasts$ci_upper[1]
  )
)

tinyplot(
  estimate ~ estimand,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  xlab = "Estimand",
  ylab = "Effect on weight change (kg)",
  main = "IPW estimates by estimand"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Effect modification with `by`

::: {.callout-important}
## IPW effect modification is a Phase 6 feature

The `by` argument stratifies the causal effect estimates that fall out of
the fitted marginal structural model (MSM). causatr currently uses an
intercept-only MSM (`Y ~ 1`) for IPW --- so predictions are flat per
intervention, and averaging them over any subset returns the same pooled
estimate. In other words, `by = "sex"` is a no-op for IPW until the MSM can
carry an interaction term (e.g. `Y ~ A + A:sex`).

This is a causatr implementation limitation, not a methodological one.
Unified effect-modification support across IPW, matching, and ICE is
planned. In the meantime, use `estimator = "gcomp"` with an `A:modifier`
term in `confounders` if you need heterogeneous treatment effects.
:::

## Tidy and glance

causatr results work with the broom ecosystem via `tidy()` and `glance()`:

```{r}
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Forest plot

The `plot()` method produces a forest plot using the `forrest` package.

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot of the IPW-estimated ATE of quitting smoking on weight change."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_ate_sw)
```

## Diagnostics

After fitting an IPW model, use `diagnose()` to assess positivity and
covariate balance. The balance diagnostics use
[cobalt](https://ngreifer.github.io/cobalt/) to compute standardised mean
differences (SMD) before and after weighting.

```{r}
diag <- diagnose(fit_ipw)
diag
```

### Love plot

A Love plot visualises covariate balance. Covariates with absolute SMD below
the threshold (dashed line at 0.1) are considered well-balanced.

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-alt: "Love plot showing covariate balance before and after IPW weighting."
plot(diag)
```

### Weight distribution

Extreme weights can inflate variance. The weight summary shows the effective
sample size (ESS) --- a large drop from the nominal sample size suggests
influential weights.

```{r}
diag$weights
```

## Summary of covered combinations

**Legend.** ✅ covered and truth-pinned in tests · 🟡 smoke test only ·
⛔ rejected with an informative error.

| Treatment | Outcome | Intervention | Estimand | Contrast | Inference | Weights | Status |
|---|---|---|---|---|---|---|---|
| Binary | Continuous | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Bootstrap | none | ✅ |
| Binary | Continuous | Static | ATT | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATC | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATT | Difference | Bootstrap | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich | survey | ✅ |
| Binary | Binary | Static | ATE | Difference | Sandwich | none | ✅ |
| Binary | Binary | Static | ATE | Difference | Bootstrap | none | ✅ |
| Binary | Binary | Static | ATE | Ratio | Sandwich | none | ✅ |
| Binary | Binary | Static | ATE | OR | Sandwich | none | ✅ |
| Binary | Binary | Static | ATE | Ratio | Bootstrap | none | ✅ |
| Binary | Continuous | Dynamic | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous | IPSI($\delta$) | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous | Shift | ATE | Difference | Sandwich | none | ✅ |
| Continuous | Continuous | Scale | ATE | Difference | Sandwich | none | ✅ |
| Categorical (k>2) | Continuous | Static | ATE | Difference | Sandwich | none | ✅ |
| Categorical (k>2) | Continuous | Dynamic | ATE | Difference | Sandwich | none | 🟡 |
| Count (Poisson) | Continuous | Shift | ATE | Difference | Sandwich | none | ✅ |
| Count (NB) | Continuous | Shift | ATE | Difference | Sandwich | none | ✅ |
| Any | --- | Threshold | --- | --- | --- | --- | ⛔ (use gcomp) |
| Continuous | --- | Static | --- | --- | --- | --- | ⛔ (use shift/scale) |
| Continuous | --- | Dynamic | --- | --- | --- | --- | ⛔ (use shift/scale) |
| Any | --- | Any (with `A:modifier`) | --- | --- | --- | --- | ⛔ (Phase 6) |
| Multivariate | --- | --- | --- | --- | --- | --- | ⛔ (Phase 7) |
| Longitudinal | --- | --- | --- | --- | --- | --- | ⛔ (use ICE) |

See [`FEATURE_COVERAGE_MATRIX.md`](https://github.com/etverse/causatr/blob/main/FEATURE_COVERAGE_MATRIX.md)
for the authoritative coverage status of every method × treatment × outcome
× intervention × variance combination.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 12: IP weighting and marginal structural models.

Kennedy EH (2019). Nonparametric causal effects based on incremental
propensity score interventions. *Journal of the American Statistical
Association* 114:645--656.

Díaz I, Williams N, Hoffman KL, Schenck EJ (2023). Non-parametric causal
effects based on longitudinal modified treatment policies. *Journal of the
American Statistical Association* 118:846--857.
