---
title: "Inverse probability weighting with causatr"
vignette: >
  %\VignetteIndexEntry{Inverse probability weighting with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Inverse probability weighting (IPW) estimates causal effects by reweighting
the observed data so that the distribution of confounders is balanced across
treatment groups. causatr delegates weight estimation to
[WeightIt](https://ngreifer.github.io/WeightIt/) and fits a weighted marginal
structural model (MSM) via `WeightIt::glm_weightit()`, which provides
M-estimation sandwich SEs that account for weight estimation uncertainty.

This vignette demonstrates IPW in causatr using the NHEFS dataset from Hernán
& Robins (2025), covering every supported combination of treatment type,
outcome type, contrast scale, and inference method for time-fixed treatments.

**Note:** IPW currently supports only `static()` interventions. For modified
treatment policies (shift, scale, threshold, dynamic), use g-computation
(see `vignette("gcomp")`).

## Setup

```{r}
#| message: false
library(causatr)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71), ]

nhefs_complete$gained_weight <- as.integer(nhefs_complete$wt82_71 > 0)

nhefs_complete$sex <- factor(nhefs_complete$sex, levels = 0:1, labels = c("Male", "Female"))
```

## Binary treatment, continuous outcome

This replicates the IPW analysis from Chapter 12 of Hernán & Robins: the
average causal effect of quitting smoking (`qsmk`) on weight change
(`wt82_71`).

### ATE with sandwich SE

```{r}
fit_ipw <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "ipw",
  estimand = "ATE"
)

res_ate_sw <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_ate_sw
```

The book reports an IPW estimate of ATE $\approx$ 3.4 kg (95% CI: 2.4, 4.5).

### ATE with bootstrap SE

```{r}
res_ate_bs <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_ate_bs
```

### ATT estimand

For IPW, the estimand is fixed at fitting time because it determines the
weights. To estimate the ATT, refit with `estimand = "ATT"`.

```{r}
fit_ipw_att <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "ipw",
  estimand = "ATT"
)

res_att <- contrast(
  fit_ipw_att,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_att
```

### ATC estimand

```{r}
fit_ipw_atc <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "ipw",
  estimand = "ATC"
)

res_atc <- contrast(
  fit_ipw_atc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_atc
```

### ATT with bootstrap SE

```{r}
res_att_bs <- contrast(
  fit_ipw_att,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_att_bs
```

### Extracting results programmatically

```{r}
coef(res_ate_sw)
confint(res_ate_sw)
vcov(res_ate_sw)
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Binary treatment, binary outcome

Using the `gained_weight` indicator as a binary outcome. With IPW, the MSM
is `Y ~ A` (a linear probability model), so `contrast()` computes marginal
risks and derives contrasts via the delta method.

### Risk difference (sandwich)

```{r}
fit_ipw_bin <- causat(
  nhefs_complete,
  outcome = "gained_weight",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "ipw"
)

res_rd <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_rd
```

### Risk difference (bootstrap)

```{r}
res_rd_bs <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rd_bs
```

### Risk ratio

```{r}
res_rr <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

### Odds ratio

```{r}
res_or <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "or",
  ci_method = "sandwich"
)
res_or
```

### Risk ratio (bootstrap)

```{r}
res_rr_bs <- contrast(
  fit_ipw_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_rr_bs
```

## Comparing estimands

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for ATE, ATT, and ATC estimated by IPW."
est_df <- data.frame(
  estimand = c("ATE", "ATT", "ATC"),
  estimate = c(
    res_ate_sw$contrasts$estimate[1],
    res_att$contrasts$estimate[1],
    res_atc$contrasts$estimate[1]
  ),
  ci_lower = c(
    res_ate_sw$contrasts$ci_lower[1],
    res_att$contrasts$ci_lower[1],
    res_atc$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    res_ate_sw$contrasts$ci_upper[1],
    res_att$contrasts$ci_upper[1],
    res_atc$contrasts$ci_upper[1]
  )
)

tinyplot(
  estimate ~ estimand,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  xlab = "Estimand",
  ylab = "Effect on weight change (kg)",
  main = "IPW estimates by estimand"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Effect modification with `by`

The `by` argument stratifies causal effect estimates by levels of a variable.
Here we examine whether the IPW-estimated effect of quitting smoking differs
by sex.

```{r}
res_by_sex <- contrast(
  fit_ipw,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich",
  by = "sex"
)
res_by_sex
```

## Tidy and glance

causatr results work with the broom ecosystem via `tidy()` and `glance()`:

```{r}
tidy(res_ate_sw)
glance(res_ate_sw)
```

## Forest plot

The `plot()` method produces a forest plot using the `forrest` package. Forest
plots are most useful when displaying multiple estimates, such as effect
modification results:

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot of the IPW-estimated effect of quitting smoking on weight change, stratified by sex."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_by_sex)
```

## Diagnostics

After fitting an IPW model, use `diagnose()` to assess positivity and
covariate balance. The balance diagnostics use
[cobalt](https://ngreifer.github.io/cobalt/) to compute standardised mean
differences (SMD) before and after weighting.

```{r}
diag <- diagnose(fit_ipw)
diag
```

### Love plot

A Love plot visualises covariate balance. Covariates with absolute SMD below
the threshold (dashed line at 0.1) are considered well-balanced.

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-alt: "Love plot showing covariate balance before and after IPW weighting."
plot(diag)
```

### Weight distribution

Extreme weights can inflate variance. The weight summary shows the effective
sample size (ESS) — a large drop from the nominal sample size suggests
influential weights.

```{r}
diag$weights
```

## Summary of covered combinations

| Treatment | Outcome | Contrast | Inference | Estimand |
|---|---|---|---|---|
| Binary | Continuous | Difference | Sandwich | ATE |
| Binary | Continuous | Difference | Bootstrap | ATE |
| Binary | Continuous | Difference | Sandwich | ATT |
| Binary | Continuous | Difference | Sandwich | ATC |
| Binary | Continuous | Difference | Bootstrap | ATT |
| Binary | Binary | Difference | Sandwich | ATE |
| Binary | Binary | Difference | Bootstrap | ATE |
| Binary | Binary | Ratio | Sandwich | ATE |
| Binary | Binary | OR | Sandwich | ATE |
| Binary | Binary | Ratio | Bootstrap | ATE |

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 12: IP weighting and marginal structural models.

Greifer N (2024). WeightIt: Weighting Methods for Covariate Balancing.
<https://ngreifer.github.io/WeightIt/>
