---
title: "Longitudinal treatments: ICE g-computation"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Longitudinal treatments: ICE g-computation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When treatment and confounders vary over time, standard regression methods give
biased estimates even when adjusted for all confounders. G-methods solve this.
causatr implements ICE (iterated conditional expectation) g-computation
[@zivich2024ice], which requires models for the outcome only --- no models for
time-varying confounders are needed.

::: {.callout-tip}
## Doubly-robust longitudinal estimation

For doubly-robust protection, use `estimator = "aipw"`. ICE-AIPW
[@bang2005doubly] augments each backward regression step with a propensity-weighted
residual correction, giving a consistent estimator if *either* the outcome
regression *or* the propensity model at each time point is correctly specified.
The same `causat()` + `contrast()` API applies; sandwich and bootstrap
variance are both supported.

The `"sandwich"` variance is the full stacked-estimating-equation
M-estimation sandwich (per-period propensity scores + per-step outcome
scores + the marginal-mean equation), so it is correct on **both balanced and
unbalanced** panels — including monotone dropout or a censoring row-filter. It
covers GLM-family and `MASS::glm.nb` outcome / propensity models and
multinomial (categorical-treatment) propensities. **Limitation:** when an
outcome or propensity nuisance is fit with a *penalised* / non-likelihood
learner (`mgcv::gam`, `betareg::betareg`), the analytic sandwich is not
available — `ci_method = "sandwich"` raises a clear error and you should use
`ci_method = "bootstrap"`, which is valid for every configuration.
:::

This vignette demonstrates ICE g-computation in causatr, covering:

- The treatment-confounder feedback problem
- ICE fitting and contrasting
- Sandwich and bootstrap inference
- Static, dynamic, and modified treatment policy interventions
- The `history` parameter (Markov order)
- Binary outcomes
- Comparison with naive regression

## Setup

```{r}
#| message: false
library(causatr)
library(tinytable)
library(tinyplot)
```

## The problem: treatment-confounder feedback

When $A_0$ affects a time-varying confounder $L_1$, and $L_1$ in turn affects
both $A_1$ and $Y$ ($A_0 \to L_1 \to A_1$ and $L_1 \to Y$), ordinary
regression cannot recover the marginal counterfactual mean
$\mathbb{E}[Y^{a_0, a_1}]$:

- **Adjusting** for $L_1$ blocks part of the causal path from $A_0$ to $Y$
  (and, depending on the DAG, can open a collider path through unmeasured
  common causes of $L_1$ and $Y$).
- **Not adjusting** leaves $L_1$-confounding of the $A_1 \to Y$ arrow.
- **ICE g-computation** handles both correctly by iterating conditional
  expectations and marginalising over the $L_1$ distribution induced by each
  intervention.

## Table 20.1: the null paradox

We reconstruct the dataset from Table 20.1 of @hernan_whatif, scaled
to 3,200 individuals observed at 2 time points. (The original table has
32,000; we use 1/10th for faster vignette rendering.) The DGP has a very
specific property: **every joint regime $(a_0, a_1)$ has the same marginal
counterfactual mean $\mathbb{E}[Y^{a_0, a_1}] = 60$**, so the true causal
contrast between any two regimes is exactly zero — the "null is true".
Crucially, the null holds only marginally: within strata of $L_1$, the
*conditional* effect of $A_0$ is $-8$. Treatment-confounder feedback is
exactly the mechanism that lets a non-zero conditional effect average out
to a zero marginal effect.

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k \in \{0, 1\}$,
binary time-varying confounder $L_1 \in \{0, 1\}$ (affected by $A_0$), and
continuous outcome $Y$ measured at $k = 1$. The causal structure is
$A_0 \to L_1 \to A_1$, $A_0 \to L_1 \to Y$, and $A_0 \to Y$ (direct).
The table is constructed so that the conditional effect of $A_0$ given $L_1$
is $-8$, but this conditional effect averages to exactly zero marginally
because $A_0$ shifts the $L_1$ distribution in a perfectly offsetting way.
The structural equations implied by the table are:

$$
\begin{aligned}
A_0 &\sim \text{Bernoulli}(p_{A_0}), \\
L_1 \mid A_0 &\sim \text{Bernoulli}\bigl(P(L_1 = 1 \mid A_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(P(A_1 = 1 \mid L_1)\bigr), \\
Y &= 84 - 8 \cdot A_0 - 32 \cdot L_1.
\end{aligned}
$$

The outcome is deterministic given $(A_0, L_1)$ and does not depend on $A_1$
directly. The key property: $P(L_1 = 1 \mid A_0 = 1) > P(L_1 = 1 \mid A_0 = 0)$,
so $A_0$ increases $L_1$, and since $L_1$ has a strong negative effect on $Y$,
the indirect harm through $L_1$ exactly cancels the direct benefit.
The known true value is $\mathbb{E}[Y^{a_0, a_1}] = 60$ for all four joint
regimes, so the ATE between any two regimes is exactly 0.

```{r}
groups <- data.frame(
  A0 = c(0, 0, 0, 0, 1, 1, 1, 1),
  L1 = c(0, 0, 1, 1, 0, 0, 1, 1),
  A1 = c(0, 1, 0, 1, 0, 1, 0, 1),
  Y  = c(84, 84, 52, 52, 76, 76, 44, 44),
  N  = c(240, 160, 240, 960, 480, 320, 160, 640)
)
rows <- lapply(seq_len(nrow(groups)), function(i) {
  g <- groups[i, ]
  off <- sum(groups$N[seq_len(i - 1)])
  data.frame(
    id = seq_len(g$N) + off,
    A0 = g$A0, L1 = g$L1, A1 = g$A1, Y = g$Y
  )
})
wide <- do.call(rbind, rows)

t0 <- data.frame(
  id = wide$id, time = 0L, A = wide$A0, L = NA_real_, Y = NA_real_
)
t1 <- data.frame(
  id = wide$id, time = 1L, A = wide$A1, L = wide$L1, Y = wide$Y
)
long <- rbind(t0, t1)
head(long, 4)
```

The data is in **person-period (long) format**: one row per individual per
time point. `L` is a time-varying confounder (only measured at time 1).

## Demonstrating bias from naive methods

@hernan_whatif, Ch. 20, uses this table to make a pointed argument:
*neither* of the two conventional regression strategies recovers the null.
Truth is $\mathbb{E}[Y^{1,1}] - \mathbb{E}[Y^{0,0}] = 0$, and in fact all
four joint regimes have the same marginal counterfactual mean of $60$.

**Strategy 1 — unadjusted.** Drop $L_1$ and hope randomisation at each
time point takes care of confounding:

```{r}
wide_sub <- wide[!is.na(wide$Y), ]
naive_unadj <- lm(Y ~ A0 + A1, data = wide_sub)
round(coef(naive_unadj), 3)
```

Both coefficients move away from zero ($A_0 \approx -1.24$,
$A_1 \approx -12.44$). The $A_1$ estimate is especially biased:
$L_1$ confounds $A_1 \to Y$, and without adjustment that confounding
leaks directly into the coefficient.

**Strategy 2 — adjust for $L_1$.** The standard epidemiological reflex:

```{r}
naive_fit   <- lm(Y ~ A0 + A1 + L1, data = wide_sub)
naive_coefs <- coef(naive_fit)
round(naive_coefs, 3)
```

Now $A_1 = 0$ (correct, because in this DGP $Y$ is a deterministic
function of $A_0$ and $L_1$), but $A_0 = -8$ — the *conditional* effect of
$A_0$ within strata of $L_1$. That is **not** the causal contrast we want.
Adjusting for $L_1$ blocks the $A_0 \to L_1 \to Y$ path and leaves only
the direct-effect residue, so $A_0$ looks like a protective effect when
in fact the marginal causal effect is zero.

**The takeaway.** One naive strategy biases $A_1$; the other biases $A_0$.
There is no regression specification on this table that jointly returns
$A_0 = 0$ and $A_1 = 0$. This is the *null paradox* that motivates
g-methods: the only way to recover the null is to marginalise over the
intervention-specific distribution of $L_1$, which is exactly what ICE
g-computation does below.

## ICE g-computation: fitting

With `causat()`, specify `id` and `time` to trigger longitudinal mode. Baseline
confounders go in `confounders`, time-varying confounders in `confounders_tv`.
For finer control, use `confounders_tv_outcome` and `confounders_tv_treatment`
to specify different time-varying covariates for the outcome and treatment
models.

```{r}
fit <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm
)
fit
```

No models are fitted at this stage. ICE defers model fitting to `contrast()`
because the sequential outcome models depend on the intervention being
evaluated.

## ICE g-computation: contrasts with sandwich SE

Compare "always treat" (A = 1 at every time) vs "never treat" (A = 0 at every
time):

```{r}
result <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
result
```

ICE correctly recovers **E[Y^{always}] = E[Y^{never}] = 60** and **ATE = 0**,
as expected from the data generating process.

### Visualising the result

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Point estimates and confidence intervals comparing naive regression to ICE g-computation. Naive estimates are biased away from zero while ICE correctly recovers ATE = 0."
naive_ci <- confint(naive_fit)
est_df <- data.frame(
  method = c("Naive (A0)", "Naive (A1)", "ICE"),
  estimate = c(
    naive_coefs["A0"], naive_coefs["A1"], result$contrasts$estimate[1]
  ),
  ci_lower = c(
    naive_ci["A0", 1], naive_ci["A1", 1], result$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    naive_ci["A0", 2], naive_ci["A1", 2], result$contrasts$ci_upper[1]
  )
)
tinyplot(
  estimate ~ method,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  ylab = "Estimated treatment effect",
  xlab = "",
  main = "ICE recovers the null effect"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Sandwich SE: stacked estimating equations

The default `ci_method = "sandwich"` uses the empirical sandwich estimator for
stacked estimating equations [@zivich2024ice]. This accounts for
uncertainty from all K sequential outcome models, not just the final model.

```{r}
result$estimates[, c("intervention", "estimate", "se")]
```

The standard errors are finite and positive, confirming that the sandwich
variance estimator is working correctly even with a large sample.

## Bootstrap SE

For robustness or when using flexible models (e.g., GAMs), bootstrap resamples
individuals (all their time points together) and re-runs the full ICE procedure.

```{r}
result_boot <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 20L
)
result_boot
```

The `parallel` and `ncpus` arguments enable parallel bootstrap replicates for
larger datasets. Four backends are supported: `"no"` (sequential), `"multicore"`
(Unix fork, default on most Linux/macOS setups), `"snow"` (cross-platform
socket cluster, both via `boot::boot()`), and `"future"` (any active
`future::plan()` — local multisession, remote `plan(cluster, ...)`, or
`future.batchtools` for HPC job arrays). `"future"` bypasses `boot::boot()` so
the plan itself owns the scheduling; `ncpus` is ignored in that case.

```{r}
#| eval: false
# boot::boot parallel (fork / socket)
result_par <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "bootstrap",
  n_boot = 20L,
  parallel = "multicore",
  ncpus = 4L
)

# future backend (any future::plan())
library(future)
plan(multisession, workers = 4)
result_future <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "bootstrap",
  n_boot = 500L,
  parallel = "future"
)
```

### Comparing sandwich and bootstrap SEs

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Point estimates and confidence intervals from sandwich and bootstrap inference for the ICE ATE estimate, showing close agreement."
se_df <- data.frame(
  method = c("Sandwich", "Bootstrap"),
  estimate = c(result$contrasts$estimate[1], result_boot$contrasts$estimate[1]),
  ci_lower = c(result$contrasts$ci_lower[1], result_boot$contrasts$ci_lower[1]),
  ci_upper = c(result$contrasts$ci_upper[1], result_boot$contrasts$ci_upper[1])
)
tinyplot(
  estimate ~ method,
  data = se_df,
  type = "pointrange",
  ymin = se_df$ci_lower,
  ymax = se_df$ci_upper,
  ylab = "ATE (always vs never)",
  xlab = "Inference method",
  main = "Sandwich vs bootstrap inference"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Dynamic interventions

ICE supports dynamic treatment regimes where treatment depends on time-varying
covariates. Here, individuals receive treatment only when L = 1 (the
confounder is present):

```{r}
result_dyn <- contrast(
  fit,
  interventions = list(
    adaptive = dynamic(
      \(data, trt) ifelse(!is.na(data$L) & data$L > 0, 1L, 0L)
    ),
    always   = static(1),
    never    = static(0)
  ),
  reference = "never",
  ci_method = "sandwich"
)
result_dyn
```

### Comparing multiple interventions

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Forest plot of marginal means under three interventions: always treat, never treat, and adaptive (treat if L=1). All are close to 60 in this null-effect example."
est <- result_dyn$estimates
tinyplot(
  estimate ~ intervention,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  ylab = "E[Y^a]",
  xlab = "Intervention",
  main = "Marginal means under different interventions"
)
abline(h = 60, lty = 2, col = "grey40")
```

All three interventions should return a marginal mean close to 60 (the
structural truth for every joint regime in this null-effect DGP).

## The `history` parameter

The `history` parameter controls the Markov order --- how many lags of treatment
and time-varying confounders enter each outcome model.

- `history = 1` (default): Include one lag of A and L (first-order Markov).
- `history = 2`: Include two lags (A_{k-1}, A_{k-2}, L_{k-1}, L_{k-2}).
- `history = Inf`: Include full treatment and confounder history up to each
  time point.

For this 2-time-point example, `history = 1` and `history = Inf` are
equivalent. With longer follow-up, higher-order histories capture longer-range
dependencies but require more data per model.

```{r}
fit_inf <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  history = Inf,
  model_fn = stats::glm
)

result_inf <- contrast(
  fit_inf,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "sandwich"
)
result_inf
```

## Simulation: ICE with a non-null effect

To show ICE working with a known non-null effect, we simulate data with
treatment-confounder feedback.

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k \in \{0, 1\}$,
binary time-varying confounder $L_k \in \{0, 1\}$, and continuous outcome $Y$.
Treatment assignment at each period depends on the current confounder via a
logistic model; the confounder at $k = 1$ is affected by prior treatment $A_0$
(treatment-confounder feedback). The structural equations are:

$$
\begin{aligned}
L_0 &\sim \text{Bernoulli}(0.5), \\
A_0 \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.5 + 0.5 \, L_0)\bigr), \\
L_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.5 + 0.8 \, A_0 + 0.3 \, L_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.5 + 0.5 \, L_1)\bigr), \\
Y \mid A_0, A_1, L_1 &= 50 + 2.5 \, A_0 + 2.5 \, A_1 - 3 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 25).
\end{aligned}
$$

The structural ATE of "always treat" vs "never treat" is **not** simply
$2.5 + 2.5 = 5$, because $A_0$ affects $L_1$ (coefficient $0.8$), and $L_1$
enters the outcome with coefficient $-3$. Under "always treat", $A_0 = 1$
increases $\mathbb{E}[L_1]$, and since $L_1$ has a negative effect on $Y$,
this indirect path partially offsets the direct benefit:

$$
\text{ATE}
  = \underbrace{(2.5 + 2.5)}_{\text{direct}}
  - 3 \bigl(\mathbb{E}[L_1 \mid A_0\!=\!1] - \mathbb{E}[L_1 \mid A_0\!=\!0]\bigr)
  \approx 5 - 3 \times 0.196
  \approx 4.41.
$$

This is the same treatment-confounder feedback that makes naive regression
biased.

```{r}
set.seed(42)
n <- 2000
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)

L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.5 + 0.5 * L0))
L1 <- rbinom(n, 1, plogis(-0.5 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.5 + 0.5 * L1))
Y <- 50 + 2.5 * A0 + 2.5 * A1 - 3 * L1 + rnorm(n, 0, 5)

sim_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_sim <- causat(
  sim_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_sim <- contrast(
  fit_sim,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "sandwich"
)
res_sim
```

The estimated ATE should be close to the true value of $\approx 4.41$.

## Binary outcome

ICE also handles binary outcomes. The first outcome model uses `binomial`
(actual 0/1 response), while subsequent pseudo-outcome models use
`quasibinomial` (fractional logistic for predicted probabilities in [0, 1]).

**DGM.** This reuses the non-null simulation above (same SEM) with the
continuous outcome binarised at $Y > 50$:

$$
\begin{aligned}
Y^* &= 50 + 2.5 \, A_0 + 2.5 \, A_1 - 3 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 25), \\
Y &= \mathbf{1}(Y^* > 50).
\end{aligned}
$$

The resulting binary outcome does not have a clean closed-form ATE on the
risk-difference scale, but the direction is positive ("always treat" increases
$P(Y = 1)$ relative to "never treat") because both treatment coefficients are
positive in the latent linear predictor.

```{r}
sim_long_bin <- sim_long
sim_long_bin$Y[!is.na(sim_long_bin$Y)] <-
  as.integer(sim_long_bin$Y[!is.na(sim_long_bin$Y)] > 50)

fit_bin <- causat(
  sim_long_bin,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  family = "binomial",
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_bin <- contrast(
  fit_bin,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_bin
```

### Risk ratio for binary longitudinal outcomes

```{r}
res_rr <- contrast(
  fit_bin,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

## Three or more time periods

The examples above use 2 time points for simplicity, but ICE scales to
any number of periods. Here we simulate 4-period data with a binary treatment
and binary outcome. The backward recursion fits 4 sequential models.

**DGM.** Four periods ($k = 1, \dots, 4$) with binary treatment $A_k$,
continuous time-varying confounder $L_k$, continuous baseline confounder $L_0$,
and binary outcome $Y$ measured at $k = 4$. The confounder evolves with
treatment feedback: at each step, $L_k$ depends on $L_{k-1}$ plus noise, and
after observing $A_k$ the next confounder shifts by $0.3 \, A_k$. The
structural equations are:

$$
\begin{aligned}
L_0 &\sim N(0, 1), \\
L_k \mid L_{k-1}, A_{k-1} &= L_{k-1} + 0.3 \, A_{k-1} + \eta_k, \quad \eta_k \sim N(0, 0.25), \\
A_k \mid L_k &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.3 + 0.4 \, L_k)\bigr), \\
Y \mid A_1, \dots, A_4, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-1.5 + 0.4 \, \textstyle\sum_{k=1}^{4} A_k + 0.3 \, L_0)\bigr).
\end{aligned}
$$

The outcome depends on cumulative treatment exposure $\sum_k A_k$ and the
baseline confounder $L_0$. Under "always treat" ($\sum A_k = 4$) the
log-odds shift by $0.4 \times 4 = 1.6$ relative to "never treat"
($\sum A_k = 0$), so the risk difference on the probability scale is positive
but does not have a simple closed form (it depends on the marginal distribution
of $L_0$).

```{r}
set.seed(99)
n_id <- 600
K <- 4

ids <- rep(seq_len(n_id), each = K)
times <- rep(seq_len(K), n_id)

L0 <- rep(rnorm(n_id), each = K)
sim_4p <- data.frame(id = ids, time = times, L0 = L0)

sim_4p$L <- NA_real_
sim_4p$A <- NA_integer_
for (i in seq_len(n_id)) {
  rows <- which(sim_4p$id == i)
  l_prev <- sim_4p$L0[rows[1]]
  for (k in seq_along(rows)) {
    sim_4p$L[rows[k]] <- l_prev + rnorm(1, 0, 0.5)
    sim_4p$A[rows[k]] <- rbinom(1, 1, plogis(-0.3 + 0.4 * sim_4p$L[rows[k]]))
    l_prev <- sim_4p$L[rows[k]] + 0.3 * sim_4p$A[rows[k]]
  }
}

sim_4p$Y <- NA_real_
for (i in seq_len(n_id)) {
  rows <- which(sim_4p$id == i)
  cum_a <- sum(sim_4p$A[rows])
  sim_4p$Y[rows[K]] <- rbinom(
    1, 1,
    plogis(-1.5 + 0.4 * cum_a + 0.3 * sim_4p$L0[rows[1]])
  )
}

fit_4p <- causat(
  sim_4p,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L0,
  confounders_tv = ~ L,
  family = "binomial",
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_4p <- contrast(
  fit_4p,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_4p
```

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot showing the risk difference between always-treat and never-treat over 4 periods."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_4p)
```

```{r}
tt(tidy(res_4p), digits = 3)
```

The ICE sandwich variance correctly propagates uncertainty through all 4
sequential models via the stacked estimating equations cascade.

## Continuous treatment with modified treatment policies

ICE also supports continuous time-varying treatments combined with the
modified-treatment-policy (MTP) interventions: `shift()`, `scale_by()`, and
`threshold()`. The intervention is applied to the current-time treatment
at each backward step; `ice_apply_intervention_long()` explicitly does
**not** recompute the lag columns from the intervened values (this would
double-count a non-static intervention — see `CLAUDE.md` for the
structural reason).

**DGM.** Two periods ($k = 0, 1$) with continuous treatment $A_k \in \mathbb{R}$,
continuous confounder $L_k$, and continuous outcome $Y$. Treatment-confounder
feedback is present: $A_0$ affects $L_1$ (coefficient $0.5$), and $L_1$ affects
$A_1$ (coefficient $0.5$). The structural equations are:

$$
\begin{aligned}
L_0 &\sim N(0, 1), \\
A_0 &= 1 + 0.4 \, L_0 + \eta_0, \quad \eta_0 \sim N(0, 1), \\
L_1 &= 0.3 \, L_0 + 0.5 \, A_0 + \eta_1, \quad \eta_1 \sim N(0, 1), \\
A_1 &= 1 + 0.5 \, L_1 + \eta_2, \quad \eta_2 \sim N(0, 1), \\
Y &= 2 + 1.5 \, A_0 + 2 \, A_1 + 0.5 \, L_0 - 0.3 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1).
\end{aligned}
$$

For a `shift(-1)` MTP (reduce each $A_k$ by 1 at each period), the
structural causal effect on $Y$ relative to the natural course accounts for
both direct and indirect paths. Shifting $A_0$ by $-1$ also shifts
$L_1$ by $-0.5$ (coefficient $0.5$ in the $L_1$ equation), which in turn
shifts the natural value of $A_1$ by $-0.25$ (coefficient $0.5$ in the
$A_1$ equation). The shifted $A_1$ is then reduced by a further $-1$ by the
MTP. The total effect on $Y$ is:

$$
1.5 \times (-1) + 2 \times (-1.25) + (-0.3) \times (-0.5) = -1.5 - 2.5 + 0.15 = -3.85.
$$

```{r}
set.seed(101)
n <- 800
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)

L0 <- rnorm(n)
A0 <- 1 + 0.4 * L0 + rnorm(n)
L1 <- 0.3 * L0 + 0.5 * A0 + rnorm(n)
A1 <- 1 + 0.5 * L1 + rnorm(n)
Y <- 2 + 1.5 * A0 + 2 * A1 + 0.5 * L0 - 0.3 * L1 + rnorm(n)

cont_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_cont <- causat(
  cont_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_mtp <- contrast(
  fit_cont,
  interventions = list(
    observed = NULL,
    shift_m1 = shift(-1),
    halved   = scale_by(0.5),
    capped   = threshold(0, 1)
  ),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_mtp
```

Structural truth for `shift(-1)` is $-3.85$ (see derivation above). The
`shift_m1` estimate should be close to this value.

## Flexible treatment terms (`treatment_form`)

By default the treatment enters every per-step outcome model as a **bare numeric**
main effect. For a binary treatment that is saturated, but for an **ordered /
ordinal dose** a single linear slope can misspecify a *non-monotone* or *kinked*
counterfactual dose-response. `treatment_form` lets the dose enter the per-step
model via a transformed term — `~ factor(A)` or `~ splines::ns(A, df)` — while the
intervention still sets the *numeric* dose column. Lag terms expand automatically
(`factor(A)` → `factor(lag1_A)`).

**DGM.** Two periods, a 3-level dose $A_k \in \{0, 1, 2\}$ confounded by an
exogenous covariate, and a deliberately **non-monotone** level effect
$f(0) = 0,\ f(1) = 3,\ f(2) = 1$ with $Y = 1 + f(A_0) + f(A_1) + 0.5\,L_0 +
\varepsilon$. Because the covariate does not depend on past treatment, the
contrast $E[Y^{\text{static}(1)}] - E[Y^{\text{static}(0)}] = 2\,(f(1) - f(0)) =
6$ exactly.

```{r}
set.seed(1)
n <- 4000
f_lvl <- c(0, 3, 1) # f(0), f(1), f(2): non-monotone
L0 <- rnorm(n)
mk_period <- function() {
  Lt <- 0.5 * L0 + rnorm(n)
  At <- pmin(2L, rpois(n, exp(-0.3 + 0.5 * Lt)))
  list(L = Lt, A = At)
}
p0 <- mk_period()
p1 <- mk_period()
Yv <- 1 + f_lvl[p0$A + 1] + f_lvl[p1$A + 1] + 0.5 * L0 + rnorm(n)

dose_long <- data.frame(
  id = rep(seq_len(n), each = 2),
  time = rep(0:1, n),
  A = c(rbind(p0$A, p1$A)),
  L = c(rbind(p0$L, p1$L)),
  Y = c(rbind(NA_real_, Yv))
)

fit_args <- list(
  data = dose_long, outcome = "Y", treatment = "A",
  confounders = ~1, confounders_tv = ~L,
  id = "id", time = "time", estimator = "gcomp", model_fn = stats::glm
)
fit_bare <- do.call(causat, fit_args)
fit_factor <- do.call(causat, c(fit_args, list(treatment_form = ~ factor(A))))

ivs <- list(a1 = static(1), a0 = static(0))
diff_of <- function(fit) {
  r <- contrast(fit, interventions = ivs, type = "difference")
  with(r$estimates, estimate[intervention == "a1"] - estimate[intervention == "a0"])
}
dose_compare <- data.frame(
  model = c("bare numeric A", "factor(A)"),
  contrast = c(diff_of(fit_bare), diff_of(fit_factor)),
  truth = 6
)
tinytable::tt(dose_compare, digits = 3)
```

The bare-numeric model fits one slope through the non-monotone response and is
far from the truth of 6; `~ factor(A)` gives each dose level its own coefficient
and recovers it to sampling error. The flexible term is **longitudinal
g-computation only** and also powers natural-history dose policies (see the
*Natural-history MTPs* vignette). Under `factor(A)` an intervention must stay
within the observed dose levels; `splines::ns(A, df)` extrapolates smoothly.

## Longitudinal IPW (sequential density-ratio weights)

ICE g-computation models the **outcome** at each period and iterates
backward. The IPW analogue models the **treatment density** at each
period and accumulates a per-id density-ratio weight, then fits a
weighted final-period MSM. Both target the same counterfactual mean
$E[Y^d]$; agreement between them is the methodological-triangulation
test causatr is built around.

The mathematical structure is the same chain rule we used for
multivariate IPW (Phase 8) but indexed by **time** rather than
treatment component. For an intervention $d$ applied at every period
$k = 0, 1, \ldots, K$:

$$
W_i = \prod_{k=0}^{K} \frac{f_k\bigl(d(A_{k,i}, L_{k,i}) \mid \bar A_{k-1,i}^{\mathrm{obs}}, \bar L_{k,i}\bigr) \cdot \lvert \mathrm{Jac}\,d^{-1} \rvert}{f_k\bigl(A_{k,i} \mid \bar A_{k-1,i}^{\mathrm{obs}}, \bar L_{k,i}\bigr)}.
$$

Both numerator and denominator condition on **observed** upstream
treatments — the sequential-MTP semantics of @robins2000msmepi and
@diaz2023lmtp. The MSM is intercept-only Hájek on
the final-period rows, weighted by the per-id cumulative product;
the (sole) coefficient is $E[Y^d]$.

```{r}
fit_ipw_long <- causat(
  cont_long,
  outcome     = "Y",
  treatment   = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id          = "id",
  time        = "time",
  estimator   = "ipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_ipw_long <- contrast(
  fit_ipw_long,
  interventions = list(observed = NULL, shift_m1 = shift(-1)),
  reference     = "observed",
  type          = "difference",
  ci_method     = "sandwich"
)
res_ipw_long
```

ICE g-comp and longitudinal IPW estimate the same target parameter,
so on a correctly-specified linear-Gaussian DGP they should agree on
the point estimate up to Monte Carlo error. Sandwich SEs differ —
IPW typically has heavier variance than g-computation when the
outcome model is correct, so the IPW SE is usually larger.

```{r}
res_mtp$contrasts
res_ipw_long$contrasts
```

**Supported interventions.** `static()` / `shift()` / `scale_by()` /
`dynamic()` are dispatched per period through the same family-aware
gates as point IPW (`check_intervention_family_compat()`). Natural
course (`NULL`) collapses every per-period weight to 1 and recovers
the observed marginal mean. `ipsi()` is supported for univariate binary
treatment via a per-period Kennedy product (see below).

**Bootstrap.** `ci_method = "bootstrap"` resamples **individuals**
(every person-period row of a sampled id travels together). Cloning
ids that are sampled multiple times mirrors the ICE bootstrap
convention so within-person dependence is preserved.

```{r}
res_ipw_boot <- contrast(
  fit_ipw_long,
  interventions = list(observed = NULL, shift_m1 = shift(-1)),
  reference     = "observed",
  type          = "difference",
  ci_method     = "bootstrap",
  n_boot        = 100
)
res_ipw_boot$contrasts
```

**Sequential positivity warning.** When any per-period weight has a
heavy tail (max > 100 by default) the engine emits a
`causatr_longitudinal_seq_positivity` warning naming the offending
period(s) and their per-period max weight. The cumulative product
loses period attribution, so this per-period diagnostic is the
primary way to identify which time step is the bottleneck.

### Stabilized weights

`stabilize = "marginal"` swaps each period's numerator density for a
**marginal** model $g_k(A_k \mid \bar A_{k-1}, V)$ that drops the
time-varying confounder $L_k$ (and optionally adds a user-supplied
baseline conditioning set $V$ via `numerator = ~ V`). The denominator
stays at the full-$L$ density. Per-period weight becomes
$g_k(d(A_k) \mid \bar A_{k-1}, V) \cdot \lvert \mathrm{Jac} \rvert /
f_k(A_k \mid \bar A_{k-1}, \bar L_k)$, which dampens the
multiplicative $L$-dependence across $K$ factors and typically
reduces the cumulative product's variance.

```{r}
fit_ipw_long_stab <- causat(
  cont_long,
  outcome     = "Y",
  treatment   = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id          = "id",
  time        = "time",
  estimator   = "ipw",
  stabilize   = "marginal",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

contrast(
  fit_ipw_long_stab,
  interventions = list(observed = NULL, shift_m1 = shift(-1)),
  reference     = "observed",
  type          = "difference",
  ci_method     = "sandwich"
)
```

By default the per-period numerator drops every confounder and
conditions only on prior treatment lags. A custom `numerator =`
formula adds baseline-only covariates on top of the lags, which
typically gives the most stable contrast estimate when the structural
treatment generation depends on baseline characteristics.

Sandwich variance treats the numerator parameters as fixed (same
nuisance-fixed convention as Phase 8e for multivariate IPW); bootstrap
refits both numerator and denominator per replicate and captures the
full uncertainty.

### Effect modification (baseline modifier)

When `confounders =` includes an `A:modifier` interaction term, the
per-period propensity formulas automatically strip the
treatment-touching interaction (the modifier main effect is kept),
and the final-period MSM expands from `Y ~ 1` to `Y ~ 1 + modifier`
so `predict()` returns stratum-specific counterfactual means.

```{r}
#| eval: false
fit_em <- causat(
  long_data,
  outcome     = "Y",
  treatment   = "A",
  confounders = ~ baseline + sex + A:sex,
  confounders_tv = ~ L,
  id          = "id",
  time        = "time",
  estimator   = "ipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)
contrast(
  fit_em,
  interventions = list(a = static(1), z = static(0)),
  reference     = "z",
  by            = "sex"
)
```

**Known limitation [@robins2000msmepi].** Under longitudinal IPW, the
modifier MUST be a **baseline** covariate. A time-varying modifier
in an MSM conditions on a post-treatment variable, which biases the
estimand via mediator and collider paths. causatr does not enforce
baseline-ness at runtime (time-varying status isn't inferable from
data); the constraint is documented only. The scientifically
correct tool for time-varying effect modification is a structural
nested model — use `estimator = "snm"` with `id =` and `time =`
(see `vignette("snm")`).

Baseline effect modification also composes with **multivariate**
longitudinal treatment: each per-period × per-component propensity
strips its own `A_k:modifier` interactions (keeping the modifier as a
confounder main effect), and the single final-period MSM still expands
to `Y ~ 1 + modifier`. Specify the interaction once per component, e.g.
`confounders = ~ baseline + sex + A1:sex + A2:sex` with
`treatment = c("A1", "A2")`.

### Incremental propensity-score interventions (IPSI)

For **binary** time-varying treatment, `ipsi(delta)` shifts the
treatment odds at every period by a factor $\delta$ rather than fixing a
counterfactual value. Kennedy's [-@kennedy2019ipsi] closed-form weight extends to a
per-period product:

$$
W_i = \prod_{t=1}^{T} \frac{\delta \, A_{t,i} + (1 - A_{t,i})}
                            {\delta \, p_{t,i} + (1 - p_{t,i})},
\qquad p_{t,i} = P(A_t = 1 \mid \bar A_{t-1}, \bar L_t).
$$

$\delta = 1$ recovers the natural course (every factor is 1); $\delta > 1$
nudges everyone toward treatment, $\delta < 1$ away from it. Because each
denominator stays in $[\min(1, \delta), \max(1, \delta)]$, the per-period
weights are bounded by construction — IPSI is a stable alternative to
deterministic "always treat" contrasts when positivity is borderline.

```{r}
set.seed(204)
n_ip <- 800
ip_id <- rep(seq_len(n_ip), each = 2)
ip_time <- rep(0:1, n_ip)
L0_ip <- rnorm(n_ip)
A0_ip <- rbinom(n_ip, 1, plogis(0.5 * L0_ip))
L1_ip <- 0.5 * L0_ip + 0.4 * A0_ip + rnorm(n_ip)
A1_ip <- rbinom(n_ip, 1, plogis(0.3 * L1_ip + 0.2 * A0_ip))
Y_ip <- 10 + 2 * (A0_ip + A1_ip) + L0_ip + L1_ip + rnorm(n_ip)
ipsi_long <- data.frame(
  id = ip_id,
  time = ip_time,
  A = as.vector(rbind(A0_ip, A1_ip)),
  L = as.vector(rbind(L0_ip, L1_ip)),
  L0 = rep(L0_ip, each = 2),
  Y = as.vector(rbind(rep(NA_real_, n_ip), Y_ip))
)

fit_ipsi_long <- causat(
  ipsi_long,
  outcome     = "Y",
  treatment   = "A",
  confounders = ~ L0,
  confounders_tv = ~ L,
  id          = "id",
  time        = "time",
  estimator   = "ipw"
)

res_ipsi <- contrast(
  fit_ipsi_long,
  interventions = list(
    down = ipsi(0.5),
    observed = ipsi(1),
    up = ipsi(2)
  ),
  ci_method = "sandwich"
)
tt(tidy(res_ipsi), digits = 3)
```

The three means interpolate monotonically: lowering the treatment odds
($\delta = 0.5$) pulls the counterfactual mean below the natural course
($\delta = 1$), raising them ($\delta = 2$) pushes it up.

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Point estimates and confidence intervals for IPSI counterfactual means at delta = 0.5, 1, and 2. The means increase monotonically with delta."
ipsi_df <- res_ipsi$estimates
ipsi_df$delta <- c(0.5, 1, 2)[match(
  ipsi_df$intervention,
  c("down", "observed", "up")
)]
ipsi_df <- ipsi_df[order(ipsi_df$delta), ]
tinyplot(
  estimate ~ factor(delta),
  data = ipsi_df,
  type = "pointrange",
  ymin = ipsi_df$ci_lower,
  ymax = ipsi_df$ci_upper,
  ylab = "Counterfactual mean E[Y(ipsi(delta))]",
  xlab = "IPSI odds multiplier (delta)",
  main = "Longitudinal IPSI interpolates with delta"
)
```

`ipsi()` under longitudinal IPW is **univariate binary only**. Multivariate
IPSI (`treatment = c("A1", "A2")` with an `ipsi()` component) and
`ipsi()` combined with `stabilize = "marginal"` are rejected with classed
errors (`causatr_longitudinal_ipsi_multivariate`,
`causatr_longitudinal_ipsi_stabilize`). Longitudinal AIPW does not support
IPSI — use `estimator = "ipw"`.

### Weight truncation

With $T$ time periods, the cumulative product $\prod_{t=1}^{T} w_t$ of
per-period density ratios can develop heavy tails. The `trim` argument on
`contrast()` winsorizes density ratios at a given quantile *before* the
cumulative product [@cole2008constructing]. `trim = 0.999` clips each
per-period ratio at its 99.9th percentile --- the default in
[`lmtp`](https://cran.r-project.org/package=lmtp).

```{r}
res_ipw_trim <- contrast(
  fit_ipw_long,
  interventions = list(shift_m1 = shift(-1), observed = NULL),
  reference     = "observed",
  type          = "difference",
  ci_method     = "sandwich",
  trim          = 0.999
)
res_ipw_trim
```

The default `trim = 1` applies no truncation. With `stabilize = "marginal"`,
stabilization and truncation compose: stabilized per-period ratios are
truncated at the quantile threshold before the cumulative product.

## Multivariate time-varying treatment

When multiple treatments vary over time (e.g. two drugs dosed jointly),
pass a character vector to `treatment =` and supply each intervention as
a named sub-list. Lag columns are built per-component by `prepare_data()`.

**DGM.** Two periods ($k = 0, 1$) with two binary treatments
$A_{1,k}, A_{2,k} \in \{0, 1\}$, a continuous confounder $L_k$, and
continuous outcome $Y$. Both treatments at $k = 0$ feed into the confounder
$L_1$, creating treatment-confounder feedback. The structural equations are:

$$
\begin{aligned}
L_0 &\sim N(0, 1), \\
A_{1,0} \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3 \, L_0)\bigr), \\
A_{2,0} \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.1 \, L_0)\bigr), \\
L_1 &= 0.3 \, L_0 + 0.4 \, A_{1,0} + 0.2 \, A_{2,0} + \eta, \quad \eta \sim N(0, 1), \\
A_{1,1} \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3 \, L_1)\bigr), \\
A_{2,1} \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.1 \, L_1)\bigr), \\
Y &= 2 + 1.5 \, A_{1,0} + 1.5 \, A_{1,1} + 0.8 \, A_{2,0} + 0.8 \, A_{2,1} + 0.5 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1).
\end{aligned}
$$

The outcome is linear in both treatments at both periods. The structural ATE
of "both on" vs "both off" includes the indirect path through $L_1$: setting
$A_{1,0} = A_{2,0} = 1$ shifts $\mathbb{E}[L_1]$ by $0.4 + 0.2 = 0.6$
relative to "both off", and $L_1$ enters $Y$ with coefficient $0.5$, adding
$0.3$ to the direct sum $2 \times (1.5 + 0.8) = 4.6$:

$$
\text{ATE} = \underbrace{4.6}_{\text{direct}} + \underbrace{0.5 \times 0.6}_{\text{indirect via } L_1} = 4.9.
$$

```{r}
set.seed(202)
n <- 600
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)
L0 <- rnorm(n)
A10 <- rbinom(n, 1, plogis(0.3 * L0))
A20 <- rbinom(n, 1, plogis(0.1 * L0))
L1 <- 0.3 * L0 + 0.4 * A10 + 0.2 * A20 + rnorm(n)
A11 <- rbinom(n, 1, plogis(0.3 * L1))
A21 <- rbinom(n, 1, plogis(0.1 * L1))
Y <- 2 + 1.5 * A10 + 1.5 * A11 + 0.8 * A20 + 0.8 * A21 + 0.5 * L1 + rnorm(n)

mv_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A1 = c(rbind(A10, A11)),
  A2 = c(rbind(A20, A21)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_mv_long <- causat(
  mv_long,
  outcome = "Y",
  treatment = c("A1", "A2"),
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_mv_long <- contrast(
  fit_mv_long,
  interventions = list(
    both_on  = list(A1 = static(1), A2 = static(1)),
    both_off = list(A1 = static(0), A2 = static(0))
  ),
  reference = "both_off",
  type = "difference"
)
res_mv_long
```

Structural truth: `both_on − both_off = 4.9` (direct effects $4.6$ plus
indirect $0.3$ via $L_1$).

### Stabilization with multivariate treatments

`stabilize = "marginal"` composes with multivariate longitudinal IPW. The
marginal numerator factorises over **both** axes: per period and per
component. Each component's numerator drops the time-varying confounders
$L$ but keeps the treatment lags and the within-period prior components
$A_{1,\dots,k-1}$, mirroring the denominator chain rule:

$$
g(\bar{A}_1, \bar{A}_2 \mid V)
  = \prod_t \prod_k g_{t,k}\bigl(A_{t,k} \mid A_{t,1:k-1}, \bar{A}_{t-1}, V\bigr).
$$

Under static interventions on binary treatments the stabilized and
unstabilized estimates coincide exactly — the indicator pins every
conditioning variable to the target, so numerator and denominator densities
cancel on the surviving rows:

```{r}
fit_mv_long_stab <- causat(
  mv_long,
  outcome = "Y",
  treatment = c("A1", "A2"),
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm,
  stabilize = "marginal"
)

res_mv_long_stab <- contrast(
  fit_mv_long_stab,
  interventions = list(
    both_on  = list(A1 = static(1), A2 = static(1)),
    both_off = list(A1 = static(0), A2 = static(0))
  ),
  reference = "both_off",
  type = "difference"
)
res_mv_long_stab
```

### Doubly-robust multivariate longitudinal AIPW

For doubly-robust protection on joint time-varying treatments, switch
`estimator = "aipw"`. The ICE-AIPW recursion [@bang2005doubly] augments
the sequential outcome models with the multivariate per-period density-ratio
weight — the same per-period $\times$ per-component chain used by the IPW
path. The estimator is consistent if **either** the sequence of outcome
models **or** the per-period propensity chain is correctly specified.

```{r}
fit_mv_long_aipw <- causat(
  mv_long,
  outcome = "Y",
  treatment = c("A1", "A2"),
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  estimator = "aipw",
  model_fn = stats::glm
)

res_mv_long_aipw <- contrast(
  fit_mv_long_aipw,
  interventions = list(
    both_on  = list(A1 = static(1), A2 = static(1)),
    both_off = list(A1 = static(0), A2 = static(0))
  ),
  reference = "both_off",
  type = "difference",
  ci_method = "sandwich"
)
tt(tidy(res_mv_long_aipw), digits = 3)
```

The point estimate recovers the same structural truth (`both_on −
both_off = 4.9`) as the IPW and g-computation fits above — under a static
contrast the multivariate IPW, g-computation, and AIPW estimands coincide
[@diaz2023lmtp]. The sandwich SE comes from a $T \times K$ block-diagonal
stacked estimating-equation system (one propensity block per period $\times$
component, plus the outcome cascade).

Two combinations are **not** supported on this path:

- `stabilize = "marginal"` is rejected for longitudinal AIPW (use
  `estimator = "ipw"` for stabilized longitudinal weights).
- Transport (`target =`) for longitudinal AIPW is not yet available.

## External weights (IPCW / survey)

External weights — survey weights or pre-computed IPCW — pass through
every backward-step model via `model_fn`'s `weights =` argument. The ICE
sandwich variance engine accounts for heterogeneous weights via the
`A_{k-1,k}` cascade gradient (see `variance-theory.qmd` §5.4 and the
inline comment in `R/variance_if.R:variance_if_ice_one`).

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k$, binary
confounder $L_k$, continuous outcome $Y$, and uninformative survey weights
drawn independently from $\text{Uniform}(0.5, 2.0)$. The structural equations
are:

$$
\begin{aligned}
L_0 &\sim \text{Bernoulli}(0.5), \\
A_0 \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_0)\bigr), \\
L_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.8 \, A_0 + 0.3 \, L_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_1)\bigr), \\
Y &= 2 + 2.5 \, A_0 + 2.5 \, A_1 - 0.5 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1), \\
w_i &\sim \text{Uniform}(0.5, 2.0).
\end{aligned}
$$

Since the weights are non-informative (independent of everything), the
structural ATE of "always treat" vs "never treat" is the same as the
unweighted case. The direct effects sum to $2.5 + 2.5 = 5$, but $A_0$
shifts $\mathbb{E}[L_1]$ by $\approx 0.19$, and $L_1$ enters $Y$ with
coefficient $-0.5$, giving a true ATE $\approx 4.90$. The purpose of this
example is to demonstrate that the ICE sandwich variance engine correctly
propagates heterogeneous weights through the backward recursion.

```{r}
set.seed(303)
n <- 500
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)
L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L0))
L1 <- rbinom(n, 1, plogis(-0.2 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L1))
Y <- 2 + 2.5 * A0 + 2.5 * A1 - 0.5 * L1 + rnorm(n)
weights_long <- runif(2 * n, 0.5, 2.0)

w_long_df <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_w <- causat(
  w_long_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  weights = weights_long,
  model_fn = stats::glm
)

res_w <- contrast(
  fit_w,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_w
```

## Censoring (IPCW-style row filter)

Pass a time-varying censoring indicator via `censoring =`. At each
backward step, ICE restricts the fit to uncensored rows; predictions are
made for the target population regardless of censoring status, so the
estimator is valid under the usual sequential-randomization assumption
on the censoring mechanism.

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k$, binary
confounder $L_k$, continuous outcome $Y$, and a censoring indicator $C_1$ at
$k = 1$ that depends on treatment and baseline confounder. About 20% of
individuals are censored at the second period. The structural equations are:

$$
\begin{aligned}
L_0 &\sim \text{Bernoulli}(0.5), \\
A_0 \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_0)\bigr), \\
L_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.8 \, A_0 + 0.3 \, L_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_1)\bigr), \\
Y^* &= 2 + 2 \, A_0 + 2 \, A_1 - 0.5 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1), \\
C_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-2 + 0.5 \, A_0 + 0.3 \, L_0)\bigr), \\
Y &= \begin{cases} Y^* & \text{if } C_1 = 0, \\ \text{NA} & \text{if } C_1 = 1. \end{cases}
\end{aligned}
$$

The structural ATE of "always treat" vs "never treat" on the full
(uncensored) outcome is $\approx 3.90$ (direct effects $2 + 2 = 4$ minus
$0.5 \times 0.19 \approx 0.10$ for the indirect path $A_0 \to L_1 \to Y$).
Under the censoring-at-random assumption
($Y^{a_0, a_1} \perp C_1 \mid A_0, L_0$), ICE with row filtering recovers
this effect by fitting the outcome model on uncensored rows and standardising
over the full population.

```{r}
set.seed(404)
n <- 500
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)
L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L0))
L1 <- rbinom(n, 1, plogis(-0.2 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L1))
Y_full <- 2 + 2 * A0 + 2 * A1 - 0.5 * L1 + rnorm(n)
# 20% censoring at t = 1 depending on A0 + L0.
C1 <- rbinom(n, 1, plogis(-2 + 0.5 * A0 + 0.3 * L0))
Y <- ifelse(C1 == 1, NA_real_, Y_full)

cens_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  C = c(rbind(0L, C1)),
  Y = c(rbind(NA_real_, Y))
)

fit_c <- causat(
  cens_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  censoring = "C",
  model_fn = stats::glm
)

res_c <- contrast(
  fit_c,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_c
```

Structural truth $\approx 3.90$ (see derivation above). The estimate should
be close, and the sandwich SE reflects the censoring-induced reduction in
effective sample size at each step.

## How ICE works

ICE g-computation [@zivich2024ice] works **backward** in time:

1. **Final time K**: Fit E[Y | A_K, L_K, baseline] among uncensored individuals.
   Predict under the intervention to get pseudo-outcomes.
2. **Each earlier time k**: Use pseudo-outcomes as the response. Fit
   E[pseudo | A_k, L_k, baseline]. Predict under intervention.
3. **First time**: The mean of the pseudo-outcomes is the marginal mean E[Y^a].

This requires **only outcome models** --- unlike forward simulation
("big g-formula"), no models for time-varying confounders are needed.

### Why ICE, not forward simulation?

| | ICE g-comp | Forward simulation |
|---|---|---|
| **Models needed** | Outcome only (at each time) | Outcome + every time-varying confounder |
| **Inference** | Sandwich (stacked EE) | Bootstrap only |
| **Speed** | Fast | Slow (many models + simulation) |
| **Model misspecification** | K models | K + p*K models |
| **Implementation** | causatr | gfoRmula |

## Key parameters

- `confounders`: Baseline (time-invariant) confounders. Enter every outcome
  model at every time step.
- `confounders_tv`: Time-varying confounders. Included with their lags
  (controlled by `history`).
- `history`: Markov order. `history = 1` (default) includes one lag of
  treatment and TV confounders. `history = Inf` includes full history.
- `censoring`: Time-varying censoring indicator (1 = censored). ICE restricts
  to uncensored at each backward step; the resulting estimator is IPCW-valid
  when the censoring mechanism is conditionally random given the covariate
  history.
- `weights`: external survey weights or pre-computed IPCW. Accepts either a
  numeric vector or a `survey::svydesign` object directly; with a
  design, the sampling weights are unpacked via `stats::weights()` and
  the design's first-stage PSU is auto-adopted as the contrast-time
  cluster. Validated upfront by `check_weights()` (rejects NA, Inf,
  negative, mis-sized) and propagated through every step of the
  backward recursion. The ICE sandwich variance engine accounts for
  the weights via the `A_{k-1,k}` cascade gradient — dropping them
  would underestimate the SE under heterogeneous weights by roughly a
  factor of two.
- `cluster`: column name in `data` identifying a design cluster (site,
  household, PSU, repeated-measures id). Preserved through
  `prepare_data()` and consumed at contrast time by the
  sum-within-cluster-then-square IF aggregator
  (`vcov_from_if(cluster = ...)`). Defaults from the fit into
  `contrast()` so users don't have to repeat it; override or switch
  off at contrast time by passing a different `cluster =` (or NULL).
  Matching rejects this argument with `causatr_bad_cluster` because
  its subclass clustering is structural.
- `parallel` / `ncpus`: four backends — `"no"`, `"multicore"`, `"snow"`
  (all via `boot::boot()`), and `"future"` (via
  `future.apply::future_lapply()` so any active `future::plan()` is
  honoured). Defaults to `getOption("boot.parallel", "no")` /
  `getOption("boot.ncpus", 1L)` so a session-wide `options(boot.parallel =
  "multicore")` flips parallelism on for every `contrast()` call.
  `ncpus` is ignored when `parallel = "future"` (the plan owns the
  worker count).

For a full enumeration of the sandwich-vs-bootstrap, continuous/binary,
external-weights, censoring, multivariate-treatment, and ipsi-rejection
combinations, see `FEATURE_COVERAGE_MATRIX.md` §"Method 4 — ICE longitudinal
g-computation" (20+ truth-pinned rows against `lmtp::lmtp_tmle` / analytic
targets / @hernan_whatif Table 20.1).

## Effect modification across time

When `confounders` includes an `A:modifier` interaction term (e.g.
`~ L0 + sex + A:sex`), ICE auto-expands the interaction across treatment
lags. At each backward step, the outcome model includes not only `A:sex`
(current period) but also `lag1_A:sex`, `lag2_A:sex`, etc. for all
available lags. This captures how the effect of treatment at every time
point is modified by the baseline variable.

Without this expansion, ICE would capture effect modification only at the
current treatment period, collapsing roughly half the intended heterogeneity
in a 2-period DGP.

**DGM.** A 2-period DGP with sex-specific treatment effects:

$$
\begin{aligned}
  L_0, \text{sex} &\sim N(0,1),\; \text{Bern}(0.5) \\
  A_0 \mid L_0 &\sim \text{Bern}(\text{expit}(0.3 L_0)) \\
  L_1 \mid A_0, L_0 &= A_0 + 0.5 L_0 + \varepsilon \\
  A_1 \mid L_1 &\sim \text{Bern}(\text{expit}(0.3 L_1)) \\
  Y \mid \bar{A}, \bar{L}, \text{sex} &= 10 + (2 + 1.5 \cdot \text{sex})(A_0 + A_1) + L_0 + L_1 + \varepsilon
\end{aligned}
$$

True always-vs-never ATE: $\text{ATE}|\text{sex}=0 = 5$,
$\text{ATE}|\text{sex}=1 = 8$.

```{r}
set.seed(42)
n <- 5000
L0 <- rnorm(n)
sex <- rbinom(n, 1, 0.5)
Ltv0 <- 0.5 * L0 + rnorm(n, 0, 0.5)
A0 <- rbinom(n, 1, plogis(0.3 * L0))
L1 <- A0 + 0.5 * L0 + rnorm(n, 0, 0.5)
A1 <- rbinom(n, 1, plogis(0.3 * L1))
Y <- 10 + (2 + 1.5 * sex) * (A0 + A1) + Ltv0 + L1 + rnorm(n)

d_em <- rbind(
  data.frame(id = 1:n, time = 0L, A = A0, L = Ltv0,
             L0 = L0, sex = sex, Y = NA_real_),
  data.frame(id = 1:n, time = 1L, A = A1, L = L1,
             L0 = L0, sex = sex, Y = Y)
)

fit_ice_em <- causat(
  d_em,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L0 + sex + A:sex,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_ice_em <- contrast(
  fit_ice_em,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich",
  by = "sex"
)
res_ice_em
```

The `A:sex` interaction is auto-expanded at the final backward step to
include `lag1_A:sex`, so both treatment periods contribute sex-specific
effects. The sandwich variance engine requires no special handling --- the
model matrices simply grow to accommodate the additional interaction columns,
and the ICE cascade gradient propagates through them.

## Stratified ICE (`stratified = "G"`)

When the outcome--treatment relationship differs structurally across a
**baseline** subgroup `G` --- a different functional form, different effect
modification, or different dispersion in each subgroup --- a single pooled
per-step outcome model is misspecified. `causat(..., stratified = "G")` fits one
outcome / pseudo-outcome model **per level of `G`** at every backward step of the
ICE recursion.

Because `G` is time-invariant and individuals never cross strata, stratified ICE
is **exactly** pooled ICE run separately on each stratum subset: the estimand ---
the marginal counterfactual mean $E[Y^{\bar d}]$ (and any `by` / `subset`
restriction) --- is unchanged. It is a *nuisance-model* choice, not an estimand
change. `G` must be a baseline, complete, low-cardinality column (time-varying,
continuous, or NA stratifiers are rejected with classed errors).

```{r}
set.seed(7)
n <- 4000
G <- rbinom(n, 1, 0.5) # baseline stratum; the treatment effect differs by G
L0 <- rnorm(n)
A0 <- rbinom(n, 1, plogis(0.3 * L0))
L1 <- 0.5 * L0 + 0.4 * A0 + rnorm(n)
A1 <- rbinom(n, 1, plogis(0.3 * L1))
Y <- 2 + (1 + 2 * G) * (A0 + A1) + L0 + L1 + rnorm(n)
d_strat <- rbind(
  data.frame(id = 1:n, time = 0L, A = A0, L = 0, G = G, L0 = L0, Y = NA_real_),
  data.frame(id = 1:n, time = 1L, A = A1, L = L1, G = G, L0 = L0, Y = Y)
)

fit_strat <- causat(
  d_strat,
  outcome = "Y", treatment = "A",
  confounders = ~ L0 + G, confounders_tv = ~ L,
  id = "id", time = "time", estimator = "gcomp",
  stratified = "G"
)

contrast(
  fit_strat,
  interventions = list(always = static(1), never = static(0)),
  reference = "never", ci_method = "sandwich", by = "G"
)$contrasts
```

The two strata show markedly different effects (here the `G = 1` always-vs-never
effect is roughly double the `G = 0` effect, by construction). Both variance
paths work: the ID-cluster bootstrap **and** the analytic per-stratum × per-time
stacked-EE sandwich (validated against `delicatessen`'s `MEstimator` to ~1e-7).
The within-stratum formula automatically drops the constant `G` term.

## Natural-history modified treatment policies (grace periods)

Some policies set the intervened treatment at time $t$ as a function of the
**natural-value history of treatment** --- grace periods, delays, and
carry-forward. These are **G-LMTPs** [@diaz2026glmtp], and they are *not*
expressible as `dynamic()` rules: the standard ICE
recursion conditions on the *observed* lagged treatment, whereas the policy
needs the *counterfactual* natural value, which under treatment-state feedback
differs from the observed value (so `dynamic(\(d, trt) d$lag1_A)` runs but
silently targets the wrong estimand). causatr estimates them with an
augmented-data sequential regression via the `grace_period()`,
`carry_forward()`, and `cap_escalation()` constructors (longitudinal
g-computation, discrete treatment, ID-cluster bootstrap). `cap_escalation(d)`
caps the natural per-period increase of an ordered / count dose at `d`, and is
consistent when the dose enters the per-step model flexibly
(`treatment_form = ~ factor(dose)`).

This family has its own article: see
[**Natural-history MTPs (grace periods)**](natural-history-mtps.html) for the
concept, a visual of how each policy assigns treatment, why `dynamic(lag1_A)` is
wrong, the augmented-data intuition, worked examples validated against the
forward-simulation truth (replicating the @diaz2026glmtp Section-6 delay
result), and the relationship to LMTPs and natural-value IPW.

## Summary of covered combinations

**Legend.** ✅ covered and truth-pinned in tests (either against Table 20.1
or against `lmtp::lmtp_tmle()`) · 🟡 smoke test only · ⛔ rejected with an
informative error.

| Treatment | Outcome | Intervention | Estimand | Contrast | Inference | Weights | Status |
|---|---|---|---|---|---|---|---|
| Binary | Continuous | Static (always vs never) | ATE | Difference | Sandwich | none | ✅ Table 20.1 |
| Binary | Continuous | Static | ATE | Difference | Bootstrap | none | ✅ Table 20.1 |
| Binary | Continuous | Dynamic (rule on L) | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich + bootstrap | survey | ✅ |
| Continuous | Continuous | Shift / Scale / Threshold | ATE | Difference | Sandwich | none | ✅ (vs lmtp) |
| Continuous | Continuous | Shift | ATE | Difference | Bootstrap | none | ✅ (vs lmtp) |
| Multivariate | Continuous | Static / Shift | ATE | Difference | Sandwich | none | ✅ |
| Binary | Binary | Static / Dynamic | ATE | Difference | Sandwich / bootstrap | none | ✅ (2 periods) |
| Binary | Binary | Static | ATE | Difference | Sandwich | none | ✅ (4 periods) |
| Binary | Binary | Static | ATE | Difference | Bootstrap (parallel) | none | ✅ |
| Binary | Binary | Static | ATE | Difference | Bootstrap (future backend) | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich (cluster-robust) | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich | censoring | ✅ |
| Binary | Continuous | Static + EM | ATE | Difference | Sandwich / bootstrap | none | ✅ |
| Continuous | Continuous | Shift + EM | ATE | Difference | Sandwich | none | ✅ |
| Multivariate | Continuous | Static + EM (`A1:sex` + `A2:sex`) | ATE | Difference | Sandwich | none / marginal | ✅ truth + vs ICE |
| Multivariate (AIPW) | Continuous / Binary | Static / Shift | ATE | Difference | Sandwich / bootstrap | none | ✅ truth + DR + vs ICE/IPW |
| Multivariate (AIPW) | Any | Static | ATE | Difference | Sandwich | marginal | ⛔ stabilize rejected |
| Binary | Continuous | IPSI($\delta$) | ATE | Difference | Sandwich / bootstrap | none | ✅ +oracle (per-period Kennedy) |
| Multivariate | Any | IPSI($\delta$) component | — | — | — | — | ⛔ binary-univariate only |
| Binary | Any | IPSI($\delta$) | ATE | — | — | marginal | ⛔ no marginal numerator |
| Binary | Continuous | Static, **stratified ICE** (`stratified = "G"`) | ATE / `by(G)` | Difference | Sandwich / bootstrap | none | ✅ vs per-stratum pooled + delicatessen |
| Binary (absorbing) | Continuous / Binary | `grace_period` / `carry_forward` (natural-history MTP) | ATE | Difference | Bootstrap / Sandwich | none / cens / ext | ✅ vs forward-MC truth [@diaz2026glmtp, §6]; sandwich validated vs Python oracle |
| Ordinal dose `{0,1,2}` | Continuous | `cap_escalation` + `treatment_form = ~ factor(dose)` | ATE | Difference | Bootstrap / Sandwich | none | ✅ tight vs forward-MC truth (gap ≈0.0015) |
| Binary | Any | Static / ATT | ATT | — | — | — | ⛔ ICE is ATE-only |

See `FEATURE_COVERAGE_MATRIX.md`
for the authoritative coverage status of every method × treatment × outcome
× intervention × variance combination.

## References

::: {#refs}
:::
