---
title: "Longitudinal treatments: ICE g-computation"
vignette: >
  %\VignetteIndexEntry{Longitudinal treatments: ICE g-computation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When treatment and confounders vary over time, standard regression methods give
biased estimates even when adjusted for all confounders. G-methods solve this.
causatr implements ICE (iterated conditional expectation) g-computation (Zivich
et al., 2024), which requires models for the outcome only --- no models for
time-varying confounders are needed.

This vignette demonstrates ICE g-computation in causatr, covering:

- The treatment-confounder feedback problem
- ICE fitting and contrasting
- Sandwich and bootstrap inference
- Static, dynamic, and modified treatment policy interventions
- The `history` parameter (Markov order)
- Binary outcomes
- Comparison with naive regression

## Setup

```{r}
#| message: false
library(causatr)
library(tinyplot)
```

## The problem: treatment-confounder feedback

When treatment at time 0 affects a confounder at time 1, which in turn affects
treatment at time 1 (A_0 -> L_1 -> A_1 -> Y), neither adjusting nor not
adjusting for L_1 gives the correct answer:

- **Adjusting** for L_1 opens a collider path (selection bias)
- **Not adjusting** leaves confounding for A_1
- **ICE g-computation** handles both correctly

## Table 20.1: a null effect with treatment-confounder feedback

We construct the dataset from Table 20.1 of Hernan & Robins (2025), scaled to
3,200 individuals observed at 2 time points with a true causal effect of zero.
(The original table has 32,000; we use 1/10th for faster vignette rendering.)

```{r}
groups <- data.frame(
  A0 = c(0, 0, 0, 0, 1, 1, 1, 1),
  L1 = c(0, 0, 1, 1, 0, 0, 1, 1),
  A1 = c(0, 1, 0, 1, 0, 1, 0, 1),
  Y  = c(84, 84, 52, 52, 76, 76, 44, 44),
  N  = c(240, 160, 240, 960, 480, 320, 160, 640)
)
rows <- lapply(seq_len(nrow(groups)), function(i) {
  g <- groups[i, ]
  off <- sum(groups$N[seq_len(i - 1)])
  data.frame(id = seq_len(g$N) + off, A0 = g$A0, L1 = g$L1, A1 = g$A1, Y = g$Y)
})
wide <- do.call(rbind, rows)

t0 <- data.frame(id = wide$id, time = 0L, A = wide$A0, L = NA_real_, Y = NA_real_)
t1 <- data.frame(id = wide$id, time = 1L, A = wide$A1, L = wide$L1, Y = wide$Y)
long <- rbind(t0, t1)
head(long, 4)
```

The data is in **person-period (long) format**: one row per individual per
time point. `L` is a time-varying confounder (only measured at time 1).

## Demonstrating bias from naive methods

A naive regression of Y on treatment adjusting for L_1 gives a biased estimate:

```{r}
wide_sub <- wide[!is.na(wide$Y), ]
naive_coefs <- coef(lm(Y ~ A0 + A1 + L1, data = wide_sub))
naive_coefs[c("A0", "A1")]
```

The true individual effects of A_0 and A_1 are both zero, but the naive
regression gives non-zero estimates due to treatment-confounder feedback.

## ICE g-computation: fitting

With `causat()`, specify `id` and `time` to trigger longitudinal mode. Baseline
confounders go in `confounders`, time-varying confounders in `confounders_tv`.

```{r}
fit <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)
fit
```

No models are fitted at this stage. ICE defers model fitting to `contrast()`
because the sequential outcome models depend on the intervention being
evaluated.

## ICE g-computation: contrasts with sandwich SE

Compare "always treat" (A = 1 at every time) vs "never treat" (A = 0 at every
time):

```{r}
result <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
result
```

ICE correctly recovers **E[Y^{always}] = E[Y^{never}] = 60** and **ATE = 0**,
as expected from the data generating process.

### Visualising the result

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Bar plot comparing naive regression estimates to ICE g-computation estimates for the effect of treatment at times 0 and 1. ICE correctly recovers ATE = 0 while naive regression gives biased estimates."
est_df <- data.frame(
  method = c("Naive (A0)", "Naive (A1)", "ICE"),
  estimate = c(naive_coefs["A0"], naive_coefs["A1"], result$contrasts$estimate[1]),
  stringsAsFactors = FALSE
)
tinyplot(
  estimate ~ method,
  data = est_df,
  type = "bar",
  ylab = "Estimated treatment effect",
  xlab = "",
  main = "ICE recovers the null effect"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Sandwich SE: stacked estimating equations

The default `ci_method = "sandwich"` uses the empirical sandwich estimator for
stacked estimating equations (Zivich et al., 2024). This accounts for
uncertainty from all K sequential outcome models, not just the final model.

```{r}
result$estimates[, c("intervention", "estimate", "se")]
```

The standard errors are finite and positive, confirming that the sandwich
variance estimator is working correctly even with a large sample.

## Bootstrap SE

For robustness or when using flexible models (e.g., GAMs), bootstrap resamples
individuals (all their time points together) and re-runs the full ICE procedure.

```{r}
result_boot <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 20L
)
result_boot
```

The `parallel` and `ncpus` arguments enable parallel bootstrap replicates for
larger datasets:

```{r}
#| eval: false
result_par <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "bootstrap",
  n_boot = 20L,
  parallel = "multicore",
  ncpus = 4L
)
```

### Comparing sandwich and bootstrap SEs

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Point estimates and confidence intervals from sandwich and bootstrap inference for the ICE ATE estimate, showing close agreement."
se_df <- data.frame(
  method = c("Sandwich", "Bootstrap"),
  estimate = c(result$contrasts$estimate[1], result_boot$contrasts$estimate[1]),
  ci_lower = c(result$contrasts$ci_lower[1], result_boot$contrasts$ci_lower[1]),
  ci_upper = c(result$contrasts$ci_upper[1], result_boot$contrasts$ci_upper[1])
)
tinyplot(
  estimate ~ method,
  data = se_df,
  type = "pointrange",
  ymin = se_df$ci_lower,
  ymax = se_df$ci_upper,
  ylab = "ATE (always vs never)",
  xlab = "Inference method",
  main = "Sandwich vs bootstrap inference"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Dynamic interventions

ICE supports dynamic treatment regimes where treatment depends on time-varying
covariates. Here, individuals receive treatment only when L = 1 (the
confounder is present):

```{r}
result_dyn <- contrast(
  fit,
  interventions = list(
    adaptive = dynamic(\(data, trt) ifelse(!is.na(data$L) & data$L > 0, 1L, 0L)),
    always   = static(1),
    never    = static(0)
  ),
  reference = "never",
  ci_method = "sandwich"
)
result_dyn
```

### Comparing multiple interventions

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Forest plot of marginal means under three interventions: always treat, never treat, and adaptive (treat if L=1). All are close to 60 in this null-effect example."
est <- result_dyn$estimates
tinyplot(
  estimate ~ intervention,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  ylab = "E[Y^a]",
  xlab = "Intervention",
  main = "Marginal means under different interventions"
)
```

## The `history` parameter

The `history` parameter controls the Markov order --- how many lags of treatment
and time-varying confounders enter each outcome model.

- `history = 1` (default): Include one lag of A and L (first-order Markov).
- `history = 2`: Include two lags (A_{k-1}, A_{k-2}, L_{k-1}, L_{k-2}).
- `history = Inf`: Include full treatment and confounder history up to each
  time point.

For this 2-time-point example, `history = 1` and `history = Inf` are
equivalent. With longer follow-up, higher-order histories capture longer-range
dependencies but require more data per model.

```{r}
fit_inf <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  history = Inf
)

result_inf <- contrast(
  fit_inf,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "sandwich"
)
result_inf
```

## Simulation: ICE with a non-null effect

To show ICE working with a known non-null effect, we simulate data with
treatment-confounder feedback and a true ATE of 5.

```{r}
set.seed(42)
n <- 2000
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)

L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.5 + 0.5 * L0))
L1 <- rbinom(n, 1, plogis(-0.5 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.5 + 0.5 * L1))
Y <- 50 + 2.5 * A0 + 2.5 * A1 - 3 * L1 + rnorm(n, 0, 5)

sim_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_sim <- causat(
  sim_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)

res_sim <- contrast(
  fit_sim,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "sandwich"
)
res_sim
```

The estimated ATE should be close to the true value of 5.

## Binary outcome

ICE also handles binary outcomes. The first outcome model uses `binomial`
(actual 0/1 response), while subsequent pseudo-outcome models use
`quasibinomial` (fractional logistic for predicted probabilities in [0, 1]).

```{r}
sim_long_bin <- sim_long
sim_long_bin$Y[!is.na(sim_long_bin$Y)] <-
  as.integer(sim_long_bin$Y[!is.na(sim_long_bin$Y)] > 50)

fit_bin <- causat(
  sim_long_bin,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  family = "binomial",
  id = "id",
  time = "time"
)

res_bin <- contrast(
  fit_bin,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_bin
```

### Risk ratio for binary longitudinal outcomes

```{r}
res_rr <- contrast(
  fit_bin,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

## How ICE works

ICE g-computation (Zivich et al., 2024) works **backward** in time:

1. **Final time K**: Fit E[Y | A_K, L_K, baseline] among uncensored individuals.
   Predict under the intervention to get pseudo-outcomes.
2. **Each earlier time k**: Use pseudo-outcomes as the response. Fit
   E[pseudo | A_k, L_k, baseline]. Predict under intervention.
3. **First time**: The mean of the pseudo-outcomes is the marginal mean E[Y^a].

This requires **only outcome models** --- unlike forward simulation
("big g-formula"), no models for time-varying confounders are needed.

### Why ICE, not forward simulation?

| | ICE g-comp | Forward simulation |
|---|---|---|
| **Models needed** | Outcome only (at each time) | Outcome + every time-varying confounder |
| **Inference** | Sandwich (stacked EE) | Bootstrap only |
| **Speed** | Fast | Slow (many models + simulation) |
| **Model misspecification** | K models | K + p*K models |
| **Implementation** | causatr | gfoRmula |

## Key parameters

- `confounders`: Baseline (time-invariant) confounders. Enter every outcome
  model at every time step.
- `confounders_tv`: Time-varying confounders. Included with their lags
  (controlled by `history`).
- `history`: Markov order. `history = 1` (default) includes one lag of
  treatment and TV confounders. `history = Inf` includes full history.
- `censoring`: Time-varying censoring indicator (1 = censored). ICE restricts
  to uncensored at each backward step.

## Summary of covered combinations

| Outcome | Contrast | Inference | Intervention | History |
|---|---|---|---|---|
| Continuous | Difference | Sandwich | Static | 1 |
| Continuous | Difference | Bootstrap | Static | 1 |
| Continuous | Difference | Sandwich | Dynamic | 1 |
| Continuous | Difference | Sandwich | Static | Inf |
| Binary | Difference | Sandwich | Static | 1 |
| Binary | Ratio | Sandwich | Static | 1 |

## References

Hernan MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapters 19--21: Time-varying treatments.

Zivich PN, Ross RK, Shook-Sa BE, Cole SR, Edwards JK (2024). Empirical
sandwich variance estimator for iterated conditional expectation
g-computation. *Statistics in Medicine* 43:5562--5572.
