---
title: "Longitudinal treatments: ICE g-computation (Chapter 21)"
vignette: >
  %\VignetteIndexEntry{Longitudinal treatments: ICE g-computation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When treatment and confounders vary over time, standard regression methods give
biased estimates even when adjusted for all confounders. G-methods solve this.
causatr implements ICE (iterated conditional expectation) g-computation (Zivich
et al., 2024), which requires models for the outcome only --- no models for
time-varying confounders are needed.

This vignette demonstrates ICE g-computation using the classic Table 20.1
example from Hernán & Robins (2025), where treatment-confounder feedback
produces a null effect that naive methods cannot recover.

## The problem: treatment-confounder feedback

When treatment at time 0 affects a confounder at time 1, which in turn affects
treatment at time 1 (A_0 -> L_1 -> A_1 -> Y), neither adjusting nor not
adjusting for L_1 gives the correct answer:

- **Adjusting** for L_1 opens a collider path (selection bias)
- **Not adjusting** leaves confounding for A_1
- **ICE g-computation** handles both correctly

## Setup: Table 20.1

```{r}
#| message: false
library(causatr)
```

We construct the dataset from Table 20.1 of Hernán & Robins (2025), which has
32,000 individuals observed at 2 time points with a true causal effect of zero.

```{r}
groups <- data.frame(
  A0 = c(0, 0, 0, 0, 1, 1, 1, 1),
  L1 = c(0, 0, 1, 1, 0, 0, 1, 1),
  A1 = c(0, 1, 0, 1, 0, 1, 0, 1),
  Y  = c(84, 84, 52, 52, 76, 76, 44, 44),
  N  = c(2400, 1600, 2400, 9600, 4800, 3200, 1600, 6400)
)
rows <- lapply(seq_len(nrow(groups)), function(i) {
  g <- groups[i, ]
  off <- sum(groups$N[seq_len(i - 1)])
  data.frame(id = seq_len(g$N) + off, A0 = g$A0, L1 = g$L1, A1 = g$A1, Y = g$Y)
})
wide <- do.call(rbind, rows)

t0 <- data.frame(id = wide$id, time = 0L, A = wide$A0, L = NA_real_, Y = NA_real_)
t1 <- data.frame(id = wide$id, time = 1L, A = wide$A1, L = wide$L1, Y = wide$Y)
long <- rbind(t0, t1)
head(long, 4)
```

The data is in **person-period (long) format**: one row per individual per
time point. `L` is a time-varying confounder (only measured at time 1).

## Demonstrating bias from naive methods

A naive regression of Y on treatment adjusting for L_1 gives a biased estimate:

```{r}
wide_sub <- wide[!is.na(wide$Y), ]
coef(lm(Y ~ A0 + A1 + L1, data = wide_sub))[c("A0", "A1")]
```

The true individual effects of A_0 and A_1 are both zero, but the naive
regression gives non-zero estimates due to treatment-confounder feedback.

## ICE g-computation: fitting

With `causat()`, specify `id` and `time` to trigger longitudinal mode. Baseline
confounders go in `confounders`, time-varying confounders in `confounders_tv`.

```{r}
fit <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)
fit
```

No models are fitted at this stage. ICE defers model fitting to `contrast()`
because the sequential outcome models depend on the intervention being
evaluated.

## ICE g-computation: contrasts

Compare "always treat" (A = 1 at every time) vs "never treat" (A = 0 at every
time):

```{r}
result <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
result
```

ICE correctly recovers **E[Y^{always}] = E[Y^{never}] = 60** and **ATE = 0**,
as expected from the data generating process.

## Variance estimation

### Sandwich SE (stacked estimating equations)

The default `ci_method = "sandwich"` uses the empirical sandwich estimator for
stacked estimating equations (Zivich et al., 2024). This accounts for
uncertainty from all K sequential outcome models, not just the final model.

```{r}
result$estimates[, c("intervention", "estimate", "se")]
```

### Bootstrap SE

For robustness or when using flexible models (e.g., GAMs), bootstrap resamples
individuals (all their time points together) and re-runs the full ICE procedure.
The `parallel` and `ncpus` arguments enable parallel bootstrap replicates:

```{r}
#| eval: false
result_boot <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 500L,
  parallel = "multicore",
  ncpus = 4L
)
```

## Dynamic interventions

ICE supports dynamic treatment regimes where treatment depends on time-varying
covariates. Here, individuals receive treatment only when L_1 = 1:

```{r}
result_dyn <- contrast(
  fit,
  interventions = list(
    adaptive = dynamic(\(data, trt) ifelse(!is.na(data$L) & data$L > 0, 1L, 0L)),
    never    = static(0)
  ),
  reference = "never",
  ci_method = "sandwich"
)
result_dyn
```

## How ICE works

ICE g-computation (Zivich et al., 2024) works **backward** in time:

1. **Final time**: Fit E[Y | A_K, L_K, baseline] among uncensored individuals.
   Predict under the intervention to get pseudo-outcomes.
2. **Each earlier time**: Use pseudo-outcomes as the response. Fit
   E[pseudo | A_k, L_k, baseline]. Predict under intervention.
3. **First time**: The mean of the pseudo-outcomes is the marginal mean E[Y^a].

This requires **only outcome models** --- unlike forward simulation
("big g-formula"), no models for time-varying confounders are needed.

## Key parameters

- `confounders`: Baseline (time-invariant) confounders. Enter every outcome
  model at every time step.
- `confounders_tv`: Time-varying confounders. Included with their lags
  (controlled by `history`).
- `history`: Markov order. `history = 1` (default) includes one lag of
  treatment and TV confounders. `history = Inf` includes full history.
- `censoring`: Time-varying censoring indicator (1 = censored). ICE restricts
  to uncensored at each backward step.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapters 19--21: Time-varying treatments.

Zivich PN, Ross RK, Shook-Sa BE, Cole SR, Edwards JK (2024). Empirical
sandwich variance estimator for iterated conditional expectation
g-computation. *Statistics in Medicine* 43:5562--5572.
