---
title: "Longitudinal treatments: ICE g-computation"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Longitudinal treatments: ICE g-computation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When treatment and confounders vary over time, standard regression methods give
biased estimates even when adjusted for all confounders. G-methods solve this.
causatr implements ICE (iterated conditional expectation) g-computation (Zivich
et al., 2024), which requires models for the outcome only --- no models for
time-varying confounders are needed.

This vignette demonstrates ICE g-computation in causatr, covering:

- The treatment-confounder feedback problem
- ICE fitting and contrasting
- Sandwich and bootstrap inference
- Static, dynamic, and modified treatment policy interventions
- The `history` parameter (Markov order)
- Binary outcomes
- Comparison with naive regression

## Setup

```{r}
#| message: false
library(causatr)
library(tinytable)
library(tinyplot)
```

## The problem: treatment-confounder feedback

When $A_0$ affects a time-varying confounder $L_1$, and $L_1$ in turn affects
both $A_1$ and $Y$ ($A_0 \to L_1 \to A_1$ and $L_1 \to Y$), ordinary
regression cannot recover the marginal counterfactual mean
$\mathbb{E}[Y^{a_0, a_1}]$:

- **Adjusting** for $L_1$ blocks part of the causal path from $A_0$ to $Y$
  (and, depending on the DAG, can open a collider path through unmeasured
  common causes of $L_1$ and $Y$).
- **Not adjusting** leaves $L_1$-confounding of the $A_1 \to Y$ arrow.
- **ICE g-computation** handles both correctly by iterating conditional
  expectations and marginalising over the $L_1$ distribution induced by each
  intervention.

## Table 20.1: the null paradox

We reconstruct the dataset from Table 20.1 of Hernan & Robins (2025), scaled
to 3,200 individuals observed at 2 time points. (The original table has
32,000; we use 1/10th for faster vignette rendering.) The DGP has a very
specific property: **every joint regime $(a_0, a_1)$ has the same marginal
counterfactual mean $\mathbb{E}[Y^{a_0, a_1}] = 60$**, so the true causal
contrast between any two regimes is exactly zero — the "null is true".
Crucially, the null holds only marginally: within strata of $L_1$, the
*conditional* effect of $A_0$ is $-8$. Treatment-confounder feedback is
exactly the mechanism that lets a non-zero conditional effect average out
to a zero marginal effect.

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k \in \{0, 1\}$,
binary time-varying confounder $L_1 \in \{0, 1\}$ (affected by $A_0$), and
continuous outcome $Y$ measured at $k = 1$. The causal structure is
$A_0 \to L_1 \to A_1$, $A_0 \to L_1 \to Y$, and $A_0 \to Y$ (direct).
The table is constructed so that the conditional effect of $A_0$ given $L_1$
is $-8$, but this conditional effect averages to exactly zero marginally
because $A_0$ shifts the $L_1$ distribution in a perfectly offsetting way.
The structural equations implied by the table are:

$$
\begin{aligned}
A_0 &\sim \text{Bernoulli}(p_{A_0}), \\
L_1 \mid A_0 &\sim \text{Bernoulli}\bigl(P(L_1 = 1 \mid A_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(P(A_1 = 1 \mid L_1)\bigr), \\
Y &= 84 - 8 \cdot A_0 - 32 \cdot L_1.
\end{aligned}
$$

The outcome is deterministic given $(A_0, L_1)$ and does not depend on $A_1$
directly. The key property: $P(L_1 = 1 \mid A_0 = 1) > P(L_1 = 1 \mid A_0 = 0)$,
so $A_0$ increases $L_1$, and since $L_1$ has a strong negative effect on $Y$,
the indirect harm through $L_1$ exactly cancels the direct benefit.
The known true value is $\mathbb{E}[Y^{a_0, a_1}] = 60$ for all four joint
regimes, so the ATE between any two regimes is exactly 0.

```{r}
groups <- data.frame(
  A0 = c(0, 0, 0, 0, 1, 1, 1, 1),
  L1 = c(0, 0, 1, 1, 0, 0, 1, 1),
  A1 = c(0, 1, 0, 1, 0, 1, 0, 1),
  Y  = c(84, 84, 52, 52, 76, 76, 44, 44),
  N  = c(240, 160, 240, 960, 480, 320, 160, 640)
)
rows <- lapply(seq_len(nrow(groups)), function(i) {
  g <- groups[i, ]
  off <- sum(groups$N[seq_len(i - 1)])
  data.frame(id = seq_len(g$N) + off, A0 = g$A0, L1 = g$L1, A1 = g$A1, Y = g$Y)
})
wide <- do.call(rbind, rows)

t0 <- data.frame(id = wide$id, time = 0L, A = wide$A0, L = NA_real_, Y = NA_real_)
t1 <- data.frame(id = wide$id, time = 1L, A = wide$A1, L = wide$L1, Y = wide$Y)
long <- rbind(t0, t1)
head(long, 4)
```

The data is in **person-period (long) format**: one row per individual per
time point. `L` is a time-varying confounder (only measured at time 1).

## Demonstrating bias from naive methods

Hernan & Robins (2025, Ch. 20) uses this table to make a pointed argument:
*neither* of the two conventional regression strategies recovers the null.
Truth is $\mathbb{E}[Y^{1,1}] - \mathbb{E}[Y^{0,0}] = 0$, and in fact all
four joint regimes have the same marginal counterfactual mean of $60$.

**Strategy 1 — unadjusted.** Drop $L_1$ and hope randomisation at each
time point takes care of confounding:

```{r}
wide_sub <- wide[!is.na(wide$Y), ]
naive_unadj <- lm(Y ~ A0 + A1, data = wide_sub)
round(coef(naive_unadj), 3)
```

Both coefficients move away from zero ($A_0 \approx -1.24$,
$A_1 \approx -12.44$). The $A_1$ estimate is especially biased:
$L_1$ confounds $A_1 \to Y$, and without adjustment that confounding
leaks directly into the coefficient.

**Strategy 2 — adjust for $L_1$.** The standard epidemiological reflex:

```{r}
naive_fit   <- lm(Y ~ A0 + A1 + L1, data = wide_sub)
naive_coefs <- coef(naive_fit)
round(naive_coefs, 3)
```

Now $A_1 = 0$ (correct, because in this DGP $Y$ is a deterministic
function of $A_0$ and $L_1$), but $A_0 = -8$ — the *conditional* effect of
$A_0$ within strata of $L_1$. That is **not** the causal contrast we want.
Adjusting for $L_1$ blocks the $A_0 \to L_1 \to Y$ path and leaves only
the direct-effect residue, so $A_0$ looks like a protective effect when
in fact the marginal causal effect is zero.

**The takeaway.** One naive strategy biases $A_1$; the other biases $A_0$.
There is no regression specification on this table that jointly returns
$A_0 = 0$ and $A_1 = 0$. This is the *null paradox* that motivates
g-methods: the only way to recover the null is to marginalise over the
intervention-specific distribution of $L_1$, which is exactly what ICE
g-computation does below.

## ICE g-computation: fitting

With `causat()`, specify `id` and `time` to trigger longitudinal mode. Baseline
confounders go in `confounders`, time-varying confounders in `confounders_tv`.

```{r}
fit <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)
fit
```

No models are fitted at this stage. ICE defers model fitting to `contrast()`
because the sequential outcome models depend on the intervention being
evaluated.

## ICE g-computation: contrasts with sandwich SE

Compare "always treat" (A = 1 at every time) vs "never treat" (A = 0 at every
time):

```{r}
result <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
result
```

ICE correctly recovers **E[Y^{always}] = E[Y^{never}] = 60** and **ATE = 0**,
as expected from the data generating process.

### Visualising the result

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Point estimates and confidence intervals comparing naive regression to ICE g-computation. Naive estimates are biased away from zero while ICE correctly recovers ATE = 0."
naive_ci <- confint(naive_fit)
est_df <- data.frame(
  method = c("Naive (A0)", "Naive (A1)", "ICE"),
  estimate = c(naive_coefs["A0"], naive_coefs["A1"], result$contrasts$estimate[1]),
  ci_lower = c(naive_ci["A0", 1], naive_ci["A1", 1], result$contrasts$ci_lower[1]),
  ci_upper = c(naive_ci["A0", 2], naive_ci["A1", 2], result$contrasts$ci_upper[1])
)
tinyplot(
  estimate ~ method,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  ylab = "Estimated treatment effect",
  xlab = "",
  main = "ICE recovers the null effect"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Sandwich SE: stacked estimating equations

The default `ci_method = "sandwich"` uses the empirical sandwich estimator for
stacked estimating equations (Zivich et al., 2024). This accounts for
uncertainty from all K sequential outcome models, not just the final model.

```{r}
result$estimates[, c("intervention", "estimate", "se")]
```

The standard errors are finite and positive, confirming that the sandwich
variance estimator is working correctly even with a large sample.

## Bootstrap SE

For robustness or when using flexible models (e.g., GAMs), bootstrap resamples
individuals (all their time points together) and re-runs the full ICE procedure.

```{r}
result_boot <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 20L
)
result_boot
```

The `parallel` and `ncpus` arguments enable parallel bootstrap replicates for
larger datasets:

```{r}
#| eval: false
result_par <- contrast(
  fit,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "bootstrap",
  n_boot = 20L,
  parallel = "multicore",
  ncpus = 4L
)
```

### Comparing sandwich and bootstrap SEs

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-alt: "Point estimates and confidence intervals from sandwich and bootstrap inference for the ICE ATE estimate, showing close agreement."
se_df <- data.frame(
  method = c("Sandwich", "Bootstrap"),
  estimate = c(result$contrasts$estimate[1], result_boot$contrasts$estimate[1]),
  ci_lower = c(result$contrasts$ci_lower[1], result_boot$contrasts$ci_lower[1]),
  ci_upper = c(result$contrasts$ci_upper[1], result_boot$contrasts$ci_upper[1])
)
tinyplot(
  estimate ~ method,
  data = se_df,
  type = "pointrange",
  ymin = se_df$ci_lower,
  ymax = se_df$ci_upper,
  ylab = "ATE (always vs never)",
  xlab = "Inference method",
  main = "Sandwich vs bootstrap inference"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Dynamic interventions

ICE supports dynamic treatment regimes where treatment depends on time-varying
covariates. Here, individuals receive treatment only when L = 1 (the
confounder is present):

```{r}
result_dyn <- contrast(
  fit,
  interventions = list(
    adaptive = dynamic(\(data, trt) ifelse(!is.na(data$L) & data$L > 0, 1L, 0L)),
    always   = static(1),
    never    = static(0)
  ),
  reference = "never",
  ci_method = "sandwich"
)
result_dyn
```

### Comparing multiple interventions

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Forest plot of marginal means under three interventions: always treat, never treat, and adaptive (treat if L=1). All are close to 60 in this null-effect example."
est <- result_dyn$estimates
tinyplot(
  estimate ~ intervention,
  data = est,
  type = "pointrange",
  ymin = est$ci_lower,
  ymax = est$ci_upper,
  ylab = "E[Y^a]",
  xlab = "Intervention",
  main = "Marginal means under different interventions"
)
abline(h = 60, lty = 2, col = "grey40")
```

All three interventions should return a marginal mean close to 60 (the
structural truth for every joint regime in this null-effect DGP).

## The `history` parameter

The `history` parameter controls the Markov order --- how many lags of treatment
and time-varying confounders enter each outcome model.

- `history = 1` (default): Include one lag of A and L (first-order Markov).
- `history = 2`: Include two lags (A_{k-1}, A_{k-2}, L_{k-1}, L_{k-2}).
- `history = Inf`: Include full treatment and confounder history up to each
  time point.

For this 2-time-point example, `history = 1` and `history = Inf` are
equivalent. With longer follow-up, higher-order histories capture longer-range
dependencies but require more data per model.

```{r}
fit_inf <- causat(
  long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  history = Inf
)

result_inf <- contrast(
  fit_inf,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "sandwich"
)
result_inf
```

## Simulation: ICE with a non-null effect

To show ICE working with a known non-null effect, we simulate data with
treatment-confounder feedback.

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k \in \{0, 1\}$,
binary time-varying confounder $L_k \in \{0, 1\}$, and continuous outcome $Y$.
Treatment assignment at each period depends on the current confounder via a
logistic model; the confounder at $k = 1$ is affected by prior treatment $A_0$
(treatment-confounder feedback). The structural equations are:

$$
\begin{aligned}
L_0 &\sim \text{Bernoulli}(0.5), \\
A_0 \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.5 + 0.5 \, L_0)\bigr), \\
L_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.5 + 0.8 \, A_0 + 0.3 \, L_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.5 + 0.5 \, L_1)\bigr), \\
Y \mid A_0, A_1, L_1 &= 50 + 2.5 \, A_0 + 2.5 \, A_1 - 3 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 25).
\end{aligned}
$$

The structural ATE of "always treat" vs "never treat" is **not** simply
$2.5 + 2.5 = 5$, because $A_0$ affects $L_1$ (coefficient $0.8$), and $L_1$
enters the outcome with coefficient $-3$. Under "always treat", $A_0 = 1$
increases $\mathbb{E}[L_1]$, and since $L_1$ has a negative effect on $Y$,
this indirect path partially offsets the direct benefit:

$$
\text{ATE}
  = \underbrace{(2.5 + 2.5)}_{\text{direct}}
  - 3 \bigl(\mathbb{E}[L_1 \mid A_0\!=\!1] - \mathbb{E}[L_1 \mid A_0\!=\!0]\bigr)
  \approx 5 - 3 \times 0.196
  \approx 4.41.
$$

This is the same treatment-confounder feedback that makes naive regression
biased.

```{r}
set.seed(42)
n <- 2000
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)

L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.5 + 0.5 * L0))
L1 <- rbinom(n, 1, plogis(-0.5 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.5 + 0.5 * L1))
Y <- 50 + 2.5 * A0 + 2.5 * A1 - 3 * L1 + rnorm(n, 0, 5)

sim_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_sim <- causat(
  sim_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)

res_sim <- contrast(
  fit_sim,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  ci_method = "sandwich"
)
res_sim
```

The estimated ATE should be close to the true value of $\approx 4.41$.

## Binary outcome

ICE also handles binary outcomes. The first outcome model uses `binomial`
(actual 0/1 response), while subsequent pseudo-outcome models use
`quasibinomial` (fractional logistic for predicted probabilities in [0, 1]).

**DGM.** This reuses the non-null simulation above (same SEM) with the
continuous outcome binarised at $Y > 50$:

$$
\begin{aligned}
Y^* &= 50 + 2.5 \, A_0 + 2.5 \, A_1 - 3 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 25), \\
Y &= \mathbf{1}(Y^* > 50).
\end{aligned}
$$

The resulting binary outcome does not have a clean closed-form ATE on the
risk-difference scale, but the direction is positive ("always treat" increases
$P(Y = 1)$ relative to "never treat") because both treatment coefficients are
positive in the latent linear predictor.

```{r}
sim_long_bin <- sim_long
sim_long_bin$Y[!is.na(sim_long_bin$Y)] <-
  as.integer(sim_long_bin$Y[!is.na(sim_long_bin$Y)] > 50)

fit_bin <- causat(
  sim_long_bin,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  family = "binomial",
  id = "id",
  time = "time"
)

res_bin <- contrast(
  fit_bin,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_bin
```

### Risk ratio for binary longitudinal outcomes

```{r}
res_rr <- contrast(
  fit_bin,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

## Three or more time periods

The examples above use 2 time points for simplicity, but ICE scales to
any number of periods. Here we simulate 4-period data with a binary treatment
and binary outcome. The backward recursion fits 4 sequential models.

**DGM.** Four periods ($k = 1, \dots, 4$) with binary treatment $A_k$,
continuous time-varying confounder $L_k$, continuous baseline confounder $L_0$,
and binary outcome $Y$ measured at $k = 4$. The confounder evolves with
treatment feedback: at each step, $L_k$ depends on $L_{k-1}$ plus noise, and
after observing $A_k$ the next confounder shifts by $0.3 \, A_k$. The
structural equations are:

$$
\begin{aligned}
L_0 &\sim N(0, 1), \\
L_k \mid L_{k-1}, A_{k-1} &= L_{k-1} + 0.3 \, A_{k-1} + \eta_k, \quad \eta_k \sim N(0, 0.25), \\
A_k \mid L_k &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.3 + 0.4 \, L_k)\bigr), \\
Y \mid A_1, \dots, A_4, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-1.5 + 0.4 \, \textstyle\sum_{k=1}^{4} A_k + 0.3 \, L_0)\bigr).
\end{aligned}
$$

The outcome depends on cumulative treatment exposure $\sum_k A_k$ and the
baseline confounder $L_0$. Under "always treat" ($\sum A_k = 4$) the
log-odds shift by $0.4 \times 4 = 1.6$ relative to "never treat"
($\sum A_k = 0$), so the risk difference on the probability scale is positive
but does not have a simple closed form (it depends on the marginal distribution
of $L_0$).

```{r}
set.seed(99)
n_id <- 600
K <- 4

ids <- rep(seq_len(n_id), each = K)
times <- rep(seq_len(K), n_id)

L0 <- rep(rnorm(n_id), each = K)
sim_4p <- data.frame(id = ids, time = times, L0 = L0)

sim_4p$L <- NA_real_
sim_4p$A <- NA_integer_
for (i in seq_len(n_id)) {
  rows <- which(sim_4p$id == i)
  l_prev <- sim_4p$L0[rows[1]]
  for (k in seq_along(rows)) {
    sim_4p$L[rows[k]] <- l_prev + rnorm(1, 0, 0.5)
    sim_4p$A[rows[k]] <- rbinom(1, 1, plogis(-0.3 + 0.4 * sim_4p$L[rows[k]]))
    l_prev <- sim_4p$L[rows[k]] + 0.3 * sim_4p$A[rows[k]]
  }
}

sim_4p$Y <- NA_real_
for (i in seq_len(n_id)) {
  rows <- which(sim_4p$id == i)
  cum_a <- sum(sim_4p$A[rows])
  sim_4p$Y[rows[K]] <- rbinom(
    1, 1,
    plogis(-1.5 + 0.4 * cum_a + 0.3 * sim_4p$L0[rows[1]])
  )
}

fit_4p <- causat(
  sim_4p,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L0,
  confounders_tv = ~ L,
  family = "binomial",
  id = "id",
  time = "time"
)

res_4p <- contrast(
  fit_4p,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_4p
```

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot showing the risk difference between always-treat and never-treat over 4 periods."
#| eval: !expr requireNamespace("forrest", quietly = TRUE)
plot(res_4p)
```

```{r}
tt(tidy(res_4p), digits = 3)
```

The ICE sandwich variance correctly propagates uncertainty through all 4
sequential models via the stacked estimating equations cascade.

## Continuous treatment with modified treatment policies

ICE also supports continuous time-varying treatments combined with the
modified-treatment-policy (MTP) interventions: `shift()`, `scale_by()`, and
`threshold()`. The intervention is applied to the current-time treatment
at each backward step; `ice_apply_intervention_long()` explicitly does
**not** recompute the lag columns from the intervened values (this would
double-count a non-static intervention — see `CLAUDE.md` for the
structural reason).

**DGM.** Two periods ($k = 0, 1$) with continuous treatment $A_k \in \mathbb{R}$,
continuous confounder $L_k$, and continuous outcome $Y$. Treatment-confounder
feedback is present: $A_0$ affects $L_1$ (coefficient $0.5$), and $L_1$ affects
$A_1$ (coefficient $0.5$). The structural equations are:

$$
\begin{aligned}
L_0 &\sim N(0, 1), \\
A_0 &= 1 + 0.4 \, L_0 + \eta_0, \quad \eta_0 \sim N(0, 1), \\
L_1 &= 0.3 \, L_0 + 0.5 \, A_0 + \eta_1, \quad \eta_1 \sim N(0, 1), \\
A_1 &= 1 + 0.5 \, L_1 + \eta_2, \quad \eta_2 \sim N(0, 1), \\
Y &= 2 + 1.5 \, A_0 + 2 \, A_1 + 0.5 \, L_0 - 0.3 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1).
\end{aligned}
$$

For a `shift(-1)` MTP (reduce each $A_k$ by 1 at each period), the
structural causal effect on $Y$ relative to the natural course accounts for
both direct and indirect paths. Shifting $A_0$ by $-1$ also shifts
$L_1$ by $-0.5$ (coefficient $0.5$ in the $L_1$ equation), which in turn
shifts the natural value of $A_1$ by $-0.25$ (coefficient $0.5$ in the
$A_1$ equation). The shifted $A_1$ is then reduced by a further $-1$ by the
MTP. The total effect on $Y$ is:

$$
1.5 \times (-1) + 2 \times (-1.25) + (-0.3) \times (-0.5) = -1.5 - 2.5 + 0.15 = -3.85.
$$

```{r}
set.seed(101)
n <- 800
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)

L0 <- rnorm(n)
A0 <- 1 + 0.4 * L0 + rnorm(n)
L1 <- 0.3 * L0 + 0.5 * A0 + rnorm(n)
A1 <- 1 + 0.5 * L1 + rnorm(n)
Y <- 2 + 1.5 * A0 + 2 * A1 + 0.5 * L0 - 0.3 * L1 + rnorm(n)

cont_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_cont <- causat(
  cont_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)

res_mtp <- contrast(
  fit_cont,
  interventions = list(
    observed = NULL,
    shift_m1 = shift(-1),
    halved   = scale_by(0.5),
    capped   = threshold(0, 1)
  ),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_mtp
```

Structural truth for `shift(-1)` is $-3.85$ (see derivation above). The
`shift_m1` estimate should be close to this value.

## Multivariate time-varying treatment

When multiple treatments vary over time (e.g. two drugs dosed jointly),
pass a character vector to `treatment =` and supply each intervention as
a named sub-list. Lag columns are built per-component by `prepare_data()`.

**DGM.** Two periods ($k = 0, 1$) with two binary treatments
$A_{1,k}, A_{2,k} \in \{0, 1\}$, a continuous confounder $L_k$, and
continuous outcome $Y$. Both treatments at $k = 0$ feed into the confounder
$L_1$, creating treatment-confounder feedback. The structural equations are:

$$
\begin{aligned}
L_0 &\sim N(0, 1), \\
A_{1,0} \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3 \, L_0)\bigr), \\
A_{2,0} \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.1 \, L_0)\bigr), \\
L_1 &= 0.3 \, L_0 + 0.4 \, A_{1,0} + 0.2 \, A_{2,0} + \eta, \quad \eta \sim N(0, 1), \\
A_{1,1} \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3 \, L_1)\bigr), \\
A_{2,1} \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.1 \, L_1)\bigr), \\
Y &= 2 + 1.5 \, A_{1,0} + 1.5 \, A_{1,1} + 0.8 \, A_{2,0} + 0.8 \, A_{2,1} + 0.5 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1).
\end{aligned}
$$

The outcome is linear in both treatments at both periods. The structural ATE
of "both on" vs "both off" includes the indirect path through $L_1$: setting
$A_{1,0} = A_{2,0} = 1$ shifts $\mathbb{E}[L_1]$ by $0.4 + 0.2 = 0.6$
relative to "both off", and $L_1$ enters $Y$ with coefficient $0.5$, adding
$0.3$ to the direct sum $2 \times (1.5 + 0.8) = 4.6$:

$$
\text{ATE} = \underbrace{4.6}_{\text{direct}} + \underbrace{0.5 \times 0.6}_{\text{indirect via } L_1} = 4.9.
$$

```{r}
set.seed(202)
n <- 600
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)
L0 <- rnorm(n)
A10 <- rbinom(n, 1, plogis(0.3 * L0))
A20 <- rbinom(n, 1, plogis(0.1 * L0))
L1 <- 0.3 * L0 + 0.4 * A10 + 0.2 * A20 + rnorm(n)
A11 <- rbinom(n, 1, plogis(0.3 * L1))
A21 <- rbinom(n, 1, plogis(0.1 * L1))
Y <- 2 + 1.5 * A10 + 1.5 * A11 + 0.8 * A20 + 0.8 * A21 + 0.5 * L1 + rnorm(n)

mv_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A1 = c(rbind(A10, A11)),
  A2 = c(rbind(A20, A21)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_mv_long <- causat(
  mv_long,
  outcome = "Y",
  treatment = c("A1", "A2"),
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)

res_mv_long <- contrast(
  fit_mv_long,
  interventions = list(
    both_on  = list(A1 = static(1), A2 = static(1)),
    both_off = list(A1 = static(0), A2 = static(0))
  ),
  reference = "both_off",
  type = "difference"
)
res_mv_long
```

Structural truth: `both_on − both_off = 4.9` (direct effects $4.6$ plus
indirect $0.3$ via $L_1$).

## External weights (IPCW / survey)

External weights — survey weights or pre-computed IPCW — pass through
every backward-step model via `model_fn`'s `weights =` argument. The ICE
sandwich variance engine accounts for heterogeneous weights via the
`A_{k-1,k}` cascade gradient (see `variance-theory.qmd` §5.4 and the
inline comment in `R/variance_if.R:variance_if_ice_one`).

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k$, binary
confounder $L_k$, continuous outcome $Y$, and uninformative survey weights
drawn independently from $\text{Uniform}(0.5, 2.0)$. The structural equations
are:

$$
\begin{aligned}
L_0 &\sim \text{Bernoulli}(0.5), \\
A_0 \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_0)\bigr), \\
L_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.8 \, A_0 + 0.3 \, L_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_1)\bigr), \\
Y &= 2 + 2.5 \, A_0 + 2.5 \, A_1 - 0.5 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1), \\
w_i &\sim \text{Uniform}(0.5, 2.0).
\end{aligned}
$$

Since the weights are non-informative (independent of everything), the
structural ATE of "always treat" vs "never treat" is the same as the
unweighted case. The direct effects sum to $2.5 + 2.5 = 5$, but $A_0$
shifts $\mathbb{E}[L_1]$ by $\approx 0.19$, and $L_1$ enters $Y$ with
coefficient $-0.5$, giving a true ATE $\approx 4.90$. The purpose of this
example is to demonstrate that the ICE sandwich variance engine correctly
propagates heterogeneous weights through the backward recursion.

```{r}
set.seed(303)
n <- 500
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)
L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L0))
L1 <- rbinom(n, 1, plogis(-0.2 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L1))
Y <- 2 + 2.5 * A0 + 2.5 * A1 - 0.5 * L1 + rnorm(n)
weights_long <- runif(2 * n, 0.5, 2.0)

w_long_df <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  Y = c(rbind(NA_real_, Y))
)

fit_w <- causat(
  w_long_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  weights = weights_long
)

res_w <- contrast(
  fit_w,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_w
```

## Censoring (IPCW-style row filter)

Pass a time-varying censoring indicator via `censoring =`. At each
backward step, ICE restricts the fit to uncensored rows; predictions are
made for the target population regardless of censoring status, so the
estimator is valid under the usual sequential-randomization assumption
on the censoring mechanism.

**DGM.** Two periods ($k = 0, 1$) with binary treatment $A_k$, binary
confounder $L_k$, continuous outcome $Y$, and a censoring indicator $C_1$ at
$k = 1$ that depends on treatment and baseline confounder. About 20% of
individuals are censored at the second period. The structural equations are:

$$
\begin{aligned}
L_0 &\sim \text{Bernoulli}(0.5), \\
A_0 \mid L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_0)\bigr), \\
L_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.8 \, A_0 + 0.3 \, L_0)\bigr), \\
A_1 \mid L_1 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-0.2 + 0.6 \, L_1)\bigr), \\
Y^* &= 2 + 2 \, A_0 + 2 \, A_1 - 0.5 \, L_1 + \varepsilon, \quad \varepsilon \sim N(0, 1), \\
C_1 \mid A_0, L_0 &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-2 + 0.5 \, A_0 + 0.3 \, L_0)\bigr), \\
Y &= \begin{cases} Y^* & \text{if } C_1 = 0, \\ \text{NA} & \text{if } C_1 = 1. \end{cases}
\end{aligned}
$$

The structural ATE of "always treat" vs "never treat" on the full
(uncensored) outcome is $\approx 3.90$ (direct effects $2 + 2 = 4$ minus
$0.5 \times 0.19 \approx 0.10$ for the indirect path $A_0 \to L_1 \to Y$).
Under the censoring-at-random assumption
($Y^{a_0, a_1} \perp C_1 \mid A_0, L_0$), ICE with row filtering recovers
this effect by fitting the outcome model on uncensored rows and standardising
over the full population.

```{r}
set.seed(404)
n <- 500
sim_id <- rep(seq_len(n), each = 2)
sim_time <- rep(0:1, n)
L0 <- rbinom(n, 1, 0.5)
A0 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L0))
L1 <- rbinom(n, 1, plogis(-0.2 + 0.8 * A0 + 0.3 * L0))
A1 <- rbinom(n, 1, plogis(-0.2 + 0.6 * L1))
Y_full <- 2 + 2 * A0 + 2 * A1 - 0.5 * L1 + rnorm(n)
# 20% censoring at t = 1 depending on A0 + L0.
C1 <- rbinom(n, 1, plogis(-2 + 0.5 * A0 + 0.3 * L0))
Y <- ifelse(C1 == 1, NA_real_, Y_full)

cens_long <- data.frame(
  id = sim_id,
  time = sim_time,
  A = c(rbind(A0, A1)),
  L = c(rbind(L0, L1)),
  C = c(rbind(0L, C1)),
  Y = c(rbind(NA_real_, Y))
)

fit_c <- causat(
  cens_long,
  outcome = "Y",
  treatment = "A",
  confounders = ~ 1,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  censoring = "C"
)

res_c <- contrast(
  fit_c,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_c
```

Structural truth $\approx 3.90$ (see derivation above). The estimate should
be close, and the sandwich SE reflects the censoring-induced reduction in
effective sample size at each step.

## How ICE works

ICE g-computation (Zivich et al., 2024) works **backward** in time:

1. **Final time K**: Fit E[Y | A_K, L_K, baseline] among uncensored individuals.
   Predict under the intervention to get pseudo-outcomes.
2. **Each earlier time k**: Use pseudo-outcomes as the response. Fit
   E[pseudo | A_k, L_k, baseline]. Predict under intervention.
3. **First time**: The mean of the pseudo-outcomes is the marginal mean E[Y^a].

This requires **only outcome models** --- unlike forward simulation
("big g-formula"), no models for time-varying confounders are needed.

### Why ICE, not forward simulation?

| | ICE g-comp | Forward simulation |
|---|---|---|
| **Models needed** | Outcome only (at each time) | Outcome + every time-varying confounder |
| **Inference** | Sandwich (stacked EE) | Bootstrap only |
| **Speed** | Fast | Slow (many models + simulation) |
| **Model misspecification** | K models | K + p*K models |
| **Implementation** | causatr | gfoRmula |

## Key parameters

- `confounders`: Baseline (time-invariant) confounders. Enter every outcome
  model at every time step.
- `confounders_tv`: Time-varying confounders. Included with their lags
  (controlled by `history`).
- `history`: Markov order. `history = 1` (default) includes one lag of
  treatment and TV confounders. `history = Inf` includes full history.
- `censoring`: Time-varying censoring indicator (1 = censored). ICE restricts
  to uncensored at each backward step; the resulting estimator is IPCW-valid
  when the censoring mechanism is conditionally random given the covariate
  history.
- `weights`: external survey weights or pre-computed IPCW. Validated upfront
  by `check_weights()` (rejects NA, Inf, negative, mis-sized) and propagated
  through every step of the backward recursion. The ICE sandwich variance
  engine accounts for the weights via the `A_{k-1,k}` cascade gradient —
  dropping them would underestimate the SE under heterogeneous weights by
  roughly a factor of two.
- `parallel` / `ncpus`: passed straight through to `boot::boot()` on the
  bootstrap path. Defaults to `getOption("boot.parallel", "no")` /
  `getOption("boot.ncpus", 1L)` so a session-wide `options(boot.parallel =
  "multicore")` flips parallelism on for every `contrast()` call.

For a full enumeration of the sandwich-vs-bootstrap, continuous/binary,
external-weights, censoring, multivariate-treatment, and ipsi-rejection
combinations, see `FEATURE_COVERAGE_MATRIX.md` §"Method 4 — ICE longitudinal
g-computation" (20+ truth-pinned rows against `lmtp::lmtp_tmle` / analytic
targets / Hernán & Robins Table 20.1).

## Summary of covered combinations

**Legend.** ✅ covered and truth-pinned in tests (either against Table 20.1
or against `lmtp::lmtp_tmle()`) · 🟡 smoke test only · ⛔ rejected with an
informative error.

| Treatment | Outcome | Intervention | Estimand | Contrast | Inference | Weights | Status |
|---|---|---|---|---|---|---|---|
| Binary | Continuous | Static (always vs never) | ATE | Difference | Sandwich | none | ✅ Table 20.1 |
| Binary | Continuous | Static | ATE | Difference | Bootstrap | none | ✅ Table 20.1 |
| Binary | Continuous | Dynamic (rule on L) | ATE | Difference | Sandwich | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich + bootstrap | survey | ✅ |
| Continuous | Continuous | Shift / Scale / Threshold | ATE | Difference | Sandwich | none | ✅ (vs lmtp) |
| Continuous | Continuous | Shift | ATE | Difference | Bootstrap | none | ✅ (vs lmtp) |
| Multivariate | Continuous | Static / Shift | ATE | Difference | Sandwich | none | ✅ |
| Binary | Binary | Static / Dynamic | ATE | Difference | Sandwich / bootstrap | none | ✅ (2 periods) |
| Binary | Binary | Static | ATE | Difference | Sandwich | none | ✅ (4 periods) |
| Binary | Binary | Static | ATE | Difference | Bootstrap (parallel) | none | ✅ |
| Binary | Continuous | Static | ATE | Difference | Sandwich | censoring | ✅ |
| Binary | Any | IPSI($\delta$) | — | — | — | — | ⛔ Phase 4 |
| Binary | Any | Static / ATT | ATT | — | — | — | ⛔ ICE is ATE-only |

See [`FEATURE_COVERAGE_MATRIX.md`](https://github.com/etverse/causatr/blob/main/FEATURE_COVERAGE_MATRIX.md)
for the authoritative coverage status of every method × treatment × outcome
× intervention × variance combination.

## References

Hernan MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapters 19--21: Time-varying treatments.

Zivich PN, Ross RK, Shook-Sa BE, Cole SR, Edwards JK (2024). Empirical
sandwich variance estimator for iterated conditional expectation
g-computation. *Statistics in Medicine* 43:5562--5572.
