---
title: "Propensity score matching with causatr"
vignette: >
  %\VignetteIndexEntry{Propensity score matching with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Propensity score matching estimates causal effects by pairing treated and
control individuals with similar propensity scores, then comparing outcomes
within matched sets. causatr delegates match construction to
[MatchIt](https://kosukeimai.github.io/MatchIt/) and fits the outcome model
on the matched sample with cluster-robust sandwich SEs (via
`sandwich::vcovCL()` on the matched-pair subclass).

This vignette demonstrates propensity score matching in causatr using the NHEFS
dataset from Hernán & Robins (2025), covering every supported combination of
treatment type, outcome type, contrast scale, and inference method for
time-fixed treatments.

**Note:** Matching currently supports only `static()` interventions and binary
treatments. For modified treatment policies, use g-computation (see
`vignette("gcomp")`).

## Setup

```{r}
#| message: false
library(causatr)
library(tinyplot)

data("nhefs")

nhefs_complete <- nhefs[!is.na(nhefs$wt82_71), ]

nhefs_complete$gained_weight <- as.integer(nhefs_complete$wt82_71 > 0)
```

## Binary treatment, continuous outcome

Matching estimate of the effect of quitting smoking on weight change, following
the approach in Chapter 15 of Hernán & Robins.

### ATT with sandwich SE

Matching naturally targets the ATT (effect on the treated). This is the default
for nearest-neighbour matching.

```{r}
fit_m <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "matching",
  estimand = "ATT"
)

res_att_sw <- contrast(
  fit_m,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_att_sw
```

### ATT with bootstrap SE

The bootstrap resamples individuals, re-matches, and refits the outcome model
on each bootstrap sample, fully accounting for matching uncertainty.

```{r}
res_att_bs <- contrast(
  fit_m,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 500L
)
res_att_bs
```

### ATE estimand

For the ATE, MatchIt performs full matching (each unit gets a weight) rather
than 1:1 nearest-neighbour matching.

```{r}
fit_m_ate <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "matching",
  estimand = "ATE"
)

res_ate_sw <- contrast(
  fit_m_ate,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_ate_sw
```

### ATE with bootstrap SE

```{r}
res_ate_bs <- contrast(
  fit_m_ate,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 500L
)
res_ate_bs
```

### ATC estimand

The ATC targets the effect among controls (those who continued smoking).

```{r}
fit_m_atc <- causat(
  nhefs_complete,
  outcome = "wt82_71",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "matching",
  estimand = "ATC"
)

res_atc <- contrast(
  fit_m_atc,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_atc
```

### Extracting results programmatically

```{r}
coef(res_att_sw)
confint(res_att_sw)
```

## Binary treatment, binary outcome

Using the `gained_weight` indicator as a binary outcome. The outcome model on
the matched sample is still `Y ~ A` (linear probability model), so
`contrast()` computes marginal risks from the weighted predictions.

### Risk difference (sandwich)

```{r}
fit_m_bin <- causat(
  nhefs_complete,
  outcome = "gained_weight",
  treatment = "qsmk",
  confounders = ~ sex + age + I(age^2) + race + factor(education) +
    smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
    factor(exercise) + factor(active) + wt71 + I(wt71^2),
  method = "matching",
  estimand = "ATT"
)

res_rd <- contrast(
  fit_m_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "sandwich"
)
res_rd
```

### Risk difference (bootstrap)

```{r}
res_rd_bs <- contrast(
  fit_m_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 500L
)
res_rd_bs
```

### Risk ratio

```{r}
res_rr <- contrast(
  fit_m_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "sandwich"
)
res_rr
```

### Odds ratio

```{r}
res_or <- contrast(
  fit_m_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "or",
  ci_method = "sandwich"
)
res_or
```

### Risk ratio (bootstrap)

```{r}
res_rr_bs <- contrast(
  fit_m_bin,
  interventions = list(quit = static(1), continue = static(0)),
  reference = "continue",
  type = "ratio",
  ci_method = "bootstrap",
  n_boot = 500L
)
res_rr_bs
```

## Comparing estimands

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Point estimates and confidence intervals for ATE, ATT, and ATC estimated by matching."
est_df <- data.frame(
  estimand = c("ATE", "ATT", "ATC"),
  estimate = c(
    res_ate_sw$contrasts$estimate[1],
    res_att_sw$contrasts$estimate[1],
    res_atc$contrasts$estimate[1]
  ),
  ci_lower = c(
    res_ate_sw$contrasts$ci_lower[1],
    res_att_sw$contrasts$ci_lower[1],
    res_atc$contrasts$ci_lower[1]
  ),
  ci_upper = c(
    res_ate_sw$contrasts$ci_upper[1],
    res_att_sw$contrasts$ci_upper[1],
    res_atc$contrasts$ci_upper[1]
  )
)

tinyplot(
  estimate ~ estimand,
  data = est_df,
  type = "pointrange",
  ymin = est_df$ci_lower,
  ymax = est_df$ci_upper,
  xlab = "Estimand",
  ylab = "Effect on weight change (kg)",
  main = "Matching estimates by estimand"
)
abline(h = 0, lty = 2, col = "grey40")
```

## Diagnostics

After fitting a matching model, use `diagnose()` to assess covariate balance
and match quality. Balance diagnostics use
[cobalt](https://ngreifer.github.io/cobalt/) to compare SMDs before and after
matching.

```{r}
diag <- diagnose(fit_m)
diag
```

### Love plot

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-alt: "Love plot showing covariate balance before and after propensity score matching."
plot(diag)
```

### Match quality

```{r}
diag$match_quality
```

## Summary of covered combinations

| Treatment | Outcome | Contrast | Inference | Estimand |
|---|---|---|---|---|
| Binary | Continuous | Difference | Sandwich | ATT |
| Binary | Continuous | Difference | Bootstrap | ATT |
| Binary | Continuous | Difference | Sandwich | ATE |
| Binary | Continuous | Difference | Bootstrap | ATE |
| Binary | Continuous | Difference | Sandwich | ATC |
| Binary | Binary | Difference | Sandwich | ATT |
| Binary | Binary | Difference | Bootstrap | ATT |
| Binary | Binary | Ratio | Sandwich | ATT |
| Binary | Binary | OR | Sandwich | ATT |
| Binary | Binary | Ratio | Bootstrap | ATT |

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 15: Outcome regression and propensity scores.

Ho DE, Imai K, King G, Stuart EA (2011). MatchIt: Nonparametric Preprocessing
for Parametric Causal Inference. *Journal of Statistical Software* 42(8):1-28.
