---
title: "Handling missing data"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Missing data is common in observational studies. causatr handles it differently
depending on *which* variable is missing (outcome, treatment, or covariates)
and *which* estimator is used. This vignette demonstrates the behaviour for
each case and shows how to verify that your analysis is robust.

## Setup

```{r}
#| message: false
library(causatr)
library(tinytable)
library(tinyplot)
```

We simulate a baseline dataset with known truth, then introduce missingness.

**Data-generating mechanism.** Two independent confounders $L_1$ and $L_2$
influence both a binary treatment $A$ and a continuous outcome $Y$. The
treatment probability follows a logistic model in $L_1$ and $L_2$, and the
outcome is linear in $A$, $L_1$, and $L_2$ with additive Gaussian noise.

$$
\begin{aligned}
L_1 &\sim \mathcal{N}(0, 1) \\
L_2 &\sim \mathcal{N}(0, 1) \\
A &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3\,L_1 + 0.2\,L_2)\bigr) \\
Y &= 2 + 3\,A + 1.5\,L_1 + 0.8\,L_2 + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

The treatment coefficient is 3 with no effect modification, so the true
average treatment effect is $\text{ATE} = E[Y^{a=1}] - E[Y^{a=0}] = 3$.

```{r}
set.seed(42)
n <- 2000
L1 <- rnorm(n)
L2 <- rnorm(n)
A <- rbinom(n, 1, plogis(0.3 * L1 + 0.2 * L2))
Y <- 2 + 3 * A + 1.5 * L1 + 0.8 * L2 + rnorm(n)
df <- data.frame(Y = Y, A = A, L1 = L1, L2 = L2)
```

## Missing outcomes (MCAR)

When outcome values are missing completely at random (MCAR), the outcome model
is fit on complete cases. `contrast()` predicts on *all* rows, then excludes
rows with NA predictions from the target-population average. Under MCAR this
is unbiased.

```{r}
df_y_miss <- df
df_y_miss$Y[sample(n, round(0.15 * n))] <- NA

fit_y <- causat(
  df_y_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2
)

res_y <- contrast(
  fit_y,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_y
```

The estimate remains close to 3 despite 15% missing outcomes.

### Comparison: complete data vs missing-outcome data

```{r}
fit_complete <- causat(df, outcome = "Y", treatment = "A", confounders = ~ L1 + L2)
res_complete <- contrast(
  fit_complete,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)

comp_df <- data.frame(
  Analysis = c("Complete data", "15% outcome MCAR"),
  Estimate = c(res_complete$contrasts$estimate[1], res_y$contrasts$estimate[1]),
  SE = c(res_complete$contrasts$se[1], res_y$contrasts$se[1]),
  CI_lower = c(res_complete$contrasts$ci_lower[1], res_y$contrasts$ci_lower[1]),
  CI_upper = c(res_complete$contrasts$ci_upper[1], res_y$contrasts$ci_upper[1])
)
tt(comp_df, digits = 3)
```

The SE increases slightly with missing data (less information), but the point
estimate stays on target.

## Missing outcomes under IPW

IPW also handles missing outcomes. The treatment model is fit on the
outcome-filtered subset (rows where Y is observed), and density-ratio weights
are computed on that same subset.

```{r}
fit_y_ipw <- causat(
  df_y_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  estimator = "ipw"
)

res_y_ipw <- contrast(
  fit_y_ipw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_y_ipw
```

## Missing outcomes under matching

Matching handles missing outcomes through MatchIt's internal mechanisms.
Rows with missing Y are excluded from the matched sample.

```{r}
fit_y_match <- causat(
  df_y_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  estimator = "matching",
  estimand = "ATT"
)

res_y_match <- contrast(
  fit_y_match,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_y_match
```

### Three-method comparison under outcome missingness

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot comparing g-comp, IPW, and matching estimates under 15% missing outcomes."
methods_df <- data.frame(
  Method = c("G-comp", "IPW", "Matching (ATT)"),
  Estimate = c(
    res_y$contrasts$estimate[1],
    res_y_ipw$contrasts$estimate[1],
    res_y_match$contrasts$estimate[1]
  ),
  CI_lower = c(
    res_y$contrasts$ci_lower[1],
    res_y_ipw$contrasts$ci_lower[1],
    res_y_match$contrasts$ci_lower[1]
  ),
  CI_upper = c(
    res_y$contrasts$ci_upper[1],
    res_y_ipw$contrasts$ci_upper[1],
    res_y_match$contrasts$ci_upper[1]
  )
)

tinyplot(
  Estimate ~ Method,
  data = methods_df,
  type = "pointrange",
  ymin = methods_df$CI_lower,
  ymax = methods_df$CI_upper,
  ylab = "ATE estimate",
  main = "Three estimators under 15% outcome MCAR"
)
abline(h = 3, lty = 2, col = "grey40")
```

All three estimators recover the truth (dashed line at 3).

## Missing covariates

When covariates have missing values, `glm(na.action = na.omit)` drops those
rows from the model fit. `contrast()` returns NA predictions for rows with
missing covariates and excludes them from the target population.

```{r}
df_l_miss <- df
df_l_miss$L1[sample(n, round(0.10 * n))] <- NA

fit_l <- causat(
  df_l_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2
)

res_l <- contrast(
  fit_l,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_l
```

Under MCAR covariate missingness, the estimate is unbiased. Under MAR,
multiple imputation is needed (planned via `causat_mice()`).

## Both outcome and covariate missing

causatr gracefully handles the case where both Y and L have missing values
(possibly on different rows):

```{r}
df_both <- df
df_both$Y[sample(n, round(0.10 * n))] <- NA
df_both$L1[sample(n, round(0.10 * n))] <- NA

fit_both <- causat(
  df_both,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2
)

res_both <- contrast(
  fit_both,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_both
```

## Missing treatment values

Treatment NAs are **not allowed** --- you cannot define a counterfactual
without knowing the observed treatment value. causatr aborts with an
informative error:
```{r}
#| error: true
df_a_miss <- df
df_a_miss$A[1:10] <- NA

causat(
  df_a_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2
)
```

If treatment is MCAR, drop the rows. If MAR, use multiple imputation before
calling `causat()`.

## Missing outcomes with continuous treatment

The same principles apply to continuous treatments. Here we show IPW with a
`shift()` intervention under outcome missingness.

**Data-generating mechanism.** A single confounder $L$ affects both a
continuous treatment $A$ and a continuous outcome $Y$. The treatment is
Gaussian with its mean depending on $L$, and the outcome is linear in $A$
and $L$.

$$
\begin{aligned}
L &\sim \mathcal{N}(0, 1) \\
A &\sim \mathcal{N}(0.5\,L,\; 1) \\
Y &= 1 + 2\,A + L + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

Under a unit additive shift intervention $A^* = A + 1$, every individual's
treatment increases by 1 and the outcome increases by $2 \times 1 = 2$.
The true causal contrast is $E[Y(A+1)] - E[Y(A)] = 2$.

```{r}
set.seed(43)
L <- rnorm(n)
A_cont <- rnorm(n, mean = 0.5 * L)
Y_cont <- 1 + 2 * A_cont + L + rnorm(n)
df_cont <- data.frame(Y = Y_cont, A = A_cont, L = L)
df_cont$Y[sample(n, round(0.10 * n))] <- NA

fit_cont_ipw <- causat(
  df_cont,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw"
)

res_cont_ipw <- contrast(
  fit_cont_ipw,
  interventions = list(up1 = shift(1), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_cont_ipw
```

Structural truth: `E[Y(A+1)] - E[Y(A)] = 2`.

## Missing data with categorical treatment

G-comp handles missing outcomes for categorical treatments.

**Data-generating mechanism.** A three-arm categorical treatment $A \in \{A, B, C\}$
is drawn independently of the confounder $L$, and the outcome depends on
both the treatment arm and $L$.

$$
\begin{aligned}
A &\sim \text{Categorical}(0.3,\; 0.4,\; 0.3) \\
L &\sim \mathcal{N}(0, 1) \\
Y &= 1 + \mathbf{1}(A = B) \cdot 1 + \mathbf{1}(A = C) \cdot 2 + 0.5\,L + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

The true contrasts relative to arm A are:
$E[Y^{B}] - E[Y^{A}] = 1$ and $E[Y^{C}] - E[Y^{A}] = 2$.

```{r}
set.seed(44)
A_cat <- factor(sample(c("A", "B", "C"), n, replace = TRUE, prob = c(0.3, 0.4, 0.3)))
L <- rnorm(n)
Y_cat <- 1 + ifelse(A_cat == "B", 1, ifelse(A_cat == "C", 2, 0)) + 0.5 * L + rnorm(n)
df_cat <- data.frame(Y = Y_cat, A = A_cat, L = L)
df_cat$Y[sample(n, round(0.10 * n))] <- NA

fit_cat <- causat(
  df_cat,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L
)

res_cat <- contrast(
  fit_cat,
  interventions = list(arm_a = static("A"), arm_b = static("B"), arm_c = static("C")),
  reference = "arm_a",
  type = "difference",
  ci_method = "sandwich"
)
res_cat
```

## Missing outcomes in longitudinal data (ICE)

ICE g-computation handles missing final-period outcomes. The backward
recursion excludes NA rows when fitting each sequential outcome model.

**Data-generating mechanism.** A two-period panel ($K = 2$) with $n = 500$
individuals. Each individual has a baseline confounder $L_0$, a time-varying
confounder $L_k$, and a binary treatment $A_k$ at each period. The outcome
$Y$ is observed only at the final period and depends on the cumulative
treatment received across both periods.

$$
\begin{aligned}
L_0 &\sim \mathcal{N}(0, 1) \\
L_k &= L_0 + \eta_k, \quad \eta_k \sim \mathcal{N}(0, 0.25) \\
A_k &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3\,L_k)\bigr), \quad k = 1, 2 \\
Y &= 2 + 1.5\,(A_1 + A_2) + 0.5\,L_0 + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

Under the "always treat" regime ($A_1 = A_2 = 1$) vs "never treat"
($A_1 = A_2 = 0$), the true ATE is $1.5 \times 2 = 3$.

```{r}
set.seed(45)
n_id <- 500
ids <- rep(seq_len(n_id), each = 2)
times <- rep(1:2, n_id)
L0 <- rep(rnorm(n_id), each = 2)
L_tv <- L0 + rnorm(n_id * 2, 0, 0.5)
A_tv <- rbinom(n_id * 2, 1, plogis(0.3 * L_tv))

Y_long <- NA_real_
for (i in seq_len(n_id)) {
  rows <- which(ids == i)
  Y_long[rows[2]] <- 2 + 1.5 * sum(A_tv[rows]) + 0.5 * L0[rows[1]] + rnorm(1)
}

long_df <- data.frame(id = ids, time = times, Y = Y_long, A = A_tv, L0 = L0, L = L_tv)
long_df$Y[sample(which(!is.na(long_df$Y)), round(0.10 * n_id))] <- NA

fit_ice_miss <- causat(
  long_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L0,
  confounders_tv = ~ L,
  id = "id",
  time = "time"
)

res_ice_miss <- contrast(
  fit_ice_miss,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_ice_miss
```

Structural truth: ATE = `1.5 * 2 = 3` (two periods of binary treatment).

## Bootstrap with missing data

Bootstrap inference correctly handles missing data --- each replicate
resamples individuals (with replacement) and refits the full pipeline:

```{r}
res_boot_miss <- contrast(
  fit_y,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_boot_miss
```

### Sandwich vs bootstrap with missing data

```{r}
boot_comp <- data.frame(
  Inference = c("Sandwich", "Bootstrap"),
  Estimate = c(res_y$contrasts$estimate[1], res_boot_miss$contrasts$estimate[1]),
  SE = c(res_y$contrasts$se[1], res_boot_miss$contrasts$se[1]),
  CI_lower = c(res_y$contrasts$ci_lower[1], res_boot_miss$contrasts$ci_lower[1]),
  CI_upper = c(res_y$contrasts$ci_upper[1], res_boot_miss$contrasts$ci_upper[1])
)
tt(boot_comp, digits = 3)
```

## Manual IPCW weights

When outcomes are missing at random (MAR) conditional on treatment and
covariates, you can supply inverse-probability-of-censoring weights (IPCW)
via the `weights` argument. Built-in IPCW is planned for Phase 14; for now,
compute the weights externally.

**Censoring mechanism.** Starting from the baseline DGM above (ATE $= 3$),
we introduce MAR outcome censoring. The probability of being censored
(outcome set to NA) depends on both the treatment $A$ and confounder $L_1$:

$$
\begin{aligned}
C_i &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-2 + 0.5\,A_i + 0.3\,L_{1,i})\bigr) \\
Y_i^{\text{obs}} &= \begin{cases} Y_i & \text{if } C_i = 0 \\ \text{NA} & \text{if } C_i = 1 \end{cases}
\end{aligned}
$$

The intercept of $-2$ keeps the censoring rate moderate. Because censoring
depends on $A$ and $L_1$, a complete-case analysis is biased; IPCW weights
$w_i = 1 / P(C_i = 0 \mid A_i, L_{1,i}, L_{2,i})$ restore consistency.
The true ATE remains 3.

```{r}
df_mar <- df
cens_prob <- plogis(-2 + 0.5 * df_mar$A + 0.3 * df_mar$L1)
df_mar$Y[runif(n) < cens_prob] <- NA

observed <- !is.na(df_mar$Y)
cens_model <- glm(observed ~ A + L1 + L2, data = df_mar, family = binomial)
ipcw <- ifelse(observed, 1 / predict(cens_model, type = "response"), 0)

fit_ipcw <- causat(
  df_mar[observed, ],
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  weights = ipcw[observed]
)

res_ipcw <- contrast(
  fit_ipcw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_ipcw
```

## Summary of missingness handling

```{r}
#| echo: false
summary_df <- data.frame(
  Variable = c("Outcome (Y)", "Outcome (Y)", "Outcome (Y)", "Outcome (Y)",
               "Treatment (A)", "Covariates (L)", "Covariates (L)", "Y + L"),
  Mechanism = c("MCAR", "MCAR", "MCAR", "MAR",
                "Any", "MCAR", "MAR", "MCAR"),
  Method = c("G-comp", "IPW", "Matching", "G-comp + IPCW",
             "Any", "G-comp", "Any", "G-comp"),
  Handling = c(
    "Complete-case model, predict on all rows",
    "Fit treatment model on observed-Y subset",
    "MatchIt excludes NA-Y rows",
    "External IPCW weights via weights=",
    "Rejected (abort with error)",
    "na.omit drops rows; predict returns NA",
    "Multiple imputation (causat_mice, planned)",
    "Combines both drop rules"
  ),
  Status = c("Supported", "Supported", "Supported", "Supported",
             "Rejected", "Supported", "Planned", "Supported")
)
tt(summary_df)
```

## Key takeaways

1. **Missing outcomes (MCAR)** are handled automatically by all estimators.
   The estimate is unbiased; the SE increases proportionally to the information
   loss.

2. **Missing treatment values** are never allowed. Drop rows (MCAR) or impute
   (MAR) before calling `causat()`.

3. **Missing covariates (MCAR)** work out-of-the-box under g-comp via
   `na.omit`. For MAR missingness, multiple imputation is needed.

4. **MAR outcome missingness** requires IPCW weights today (manual via
   `weights =`). Built-in IPCW is planned for Phase 14.

5. **Bootstrap** handles all missingness patterns correctly by resampling
   the full pipeline.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 12 (IP weighting under censoring) and Chapter 19 (censoring in
longitudinal studies).

Little RJA, Rubin DB (2019). *Statistical Analysis with Missing Data*. 3rd ed.
Wiley.
