---
title: "Handling missing data"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Missing data is common in observational studies. causatr handles it differently
depending on *which* variable is missing (outcome, treatment, or covariates)
and *which* estimator is used. This vignette demonstrates the behaviour for
each case and shows how to verify that your analysis is robust.

## Setup

```{r}
#| message: false
library(causatr)
library(tinytable)
library(tinyplot)
```

We simulate a baseline dataset with known truth, then introduce missingness.

**Data-generating mechanism.** Two independent confounders $L_1$ and $L_2$
influence both a binary treatment $A$ and a continuous outcome $Y$. The
treatment probability follows a logistic model in $L_1$ and $L_2$, and the
outcome is linear in $A$, $L_1$, and $L_2$ with additive Gaussian noise.

$$
\begin{aligned}
L_1 &\sim \mathcal{N}(0, 1) \\
L_2 &\sim \mathcal{N}(0, 1) \\
A &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3\,L_1 + 0.2\,L_2)\bigr) \\
Y &= 2 + 3\,A + 1.5\,L_1 + 0.8\,L_2 + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

The treatment coefficient is 3 with no effect modification, so the true
average treatment effect is $\text{ATE} = E[Y^{a=1}] - E[Y^{a=0}] = 3$.

```{r}
set.seed(42)
n <- 2000
L1 <- rnorm(n)
L2 <- rnorm(n)
A <- rbinom(n, 1, plogis(0.3 * L1 + 0.2 * L2))
Y <- 2 + 3 * A + 1.5 * L1 + 0.8 * L2 + rnorm(n)
df <- data.frame(Y = Y, A = A, L1 = L1, L2 = L2)
```

## Missing outcomes (MCAR)

When outcome values are missing completely at random (MCAR), the outcome model
is fit on complete cases. `contrast()` predicts on *all* rows, then excludes
rows with NA predictions from the target-population average. Under MCAR this
is unbiased.

```{r}
df_y_miss <- df
df_y_miss$Y[sample(n, round(0.15 * n))] <- NA

fit_y <- causat(
  df_y_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  model_fn = stats::glm
)

res_y <- contrast(
  fit_y,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_y
```

The estimate remains close to 3 despite 15% missing outcomes.

### Comparison: complete data vs missing-outcome data

```{r}
fit_complete <- causat(
  df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  model_fn = stats::glm
)
res_complete <- contrast(
  fit_complete,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)

comp_df <- data.frame(
  Analysis = c("Complete data", "15% outcome MCAR"),
  Estimate = c(res_complete$contrasts$estimate[1], res_y$contrasts$estimate[1]),
  SE = c(res_complete$contrasts$se[1], res_y$contrasts$se[1]),
  CI_lower = c(res_complete$contrasts$ci_lower[1], res_y$contrasts$ci_lower[1]),
  CI_upper = c(res_complete$contrasts$ci_upper[1], res_y$contrasts$ci_upper[1])
)
tt(comp_df, digits = 3)
```

The SE increases slightly with missing data (less information), but the point
estimate stays on target.

## Missing outcomes under IPW

IPW also handles missing outcomes. The treatment model is fit on the
outcome-filtered subset (rows where Y is observed), and density-ratio weights
are computed on that same subset.

```{r}
fit_y_ipw <- causat(
  df_y_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  estimator = "ipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_y_ipw <- contrast(
  fit_y_ipw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_y_ipw
```

## Missing outcomes under matching

Matching handles missing outcomes through MatchIt's internal mechanisms.
Rows with missing Y are excluded from the matched sample.

```{r}
fit_y_match <- causat(
  df_y_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  estimator = "matching",
  estimand = "ATT"
)

res_y_match <- contrast(
  fit_y_match,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_y_match
```

### Three-method comparison under outcome missingness

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot comparing g-comp, IPW, and matching estimates under 15% missing outcomes."
methods_df <- data.frame(
  Method = c("G-comp", "IPW", "Matching (ATT)"),
  Estimate = c(
    res_y$contrasts$estimate[1],
    res_y_ipw$contrasts$estimate[1],
    res_y_match$contrasts$estimate[1]
  ),
  CI_lower = c(
    res_y$contrasts$ci_lower[1],
    res_y_ipw$contrasts$ci_lower[1],
    res_y_match$contrasts$ci_lower[1]
  ),
  CI_upper = c(
    res_y$contrasts$ci_upper[1],
    res_y_ipw$contrasts$ci_upper[1],
    res_y_match$contrasts$ci_upper[1]
  )
)

tinyplot(
  Estimate ~ Method,
  data = methods_df,
  type = "pointrange",
  ymin = methods_df$CI_lower,
  ymax = methods_df$CI_upper,
  ylab = "ATE estimate",
  main = "Three estimators under 15% outcome MCAR"
)
abline(h = 3, lty = 2, col = "grey40")
```

All three estimators recover the truth (dashed line at 3).

## Missing covariates

When covariates have missing values, `glm(na.action = na.omit)` drops those
rows from the model fit. `contrast()` returns NA predictions for rows with
missing covariates and excludes them from the target population.

```{r}
df_l_miss <- df
df_l_miss$L1[sample(n, round(0.10 * n))] <- NA

fit_l <- causat(
  df_l_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  model_fn = stats::glm
)

res_l <- contrast(
  fit_l,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_l
```

Under MCAR covariate missingness, the estimate is unbiased. Under MAR, the
complete-case fit is *biased* — the missing rows are not a random sample of the
data — and multiple imputation is needed. The next section walks through that
workflow with `causat_mice()`.

## Multiple imputation for MAR covariates and treatment

When covariates (or treatment) are missing **at random** (MAR) rather than
completely at random, dropping the incomplete rows distorts the analysis: the
observed subset is no longer representative. Multiple imputation (MI) fixes this
by filling in the missing values from their predictive distribution several
times, running the analysis on each completed dataset, and pooling. causatr's
`causat_mice()` is the *analysis-and-pooling* step: you impute upstream with
`mice::mice()`, then hand the `mids` object to `causat_mice()`, which runs
`causat()` + `contrast()` on every completed dataset and pools the results into
a single `causatr_result`.

```{r}
#| label: mi-setup
# Multiple imputation needs the mice package (a Suggested dependency). Every
# chunk in this section is guarded so the vignette still renders without it.
has_mice <- requireNamespace("mice", quietly = TRUE)
```

**A MAR covariate.** We simulate a confounder `L` whose missingness depends on
the *observed* treatment `A` (so it is MAR, not MCAR) — `L` is recorded less
often for the untreated. The true ATE is 3.

```{r}
#| eval: !expr has_mice
set.seed(2024)
n_mi <- 2000
L <- rnorm(n_mi)
A <- rbinom(n_mi, 1, plogis(0.5 * L))
Y <- 2 + 3 * A + 1.5 * L + rnorm(n_mi) # true ATE = 3
# L is observed more often among the treated -> missing more often among
# controls: missingness depends on the observed A (MAR, not MCAR).
R_L <- rbinom(n_mi, 1, plogis(0.6 + 0.7 * A))
df_full <- data.frame(Y = Y, A = A, L = L)
df_mar_l <- df_full
df_mar_l$L[R_L == 0] <- NA
round(mean(is.na(df_mar_l$L)), 2) # fraction of L missing
```

Helper to read a difference contrast as a signed magnitude with a symmetric
interval (the reference direction is a display convention; the true effect is
3):

```{r}
#| eval: !expr has_mice
ate_row <- function(res, label) {
  e <- res$contrasts$estimate[1]
  half <- (res$contrasts$ci_upper[1] - res$contrasts$ci_lower[1]) / 2
  data.frame(Method = label, Estimate = abs(e),
             CI_lower = abs(e) - half, CI_upper = abs(e) + half)
}
```

**Complete-case analysis discards the incomplete rows.** A subtlety worth being
honest about: under g-computation with a *correctly specified* outcome model,
covariate missingness that depends only on modelled variables (here `A`) leaves
the complete-case estimate approximately *valid* — so it will land near the
truth below. But that is a fragile special case, not a general guarantee:

- it throws away every partially observed row (here ~30%), and
- it is not even an option for `estimator = "ipw"` or `"matching"`, which abort
  on rows with a missing confounder, and
- it breaks once missingness depends on the **outcome**, which a covariate-only
  model cannot absorb.

Multiple imputation sidesteps all three by using every row and propagating the
imputation uncertainty into the standard error.

```{r}
#| eval: !expr has_mice
cc_fit <- causat(df_mar_l, outcome = "Y", treatment = "A",
                 confounders = ~L, model_fn = stats::glm)
cc_res <- contrast(cc_fit, interventions = list(a1 = static(1), a0 = static(0)),
                   reference = "a0", type = "difference", ci_method = "sandwich")
cc_res
```

**Multiple imputation uses every row.** Impute `L` with `mice()` (the default
predictor matrix carries `A` *and* `Y`, which is exactly what we want — see the
congeniality note below), then pool with `causat_mice()`:

```{r}
#| eval: !expr has_mice
library(mice)
imp <- mice(df_mar_l, m = 20, printFlag = FALSE)

mi_res <- causat_mice(
  imp,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  interventions = list(a1 = static(1), a0 = static(0)),
  estimator = "gcomp",
  type = "difference"
)
mi_res
```

The pooled standard error reflects both the within-imputation sampling
variability and the between-imputation uncertainty, via Rubin's (1987) rules
with Barnard-Rubin (1999) degrees of freedom. The per-imputation diagnostics
(including the fraction of missing information) live in the `"mi_details"`
attribute:

```{r}
#| eval: !expr has_mice
attr(mi_res, "mi_details")$contrasts
```

Be transparent about what MI does and does not buy here. Complete-case is valid
in this design, and the two estimates agree closely — MI does not magically
sharpen a valid analysis. It does still use the partially observed rows (so its
interval is, if anything, a touch narrower than complete-case's), and the `fmi`
column reports how much information the missingness costs. The honest caveat
runs the *other* way: MI is only as good as its imputation model. A misspecified
imputation can bias the causal estimate **worse** than complete-case — we
demonstrate exactly that below by dropping `Y` from the predictor matrix. MI's
decisive *advantage*, meanwhile, shows up when complete-case is **biased** — the
next design.

### When complete-case fails: outcome-dependent missingness

Complete-case validity above hinged on `L` being missing as a function of `A`
alone — a *modelled* covariate. Now make missingness depend on the **outcome**,
and in *opposite directions* across treatment arms: among the treated `L` is
recorded when `Y` is low, among controls when `Y` is high. The outcome
regression conditions on covariates, not on its own selection, so the
complete-case fit cannot undo this distortion and its ATE is badly biased. MI,
which imputes `L` from `(A, Y)` over the whole sample, recovers the truth.

```{r}
#| eval: !expr has_mice
# Reusable complete-case (na.omit) + MI g-computation on a partially observed
# data frame, each summarised by ate_row().
cc_mi_gcomp <- function(dmar, m = 20) {
  ints <- list(a1 = static(1), a0 = static(0))
  cc <- contrast(
    causat(na.omit(dmar), outcome = "Y", treatment = "A",
           confounders = ~L, model_fn = stats::glm),
    interventions = ints, reference = "a0", ci_method = "sandwich"
  )
  imp_d <- mice(dmar, m = m, printFlag = FALSE)
  mi <- causat_mice(imp_d, outcome = "Y", treatment = "A", confounders = ~L,
                    interventions = ints, estimator = "gcomp")
  list(cc = cc, mi = mi)
}

set.seed(3)
n2 <- 4000
L2 <- rnorm(n2)
A2 <- rbinom(n2, 1, plogis(0.5 * L2))
Y2 <- 2 + 3 * A2 + 1.5 * L2 + rnorm(n2) # true ATE = 3
z2 <- (Y2 - mean(Y2)) / sd(Y2)
# Observe L with arm-flipped dependence on Y: MAR on (A, Y), unmodellable by a
# covariate-only outcome regression fit on complete cases.
R2 <- rbinom(n2, 1, plogis(1.0 + 1.8 * ifelse(A2 == 1, -z2, z2)))
df2_full <- data.frame(Y = Y2, A = A2, L = L2)
df2 <- df2_full
df2$L[R2 == 0] <- NA
fit2 <- cc_mi_gcomp(df2)
fit2$cc$contrasts
fit2$mi$contrasts
```

Side by side, with the (unobservable) full-data oracle for reference. Design 1
is the benign covariate-MAR case (complete-case valid, MI slightly wider);
Design 2 is the outcome-dependent case (complete-case biased, MI on target):

```{r}
#| eval: !expr has_mice
oracle1 <- contrast(causat(df_full, "Y", "A", confounders = ~L, model_fn = stats::glm),
                    interventions = list(a1 = static(1), a0 = static(0)),
                    reference = "a0", ci_method = "sandwich")
oracle2 <- contrast(causat(df2_full, "Y", "A", confounders = ~L, model_fn = stats::glm),
                    interventions = list(a1 = static(1), a0 = static(0)),
                    reference = "a0", ci_method = "sandwich")

d1 <- "Design 1: covariate-MAR (CC valid)"
d2 <- "Design 2: outcome-dependent (CC biased)"
compare <- rbind(
  cbind(Design = d1, rbind(ate_row(oracle1, "Oracle"),
        ate_row(cc_res, "Complete-case"), ate_row(mi_res, "MI"))),
  cbind(Design = d2, rbind(ate_row(oracle2, "Oracle"),
        ate_row(fit2$cc, "Complete-case"), ate_row(fit2$mi, "MI")))
)
compare$Method <- factor(compare$Method,
                         levels = c("Oracle", "Complete-case", "MI"))
tt(compare, digits = 3)
```

```{r}
#| eval: !expr has_mice
#| fig-alt: "Pointrange plot of ATE estimates with confidence intervals. Two coloured series (Design 1 and Design 2) at each of Oracle, Complete-case, and MI. In Design 1 all three sit on the dashed truth line at 3. In Design 2 the complete-case point falls well below 3 while Oracle and MI remain on it."
tinyplot(
  Estimate ~ Method | Design,
  data = compare,
  type = "pointrange",
  ymin = compare$CI_lower,
  ymax = compare$CI_upper,
  ylab = "ATE estimate",
  xlab = "",
  main = "Complete-case vs MI across two missingness designs"
)
abline(h = 3, lty = 2, col = "grey40")
```

Design 2 is where MI earns its keep: complete-case misses the truth by a wide
margin, MI lands on it. Design 1 is the honest counterweight — when
complete-case is valid, MI buys nothing and costs a little precision.

`causat_mice()` is **estimator-agnostic** — swap `estimator = "ipw"`,
`"aipw"`, `"matching"`, or `"snm"`, change the outcome family, the contrast
scale, add `by =` stratification, or go longitudinal with `id =`/`time =`, and
the same pooling applies. For `"ipw"` and `"matching"` MI is not just better but
*necessary*: those engines abort on a missing confounder, so the only
alternatives are manual listwise deletion or imputation. MAR *treatment* is
handled the same way: impute `A` (as a factor, so `mice` uses logistic
regression) and pass the `mids` object on.

### Boot MI for inference that does not assume congeniality

The default `pool_method = "rubin"` is exactly valid when the imputation and
analysis models are *congenial*. Causal estimands are often uncongenial with the
`mice` model, and under uncongeniality Rubin's variance can be biased in either
direction (Meng 1994; Bartlett & Hughes 2020) — not reliably conservative.
`pool_method = "boot_mi"` (von Hippel 2020) replaces Rubin's variance
decomposition with a bootstrap-then-impute resampling variance:

```{r}
#| eval: !expr has_mice
boot_res <- causat_mice(
  imp,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  interventions = list(a1 = static(1), a0 = static(0)),
  estimator = "gcomp",
  type = "difference",
  pool_method = "boot_mi",
  B = 50, # raise to ~200-1000 in practice; small here for a fast vignette
  M = 2,
  seed = 1
)
boot_res
```

Both pooling rules are only as good as the consistency of the underlying
estimator — the next note explains why that puts the outcome in the imputation
model.

### Congeniality, consistency, and why Y must be a predictor

A bootstrap (and Rubin's rules) corrects the *variance*, not *bias* in the point
estimate. Both pooling rules attain approximately nominal coverage **only when
the causal estimator stays consistent** (Bartlett & Hughes 2020). The single
most important way to keep it consistent is to **include the outcome `Y` as a
predictor in the imputation model** (van Buuren 2018, ch. 6): omitting it
produces biased imputed covariates, which biases the causal estimate — and no
pooling rule can repair that.

Design 2 makes the stakes concrete. Re-impute it with `Y` *excluded* from the
predictor matrix — exactly the misspecification the missingness mechanism
punishes — and the MI estimate is biased *further* from the truth than even
complete-case. `causat_mice()` flags it: it warns when an analysis variable is
absent from, or unused by, the imputation model.

```{r}
#| eval: !expr has_mice
#| warning: true
pred <- mice::make.predictorMatrix(df2)
pred[, "Y"] <- 0 # deliberately drop Y from the predictor matrix
imp_noY <- mice(df2, predictorMatrix = pred, m = 20, printFlag = FALSE)

# The warning fires: Y is present in the data but never used to impute L.
mi_noY <- causat_mice(
  imp_noY,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  interventions = list(a1 = static(1), a0 = static(0)),
  estimator = "gcomp"
)
```

```{r}
#| eval: !expr has_mice
#| echo: false
data.frame(
  Method = c("Complete-case", "MI (Y included)", "MI (Y excluded)"),
  Estimate = round(c(abs(fit2$cc$contrasts$estimate[1]),
                     abs(fit2$mi$contrasts$estimate[1]),
                     abs(mi_noY$contrasts$estimate[1])), 3)
) |> tt()
```

So MI is not automatically safe: done well (with `Y`) it recovers the truth that
complete-case missed; done badly (without `Y`) it can be *worse* than
complete-case. The imputation model matters.

Do **not** impute `Y` itself: missing outcomes belong to inverse-probability-of-
censoring weighting (`ipcw = TRUE`, shown below) or complete-case analysis, not
MI. Including `Y` as a *predictor* (good) is the opposite of *imputing* `Y`
(wrong tool).

## Both outcome and covariate missing

causatr gracefully handles the case where both Y and L have missing values
(possibly on different rows):

```{r}
df_both <- df
df_both$Y[sample(n, round(0.10 * n))] <- NA
df_both$L1[sample(n, round(0.10 * n))] <- NA

fit_both <- causat(
  df_both,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  model_fn = stats::glm
)

res_both <- contrast(
  fit_both,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_both
```

## Missing treatment values

Treatment NAs are **not allowed** --- you cannot define a counterfactual
without knowing the observed treatment value. causatr aborts with an
informative error:
```{r}
#| error: true
df_a_miss <- df
df_a_miss$A[1:10] <- NA

causat(
  df_a_miss,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  model_fn = stats::glm
)
```

If treatment is MCAR, drop the rows. If MAR, impute `A` with `mice()` (store it
as a factor so `mice` uses logistic regression) and pool with `causat_mice()`,
exactly as for a MAR covariate above.

## Missing outcomes with continuous treatment

The same principles apply to continuous treatments. Here we show IPW with a
`shift()` intervention under outcome missingness.

**Data-generating mechanism.** A single confounder $L$ affects both a
continuous treatment $A$ and a continuous outcome $Y$. The treatment is
Gaussian with its mean depending on $L$, and the outcome is linear in $A$
and $L$.

$$
\begin{aligned}
L &\sim \mathcal{N}(0, 1) \\
A &\sim \mathcal{N}(0.5\,L,\; 1) \\
Y &= 1 + 2\,A + L + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

Under a unit additive shift intervention $A^* = A + 1$, every individual's
treatment increases by 1 and the outcome increases by $2 \times 1 = 2$.
The true causal contrast is $E[Y(A+1)] - E[Y(A)] = 2$.

```{r}
set.seed(43)
L <- rnorm(n)
A_cont <- rnorm(n, mean = 0.5 * L)
Y_cont <- 1 + 2 * A_cont + L + rnorm(n)
df_cont <- data.frame(Y = Y_cont, A = A_cont, L = L)
df_cont$Y[sample(n, round(0.10 * n))] <- NA

fit_cont_ipw <- causat(
  df_cont,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_cont_ipw <- contrast(
  fit_cont_ipw,
  interventions = list(up1 = shift(1), observed = NULL),
  reference = "observed",
  type = "difference",
  ci_method = "sandwich"
)
res_cont_ipw
```

Structural truth: `E[Y(A+1)] - E[Y(A)] = 2`.

## Missing data with categorical treatment

G-comp handles missing outcomes for categorical treatments.

**Data-generating mechanism.** A three-arm categorical treatment $A \in \{A, B, C\}$
is drawn independently of the confounder $L$, and the outcome depends on
both the treatment arm and $L$.

$$
\begin{aligned}
A &\sim \text{Categorical}(0.3,\; 0.4,\; 0.3) \\
L &\sim \mathcal{N}(0, 1) \\
Y &= 1 + \mathbf{1}(A = B) \cdot 1 + \mathbf{1}(A = C) \cdot 2 + 0.5\,L + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

The true contrasts relative to arm A are:
$E[Y^{B}] - E[Y^{A}] = 1$ and $E[Y^{C}] - E[Y^{A}] = 2$.

```{r}
set.seed(44)
A_cat <- factor(
  sample(c("A", "B", "C"), n, replace = TRUE, prob = c(0.3, 0.4, 0.3))
)
L <- rnorm(n)
Y_cat <- 1 +
  ifelse(A_cat == "B", 1, ifelse(A_cat == "C", 2, 0)) +
  0.5 * L + rnorm(n)
df_cat <- data.frame(Y = Y_cat, A = A_cat, L = L)
df_cat$Y[sample(n, round(0.10 * n))] <- NA

fit_cat <- causat(
  df_cat,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  model_fn = stats::glm
)

res_cat <- contrast(
  fit_cat,
  interventions = list(
    arm_a = static("A"),
    arm_b = static("B"),
    arm_c = static("C")
  ),
  reference = "arm_a",
  type = "difference",
  ci_method = "sandwich"
)
res_cat
```

## Missing outcomes in longitudinal data (ICE)

ICE g-computation handles missing final-period outcomes. The backward
recursion excludes NA rows when fitting each sequential outcome model.

**Data-generating mechanism.** A two-period panel ($K = 2$) with $n = 500$
individuals. Each individual has a baseline confounder $L_0$, a time-varying
confounder $L_k$, and a binary treatment $A_k$ at each period. The outcome
$Y$ is observed only at the final period and depends on the cumulative
treatment received across both periods.

$$
\begin{aligned}
L_0 &\sim \mathcal{N}(0, 1) \\
L_k &= L_0 + \eta_k, \quad \eta_k \sim \mathcal{N}(0, 0.25) \\
A_k &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(0.3\,L_k)\bigr), \quad k = 1, 2 \\
Y &= 2 + 1.5\,(A_1 + A_2) + 0.5\,L_0 + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, 1)
\end{aligned}
$$

Under the "always treat" regime ($A_1 = A_2 = 1$) vs "never treat"
($A_1 = A_2 = 0$), the true ATE is $1.5 \times 2 = 3$.

```{r}
set.seed(45)
n_id <- 500
ids <- rep(seq_len(n_id), each = 2)
times <- rep(1:2, n_id)
L0 <- rep(rnorm(n_id), each = 2)
L_tv <- L0 + rnorm(n_id * 2, 0, 0.5)
A_tv <- rbinom(n_id * 2, 1, plogis(0.3 * L_tv))

Y_long <- NA_real_
for (i in seq_len(n_id)) {
  rows <- which(ids == i)
  Y_long[rows[2]] <- 2 + 1.5 * sum(A_tv[rows]) + 0.5 * L0[rows[1]] + rnorm(1)
}

long_df <- data.frame(
  id = ids, time = times, Y = Y_long, A = A_tv, L0 = L0, L = L_tv
)
long_df$Y[sample(which(!is.na(long_df$Y)), round(0.10 * n_id))] <- NA

fit_ice_miss <- causat(
  long_df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L0,
  confounders_tv = ~ L,
  id = "id",
  time = "time",
  model_fn = stats::glm
)

res_ice_miss <- contrast(
  fit_ice_miss,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_ice_miss
```

Structural truth: ATE = `1.5 * 2 = 3` (two periods of binary treatment).

## Bootstrap with missing data

Bootstrap inference correctly handles missing data --- each replicate
resamples individuals (with replacement) and refits the full pipeline:

```{r}
res_boot_miss <- contrast(
  fit_y,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "bootstrap",
  n_boot = 50L
)
res_boot_miss
```

### Sandwich vs bootstrap with missing data

```{r}
boot_comp <- data.frame(
  Inference = c("Sandwich", "Bootstrap"),
  Estimate = c(
    res_y$contrasts$estimate[1], res_boot_miss$contrasts$estimate[1]
  ),
  SE = c(res_y$contrasts$se[1], res_boot_miss$contrasts$se[1]),
  CI_lower = c(
    res_y$contrasts$ci_lower[1], res_boot_miss$contrasts$ci_lower[1]
  ),
  CI_upper = c(
    res_y$contrasts$ci_upper[1], res_boot_miss$contrasts$ci_upper[1]
  )
)
tt(boot_comp, digits = 3)
```

## Censoring modes: row filter vs built-in IPCW

causatr offers two ways to handle outcome censoring via the `censoring =`
argument:

1. **Row filter** (`censoring = "C"`, `ipcw = FALSE` — the default): censored
   rows are dropped before model fitting. No censoring model is fit. Under
   g-computation with a correctly specified outcome model this is sufficient
   for MCAR and MAR censoring (the outcome model adjusts for censoring
   predictors). Under IPW this is **biased** for MAR because the treatment
   model cannot correct for differential censoring.

2. **Built-in IPCW** (`censoring = "C"`, `ipcw = TRUE`): causatr fits an
   internal censoring model $P(C = 0 \mid A, L)$ and computes stabilized IPCW
   weights. These weights are composed multiplicatively with any external
   `weights =`. For g-comp this provides doubly-robust protection (correct if
   *either* the outcome model *or* the censoring model is well-specified). For
   IPW it is essential under MAR. The sandwich SE fully accounts for
   censoring-model estimation uncertainty (Channel 2 correction).

::: {.callout-tip}
**Custom censoring model.** By default the censoring model uses `stats::glm`.
Pass `censoring_model_fn = mgcv::gam` (or any function with the standard
`(formula, data, family, weights, ...)` signature) to use a more flexible
model for the censoring mechanism. Use `confounders_censoring` to specify
a different covariate set for the censoring model than for the outcome or
treatment models (e.g. when censoring depends on a different subset of
covariates).
:::

## Built-in IPCW (`ipcw = TRUE`)

When outcomes are missing at random (MAR) conditional on treatment and
covariates, causatr can fit a censoring model $P(C = 0 \mid A, L)$
internally and compute stabilized IPCW weights. This requires a binary
censoring indicator column (1 = censored, 0 = observed) and `ipcw = TRUE`.

**Censoring mechanism.** Starting from the baseline DGM above (ATE $= 3$),
we introduce MAR outcome censoring. The probability of being censored
depends on both the treatment $A$ and confounder $L_1$:

$$
\begin{aligned}
C_i &\sim \text{Bernoulli}\bigl(\text{logit}^{-1}(-2 + 0.5\,A_i + 0.3\,L_{1,i})\bigr) \\
Y_i^{\text{obs}} &= \begin{cases} Y_i & \text{if } C_i = 0 \\ \text{NA} & \text{if } C_i = 1 \end{cases}
\end{aligned}
$$

The intercept of $-2$ keeps the censoring rate moderate. Because censoring
depends on $A$ and $L_1$, a complete-case analysis is biased; IPCW weights
restore consistency. The true ATE remains 3.

```{r}
df_mar <- df
cens_prob <- plogis(-2 + 0.5 * df_mar$A + 0.3 * df_mar$L1)
C <- as.integer(runif(n) < cens_prob)
df_mar$C <- C
df_mar$Y[C == 1L] <- NA
```

### G-computation with IPCW

```{r}
fit_ipcw <- causat(
  df_mar,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  censoring = "C",
  ipcw = TRUE,
  model_fn = stats::glm
)

res_ipcw <- contrast(
  fit_ipcw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_ipcw
```

### IPW with IPCW

For IPW, IPCW is essential under MAR censoring --- without it, the
density-ratio weights cannot correct for the selection bias introduced
by differential censoring.

```{r}
fit_ipcw_ipw <- causat(
  df_mar,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  estimator = "ipw",
  censoring = "C",
  ipcw = TRUE,
  model_fn = stats::glm,
  propensity_model_fn = stats::glm
)

res_ipcw_ipw <- contrast(
  fit_ipcw_ipw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_ipcw_ipw
```

### Comparison: complete-case vs IPCW

```{r}
fit_naive <- causat(
  df_mar,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  censoring = "C",
  model_fn = stats::glm
)
res_naive <- contrast(
  fit_naive,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)

comp_ipcw <- data.frame(
  Analysis = c("Complete-case (naive)", "G-comp + IPCW", "IPW + IPCW"),
  Estimate = c(
    res_naive$contrasts$estimate[1],
    res_ipcw$contrasts$estimate[1],
    res_ipcw_ipw$contrasts$estimate[1]
  ),
  SE = c(
    res_naive$contrasts$se[1],
    res_ipcw$contrasts$se[1],
    res_ipcw_ipw$contrasts$se[1]
  )
)
tt(comp_ipcw, digits = 3)
```

The IPCW estimates should be closer to the truth (ATE = 3) than naive
complete-case analysis:

```{r}
#| label: fig-ipcw-comparison
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Pointrange plot comparing naive complete-case, g-comp + IPCW, and IPW + IPCW estimates under MAR censoring. A dashed line marks the true ATE of 3. IPCW-corrected estimates are closer to the truth."
ipcw_plot_df <- data.frame(
  Method = factor(
    c("Naive (complete-case)", "G-comp + IPCW", "IPW + IPCW"),
    levels = c("Naive (complete-case)", "G-comp + IPCW", "IPW + IPCW")
  ),
  Estimate = c(
    res_naive$contrasts$estimate[1],
    res_ipcw$contrasts$estimate[1],
    res_ipcw_ipw$contrasts$estimate[1]
  ),
  CI_lower = c(
    res_naive$contrasts$ci_lower[1],
    res_ipcw$contrasts$ci_lower[1],
    res_ipcw_ipw$contrasts$ci_lower[1]
  ),
  CI_upper = c(
    res_naive$contrasts$ci_upper[1],
    res_ipcw$contrasts$ci_upper[1],
    res_ipcw_ipw$contrasts$ci_upper[1]
  )
)

tinyplot::tinyplot(
  Estimate ~ Method,
  data = ipcw_plot_df,
  type = "pointrange",
  ymin = ipcw_plot_df$CI_lower,
  ymax = ipcw_plot_df$CI_upper,
  pch = 19,
  ylab = "ATE estimate (95% CI)",
  main = "MAR censoring: naive vs IPCW correction"
)
abline(h = 3, lty = 2, col = "grey40")
```

### Diagnostics

```{r}
diag_ipcw <- diagnose(fit_ipcw)
diag_ipcw
```

The censoring panel shows IPCW weight summaries and the censoring model's
positivity (predicted $P(C = 0 \mid A, L)$ quantiles). Watch for extreme
weights or near-zero uncensoring probabilities.

## Manual IPCW via `weights =`

For advanced users who want full control over the censoring model, you can
compute IPCW weights externally and pass them via `weights =`:

```{r}
observed <- !is.na(df$Y) | (df_mar$C == 0L)
cens_model <- glm(observed ~ A + L1 + L2, data = df_mar, family = binomial)
ipcw_manual <- ifelse(observed, 1 / predict(cens_model, type = "response"), 0)

fit_manual_ipcw <- causat(
  df_mar[observed, ],
  outcome = "Y",
  treatment = "A",
  confounders = ~ L1 + L2,
  weights = ipcw_manual[observed],
  model_fn = stats::glm
)

res_manual_ipcw <- contrast(
  fit_manual_ipcw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_manual_ipcw
```

Note: with manual IPCW, the sandwich SE does *not* account for censoring-model
estimation uncertainty (the Channel 2 correction). Use `ipcw = TRUE` for
correct standard errors, or `ci_method = "bootstrap"` with manual weights.

## Summary of missingness handling

```{r}
#| echo: false
summary_df <- data.frame(
  Variable = c("Outcome (Y)", "Outcome (Y)", "Outcome (Y)",
               "Outcome (Y)", "Outcome (Y)",
               "Treatment (A)", "Treatment (A)",
               "Covariates (L)", "Covariates (L)", "Y + L"),
  Mechanism = c("MCAR", "MCAR", "MCAR", "MAR", "MAR",
                "NA to causat()", "MAR", "MCAR", "MAR", "MCAR"),
  Method = c("G-comp", "IPW", "Matching",
             "Any + ipcw=TRUE", "Any + manual weights",
             "Any", "Any", "G-comp", "Any", "G-comp"),
  Handling = c(
    "Complete-case model, predict on all rows",
    "Fit treatment model on observed-Y subset",
    "MatchIt excludes NA-Y rows",
    "Built-in IPCW via censoring= + ipcw=TRUE",
    "External IPCW weights via weights=",
    "Rejected (abort with error)",
    "Impute A, then causat_mice()",
    "na.omit drops rows; predict returns NA",
    "Multiple imputation via causat_mice()",
    "Combines both drop rules"
  ),
  Status = c("Supported", "Supported", "Supported",
             "Supported", "Supported",
             "Rejected", "Supported", "Supported", "Supported", "Supported")
)
tt(summary_df)
```

## Key takeaways

1. **Missing outcomes (MCAR)** are handled automatically by all estimators.
   The estimate is unbiased; the SE increases proportionally to the information
   loss.

2. **Missing treatment values** cannot be passed to `causat()` directly. Drop
   rows (MCAR) or impute with `mice()` + `causat_mice()` (MAR).

3. **Missing covariates (MCAR)** work out-of-the-box under g-comp via
   `na.omit`. For MAR missingness, use multiple imputation via `causat_mice()`
   (Rubin's rules by default, Boot MI when congeniality is in doubt). Always
   include the outcome as a *predictor* in the imputation model.

4. **MAR outcome missingness** is handled by `ipcw = TRUE`, which fits a
   censoring model and computes stabilized IPCW weights automatically.
   Advanced users can also supply manual IPCW weights via `weights =`.

5. **Bootstrap** handles all missingness patterns correctly by resampling
   the full pipeline.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 12 (IP weighting under censoring) and Chapter 19 (censoring in
longitudinal studies).

Little RJA, Rubin DB (2019). *Statistical Analysis with Missing Data*. 3rd ed.
Wiley.

Rubin DB (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.

Barnard J, Rubin DB (1999). Small-sample degrees of freedom with multiple
imputation. *Biometrika* 86(4):948-955.

Meng XL (1994). Multiple-imputation inferences with uncongenial sources of
input. *Statistical Science* 9(4):538-558.

van Buuren S (2018). *Flexible Imputation of Missing Data*. 2nd ed. CRC Press.

Bartlett JW, Hughes RA (2020). Bootstrap inference for multiple imputation under
uncongeniality and misspecification. *Statistical Methods in Medical Research*
29(12):3533-3546.

von Hippel PT (2020). How many imputations do you need? A two-stage calculation
using a quadratic rule. *Sociological Methods & Research* 49(3):699-718.
