---
title: "Structural nested mean models with causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Structural nested mean models with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Structural nested mean models (SNMMs) are the third leg of the Robins
triangle — alongside g-formula (gcomp/ICE) and inverse probability weighting
(IPW-MSM). They estimate **blip parameters** that describe how treatment at
each time point causally shifts the outcome, conditional on covariates.

The headline feature: SNMMs correctly handle **time-varying effect
modification** — modifiers that are themselves affected by treatment. IPW-MSM
and matching require modifiers to be baseline (pre-treatment) variables;
conditioning on a post-treatment modifier in an MSM opens collider paths and
biases the estimand (Robins 2000). SNMMs identify the blip under
treatment-model correctness alone, so post-treatment modifiers are valid
(Vansteelandt & Joffe 2014).

## When must you use SNMs?

When your research question involves **effect modification by a variable
that is itself affected by treatment**, the standard estimators produce biased
results. This section demonstrates the problem with a simulated DGP and shows
that SNM g-estimation recovers the truth.

### The DGP

```{r}
#| label: dgp-tv-modifier
library(causatr)

set.seed(1801)
n <- 5000
L <- rnorm(n)
A <- rnorm(n, mean = 0.5 * L, sd = 1)

# M is post-treatment: it depends on A
M <- 0.3 * A + 0.5 * L + rnorm(n, sd = sqrt(0.5))

# Y depends on A, L, and the A*M interaction
Y <- 2 + 3 * A + 1.5 * L + 2 * A * M + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)
```

The structural blip is $\gamma(a, m; \psi) = a \cdot (\psi_0 + \psi_M \cdot m)$
with truth $\psi_0 = 3$ and $\psi_M = 2$.

### IPW-MSM: biased

When we include M as an effect modifier in an IPW-MSM, the estimator
conditions on a descendant of A. This opens a non-causal path and produces
severely biased estimates:

```{r}
#| label: ipw-biased
fit_ipw <- causat(d, outcome = "Y", treatment = "A",
                  confounders = ~ L + A:M, estimator = "ipw")
r_ipw <- contrast(fit_ipw,
                  interventions = list(a1 = shift(1), a0 = shift(0)),
                  type = "difference")
r_ipw$contrasts
```

The IPW estimate is far from the truth (the averaged structural blip is
approximately $3 + 2 \times \bar{M} \approx 3 + 2 \times 0.25 = 3.5$).

### SNM g-estimation: correct

The SNM identifies the blip using the treatment residual as an instrument.
With a treatment-free model to absorb the $L \to Y$ pathway, the structural
blip parameters are recovered:

```{r}
#| label: snm-correct
fit_snm <- causat(d, outcome = "Y", treatment = "A",
                  confounders_outcome = ~ L + A:M,
                  confounders_treatment = ~ L,
                  estimator = "snm",
                  treatment_free = ~ L)

# Blip parameter table
r_blip <- contrast(fit_snm, ci_method = "sandwich")
r_blip$estimates
```

The estimated $\hat\psi_0 \approx 3$ and $\hat\psi_M \approx 2$ match the
truth closely, with tight standard errors.

### Averaged blip effect

To get a single summary number (like an ATE), use `treatment_values`:

```{r}
#| label: snm-averaged
r_avg <- contrast(fit_snm, treatment_values = c(0, 1))
r_avg$estimates
```

## Basic usage

### Point treatment with effect modification

```{r}
#| label: basic-em
set.seed(42)
n <- 2000
L <- rnorm(n)
sex <- rbinom(n, 1, 0.5)
A <- rbinom(n, 1, plogis(0.3 * L + 0.2 * sex))
Y <- 1 + 2 * A + L + 1.5 * A * sex + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, sex = sex)

fit <- causat(d, outcome = "Y", treatment = "A",
              confounders_outcome = ~ L + sex + A:sex,
              confounders_treatment = ~ L + sex,
              estimator = "snm",
              treatment_free = ~ L + sex)

contrast(fit, ci_method = "sandwich")
```

The blip parameters are: $\psi_{\text{intercept}}$ (treatment effect when
`sex = 0`) and $\psi_{\text{sex}}$ (additional effect when `sex = 1`).

### Continuous treatment

SNMs work naturally with continuous treatments — the canonical SNMM case.
The treatment model is a Gaussian GLM by default:

```{r}
#| label: continuous-treatment
set.seed(101)
n <- 2000
L <- rnorm(n)
A <- rnorm(n, mean = 0.5 * L, sd = 1)
M <- as.numeric(L > 0)
Y <- 2 + 3 * A + 1.5 * L + 2 * A * M + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)

fit <- causat(d, outcome = "Y", treatment = "A",
              confounders_outcome = ~ L + A:M,
              confounders_treatment = ~ L,
              estimator = "snm")

contrast(fit, ci_method = "sandwich")
```

### Treatment-free model for efficiency

The `treatment_free` argument specifies an outcome model $E[Y \mid L]$ that
absorbs the $L \to Y$ variance, improving efficiency (Vansteelandt & Joffe
2014). It does not change the point estimates when the blip is correctly
specified — only the standard errors shrink:

```{r}
#| label: tf-efficiency
# Without treatment-free model
r_no_tf <- contrast(fit, ci_method = "sandwich")

# With treatment-free model
fit_tf <- causat(d, outcome = "Y", treatment = "A",
                 confounders_outcome = ~ L + A:M,
                 confounders_treatment = ~ L,
                 estimator = "snm",
                 treatment_free = ~ L)
r_tf <- contrast(fit_tf, ci_method = "sandwich")

# Compare SEs
data.table::data.table(
  parameter = r_no_tf$estimates$parameter,
  se_without_tf = r_no_tf$estimates$se,
  se_with_tf = r_tf$estimates$se,
  reduction_pct = round(100 * (1 - r_tf$estimates$se / r_no_tf$estimates$se), 1)
)
```

## Interpretation

SNM blip parameters have a direct causal interpretation:

- **$\psi_{\text{intercept}}$**: the causal effect of a one-unit treatment
  increase in the reference stratum (all modifiers = 0).
- **$\psi_{\text{modifier}}$**: the additional causal effect per unit of the
  modifier. This is the effect modification parameter.

The averaged blip effect $\Delta = (a_1 - a_0) \cdot (\hat\psi_0 + \sum_j \hat\psi_j \bar{m}_j)$
is an ATE-like summary that averages the blip over the observed modifier
distribution.

## Key differences from other estimators

| Feature | gcomp | IPW-MSM | SNM |
|---|---|---|---|
| Time-varying EM | Requires correct high-dimensional outcome model | Biased (conditions on post-treatment) | Correct (headline use case) |
| Identification | Outcome model | Treatment model | Treatment model |
| Output | Counterfactual means | Counterfactual means | Blip parameters |
| `interventions =` | Yes | Yes | No (`treatment_values =` instead) |
| Near-positivity | Extrapolation risk | Weight explosion | Degrades gracefully |

## References

- Robins JM (1994). Correcting for non-compliance using structural nested mean
  models. *Comm Stat Theory Methods* 23:2379--2412.
- Vansteelandt S, Joffe MM (2014). Structural nested models and g-estimation:
  The partially realized promise. *Stat Sci* 29:707--731.
- Robins JM (2000). Marginal structural models versus structural nested models
  as tools for causal inference. In *Statistical Models in Epidemiology*.
