---
title: "Structural nested mean models with causatr"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Structural nested mean models with causatr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Structural nested mean models (SNMMs) are the third leg of the Robins
triangle — alongside g-formula (gcomp/ICE) and inverse probability weighting
(IPW-MSM). They estimate **blip parameters** that describe how treatment at
each time point causally shifts the outcome, conditional on covariates.

The headline feature: SNMMs correctly handle **time-varying effect
modification** — modifiers that are themselves affected by treatment. IPW-MSM
and matching require modifiers to be baseline (pre-treatment) variables;
conditioning on a post-treatment modifier in an MSM opens collider paths and
biases the estimand (Robins 2000). SNMMs identify the blip under
treatment-model correctness alone, so post-treatment modifiers are valid
(Vansteelandt & Joffe 2014).

## What SNMs estimate vs what MSMs estimate

Before diving into settings where only SNMs work, it helps to see what makes
them *different* — even in settings where both estimators are perfectly valid.

The core distinction (Robins 2000):

- **MSMs** answer: "What would the population mean outcome be if everyone
  received treatment $a$?" The estimand is $E[Y(a)]$, a marginal
  counterfactual mean.
- **SNMs** answer: "What is the causal effect of the treatment actually
  received, as a function of covariates?" The estimand is the **blip function**
  $\gamma(a, \ell; \psi) = E[Y(a) - Y(0) \mid A = a, L = \ell]$, a
  conditional treatment effect structure.

Both encode the same underlying causal truth, but through different lenses.
MSMs give you summary-level potential outcomes under hypothetical regimes.
SNMs give you the *architecture* of the treatment effect — how it varies
across covariate strata.

### A shared DGP

Consider a simple point treatment with a baseline binary modifier $M$:

```{r}
#| label: dgp-shared
library(causatr)

set.seed(1800)
n <- 5000
L <- rnorm(n)
M <- rbinom(n, 1, 0.5)
A <- rbinom(n, 1, plogis(0.3 * L + 0.2 * M))
Y <- 1 + 2 * A + 3 * A * M + 0.5 * L + 0.8 * M + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)
```

The structural truth:

- Blip: $\gamma(A, M; \psi) = A \cdot (\psi_0 + \psi_M \cdot M)$ with
  $\psi_0 = 2$ and $\psi_M = 3$.
- When $M = 0$: the treatment effect is $2$.
- When $M = 1$: the treatment effect is $2 + 3 = 5$.
- Marginal ATE: $\psi_0 + \psi_M \cdot P(M = 1) = 2 + 3 \times 0.5 = 3.5$.

All three numbers are correct descriptions of the same DGP. The question is
which framing you want.

### The MSM question: what happens under an intervention?

G-computation and IPW-MSM both compute counterfactual means $\hat{E}[Y(a)]$
for each intervention regime, then contrast them. The output is an ATE — a
single summary number:

```{r}
#| label: gcomp-ate
fit_gc <- causat(d, outcome = "Y", treatment = "A",
                 confounders = ~ L + M + A:M, estimator = "gcomp",
                 model_fn = stats::glm)
r_gc <- contrast(fit_gc,
                 interventions = list(a1 = static(1), a0 = static(0)),
                 type = "difference", ci_method = "sandwich")
r_gc$contrasts
```

The g-comp estimate gives $\widehat{\text{ATE}} \approx 3.5$, the correct
marginal treatment effect. This answers: "On average, how much does treatment
help?"

To uncover subgroup-specific effects, the MSM/g-comp approach can be
stratified with `by`:

```{r}
#| label: gcomp-by
r_gc_by <- contrast(fit_gc,
                    interventions = list(a1 = static(1), a0 = static(0)),
                    type = "difference", ci_method = "sandwich",
                    by = "M")
r_gc_by$contrasts
```

Now we see the subgroup-specific ATEs: $\approx 2$ when $M = 0$ and $\approx
5$ when $M = 1$. But these are still counterfactual mean differences — the
MSM framework has no single parameter that directly represents the
modification effect. You need to run the estimator twice (or use `by`) and
then compare the subgroup estimates by eye. The effect modification itself is
a *derived* quantity, not a model parameter.

### The SNM question: what is the structure of the treatment effect?

The SNM directly estimates the blip parameters $\psi$ that describe how
treatment modifies the outcome as a function of covariates:

```{r}
#| label: snm-blip
fit_snm <- causat(d, outcome = "Y", treatment = "A",
                  confounders_outcome = ~ L + M + A:M,
                  confounders_treatment = ~ L + M,
                  estimator = "snm",
                  treatment_free = ~ L + M)
r_snm <- contrast(fit_snm, ci_method = "sandwich")
r_snm$estimates
```

The SNM output is not a counterfactual mean or a contrast between regimes.
It is the **effect modification structure itself**: $\hat\psi_0 \approx 2$
(effect when $M = 0$) and $\hat\psi_M \approx 3$ (additional effect when
$M = 1$). These are directly interpretable:

- The treatment helps everyone ($\psi_0 > 0$), but it helps much more when
  $M = 1$ ($\psi_M > 0$, and the $z$-test on $\psi_M$ is the direct test of
  effect modification).
- No post-hoc stratification or subgroup comparisons are needed — the blip
  parameters *are* the effect modification estimates.

### Connecting the two

The MSM's marginal ATE and the SNM's blip parameters encode the same causal
truth. The averaged blip, $\Delta = (a_1 - a_0) \cdot (\hat\psi_0 + \hat\psi_M
\cdot \bar{M})$, recovers the MSM's ATE:

```{r}
#| label: averaged-blip
r_avg <- contrast(fit_snm, treatment_values = c(0, 1))
r_avg$estimates
```

The averaged blip estimate ($\approx 3.5$) matches the g-comp estimate. The
SNM decomposes this average into its components; the MSM reports the
composite directly.

The following figure shows all six estimates side by side. The g-comp
(MSM) panel reports counterfactual mean differences at three levels of
aggregation; the SNM (blip) panel reports the structural effect parameters.
Both recover the truth (dashed lines), but through different lenses:

```{r}
#| label: fig-msm-vs-snm
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Pointrange plot comparing g-comp (MSM) and SNM estimates from the same DGP. The MSM panel shows marginal ATE and subgroup-specific ATEs; the SNM panel shows blip parameters and the averaged blip. All estimates bracket the true values."
comp_df <- data.frame(
  estimator = factor(
    rep(c("G-comp (MSM)", "SNM (blip)"), each = 3),
    levels = c("G-comp (MSM)", "SNM (blip)")
  ),
  parameter = factor(
    c("Marginal ATE", "ATE | M = 0", "ATE | M = 1",
      "Averaged blip", "psi_0 (intercept)", "psi_M (modifier)"),
    levels = c("psi_M (modifier)", "psi_0 (intercept)", "Averaged blip",
               "ATE | M = 1", "ATE | M = 0", "Marginal ATE")
  ),
  estimate = c(-r_gc$contrasts$estimate[1],
               -r_gc_by$contrasts$estimate[1],
               -r_gc_by$contrasts$estimate[2],
               r_avg$estimates$estimate[1],
               r_snm$estimates$estimate[1],
               r_snm$estimates$estimate[2]),
  ci_lower = c(-r_gc$contrasts$ci_upper[1],
               -r_gc_by$contrasts$ci_upper[1],
               -r_gc_by$contrasts$ci_upper[2],
               r_avg$estimates$ci_lower[1],
               r_snm$estimates$ci_lower[1],
               r_snm$estimates$ci_lower[2]),
  ci_upper = c(-r_gc$contrasts$ci_lower[1],
               -r_gc_by$contrasts$ci_lower[1],
               -r_gc_by$contrasts$ci_lower[2],
               r_avg$estimates$ci_upper[1],
               r_snm$estimates$ci_upper[1],
               r_snm$estimates$ci_upper[2]),
  truth = c(3.5, 2, 5, 3.5, 2, 3)
)

tinyplot::tinyplot(
  estimate ~ parameter | estimator,
  data = comp_df,
  type = "pointrange",
  ymin = comp_df$ci_lower,
  ymax = comp_df$ci_upper,
  pch = 19,
  ylab = "Estimate (95% CI)",
  xlab = "",
  main = "Same DGP, different questions",
  palette = "dark2",
  axes = "l"
)
```

The subgroup-specific MSM estimates ($\text{ATE} \mid M = 0$ and $\text{ATE}
\mid M = 1$) correspond to the blip evaluated at each modifier level. But the
mapping is one-directional: you can always aggregate blip parameters into
an ATE, but you cannot decompose an ATE into blip parameters without
specifying a blip model.

### When does framing matter?

In this simple baseline-modifier setting, both estimators work and the choice
is about what you want to report:

| Goal | Better tool |
|---|---|
| Population-level treatment policy decision | MSM (counterfactual means under regimes) |
| Understanding *who benefits most* from treatment | SNM (blip parameters directly quantify heterogeneity) |
| Formal test of effect modification | SNM ($z$-test on $\psi_M$) |
| Dose-response under a specific intervention | MSM with `shift()` or `scale_by()` |
| Optimal treatment rule (DTR) estimation | SNM (blip sign determines optimal assignment) |

The next section shows where the SNM is not just convenient but
**necessary**: when the modifier is itself affected by treatment. There, the
MSM cannot include $M$ without introducing collider bias, while the SNM
handles it natively.

## When must you use SNMs?

When your research question involves **effect modification by a variable
that is itself affected by treatment**, the standard estimators produce biased
results. This section demonstrates the problem with a simulated DGP and shows
that SNM g-estimation recovers the truth.

### The DGP

```{r}
#| label: dgp-tv-modifier
set.seed(1801)
n <- 5000
L <- rnorm(n)
A <- rnorm(n, mean = 0.5 * L, sd = 1)

# M is post-treatment: it depends on A
M <- 0.3 * A + 0.5 * L + rnorm(n, sd = sqrt(0.5))

# Y depends on A, L, and the A*M interaction
Y <- 2 + 3 * A + 1.5 * L + 2 * A * M + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)
```

The structural blip is $\gamma(a, m; \psi) = a \cdot (\psi_0 + \psi_M \cdot m)$
with truth $\psi_0 = 3$ and $\psi_M = 2$.

### IPW-MSM: biased

When we include M as an effect modifier in an IPW-MSM, the estimator
conditions on a descendant of A. This opens a non-causal path and produces
severely biased estimates:

```{r}
#| label: ipw-biased
fit_ipw <- causat(d, outcome = "Y", treatment = "A",
                  confounders = ~ L + A:M, estimator = "ipw")
r_ipw <- contrast(fit_ipw,
                  interventions = list(a1 = shift(1), a0 = shift(0)),
                  type = "difference")
r_ipw$contrasts
```

The IPW estimate is far from the truth (the averaged structural blip is
approximately $3 + 2 \times \bar{M} \approx 3 + 2 \times 0.25 = 3.5$).

### SNM g-estimation: correct

The SNM identifies the blip using the treatment residual as an instrument.
With a treatment-free model to absorb the $L \to Y$ pathway, the structural
blip parameters are recovered:

```{r}
#| label: snm-correct
fit_snm <- causat(d, outcome = "Y", treatment = "A",
                  confounders_outcome = ~ L + A:M,
                  confounders_treatment = ~ L,
                  estimator = "snm",
                  treatment_free = ~ L)

# Blip parameter table
r_blip <- contrast(fit_snm, ci_method = "sandwich")
r_blip$estimates
```

The estimated $\hat\psi_0 \approx 3$ and $\hat\psi_M \approx 2$ match the
truth closely, with tight standard errors.

### Averaged blip effect

To get a single summary number (like an ATE), use `treatment_values`:

```{r}
#| label: snm-averaged
r_avg <- contrast(fit_snm, treatment_values = c(0, 1))
r_avg$estimates
```

### Visualizing the bias

The contrast between the two approaches is clearest in a figure. The IPW-MSM
estimate ($\approx 6.5$) is far from the true averaged blip ($\approx 3.5$,
dashed line), while the SNM recovers the truth:

```{r}
#| label: fig-bias-comparison
#| fig-width: 6
#| fig-height: 4
#| fig-alt: "Pointrange plot showing the IPW-MSM estimate (severely biased at ~6.5) versus the SNM averaged blip (correct at ~3.5). A dashed line marks the true value."
bias_df <- data.frame(
  Method = factor(
    c("IPW-MSM (biased)", "SNM (correct)"),
    levels = c("IPW-MSM (biased)", "SNM (correct)")
  ),
  estimate = c(-r_ipw$contrasts$estimate[1],
               r_avg$estimates$estimate[1]),
  ci_lower = c(-r_ipw$contrasts$ci_upper[1],
               r_avg$estimates$ci_lower[1]),
  ci_upper = c(-r_ipw$contrasts$ci_lower[1],
               r_avg$estimates$ci_upper[1])
)

# Approximate truth: 3 + 2 * mean(M) where M = 0.3*A + 0.5*L + noise
avg_blip_truth <- 3 + 2 * mean(d$M)

tinyplot::tinyplot(
  estimate ~ Method,
  data = bias_df,
  type = "pointrange",
  ymin = bias_df$ci_lower,
  ymax = bias_df$ci_upper,
  pch = 19,
  ylab = "Effect estimate (95% CI)",
  main = "Post-treatment effect modification:\nIPW-MSM vs SNM"
)
abline(h = avg_blip_truth, lty = 2, col = "grey40")
```

## Basic usage

### Point treatment with effect modification

```{r}
#| label: basic-em
set.seed(42)
n <- 2000
L <- rnorm(n)
sex <- rbinom(n, 1, 0.5)
A <- rbinom(n, 1, plogis(0.3 * L + 0.2 * sex))
Y <- 1 + 2 * A + L + 1.5 * A * sex + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, sex = sex)

fit <- causat(d, outcome = "Y", treatment = "A",
              confounders_outcome = ~ L + sex + A:sex,
              confounders_treatment = ~ L + sex,
              estimator = "snm",
              treatment_free = ~ L + sex)

contrast(fit, ci_method = "sandwich")
```

The blip parameters are: $\psi_{\text{intercept}}$ (treatment effect when
`sex = 0`) and $\psi_{\text{sex}}$ (additional effect when `sex = 1`).

### Continuous treatment

SNMs work naturally with continuous treatments — the canonical SNMM case.
The treatment model is a Gaussian GLM by default:

```{r}
#| label: continuous-treatment
set.seed(101)
n <- 2000
L <- rnorm(n)
A <- rnorm(n, mean = 0.5 * L, sd = 1)
M <- as.numeric(L > 0)
Y <- 2 + 3 * A + 1.5 * L + 2 * A * M + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)

fit <- causat(d, outcome = "Y", treatment = "A",
              confounders_outcome = ~ L + A:M,
              confounders_treatment = ~ L,
              estimator = "snm")

contrast(fit, ci_method = "sandwich")
```

### Treatment-free model for efficiency

The `treatment_free` argument specifies an outcome model $E[Y \mid L]$ that
absorbs the $L \to Y$ variance, improving efficiency (Vansteelandt & Joffe
2014). It does not change the point estimates when the blip is correctly
specified — only the standard errors shrink:

```{r}
#| label: tf-efficiency
# Without treatment-free model
r_no_tf <- contrast(fit, ci_method = "sandwich")

# With treatment-free model
fit_tf <- causat(d, outcome = "Y", treatment = "A",
                 confounders_outcome = ~ L + A:M,
                 confounders_treatment = ~ L,
                 estimator = "snm",
                 treatment_free = ~ L)
r_tf <- contrast(fit_tf, ci_method = "sandwich")

# Compare SEs
data.table::data.table(
  parameter = r_no_tf$estimates$parameter,
  se_without_tf = r_no_tf$estimates$se,
  se_with_tf = r_tf$estimates$se,
  reduction_pct = round(100 * (1 - r_tf$estimates$se / r_no_tf$estimates$se), 1)
)
```

## Interpretation

SNM blip parameters have a direct causal interpretation:

- **$\psi_{\text{intercept}}$**: the causal effect of a one-unit treatment
  increase in the reference stratum (all modifiers = 0).
- **$\psi_{\text{modifier}}$**: the additional causal effect per unit of the
  modifier. This is the effect modification parameter.

The averaged blip effect $\Delta = (a_1 - a_0) \cdot (\hat\psi_0 + \sum_j \hat\psi_j \bar{m}_j)$
is an ATE-like summary that averages the blip over the observed modifier
distribution.

## Key differences from other estimators

| Feature | gcomp | IPW-MSM | SNM |
|---|---|---|---|
| Time-varying EM | Requires correct high-dimensional outcome model | Biased (conditions on post-treatment) | Correct (headline use case) |
| Identification | Outcome model | Treatment model | Treatment model |
| Output | Counterfactual means | Counterfactual means | Blip parameters |
| `interventions =` | Yes | Yes | No (`treatment_values =` instead) |
| Near-positivity | Extrapolation risk | Weight explosion | Degrades gracefully |

## Longitudinal SNM

SNMs extend naturally to time-varying treatments. The backward-sequential
g-estimation algorithm (Robins 1994) estimates per-stage blip parameters
$\psi_k$ by iterating from the last time point backward:

1. At stage $K$, solve the blip equation for $\psi_K$ using the observed
   outcome $Y$.
2. At stage $k < K$, replace $Y$ with the **backward-transformed outcome**
   $H_k = Y - \sum_{j > k} \gamma_j(A_j, \bar{L}_j; \hat\psi_j)$ — the
   hypothetical outcome under no treatment from $k+1$ onward.

### Two-period DGP with time-varying effect modification

This is the headline use case: the modifier $M_1 = \mathbf{1}\{L_1 > 0\}$
is post-treatment (it depends on $A_0$), so IPW-MSM cannot include it
without collider bias. The SNM identifies the per-stage blip correctly.

```{r}
#| label: longitudinal-snm
set.seed(18)
n <- 5000

L0 <- rnorm(n)
A0 <- rbinom(n, 1, plogis(0.3 * L0))
L1 <- 0.5 * L0 + 0.3 * A0 + rnorm(n, sd = sqrt(0.5))
M1 <- as.numeric(L1 > 0)
A1 <- rbinom(n, 1, plogis(0.3 * L1 + 0.2 * A0))
Y <- 2 + 1 * A0 + 2 * A1 + 2 * A1 * M1 + 1.5 * L0 + 0.5 * L1 + rnorm(n)

d_long <- data.table::rbindlist(list(
  data.table::data.table(id = seq_len(n), time = 0L,
    A = A0, L = L0, M = as.numeric(L0 > 0), Y = NA_real_),
  data.table::data.table(id = seq_len(n), time = 1L,
    A = A1, L = L1, M = M1, Y = Y)
))

fit_long <- causat(d_long, outcome = "Y", treatment = "A",
  confounders_outcome = ~ A:M,
  confounders_tv = ~ L,
  id = "id", time = "time", type = "longitudinal",
  estimator = "snm",
  treatment_free = ~ L)

result_long <- contrast(fit_long, ci_method = "sandwich")
result_long$estimates
```

The truth is $\psi_{0,\text{intercept}} = 1$, $\psi_{1,\text{intercept}} = 2$,
$\psi_{1,M} = 2$. The SNM recovers these per-stage blip parameters with
backward-sequential g-estimation and cluster-robust sandwich standard errors.

The forest plot shows all per-stage blip parameters with 95% CIs:

```{r}
#| label: fig-longitudinal-blip
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot of longitudinal SNM blip parameters. Three parameters are shown: psi_0 intercept near 1, psi_1 intercept near 2, and psi_1_M near 2, all with tight confidence intervals around the true values."
plot(result_long)
```

## Categorical treatment

SNMs work with categorical treatments ($k > 2$ levels). The treatment model is
a multinomial logistic regression via `nnet::multinom()`. The blip expands to
$(K-1)$ level-specific parameters (relative to the reference level):

```{r}
#| label: categorical-treatment
set.seed(1803)
n <- 3000
L <- rnorm(n)

# Multinomial probabilities from softmax
eta1 <- 0.2 + 0.5 * L
eta2 <- -0.3 + 0.3 * L
denom <- 1 + exp(eta1) + exp(eta2)
A_num <- sapply(seq_len(n), function(i) {
  sample(0:2, 1, prob = c(1 / denom[i], exp(eta1[i]) / denom[i],
                           exp(eta2[i]) / denom[i]))
})
A <- factor(A_num, levels = c("0", "1", "2"))
Y <- 2 + 0.8 * (A_num == 1) + 1.5 * (A_num == 2) + L + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L)

fit_cat <- causat(d, outcome = "Y", treatment = "A",
                  confounders = ~ L,
                  estimator = "snm",
                  propensity_model_fn = nnet::multinom)

contrast(fit_cat, ci_method = "sandwich")
```

The truth is $\psi_1 = 0.8$ (level 1 vs 0) and $\psi_2 = 1.5$ (level 2 vs 0).
Note that `treatment_values` is not available for categorical SNMs — the
per-level blip parameters are the estimand.

## Count treatment

Poisson or negative-binomial treatments use the appropriate propensity model.
The residual $R = A - \hat\lambda$ is used in the g-estimating equation:

```{r}
#| label: count-treatment
set.seed(1804)
n <- 3000
L <- rnorm(n)
A <- rpois(n, lambda = exp(0.5 + 0.3 * L))
M <- as.numeric(L > 0)
Y <- 2 + 0.5 * A + 0.3 * A * M + L + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)

fit_count <- causat(d, outcome = "Y", treatment = "A",
                    confounders_outcome = ~ L + A:M,
                    confounders_treatment = ~ L,
                    estimator = "snm",
                    propensity_family = "poisson")

contrast(fit_count, ci_method = "sandwich")
```

The truth is $\psi_0 = 0.5$ and $\psi_M = 0.3$.

## Bootstrap confidence intervals

All SNM variants support bootstrap inference via `ci_method = "bootstrap"`.
For longitudinal SNMs, the bootstrap uses ID-clustered resampling:

```{r}
#| label: bootstrap-ci
set.seed(1805)
n <- 2000
L <- rnorm(n)
A <- rnorm(n, mean = 0.5 * L, sd = 1)
M <- 0.3 * A + 0.5 * L + rnorm(n, sd = sqrt(0.5))
Y <- 2 + 3 * A + 1.5 * L + 2 * A * M + rnorm(n)
d <- data.table::data.table(Y = Y, A = A, L = L, M = M)

fit <- causat(d, outcome = "Y", treatment = "A",
              confounders_outcome = ~ L + A:M,
              confounders_treatment = ~ L,
              estimator = "snm",
              treatment_free = ~ L)

r_boot <- contrast(fit, ci_method = "bootstrap", n_boot = 500)
r_boot$estimates
```

## Inspecting results

The standard S3 methods work with SNM results. `print()` displays a
"Blip parameters" header, `summary()` shows the blip variance-covariance
matrix, and `plot()` produces a forest plot with a zero reference line
(blip = 0 means no treatment effect):

```{r}
#| label: s3-methods
#| fig-width: 7
#| fig-height: 4
fit_demo <- causat(d, outcome = "Y", treatment = "A",
                   confounders_outcome = ~ L + A:M,
                   confounders_treatment = ~ L,
                   estimator = "snm",
                   treatment_free = ~ L)
r_demo <- contrast(fit_demo, ci_method = "sandwich")

# Forest plot: reference line at 0
plot(r_demo)
```

The `tidy()` method returns a tidy data frame compatible with `broom`:

```{r}
#| label: tidy-method
# Use which = "means" for blip parameters (SNM Path A has no contrasts)
tidy(r_demo, which = "means")
```

## References

- Robins JM (1994). Correcting for non-compliance using structural nested mean
  models. *Comm Stat Theory Methods* 23:2379--2412.
- Vansteelandt S, Joffe MM (2014). Structural nested models and g-estimation:
  The partially realized promise. *Stat Sci* 29:707--731.
- Robins JM (2000). Marginal structural models versus structural nested models
  as tools for causal inference. In *Statistical Models in Epidemiology*.
