---
title: "Causal survival analysis (Chapter 17)"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Causal survival analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette covers causal survival analysis in causatr. The current
implementation (Phase 7 scaffolding) exposes a fit-only path via
`causat_survival()`; the survival-curve contrast step (needed to produce
counterfactual survival curves, risk differences, and risk ratios under each
intervention) is still in development and currently aborts with an
informative error. Competing-risks analysis is also reserved for Phase 7.

The intended end-to-end workflow, once the contrast path lands, is a direct
translation of Chapter 17 of Hernán & Robins (2025): a discrete-time hazard
model fit by pooled logistic regression, followed by an individual-level
prediction sweep to obtain S^a(t) = prod(1 - h^a(m)) under each intervention.

## Setup

```{r}
#| message: false
library(causatr)
data("nhefs")
```

The NHEFS dataset includes a binary event indicator `death` (1 if the
individual died during follow-up) and a continuous `yrdth` column with the
year of death. For the pooled-logistic path we need person-period data with
one row per (individual, time interval). `to_person_period()` performs the
wide-to-long conversion.

## Fitting the pooled logistic hazard model

`causat_survival()` converts data to person-period format (if not already),
fits a pooled logistic regression for the discrete hazard
$\text{logit Pr}[D_{k+1} = 1 | \text{survived to } k, A, L]$, and stores the
fitted model in a `causatr_fit` object of `type = "survival"`.

```{r}
#| eval: false
fit_surv <- causat_survival(
  nhefs_person_period,
  outcome = "death_indicator",
  treatment = "qsmk",
  confounders = ~ sex + age + race + factor(education) + smokeintensity +
    smokeyrs + factor(exercise) + factor(active) + wt71,
  id = "seqn",
  time = "interval",
  time_formula = ~ splines::ns(time, 4)
)
fit_surv
```

Key arguments:

- `time_formula`: the function of time that enters the hazard model. Default
  `~ splines::ns(time, 4)` fits a smooth baseline hazard. Use
  `~ factor(time)` for a fully saturated (non-parametric) baseline.
- `censoring`: a time-varying censoring indicator column, used to drop rows
  where the individual has been lost to follow-up.
- `weights`: pre-computed IPCW or survey weights, validated upfront by
  `check_weights()`.
- `competing`: reserved for competing-risks analysis. Supplying a non-`NULL`
  value currently aborts — the plain pooled-logistic model it would wrap
  is biased under competing events, so causatr refuses to fit it until the
  proper sub-distribution / Aalen–Johansen path is implemented in Phase 7.

Under the hood, `causat_survival()` builds the risk set by dropping rows
after the first event for each individual (`prev_event > 0`), merges with
the censoring mask, and fits the weighted logistic model on the surviving
at-risk rows.

## Contrasts (not yet implemented)

The survival-curve contrast step would take the fitted hazard model and
compute, for each named intervention:

1. `apply_intervention()` on each (person, time) cell in the risk set.
2. `predict(fit_surv$model, newdata = ..., type = "response")` to get
   counterfactual discrete hazards $\hat h_i^a(k)$.
3. Cumulative product within each individual to obtain
   $\hat S_i^a(k) = \prod_{m \le k}(1 - \hat h_i^a(m))$.
4. Average across individuals in the target population to get $\hat S^a(k)$.
5. Contrasts across interventions: risk difference
   $(1 - \hat S^{a_1}(t)) - (1 - \hat S^{a_0}(t))$, risk ratio, etc., with
   the delta-method variance propagated from the hazard model's vcov.

Today, attempting `contrast(fit_surv, ...)` returns an informative error
pointing to this open scope. Track progress in `PHASE_7_SURVIVAL.md`.

```{r}
#| eval: false
# Will abort until Phase 7 contrast support lands:
result <- contrast(
  fit_surv,
  interventions = list(quit = static(1), continue = static(0)),
  type = "difference"
)
```

## Workaround: pooled logistic fit diagnostics today

Even without the contrast step, the fitted hazard model is a plain `glm()`
object and can be inspected directly:

```{r}
#| eval: false
summary(fit_surv$model)
coef(fit_surv$model)
```

For a quick sanity check before the Phase 7 contrast path is wired up, you
can extract the predicted hazard under each intervention manually, then
compute the cumulative survival in base R. This is exactly the loop that
will eventually move into `contrast()`.

## Tracking Phase 7

The survival path is tracked in:

- `PHASE_7_SURVIVAL.md` — design notes and the outstanding work list.
- `FEATURE_COVERAGE_MATRIX.md` §"Method 5 — Survival" — current fit-only
  coverage and the contrast rows marked as ❌ / GAP.
- `test-simulation.R` and `test-s3-methods.R` — fit-only smoke tests that
  guard the current scaffolding.

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 17: Causal survival analysis.

D'Agostino RB, Lee ML, Belanger AJ, Cupples LA, Anderson K, Kannel WB (1990).
Relation of pooled logistic regression to time dependent Cox regression
analysis: the Framingham Heart Study. *Statistics in Medicine* 9:1501–1515.
