---
title: "Transportability and generalizability"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Transportability and generalizability}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Causal effects estimated by `causat()` are, by default, effects in the
**study sample** --- the rows of `data`. In applied work the scientific
question is often about a different population: a clinic, country, or
cohort that differs in covariate distribution from the study. Moving the
estimand from study to target is the job of **transportability** (or
**generalizability**) weights.

This vignette shows how to transport causal estimates to a target
population using the `target =` argument.

## Setup

```{r}
#| message: false
library(causatr)
```

### Simulated data

We simulate a mixed dataset with study rows ($S = 1$, full data) and
target rows ($S = 0$, covariates only):

```{r}
set.seed(42)
n <- 3000
L <- rnorm(n)
S <- rbinom(n, 1, plogis(-0.5 + 1.0 * L))
A <- ifelse(S == 1L, rbinom(n, 1, plogis(0.2 + 0.3 * L)), NA_integer_)
Y <- ifelse(S == 1L, 2 + 3 * A + 1.5 * L + 1.0 * A * L + rnorm(n), NA_real_)

df <- data.frame(Y = Y, A = A, L = L, S = S)
head(df)
table(df$S)
```

The true ATE in the **study** population is 3 (constant treatment effect
plus an interaction $A \times L$ whose average depends on the
distribution of $L$ among the study rows). Because the sampling model
shifts the $L$-distribution, the target-population ATE differs from the
study ATE.

## The `target =` argument

Pass the name of the binary selection indicator ($S$) to `target =`:

```{r}
fit_gc <- causat(
  df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  target = "S"
)
fit_gc
```

This tells causatr to:

1. Fit the outcome model on study rows ($S = 1$) only.
2. Fit a **sampling model** $P(S = 1 \mid L)$ on all rows.
3. At contrast time, standardise predictions over target-population
   covariates (not study covariates).

### Contrasting

```{r}
res_gc <- contrast(
  fit_gc,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0"
)
res_gc
```

Compare with the study-only estimate (no `target =`):

```{r}
fit_study <- causat(
  df[df$S == 1, ],
  outcome = "Y",
  treatment = "A",
  confounders = ~ L
)

res_study <- contrast(
  fit_study,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0"
)
res_study
```

The two estimates differ because the target population has a different
distribution of $L$ than the study sample.

## Transportability vs generalizability

The `target_subset` argument controls which rows define the target:

- `target_subset = "target"` (default): the target population is $S = 0$
  only. This is **transportability** --- the study is external to the
  target.
- `target_subset = "all"`: the target population is $S = 0 \cup S = 1$
  (the full combined sample). This is **generalizability** --- the study
  is a biased subsample of the target.

```{r}
fit_gen <- causat(
  df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  target = "S",
  target_subset = "all"
)

res_gen <- contrast(
  fit_gen,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0"
)
res_gen
```

## Three estimators

All three point-treatment estimators support `target =`:

### IPW transport

Under IPW, the treatment density-ratio weights are multiplied by
sampling weights $w^S_i = [1 - \hat P(S = 1 \mid L_i)] / \hat P(S = 1 \mid L_i)$:

```{r}
fit_ipw <- causat(
  df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "ipw",
  target = "S"
)

res_ipw <- contrast(
  fit_ipw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0"
)
res_ipw
```

### AIPW transport (doubly robust)

AIPW combines the outcome model and treatment weights with sampling
weights, providing consistency if either the outcome model or the
treatment model is correctly specified:

```{r}
fit_aipw <- causat(
  df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "aipw",
  target = "S"
)

res_aipw <- contrast(
  fit_aipw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0"
)
res_aipw
```

### Comparison

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Forest plot comparing gcomp, IPW, and AIPW transport estimates."
methods_df <- data.frame(
  Method = c("G-comp", "IPW", "AIPW"),
  Estimate = c(
    res_gc$contrasts$estimate[1],
    res_ipw$contrasts$estimate[1],
    res_aipw$contrasts$estimate[1]
  ),
  CI_lower = c(
    res_gc$contrasts$ci_lower[1],
    res_ipw$contrasts$ci_lower[1],
    res_aipw$contrasts$ci_lower[1]
  ),
  CI_upper = c(
    res_gc$contrasts$ci_upper[1],
    res_ipw$contrasts$ci_upper[1],
    res_aipw$contrasts$ci_upper[1]
  )
)

tinyplot::tinyplot(
  Estimate ~ Method,
  data = methods_df,
  type = "pointrange",
  ymin = methods_df$CI_lower,
  ymax = methods_df$CI_upper,
  ylab = "Target-population ATE",
  main = "Transport estimates: three methods"
)
```

## Diagnostics

`diagnose()` includes a **sampling model panel** for transport fits,
reporting the sampling-score distribution, sampling-weight summary, and
extreme-weight flags:

```{r}
diag <- diagnose(fit_ipw)
diag
```

Key diagnostics to check:

- **Sampling positivity**: $P(S = 1 \mid L)$ should not be too close to
  0 or 1. The `p_study_min` and `p_study_max` quantiles flag overlap
  violations.
- **Extreme weights**: large sampling weights indicate regions of $L$
  where the study underrepresents the target, leading to instability.
- **ESS**: the effective sample size after sampling reweighting. Low ESS
  signals that a few observations drive the estimate.

## Identification assumptions

Transport estimates require four assumptions (Dahabreh et al. 2020):

1. **Conditional exchangeability in the study:**
   $Y^a \perp A \mid L, S = 1$.
2. **Treatment positivity in the study:**
   $0 < P(A = a \mid L, S = 1) < 1$.
3. **Exchangeability over populations (transportability):**
   $E[Y^a \mid L, S = 1] = E[Y^a \mid L, S = 0]$ --- the conditional
   potential-outcome mean is the same across populations given $L$.
4. **Sampling positivity:**
   $P(S = 1 \mid L) > 0$ on the target support.

Assumptions 1--2 are the standard causal assumptions (checked via
`diagnose()` balance and positivity panels). Assumption 3 is the
transportability leap --- it is not directly testable but requires
domain knowledge. Assumption 4 is checked by the sampling diagnostics.

## Matching is not supported

Matching + transportability is a hard rejection:

```{r}
#| error: true
causat(
  df,
  outcome = "Y",
  treatment = "A",
  confounders = ~ L,
  estimator = "matching",
  target = "S"
)
```

This is by design: MatchIt's match-weight pathway does not compose
cleanly with external sampling weights.

## References

- Dahabreh IJ, Robertson SE, Steingrimsson JA, Stuart EA, Hernán MA
  (2020). Extending inferences from a randomized trial to a new target
  population. *Stat Med* 39:1999--2014.
- Cole SR, Stuart EA (2010). Generalizing evidence from randomized
  clinical trials to target populations. *Am J Epidemiol* 172:107--115.
- Hernán MA, VanderWeele TJ (2017). Causal inference under multiple
  versions of treatment. *J Causal Inf* 5(2):20160014.
