---
title: "Validation against reference implementations"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Validation against reference implementations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

causatr is built around a single influence-function engine that reproduces the
sandwich-variance construction used by several purpose-built R packages. This
vignette runs side-by-side comparisons against those packages so you can see
the agreement (and disagreement, where it's expected) directly. Each section
fits the same causal quantity with causatr and with a reference implementation
and prints a two-row table of the point estimate, standard error, and 95% CI.

**What counts as "matching".** Some references implement the *same* estimator
and should agree to high precision:

- `stdReg2::standardize_glm()` uses the same g-formula influence function as
  causatr's point-gcomp engine, so the point estimate should match to ~1e-6 and
  the SE to within rounding of the covariance inverse.
- `WeightIt::glm_weightit()` uses the same M-estimator sandwich as causatr's
  IPW Branch A, so the SE should match within rounding.
- `marginaleffects::avg_comparisons()` on a `MatchIt::matchit()` output with
  `vcov = ~subclass` uses the same cluster-robust aggregation as causatr's
  matching branch, so point estimate and SE should match closely.

Other references implement a *different* estimator that targets the same
causal quantity. In those cases we expect agreement on the point estimate (up
to finite-sample noise) but not on the SE construction:

- `lmtp::lmtp_tmle()` is a TMLE, not a plug-in g-formula. It targets the same
  causal parameter but has a different finite-sample distribution; agreement
  is a sanity check, not a correctness proof.

The packages used here are all in `Suggests`; every chunk below is guarded
with `requireNamespace()` so the vignette still builds if any of them is
missing.

## Setup

```{r}
#| message: false
library(causatr)
library(tinytable)

compare_row <- function(label, result, name, ref_name,
                        ref_est, ref_se, level = 0.95) {
  con <- result$contrasts
  est <- result$estimates
  if (name %in% con$comparison) {
    row <- con[con$comparison == name, ]
  } else if (name %in% est$intervention) {
    row <- est[est$intervention == name, ]
  } else {
    stop(
      "name '", name,
      "' not found in result$contrasts$comparison or result$estimates$intervention"
    )
  }
  stopifnot(
    length(ref_est) == 1L && is.finite(ref_est),
    length(ref_se) == 1L && is.finite(ref_se)
  )
  z <- stats::qnorm(1 - (1 - level) / 2)

  # Three rows: causatr, reference, and a row of absolute differences.
  # The diff row is the whole point of the vignette — it makes
  # "do the two implementations agree?" visually obvious without making
  # the reader compare two numbers to 6 decimal places.
  est_c <- row$estimate
  se_c  <- row$se
  ci_lc <- row$ci_lower
  ci_uc <- row$ci_upper
  ci_lr <- ref_est - z * ref_se
  ci_ur <- ref_est + z * ref_se

  dt <- data.table::data.table(
    source   = c("causatr", ref_name, "abs diff"),
    estimate = c(est_c, ref_est, abs(est_c - ref_est)),
    se       = c(se_c,  ref_se,  abs(se_c  - ref_se)),
    ci_lower = c(ci_lc, ci_lr,   abs(ci_lc - ci_lr)),
    ci_upper = c(ci_uc, ci_ur,   abs(ci_uc - ci_ur))
  )

  tinytable::tt(dt, caption = label) |>
    tinytable::format_tt(
      j = 2:5, digits = 4, num_fmt = "decimal"
    ) |>
    tinytable::style_tt(i = 3, italic = TRUE, color = "#888")
}
```

`compare_row()` pulls an estimate from a `causatr_result` (either an
intervention mean or a contrast by name) and pairs it with a reference
estimate + SE under a normal-approximation CI. The output is a
three-row tinytable: causatr, the reference, and the absolute
discrepancy on each column, so agreement (or the lack of it) is
immediately legible.

We use two small simulated DGPs for reproducibility. NHEFS would work too, but
inline simulation keeps the vignette self-contained and cheap to render.

```{r}
simulate_point <- function(n = 2000, seed = 1L, binary_outcome = FALSE) {
  set.seed(seed)
  L <- stats::rnorm(n)
  A <- stats::rbinom(n, 1L, stats::plogis(0.4 * L))
  if (binary_outcome) {
    Y <- stats::rbinom(n, 1L, stats::plogis(-0.5 + 0.8 * A + 0.3 * L))
  } else {
    Y <- 1 + 2 * A + 0.5 * L + stats::rnorm(n, sd = 0.8)
  }
  data.frame(L = L, A = A, Y = Y)
}

simulate_longitudinal <- function(n = 500L, K = 3L, seed = 2L) {
  set.seed(seed)
  L <- matrix(0, n, K)
  A <- matrix(0L, n, K)
  for (k in seq_len(K)) {
    L[, k] <- stats::rnorm(n) +
      if (k == 1L) 0 else 0.3 * A[, k - 1L]
    A[, k] <- stats::rbinom(n, 1L, stats::plogis(0.3 * L[, k]))
  }
  Y <- stats::rbinom(n, 1L, stats::plogis(-1 + 0.4 * rowSums(A) + 0.2 * L[, K]))
  wide <- data.table::data.table(id = seq_len(n))
  for (k in seq_len(K)) {
    wide[[paste0("L", k)]] <- L[, k]
    wide[[paste0("A", k)]] <- A[, k]
  }
  wide$Y <- Y
  wide[]
}

simulate_longitudinal_continuous <- function(n = 400L, K = 2L, seed = 3L) {
  set.seed(seed)
  L <- matrix(0, n, K)
  A <- matrix(0, n, K)
  for (k in seq_len(K)) {
    L[, k] <- stats::rnorm(n) +
      if (k == 1L) 0 else 0.2 * A[, k - 1L]
    A[, k] <- stats::rnorm(n, mean = 0.3 * L[, k], sd = 1)
  }
  Y <- 1 + rowSums(A) + 0.3 * L[, K] + stats::rnorm(n, sd = 0.5)
  wide <- data.table::data.table(id = seq_len(n))
  for (k in seq_len(K)) {
    wide[[paste0("L", k)]] <- L[, k]
    wide[[paste0("A", k)]] <- A[, k]
  }
  wide$Y <- Y
  wide[]
}
```

## 1. Point g-computation vs `stdReg2`

`stdReg2::standardize_glm()` implements the parametric g-formula with the same
influence-function sandwich variance that causatr uses in its point-gcomp
branch. Agreement here should be exact up to rounding.

### 1a. Continuous outcome, risk difference

```{r}
df_pt <- simulate_point(n = 2000, binary_outcome = FALSE)

fit_gcomp <- causat(
  df_pt,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  estimator = "gcomp"
)
res_gcomp <- contrast(
  fit_gcomp,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_gcomp
```

```{r}
#| eval: !expr requireNamespace("stdReg2", quietly = TRUE)
std_fit <- stdReg2::standardize_glm(
  Y ~ A + L,
  data = df_pt,
  values = list(A = c(0, 1)),
  contrasts = "difference",
  reference = 0
)
std_res <- generics::tidy(std_fit)
std_diff <- std_res[std_res$contrast == "difference" & std_res$A == 1, ]

compare_row(
  "ATE (risk difference)", res_gcomp,
  name = "a1 vs a0",
  ref_name = "stdReg2",
  ref_est = std_diff$Estimate,
  ref_se = std_diff$Std.Error
)
```

### 1b. Binary outcome, risk difference

```{r}
df_pt_bin <- simulate_point(n = 2000, binary_outcome = TRUE)

fit_bin <- causat(
  df_pt_bin,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  family = binomial(),
  estimator = "gcomp"
)
res_bin_rd <- contrast(
  fit_bin,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_bin_rd
```

```{r}
#| eval: !expr requireNamespace("stdReg2", quietly = TRUE)
std_bin <- stdReg2::standardize_glm(
  Y ~ A + L,
  data = df_pt_bin,
  family = "binomial",
  values = list(A = c(0, 1)),
  contrasts = "difference",
  reference = 0
)
std_bin_res <- generics::tidy(std_bin)
std_bin_diff <- std_bin_res[std_bin_res$contrast == "difference" & std_bin_res$A == 1, ]

compare_row(
  "Risk difference", res_bin_rd,
  name = "a1 vs a0",
  ref_name = "stdReg2",
  ref_est = std_bin_diff$Estimate,
  ref_se = std_bin_diff$Std.Error
)
```

### 1c. Binary outcome, risk ratio

`stdReg2` also supports `contrasts = "ratio"` and returns the SE on the
ratio scale using the same influence-function approach causatr uses
internally. Note that causatr's CI is built on the *log* scale
(`exp(log(est) ± z * se/est)`) while `stdReg2`'s CI is linear on the ratio
scale, so the CI bounds will differ slightly even when the point estimate
and SE agree to rounding.

```{r}
res_bin_rr <- contrast(
  fit_bin,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "ratio",
  ci_method = "sandwich"
)
res_bin_rr
```

```{r}
#| eval: !expr requireNamespace("stdReg2", quietly = TRUE)
std_rr <- stdReg2::standardize_glm(
  Y ~ A + L,
  data = df_pt_bin,
  family = "binomial",
  values = list(A = c(0, 1)),
  contrasts = "ratio",
  reference = 0
)
std_rr_res <- generics::tidy(std_rr)
std_rr_row <- std_rr_res[std_rr_res$contrast == "ratio" & std_rr_res$A == 1, ]

compare_row(
  "Risk ratio", res_bin_rr,
  name = "a1 vs a0",
  ref_name = "stdReg2",
  ref_est = std_rr_row$Estimate,
  ref_se = std_rr_row$Std.Error
)
```

## 2. IPW vs `WeightIt::glm_weightit`

causatr's IPW Branch A wraps `WeightIt::glm_weightit` for the propensity
fit and applies the Mparts correction to the marginal structural model.
Comparing against `glm_weightit` directly verifies that the bookkeeping
matches: point estimate and SE should agree.

```{r}
fit_ipw <- causat(
  df_pt,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  estimator = "ipw"
)
res_ipw <- contrast(
  fit_ipw,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_ipw
```

```{r}
#| eval: !expr requireNamespace("WeightIt", quietly = TRUE)
w_ref <- WeightIt::weightit(
  A ~ L,
  data = df_pt,
  method = "glm",
  estimand = "ATE"
)
msm_ref <- WeightIt::glm_weightit(
  Y ~ A,
  data = df_pt,
  weightit = w_ref
)
coef_msm <- stats::coef(msm_ref)
vcov_msm <- stats::vcov(msm_ref)
ref_est <- unname(coef_msm["A"])
ref_se <- sqrt(vcov_msm["A", "A"])

compare_row(
  "ATE (IPW)", res_ipw,
  name = "a1 vs a0",
  ref_name = "WeightIt (glm_weightit)",
  ref_est = ref_est,
  ref_se = ref_se
)
```

With a saturated `Y ~ A` MSM, the coefficient on `A` *is* the marginal
contrast, so reading it off `glm_weightit` gives the reference directly. If
you wanted to use a richer MSM (e.g. to coarsen a multi-level treatment), you
would standardize predictions from `glm_weightit` via `marginaleffects`
instead — the same logic, just with an extra delta-method step.

## 3. Matching vs `marginaleffects` on `MatchIt`

For matching, the canonical post-match estimator is to refit the outcome on
the matched subset with cluster-robust SE on the subclass. causatr's matching
branch does exactly this internally; `marginaleffects::avg_comparisons()` with
`vcov = ~subclass` reproduces the same construction.

```{r}
fit_match <- causat(
  df_pt,
  outcome = "Y",
  treatment = "A",
  confounders = ~L,
  estimator = "matching",
  estimand = "ATT"
)
res_match <- contrast(
  fit_match,
  interventions = list(a1 = static(1), a0 = static(0)),
  reference = "a0",
  type = "difference",
  ci_method = "sandwich"
)
res_match
```

```{r}
#| eval: !expr (requireNamespace("MatchIt", quietly = TRUE) && requireNamespace("marginaleffects", quietly = TRUE))
m_ref <- MatchIt::matchit(
  A ~ L,
  data = df_pt,
  method = "nearest",
  estimand = "ATT"
)
md_ref <- MatchIt::match.data(m_ref)
lm_ref <- stats::lm(Y ~ A, data = md_ref, weights = weights)
me_match <- marginaleffects::avg_comparisons(
  lm_ref,
  variables = list(A = c(0, 1)),
  vcov = ~subclass,
  wts = "weights",
  newdata = subset(md_ref, A == 1)
)
compare_row(
  "ATT (matching)", res_match,
  name = "a1 vs a0",
  ref_name = "marginaleffects+MatchIt",
  ref_est = me_match$estimate,
  ref_se = me_match$std.error
)
```

The `newdata = subset(md_ref, A == 1)` argument tells `marginaleffects` to
average over the treated-subject target population, which is what
`estimand = "ATT"` means.

## 4. Longitudinal ICE vs `lmtp::lmtp_tmle`

`lmtp::lmtp_tmle()` is a TMLE that targets the same parameter as ICE
g-computation but with a different finite-sample bias-variance profile.
We expect agreement on the point estimate up to sampling noise (both methods
are consistent for the same target), but not exact agreement on the SE.

### 4a. Binary treatment, binary outcome, "always" vs "never"

```{r}
wide_bin <- simulate_longitudinal(n = 500L, K = 3L)
long_bin <- to_person_period(
  wide_bin,
  id = "id",
  time_varying = list(L = c("L1", "L2", "L3"), A = c("A1", "A2", "A3")),
  time_invariant = "Y"
)

fit_ice <- causat(
  long_bin,
  outcome = "Y",
  treatment = "A",
  confounders = ~1,
  confounders_tv = ~L,
  family = binomial(),
  id = "id",
  time = "time",
  estimator = "gcomp",
  type = "longitudinal",
  history = 1L
)
res_ice <- contrast(
  fit_ice,
  interventions = list(always = static(1), never = static(0)),
  reference = "never",
  type = "difference",
  ci_method = "sandwich"
)
res_ice
```

```{r}
#| eval: !expr requireNamespace("lmtp", quietly = TRUE)
set.seed(42)
lmtp_always <- tryCatch(
  lmtp::lmtp_tmle(
    data = as.data.frame(wide_bin),
    trt = c("A1", "A2", "A3"),
    outcome = "Y",
    time_vary = list("L1", "L2", "L3"),
    shift = function(data, trt) rep(1, nrow(data)),
    outcome_type = "binomial",
    learners_outcome = "SL.glm",
    learners_trt = "SL.glm",
    folds = 2
  ),
  error = function(e) NULL
)
lmtp_never <- tryCatch(
  lmtp::lmtp_tmle(
    data = as.data.frame(wide_bin),
    trt = c("A1", "A2", "A3"),
    outcome = "Y",
    time_vary = list("L1", "L2", "L3"),
    shift = function(data, trt) rep(0, nrow(data)),
    outcome_type = "binomial",
    learners_outcome = "SL.glm",
    learners_trt = "SL.glm",
    folds = 2
  ),
  error = function(e) NULL
)

if (!is.null(lmtp_always) && !is.null(lmtp_never)) {
  lmtp_diff <- lmtp::lmtp_contrast(lmtp_always, ref = lmtp_never, type = "additive")
  ref_row <- lmtp_diff$estimates
  compare_row(
    "Always vs never (RD)", res_ice,
    name = "always vs never",
    ref_name = "lmtp_tmle",
    ref_est = ref_row$estimate,
    ref_se = ref_row$std.error
  )
} else {
  message("lmtp fit failed; skipping comparison row.")
}
```

### 4b. Continuous treatment, shift intervention

```{r}
wide_cont <- simulate_longitudinal_continuous(n = 400L, K = 2L)
long_cont <- to_person_period(
  wide_cont,
  id = "id",
  time_varying = list(L = c("L1", "L2"), A = c("A1", "A2")),
  time_invariant = "Y"
)

fit_ice_shift <- causat(
  long_cont,
  outcome = "Y",
  treatment = "A",
  confounders = ~1,
  confounders_tv = ~L,
  family = gaussian(),
  id = "id",
  time = "time",
  estimator = "gcomp",
  type = "longitudinal",
  history = 1L
)
res_ice_shift <- contrast(
  fit_ice_shift,
  interventions = list(plus1 = shift(1), zero = shift(0)),
  reference = "zero",
  type = "difference",
  ci_method = "sandwich"
)
res_ice_shift
```

```{r}
#| eval: !expr requireNamespace("lmtp", quietly = TRUE)
set.seed(43)
shift_up <- function(data, trt) data[[trt]] + 1
shift_id <- function(data, trt) data[[trt]]

lmtp_up <- tryCatch(
  lmtp::lmtp_tmle(
    data = as.data.frame(wide_cont),
    trt = c("A1", "A2"),
    outcome = "Y",
    time_vary = list("L1", "L2"),
    shift = shift_up,
    outcome_type = "continuous",
    learners_outcome = "SL.glm",
    learners_trt = "SL.glm",
    folds = 2
  ),
  error = function(e) NULL
)
lmtp_id <- tryCatch(
  lmtp::lmtp_tmle(
    data = as.data.frame(wide_cont),
    trt = c("A1", "A2"),
    outcome = "Y",
    time_vary = list("L1", "L2"),
    shift = shift_id,
    outcome_type = "continuous",
    learners_outcome = "SL.glm",
    learners_trt = "SL.glm",
    folds = 2
  ),
  error = function(e) NULL
)

if (!is.null(lmtp_up) && !is.null(lmtp_id)) {
  lmtp_diff <- lmtp::lmtp_contrast(lmtp_up, ref = lmtp_id, type = "additive")
  ref_row <- lmtp_diff$estimates
  compare_row(
    "Shift(+1) vs identity", res_ice_shift,
    name = "plus1 vs zero",
    ref_name = "lmtp_tmle",
    ref_est = ref_row$estimate,
    ref_se = ref_row$std.error
  )
} else {
  message("lmtp fit failed; skipping comparison row.")
}
```

The structural effect of `shift(+1)` on this DGP is exactly 2 (one unit per
period, two periods). Both estimators should recover this up to sampling
noise; the two SEs correspond to different estimators (ICE plug-in vs TMLE)
and are expected to differ in the second decimal place.

## 5. Other references

Several additional R packages would extend this validation suite:

- **`gfoRmula`** implements the parametric g-formula on wide longitudinal
  data and is the most direct peer of causatr's ICE engine. Its intervention
  specification is more verbose (lists of `intvars`, `interventions`,
  `int_descript`), but a comparison on static "always vs never" would make
  a natural addition to section 4a.
- **`survey::svyglm()`** (combined with a user-supplied weights column) is
  the standard reference for survey-weighted IPW. Since causatr validates
  survey-weight handling in its tests, a vignette comparison would mostly
  duplicate that coverage.
- **Survival validation** will land once `causat_survival()`'s contrast path
  is implemented in Phase 6; the natural references are
  `survival::survfit()` for the unadjusted Kaplan-Meier sanity check and
  `gfoRmula::gformula_survival()` for the adjusted g-formula reference.

If you'd like to contribute a reference comparison for one of these (or for a
package not listed here), open an issue — the pattern is the same as in
sections 1–4: fit the same quantity two ways, pass both to `compare_row()`,
and print the resulting two-row table.
