---
title: "Variance estimation for causal effects"
toc-depth: 4
vignette: >
  %\VignetteIndexEntry{Variance estimation for causal effects}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
mjx-container[jax="CHTML"][display="true"] {
  overflow-x: auto;
  overflow-y: visible;
  padding: 0.5em 0;
}
</style>
```

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette develops the theory behind variance estimation for causal effect
estimators. The core message is simple: **there is one method** — the influence
function (IF) — applied uniformly to g-computation, IPW, and matching. What
varies across methods is only the estimating equations that define the
estimator; the variance formula is always the same.

We build up from first principles: M-estimation gives us
$\text{Var}(\hat\beta)$ (Section 1), the IF gives us a per-individual
decomposition (Section 2), and the two-channel IF gives us
$\text{Var}(\hat\mu)$ for any causal estimand (Section 3). Section 4 applies
this to the three estimation methods. Section 5 extends to chained estimators
(ICE g-computation for longitudinal data).

```{r}
#| label: setup-data
#| code-fold: true
#| code-summary: "Running example: simulated data"
set.seed(42)
n <- 200
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- rbinom(n, 1, plogis(-1 + 0.8 * A + 0.6 * L))
dat <- data.frame(Y = Y, A = A, L = L)

fit <- glm(Y ~ A + L, data = dat, family = binomial)
beta_hat <- coef(fit)
```

```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "The variance estimation pipeline for causal effects. The influence function (IF) is the one method: write down the estimating equations for the full system (nuisance models + target estimand), compute each individual's IF, then square and sum. The delta method (J V_β J^T) is a shortcut that captures only parameter uncertainty — it's exact when predictions are constant per intervention, but underestimates otherwise."
flowchart TB
  EE["Complete estimating equations<br/>(nuisance models + estimand)"] -->|"linearise<br/>(Sec 1–2)"| IF["IF_i per individual<br/>= direct + model correction"]
  IF -->|"Var = (1/n²) Σ IF²<br/>(always correct)"| VM["Var(μ̂)"]

  EE -->|"shortcut:<br/>J V_β Jᵀ"| SC{"Predictions<br/>constant?"}
  SC -->|"Yes<br/>(saturated MSM,<br/>static intervention)"| VM
  SC -->|"No<br/>(g-comp,<br/>non-saturated MSM)"| UNDER["Underestimates ❌"]

  style EE fill:#f0f4ff,stroke:#3182BD
  style IF fill:#fff5eb,stroke:#E6550D
  style VM fill:#e8f5e9,stroke:#31A354
  style SC fill:#fff5eb,stroke:#E6550D
  style UNDER fill:#fee,stroke:#c00
```


## 1. The sandwich estimator for parametric models

This section derives $\text{Var}(\hat\beta)$ — the variance of model
parameters. This is a building block, not the final answer. Section 3 shows how
to get from $\text{Var}(\hat\beta)$ to $\text{Var}(\hat\mu)$.

### 1.1 What is an estimating equation?

The starting point is a simple idea: many statistical estimators can be defined
as the solution to an equation of the form

$$
\sum_{i=1}^n \psi(O_i; \beta) = 0
$$

where $O_i$ is the observed data for individual $i$, $\beta$ is the parameter
we want to estimate, and $\psi$ is a function — called the **estimating
function** — that maps each observation and a candidate parameter value to a
vector. The estimator $\hat\beta$ is the value of $\beta$ that makes this sum
equal to zero.

But what *is* $\psi$, concretely? And where does it come from?

### 1.2 Deriving $\psi$: three routes to the same place

#### Route 1: Score equations (maximum likelihood)

The most familiar source of $\psi$ is maximum likelihood. If observations have
log-likelihood $\ell_i(\beta) = \log f(O_i; \beta)$, the MLE maximises
$\sum_i \ell_i(\beta)$. Setting the derivative to zero gives:

$$
\sum_{i=1}^n \underbrace{\frac{\partial \ell_i(\beta)}{\partial \beta}}_
{\psi(O_i;\beta)} = 0
$$

So $\psi$ is the **score function** — the gradient of the individual
log-likelihood. For a GLM with outcome $Y_i$, covariates $X_i$, linear
predictor $\eta_i = X_i^T\beta$, and mean function $\mu_i = g^{-1}(\eta_i)$:

$$
\psi(O_i; \beta) =
\underset{\color{#E6550D}{\text{residual}}}{\color{#E6550D}{(Y_i - \mu_i)}}
\;\cdot\;
\underset{\color{#3182BD}{\text{scale factor}}}
{\color{#3182BD}{\frac{d\mu/d\eta}{V(\mu_i)}}}
\;\cdot\;
\underset{\color{#31A354}{\text{direction}}}{\color{#31A354}{X_i}}
$$

where $V(\mu_i)$ is the variance function of the exponential family
($V(\mu) = \mu(1-\mu)$ for binomial, $V(\mu) = \mu$ for Poisson, $V(\mu) =
1$ for Gaussian — note that $\sigma^2$ enters separately as the dispersion
parameter $\phi$, not as $V(\mu)$). Reading the three factors:

- $\color{#E6550D}{(Y_i - \mu_i)}$ — **Residual**: how far the prediction is
  from the observation. This is zero on average at the truth, which is why
  $E[\psi_i] = 0$.
- $\color{#3182BD}{(d\mu/d\eta) / V(\mu_i)}$ — **Scale factor**: combines the
  link derivative (converting from $\eta$-scale to $\mu$-scale) with the
  variance weight (downweighting noisy observations). For canonical links this
  product simplifies (see below).
- $\color{#31A354}{X_i}$ — **Direction**: the covariates tell us *which*
  parameters this observation informs. An observation with large $|X_{ij}|$
  provides more information about $\beta_j$.

::: {.callout-note}
### Example: logistic regression
For logistic regression, $\mu_i = \text{expit}(\eta_i)$, so $d\mu/d\eta =
\mu_i(1-\mu_i)$ and $V(\mu_i) = \mu_i(1 - \mu_i)$. The scale factor cancels
to $1$, giving:

$$\psi(O_i;\beta) = (Y_i - \mu_i) \cdot X_i$$

The estimating equation $\sum_i (Y_i - \mu_i) X_i = 0$ says: at the MLE, the
covariates are uncorrelated with the residuals.
:::

#### Route 2: Moment conditions (method of moments)

Not every estimating equation comes from a likelihood. The **generalised method
of moments** (GMM) starts from a theoretical property of the true parameter
$\beta_0$ — a *moment condition* — and turns it into $\psi$.

For example, if we know that $E[X_i(Y_i - X_i^T\beta_0)] = 0$, we can define:

$$
\psi(O_i; \beta) = X_i(Y_i - X_i^T\beta)
$$

Setting $\sum_i \psi = 0$ gives the OLS normal equations. No distributional
assumption was needed — just the moment condition $E[X_i \varepsilon_i] = 0$.

This is the key insight of M-estimation: the estimating function $\psi$ does not
have to be a score function. It just needs to satisfy $E[\psi(O_i; \beta_0)] =
0$ at the true parameter value (the **unbiasedness** condition). This is what
guarantees consistency of $\hat\beta$.

#### Route 3: Quasi-likelihood

Quasi-likelihood provides a middle ground. We specify only the first two moments
of $Y_i \mid X_i$ — the mean function $\mu_i = g^{-1}(X_i^T\beta)$ and a
variance function $V(\mu_i)$ — without assuming a full distributional form. The
quasi-score is:

$$
\psi(O_i; \beta) = \frac{(Y_i - \mu_i)}{V(\mu_i)} \cdot
\frac{d\mu_i}{d\eta_i} \cdot X_i
$$

This looks identical to the GLM score, but now $V(\mu)$ is a *working*
variance — it might be wrong. The sandwich estimator (Section 1.4) protects
against this misspecification.

### 1.3 The general M-estimation framework

All three routes produce estimators of the same form: find $\hat\beta$ solving
$\sum_i \psi(O_i; \hat\beta) = 0$. The M-estimation framework (Huber, 1964;
Stefanski & Boos, 2002) unifies them under two conditions:

1. **Unbiasedness**: $E[\psi(O_i; \beta_0)] = 0$ at the true $\beta_0$. This
   guarantees that the estimating equation has the right solution in
   expectation.
2. **Regularity**: $\psi$ is smooth enough in $\beta$ for a Taylor expansion.

Given these, $\hat\beta$ is consistent and asymptotically normal, and its
variance has the sandwich form — regardless of whether $\psi$ came from a
likelihood, a moment condition, or a quasi-likelihood.

::: {.callout-tip}
### Why "M-estimation"?
The "M" stands for "maximum" (as in maximum-likelihood), but the framework is
much broader. Any estimator defined as the root of a sum of functions of the
data and parameters is an M-estimator — this includes MLE, weighted least
squares, quantile regression, robust regression (Huber loss), generalised
estimating equations, and the causal estimators in Sections 4–5 (Stefanski &
Boos, 2002).
:::

### 1.4 Asymptotic variance of M-estimators

We want $\text{Var}(\hat\beta)$. The derivation has three steps: linearise
$\hat\beta$ around the truth, identify the variance of the linear
approximation, then estimate it from data.

#### Step 1: Linearise via Taylor expansion

We know that $\hat\beta$ solves $\sum_i \psi(O_i; \hat\beta) = 0$. We also
know that the true $\beta_0$ satisfies $E[\psi(O_i; \beta_0)] = 0$, but the
finite-sample sum $\sum_i \psi(O_i; \beta_0)$ is generally *not* zero — it is
a random quantity that fluctuates around zero. The idea is to relate these two
facts via a Taylor expansion.

Expand $\sum_i \psi(O_i; \hat\beta)$ around $\beta_0$, keeping only the
first-order term (valid for large $n$ by regularity):

$$
\underbrace{\sum_i \psi(O_i; \hat\beta)}_{= \, 0 \text{ (by definition)}}
\approx \sum_i \psi(O_i; \beta_0) +
\left[\sum_i \frac{\partial \psi(O_i;\beta_0)}
{\partial \beta^T}\right]
(\hat\beta - \beta_0)
$$

Rearranging for the estimation error $\hat\beta - \beta_0$:

$$
\hat\beta - \beta_0 \approx
-\left[\sum_i \frac{\partial \psi(O_i;
\beta_0)}{\partial \beta^T}\right]^{-1}
\sum_i \psi(O_i; \beta_0)
$$

Dividing numerator and denominator by $n$ and labelling the two factors:

$$
\underset{p \times 1}{\hat\beta - \beta_0}
\;\approx\;
\underbrace{\left[-\frac{1}{n}\sum_i \frac{\partial \psi_i}
{\partial \beta^T}\right]^{-1}}_{\to\;
\color{#3182BD}{A^{-1}} \;(p \times p)}
\;\cdot\;
\underbrace{\frac{1}{n}\sum_i \psi(O_i; \beta_0)}_{\text{mean zero,
covariance } \color{#E6550D}{B/n}}
$$

This is the key linearisation: the estimation error is approximately the bread
inverse times the average score. The first factor converges to the constant
matrix $\color{#3182BD}{A^{-1}}$; the second is a sum of iid mean-zero random
$p$-vectors whose covariance we call $\color{#E6550D}{B}$.

#### Step 2: Identify the variance

Name the two matrices that appeared in Step 1:

- **Bread** $\color{#3182BD}{A}$: the expected negative Jacobian of the
  estimating function ($p \times p$):

$$
\color{#3182BD}{A}
\;\equiv\; -E\!\left[\frac{\partial \psi(O_i; \beta_0)}
{\partial \beta^T}\right]
\qquad (p \times p)
$$

For a GLM, $\color{#3182BD}{A} = E[w_i X_i X_i^T]$ where $w_i =
(d\mu/d\eta)^2 / V(\mu)$ are the IWLS working weights — essentially the
curvature of the (quasi-)log-likelihood.

- **Meat** $\color{#E6550D}{B}$: the covariance of the estimating function
  ($p \times p$):

$$
\color{#E6550D}{B}
\;\equiv\; E\!\left[\psi(O_i; \beta_0)\,\psi(O_i; \beta_0)^T\right]
\qquad (p \times p)
$$

This measures how much the individual scores $\psi_i$ vary across observations.
It appeared in Step 1 as the covariance of the second factor: $\text{Var}\!
\left(\frac{1}{n}\sum_i \psi_i\right) = \frac{1}{n}\color{#E6550D}{B}$.

Now propagate through the linearisation. Taking $\text{Var}(\cdot)$ of both
sides of the Step 1 equation:

$$
\underset{p \times p}{\text{Var}(\hat\beta)}
\;\approx\;
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\cdot\;
\underset{p \times p}{\frac{\color{#E6550D}{B}}{n}}
\;\cdot\;
\underset{p \times p}{\color{#3182BD}{(A^{-1})^T}}
\;=\;
\frac{1}{n}\,
\color{#3182BD}{A^{-1}}\,
\color{#E6550D}{B}\,
\color{#3182BD}{(A^{-1})^T}
$$

#### Step 3: Estimate from data

Replace the population expectations $A$ and $B$ with sample averages:

$$
\color{#3182BD}{\hat{A}}
= -\frac{1}{n}\sum_i
\frac{\partial \psi(O_i; \hat\beta)}{\partial \beta^T}
\qquad (p \times p)
$$

$$
\color{#E6550D}{\hat{B}}
= \frac{1}{n}\sum_i
\psi(O_i; \hat\beta)\,\psi(O_i; \hat\beta)^T
\qquad (p \times p)
$$

The **sandwich estimator** is:

$$
\underset{p \times p}{\widehat{\text{Var}}(\hat\beta)}
= \frac{1}{n}\;
\underset{p \times p}{\color{#3182BD}{\hat{A}^{-1}}}
\;
\underset{p \times p}{\color{#E6550D}{\hat{B}}}
\;
\underset{p \times p}{\color{#3182BD}{(\hat{A}^{-1})^T}}
$$ {#eq-sandwich}

so called because the meat $\color{#E6550D}{\hat B}$ is "sandwiched" between
two slices of bread $\color{#3182BD}{\hat A^{-1}}$.

::: {.callout-important}
### Why is this robust?
The bread $\color{#3182BD}{A}$ and meat $\color{#E6550D}{B}$ are estimated
*separately*. Under correct model specification (e.g. the GLM variance function
is right), $A = B$ and the sandwich simplifies to $\frac{1}{n} A^{-1}$ — the
usual model-based variance. But if the variance function is wrong, $A \neq B$:
the bread still captures the curvature correctly (it only depends on the mean
model), while the meat captures the *actual* variability of the scores. The
sandwich remains consistent regardless — this is the Huber–White robustness
property (Huber, 1967; White, 1982).
:::

### 1.5 Pseudo-code: sandwich variance for a GLM

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Pseudo-code: sandwich_variance()"
sandwich_variance <- function(X, Y, beta_hat, family) {
  n <- length(Y)
  eta <- X %*% beta_hat
  mu <- family$linkinv(eta)
  mu_eta <- family$mu.eta(eta)    # dmu/deta
  V <- family$variance(mu)        # Var(Y|X) under the model

  # Bread: A = (1/n) X^T W X where W = diag((dmu/deta)^2 / V(mu))
  W <- mu_eta^2 / V
  A <- (1 / n) * crossprod(X, X * W)

  # Meat: B = (1/n) sum psi_i psi_i^T
  residuals <- Y - mu
  score_weights <- mu_eta / V
  psi <- X * (residuals * score_weights)  # n x p matrix of score contributions
  B <- (1 / n) * crossprod(psi)

  # Sandwich: (1/n) A^{-1} B (A^{-1})^T
  A_inv <- solve(A)
  V_beta <- (1 / n) * A_inv %*% B %*% t(A_inv)
  V_beta
}
```

::: {.callout-tip}
### In practice
You almost never code this yourself. In R, `sandwich::sandwich(model)` computes
$\hat A^{-1} \hat B (\hat A^{-1})^T$ from any `glm` object. The `sandwich`
package (Zeileis, 2004, 2006) provides variants for clustered data
(`vcovCL()`), panel data (`vcovPL()`), and heteroscedasticity-autocorrelation
(`vcovHAC()`).
:::

### 1.6 Concrete example: linear regression

Let us work through the sandwich for OLS step by step.

**Model**: $Y_i = X_i^T\beta + \varepsilon_i$ where $E[\varepsilon_i \mid X_i]
= 0$.

**Step 1: Write down $\psi$.**
The OLS estimator minimises $\sum_i (Y_i - X_i^T\beta)^2$. Setting the
gradient to zero gives $\sum_i X_i(Y_i - X_i^T\beta) = 0$, so:

$$
\psi(O_i; \beta) = X_i(Y_i - X_i^T\beta)
\qquad (p \times 1)
$$

At $\beta_0$: $E[\psi_i] = E[X_i \varepsilon_i] = 0$ by exogeneity.

**Step 2: Compute the bread $\color{#3182BD}{A}$.**

$$
\frac{\partial \psi(O_i; \beta)}{\partial \beta^T}
= \frac{\partial}{\partial \beta^T}\left[X_i(Y_i - X_i^T\beta)\right]
= -X_i X_i^T
\qquad (p \times p)
$$

So $\color{#3182BD}{A} = E[X_i X_i^T]$, estimated by
$\color{#3182BD}{\hat{A}} = \frac{1}{n} X^T X$.

**Step 3: Compute the meat $\color{#E6550D}{B}$.**
Since $\psi_i = X_i \hat e_i$, the outer product is $\psi_i\psi_i^T =
\hat e_i^2 X_iX_i^T$:

$$
\color{#E6550D}{\hat{B}} = \frac{1}{n}\sum_i \hat{e}_i^2 \, X_i X_i^T
\qquad (p \times p)
$$

The meat depends on the *squared residuals*, not on any assumed variance. This
is the source of robustness.

**Step 4: Assemble the sandwich.**

$$
\underset{p \times p}{\widehat{\text{Var}}(\hat\beta)}
= \frac{1}{n}\,
\color{#3182BD}{\hat{A}^{-1}}\;
\color{#E6550D}{\hat{B}}\;
\color{#3182BD}{(\hat{A}^{-1})^T}
$$

Substituting:

$$
= \underset{p \times p}{\color{#3182BD}{(X^TX)^{-1}}}
\;\;
\underset{p \times p}{\color{#E6550D}{\left[\sum_i
\hat{e}_i^2\,X_iX_i^T\right]}}
\;\;
\underset{p \times p}{\color{#3182BD}{(X^TX)^{-1}}}
$$

This is the *HC0* heteroscedasticity-consistent estimator (White, 1980).

**What if homoscedasticity holds?** If $\text{Var}(\varepsilon_i \mid X_i) =
\sigma^2$ for all $i$, then $E[\hat{e}_i^2 X_i X_i^T] = \sigma^2 E[X_i
X_i^T]$, so $\color{#E6550D}{B} = \sigma^2 \color{#3182BD}{A}$ and the
sandwich simplifies to $\frac{\sigma^2}{n} \color{#3182BD}{A^{-1}}
= \hat\sigma^2(X^TX)^{-1}$, the usual OLS variance.


## 2. The influence function

Section 1 gave us $\text{Var}(\hat\beta)$ as a matrix product (the sandwich).
The **influence function** (IF) gives the same answer but expressed as a
per-individual contribution. This is more than a curiosity — IFs are the tool
that lets us compute $\text{Var}(\hat\mu)$ for causal estimands (Section 3) and
for multi-model pipelines (Section 5).

### 2.1 The idea: how much does each observation matter?

Imagine removing observation $i$ from the dataset and re-estimating
$\hat\beta$. The change $\hat\beta_{(-i)} - \hat\beta$ tells you how much
observation $i$ "influenced" the estimate. The **influence function** (IF) is
the first-order approximation to this quantity, scaled by $n$:

$$
\text{IF}_i \approx -n(\hat\beta_{(-i)} - \hat\beta)
$$

But we do not need to actually refit the model $n$ times. For M-estimators,
the IF can be computed in closed form.

### 2.2 Deriving the IF from the linearisation

In Section 1.4, Step 1, we derived the linearisation:

$$
\underset{p \times 1}{\hat\beta - \beta_0}
\;\approx\;
\frac{1}{n}\sum_{i=1}^n
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\;
\underset{p \times 1}{\psi(O_i; \beta_0)}
$$

Each summand $\color{#3182BD}{A^{-1}}\psi_i$ is the contribution of
observation $i$ to the total estimation error. This is exactly the **influence
function**:

$$
\underset{p \times 1}{\text{IF}_i(\hat\beta)}
\;=\;
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\cdot\;
\underset{p \times 1}{\psi(O_i; \hat\beta)}
$$ {#eq-if}

Reading each factor:

- $\psi_i$ ($p \times 1$): the score from Section 1.2 — how much observation
  $i$'s data "disagrees" with the current $\hat\beta$.
- $\color{#3182BD}{A^{-1}}$ ($p \times p$): the bread inverse from Section
  1.4 — converts from score-space to parameter-space, accounting for curvature.
- $\text{IF}_i$ ($p \times 1$): the resulting shift in $\hat\beta$ attributable
  to observation $i$ (times $n$).

No new derivation is needed. The IF *is* the linearisation, read term-by-term.

### 2.3 IF = sandwich (the key equivalence)

**Property 1: Linearisation.** The IF lets us write a complex estimator as a
simple average:

$$
\hat\beta - \beta_0 \approx \frac{1}{n}\sum_i \text{IF}_i
$$

**Property 2: Variance recovery.** Since $E[\text{IF}_i] = 0$:

$$
\text{Var}(\hat\beta) \approx
\frac{1}{n^2}\sum_i \text{IF}_i \, \text{IF}_i^T
$$

Expanding $\text{IF}_i \text{IF}_i^T = \color{#3182BD}{A^{-1}} \psi_i
\psi_i^T \color{#3182BD}{(A^{-1})^T}$ and factoring:

$$
\frac{1}{n^2}\sum_i \color{#3182BD}{A^{-1}} \psi_i\psi_i^T
\color{#3182BD}{(A^{-1})^T}
= \frac{1}{n} \color{#3182BD}{A^{-1}}
\underbrace{\frac{1}{n}\sum_i \psi_i\psi_i^T}_
{\color{#E6550D}{B}\text{ (the meat)}}
\color{#3182BD}{(A^{-1})^T}
$$

This is the sandwich from Section 1.4.

::: {.callout-note}
### The key equivalence
$$
\underbrace{\frac{1}{n^2}\sum_i \text{IF}_i\,\text{IF}_i^T}_
{\text{IF-based variance}}
\;=\;
\underbrace{\frac{1}{n}\,\color{#3182BD}{A^{-1}}\,\color{#E6550D}{B}\,
\color{#3182BD}{(A^{-1})^T}}_{\text{sandwich variance}}
$$ {#eq-if-sandwich}

The sandwich operates on matrices. The IF operates on per-individual vectors.
Same answer, different representation. We will use the IF representation from
here on, because it extends cleanly to multi-step estimators (Sections 3–5).
:::

### 2.4 Example: IF for a sample mean

$\hat\mu = \frac{1}{n}\sum_i Y_i$. Estimating equation: $\psi(O_i; \mu)
= Y_i - \mu$. Bread: $A = 1$. IF: $\text{IF}_i = Y_i - \hat\mu$. Variance:
$\frac{1}{n^2}\sum_i(Y_i - \hat\mu)^2 = s^2/n$.


## 3. Variance of a causal estimand

### 3.1 The problem

The sandwich (Section 1) gives $\text{Var}(\hat\beta)$. But the causal
estimand is typically a nonlinear function of $\hat\beta$ **and** the data.
For instance, the g-computation marginal mean under intervention $a$ is:

$$
\hat\mu_a = \frac{1}{n}\sum_{i=1}^n
g^{-1}(X_i^{*T} \hat\beta)
$$

where $X_i^*$ is individual $i$'s covariate vector with treatment set to $a$,
and $g^{-1}(\cdot)$ is the inverse link function.

This quantity depends on **two** random things:

1. $\hat\beta$ — the estimated model parameters (random because of finite
   data).
2. $\{X_i\}$ — the covariate distribution (random because a different sample
   gives different covariates, hence a different average of predictions).

A complete variance estimator must account for both sources. As we will see,
the standard "delta method" shortcut only captures source (1).

### 3.2 The two-channel influence function

The g-computation marginal mean is a **two-step** estimator: first estimate
$\hat\beta$, then compute $\hat\mu_a = \frac{1}{n}\sum_i g^{-1}(X_i^{*T}
\hat\beta)$. Individual $i$ affects $\hat\mu_a$ through two channels.

#### Channel 1: Direct contribution (covariate sampling)

Even if $\hat\beta$ were fixed, observation $i$'s prediction
$g^{-1}(X_i^{*T}\hat\beta)$ is one of the $n$ terms being averaged. A
different sample gives a different set of predictions. Individual $i$'s direct
contribution to the estimation error is:

$$
\text{IF}_i^{\text{direct}} = g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a
\qquad (\text{scalar})
$$

This is the same structure as the sample mean IF from Section 2.4 — the
deviation of one term from the average.

#### Channel 2: Model correction (parameter uncertainty)

Observation $i$ also affects $\hat\mu_a$ indirectly by shifting $\hat\beta$.
From Section 2.2, $i$'s influence on $\hat\beta$ is approximately
$\frac{1}{n}\color{#3182BD}{A^{-1}}\psi_i$. This shift changes $\hat\mu_a$ by:

$$
\delta\hat\mu_a \;\approx\;
\underset{1 \times p}{\color{#31A354}{J_a}}
\;\cdot\;
\underset{p \times 1}{\color{#3182BD}{A^{-1}}\psi_i}
$$

where $\color{#31A354}{J_a} = \frac{\partial \hat\mu_a}{\partial \beta^T}$ is
the $1 \times p$ **Jacobian** — how the marginal mean changes as each
coefficient shifts:

$$
\underset{1 \times p}{\color{#31A354}{J_a}}
= \frac{1}{n} \sum_{i=1}^n
\underbrace{\frac{d\mu}{d\eta}\bigg|_{\eta_i^*}}_{\text{scalar}}
\;\cdot\;
\underbrace{X_i^{*T}}_{1 \times p}
$$

The link derivatives for common GLMs:

| Family | Link | $d\mu/d\eta$ | Effect |
|---|---|---|---|
| Gaussian | Identity | $1$ | All observations contribute equally |
| Binomial | Logit | $\mu(1 - \mu)$ | Observations near $\mu = 0.5$ contribute most |
| Poisson | Log | $\mu$ | Observations with larger means contribute more |

#### Combining both channels

The full IF for the marginal mean $\hat\mu_a$ is:

$$
\text{IF}_i(\hat\mu_a) =
\underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_
{\text{Channel 1: covariate sampling}}
\;+\;
\underbrace{
\underset{1 \times p}{\color{#31A354}{J_a}}
\;\;
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\;
\underset{p \times 1}{\psi_i}
}_{\text{Channel 2: parameter uncertainty}}
$$ {#eq-if-gcomp}

And the variance is:

$$
\widehat{\text{Var}}(\hat\mu_a) = \frac{1}{n^2}\sum_{i=1}^n
\text{IF}_i(\hat\mu_a)^2
$$ {#eq-var-if}

For two interventions $a$ and $b$, the covariance is:

$$
\widehat{\text{Cov}}(\hat\mu_a, \hat\mu_b)
= \frac{1}{n^2}\sum_{i=1}^n \text{IF}_i^{(a)} \cdot \text{IF}_i^{(b)}
$$

This gives the full $k \times k$ covariance matrix needed for contrasts (risk
difference, risk ratio, etc.).

::: {.callout-note}
### Why the covariance matters
Both $\hat\mu_1$ and $\hat\mu_0$ are functions of the *same* data, so they are
correlated. This covariance typically *reduces* the variance of the difference.
Ignoring it (computing variances separately and adding them) would overestimate
the SE.
:::

### 3.3 The delta method shortcut

A widely used shortcut computes the Jacobian $\color{#31A354}{J}$ and the
parameter sandwich $V_\beta$ separately, then propagates via:

$$
\underset{k \times k}{\text{Var}(\hat\mu)}
\;\approx\;
\underset{k \times p}{\color{#31A354}{J}}
\;\;
\underset{p \times p}{V_\beta}
\;\;
\underset{p \times k}{\color{#31A354}{J^T}}
$$ {#eq-delta}

This is the **multivariate delta method**. To see what it captures, expand the
IF-based variance (@eq-var-if) using the two-channel decomposition
(@eq-if-gcomp):

$$
\frac{1}{n^2}\sum_i \text{IF}_i^2
= \underbrace{\frac{1}{n^2}\sum_i
\left(\text{Ch.1}_i\right)^2}_
{\text{covariate sampling}}
+ \underbrace{\frac{1}{n^2}\sum_i
\left(\text{Ch.2}_i\right)^2}_{
= \; \color{#31A354}{J} \, V_\beta \,
\color{#31A354}{J^T}}
+ \underbrace{\frac{2}{n^2}\sum_i
\text{Ch.1}_i \cdot \text{Ch.2}_i}_
{\text{cross-term}}
$$

The delta method gives only the middle term — **Channel 2 squared**. It misses
the covariate sampling variance (Channel 1 squared) and the cross-term. All
three terms are $O(1/n)$, so the delta method systematically underestimates.

::: {.callout-important}
### The delta method treats covariates as fixed
$\color{#31A354}{J} V_\beta \color{#31A354}{J^T}$ answers the question: "given
these specific covariates $\{X_i\}$, how does uncertainty in $\hat\beta$
propagate to $\hat\mu_a$?" It conditions on the covariate distribution. But
$\{X_i\}$ are random — a different sample gives different covariates, hence
a different average of predictions even for fixed $\beta$. The full IF captures
both sources; the delta method captures only one.
:::

### 3.4 When the shortcut is exact

The delta method underestimates whenever Channel 1 is nonzero. Channel 1 is:

$$
\text{Ch.1}_i = g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a
$$

This is zero for all $i$ if and only if **every individual gets the same
prediction** — i.e. $g^{-1}(X_i^{*T}\hat\beta)$ does not depend on $i$.
This happens when:

1. **The model is saturated in the treatment** (e.g. $Y \sim A$ for binary
   treatment) **and** the intervention is static ($A = a$ for everyone). Then
   all individuals share the same $X^*$ (only the intercept and treatment
   indicator matter), so predictions are identical.

2. **The model has no covariates** (e.g. $Y \sim A$). Same reasoning.

The shortcut **fails** when predictions vary across individuals:

- **G-computation** with confounders: $Y \sim A + L$. Predictions
  $g^{-1}(\hat\beta_0 + \hat\beta_1 a + \hat\beta_2 L_i)$ vary with $L_i$
  across individuals. Channel 1 is nonzero.

- **Non-saturated MSM**: $Y \sim A$ for continuous $A$ is a linear model, not
  saturated. With non-static interventions (e.g. shift each person's treatment
  by $+\delta$), different individuals get different $A$ values, hence different
  predictions.

| Situation | Predictions constant? | Delta method exact? |
|---|---|---|
| G-comp with confounders ($Y \sim A + L$) | No | **No** — underestimates |
| IPW/matching, binary $A$, saturated MSM ($Y \sim A$), static intervention | Yes | Yes |
| IPW/matching, continuous $A$, linear MSM ($Y \sim A$), static intervention | Yes | Yes |
| IPW, continuous $A$, non-static intervention (shift, MTP) | No | **No** — underestimates |


### 3.5 The IF is the stacked sandwich, solved analytically

The two-channel formula (@eq-if-gcomp) may seem to appear from thin air. In
fact it is the sandwich (@eq-sandwich) applied to the **complete** estimating
equations — model score plus mean equation — with the block-triangular bread
inverted by back-substitution. This section shows the derivation.

#### The complete estimating equations

G-computation has two estimating equations per individual:

$$
\underset{(p+1) \times 1}{\Psi(O_i;\,\beta,\,\mu_a)}
= \begin{pmatrix}
\underset{p \times 1}{\psi_\beta(O_i;\,\beta)} \\[4pt]
\underset{1 \times 1}{\psi_\mu(O_i;\,\beta,\,\mu_a)}
\end{pmatrix}
= \begin{pmatrix}
\text{GLM score (Section 1.2)} \\[4pt]
g^{-1}(X_i^{*T}\beta) - \mu_a
\end{pmatrix}
$$

The first block $\psi_\beta$ defines the model. The second block $\psi_\mu$
defines the estimand: $\hat\mu_a$ is the value of $\mu_a$ that makes $\sum_i
\psi_\mu = 0$, i.e. $\hat\mu_a = \frac{1}{n}\sum_i g^{-1}(X_i^{*T}
\hat\beta)$.

#### The block-triangular bread

The bread matrix of the stacked system is:

$$
\underset{(p+1) \times (p+1)}{A}
= -\frac{1}{n}\sum_i \frac{\partial\Psi_i}{\partial(\beta,\mu_a)^T}
= \begin{pmatrix}
\underset{p \times p}{\color{#3182BD}{A_{\beta\beta}}} &
\underset{p \times 1}{0} \\[6pt]
\underset{1 \times p}{\color{#E6550D}{A_{\mu\beta}}} &
\underset{1 \times 1}{\color{#E6550D}{A_{\mu\mu}}}
\end{pmatrix}
$$

where:

- $\color{#3182BD}{A_{\beta\beta}} = -\frac{1}{n}\sum_i
  \frac{\partial\psi_\beta}{\partial\beta^T}$ — the GLM Hessian
  (Section 1.4).
- $\color{#E6550D}{A_{\mu\beta}} = -\frac{1}{n}\sum_i
  \frac{\partial\psi_\mu}{\partial\beta^T}
  = -\frac{1}{n}\sum_i \frac{d\mu}{d\eta}\big|_{\eta_i^*} X_i^{*T}
  = -\color{#31A354}{J_a}$ — the negative Jacobian from Section 3.2.
- $\color{#E6550D}{A_{\mu\mu}} = -\frac{\partial\psi_\mu}{\partial\mu_a}
  = 1$.
- Upper-right = $0$: the GLM score does not depend on $\mu_a$.

The block-triangular structure arises because $\psi_\beta$ depends only on
$\beta$ (not on $\mu_a$), while $\psi_\mu$ depends on both $\beta$ and
$\mu_a$.

#### Inverting by back-substitution

For a $2 \times 2$ block-lower-triangular matrix, the inverse is:

$$
A^{-1} = \begin{pmatrix}
\color{#3182BD}{A_{\beta\beta}^{-1}} & 0 \\[4pt]
-A_{\mu\mu}^{-1} \, \color{#E6550D}{A_{\mu\beta}} \,
\color{#3182BD}{A_{\beta\beta}^{-1}} &
A_{\mu\mu}^{-1}
\end{pmatrix}
$$

The IF for the full parameter $(\beta, \mu_a)$ is $A^{-1}\Psi_i$. Extracting
the $\mu_a$ row (the last row of $A^{-1}$ times $\Psi_i$):

$$
\text{IF}_i(\hat\mu_a)
= \underbrace{A_{\mu\mu}^{-1}\;\psi_{\mu,i}}_{\text{Channel 1}}
\;+\;
\underbrace{(-A_{\mu\mu}^{-1})\;
\color{#E6550D}{A_{\mu\beta}}\;
\color{#3182BD}{A_{\beta\beta}^{-1}}\;
\psi_{\beta,i}}_{\text{Channel 2}}
$$

Substituting $A_{\mu\mu} = 1$, $A_{\mu\beta} = -\color{#31A354}{J_a}$, and
$\psi_{\mu,i} = g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a$:

$$
\text{IF}_i(\hat\mu_a) =
\underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_
{\text{Channel 1}}
\;+\;
\underbrace{
\color{#31A354}{J_a}\;
\color{#3182BD}{A_{\beta\beta}^{-1}}\;
\psi_{\beta,i}}_{\text{Channel 2}}
$$

This is exactly @eq-if-gcomp. The bread $\color{#3182BD}{A_{\beta\beta}^{-1}}$
is `model_bread_inv()` = $(X^TWX)^{-1}$. The score $\psi_{\beta,i}$ is
$r_i \cdot (d\mu/d\eta) / V(\mu_i) \cdot X_i$. The Jacobian
$\color{#31A354}{J_a}$ projects the $p$-vector result to a scalar.

::: {.callout-note}
### The meat is hidden, not absent
The IF formula computes $\color{#3182BD}{A^{-1}}\psi_i$ for each individual,
then squares: $\text{Var} = \frac{1}{n^2}\sum_i \text{IF}_i^2$. Expanding:

$$
\frac{1}{n^2}\sum_i \text{IF}_i^2
= \frac{1}{n}\;\color{#3182BD}{A^{-1}}
\underbrace{\frac{1}{n}\sum_i \Psi_i \Psi_i^T}_
{\color{#E6550D}{B} \text{ (the meat)}}
\color{#3182BD}{(A^{-1})^T}
$$

This is the sandwich (@eq-sandwich) for the stacked system. The IF avoids
forming the $(p+1) \times (p+1)$ meat matrix $\color{#E6550D}{B}$ — it
computes the same answer by summing scalar products.
:::

### 3.6 Target subpopulations (ATT, ATU, custom subset)

Sections 3.1–3.5 define the marginal mean as an average over the **full
sample**: $\hat\mu_a = (1/n)\sum_{i=1}^n g^{-1}(X_i^{*T}\hat\beta)$.
Many causal estimands — the ATT, the ATU, or an estimand on a
user-supplied subset — instead average only over a **target
subpopulation** $T \subseteq \{1,\ldots,n\}$ of size $n_t$:

$$
\hat\mu_a \;=\; \frac{1}{n_t}\sum_{i\in T} g^{-1}(X_i^{*T}\hat\beta)
$$

The stacked estimating equations need one small change to accommodate
this: the mean equation is now indicator-weighted. Let $t_i = \mathbb{1}[i \in T]$ and write

$$
\psi_{\mu,i}(\beta,\mu_a) \;=\; t_i \cdot \bigl[g^{-1}(X_i^{*T}\beta) - \mu_a\bigr]
$$

so that $\sum_i \psi_{\mu,i} = 0$ still recovers $\hat\mu_a$.

Redoing the bread calculation from Section 3.5 with this indicator-weighted mean equation:

- $\color{#E6550D}{A_{\mu\mu}} = -\frac{1}{n}\sum_i \frac{\partial\psi_{\mu,i}}{\partial\mu_a} = -\frac{1}{n}\sum_i(-t_i) = \frac{n_t}{n}$, so $A_{\mu\mu}^{-1} = \frac{n}{n_t}$.
- $\color{#E6550D}{A_{\mu\beta}} = -\frac{1}{n}\sum_i t_i \cdot \frac{d\mu}{d\eta}\Big|_{\eta_i^*} X_i^{*T} = -\frac{n_t}{n}\,\color{#31A354}{J_a}$, where $\color{#31A354}{J_a} = \frac{1}{n_t}\sum_{i\in T}\frac{d\mu}{d\eta}\big|_{\eta_i^*} X_i^{*T}$ is the Jacobian averaged over the target.

Plugging into the back-substitution formula for the last row of $A^{-1}$:

$$
\text{IF}_i(\hat\mu_a)
\;=\; -A_{\mu\mu}^{-1}\,\color{#E6550D}{A_{\mu\beta}}\,\color{#3182BD}{A_{\beta\beta}^{-1}}\,\psi_{\beta,i}
\;+\; A_{\mu\mu}^{-1}\,\psi_{\mu,i}
$$

Substituting $A_{\mu\mu}^{-1} = n/n_t$ and $A_{\mu\beta} = -(n_t/n)\,\color{#31A354}{J_a}$:

$$
\text{IF}_i(\hat\mu_a)
\;=\; \underbrace{\frac{n}{n_t}\cdot t_i \cdot (g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a)}_
{\text{Channel 1 (rescaled)}}
\;+\; \underbrace{\color{#31A354}{J_a}\,\color{#3182BD}{A_{\beta\beta}^{-1}}\,\psi_{\beta,i}}_
{\text{Channel 2 (unchanged)}}
$$ {#eq-if-gcomp-target}

Two things to notice:

1. **Channel 1 picks up a factor of $n/n_t$** and is nonzero only for
   individuals inside the target. Without this factor the variance is
   underestimated by $(n_t/n)^2$ — a serious undercount when the target
   is a proper subset (e.g., the ATT with only ~40% treated).
2. **Channel 2 is unchanged.** The $n_t/n$ in $A_{\mu\beta}$ cancels with
   the $n/n_t$ in $A_{\mu\mu}^{-1}$, so the Channel 2 formula from
   Section 3.2 carries over verbatim — $\color{#31A354}{J_a}$ is just
   averaged over the target population instead of the full sample, and
   $\psi_{\beta,i}$ still runs over every individual the outcome model
   used (inside **and** outside the target), because the outcome model
   was fit on all those rows.

The variance is still
$\widehat{\text{Var}}(\hat\mu_a) = (1/n^2)\sum_{i=1}^n \text{IF}_i^2$,
summed over all $n$ individuals. Individuals outside the target have
$\text{Ch.1}_i = 0$ and contribute only through Channel 2.

Setting $T = \{1,\ldots,n\}$ (so $n_t = n$, $t_i \equiv 1$) recovers
@eq-if-gcomp exactly. The full-sample case is the special case of
@eq-if-gcomp-target where the scaling factor is $1$.

### 3.7 Pseudo-code: the full IF for a marginal mean

This incorporates the target subpopulation scaling from Section 3.6
and the correct non-canonical link handling for the score (multiply
response residuals by $\mu_\eta/V$; see Section 1.2 route 3).

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Pseudo-code: gcomp_influence()"
gcomp_influence <- function(model, data_a, target_idx) {
  n <- nrow(data_a)
  beta_hat <- coef(model)
  pred_terms <- delete.response(terms(model))
  family <- model$family

  # Predictions under intervention (target rows only)
  X_star <- model.matrix(pred_terms, data = data_a[target_idx, ])
  eta_star <- as.numeric(X_star %*% beta_hat)
  preds <- family$linkinv(eta_star)
  mu_hat <- mean(preds)
  n_t <- length(preds)

  # Channel 1: (n/n_t) * t_i * (pred_i - mu_hat), zero outside target
  IF_direct <- rep(0, n)
  IF_direct[target_idx] <- preds - mu_hat

  # Channel 2 gradient: J_a = (1/n_t) sum_{target} (dmu/deta) * X_star
  mu_eta_star <- family$mu.eta(eta_star)
  J <- as.numeric(crossprod(X_star, mu_eta_star)) / n_t  # p x 1

  # Bread inverse (X'WX)^{-1} with IWLS working weights
  X_fit <- model.matrix(model)
  eta_fit <- model$linear.predictors
  mu_eta_fit <- family$mu.eta(eta_fit)
  V_fit <- family$variance(fitted(model))
  W <- mu_eta_fit^2 / V_fit
  XtWX_inv <- solve(crossprod(X_fit, X_fit * W))
  h <- as.numeric(XtWX_inv %*% J)  # p x 1

  # Per-observation directional sensitivity and SCORE residual.
  # Score residual = response residual * (dmu/deta) / V(mu), so that
  # psi_i = X_i * r_i * (mu_eta/V). For canonical links mu_eta/V = 1
  # and this reduces to the response residual.
  d_fit <- as.numeric(X_fit %*% h)
  r_response <- residuals(model, type = "response")
  r_score <- r_response * mu_eta_fit / V_fit
  IF_correction <- d_fit * r_score                         # n_fit x 1

  # Combine: full IF = (n/n_t) * Ch.1 + n * Ch.2
  # Factor of n on Ch.2 from A_{beta,beta}^{-1} = n * (X'WX)^{-1}
  # Factor of (n/n_t) on Ch.1 from A_{mu,mu}^{-1} = n/n_t (Sec. 3.6)
  IF <- (n / n_t) * IF_direct
  IF[seq_along(IF_correction)] <- IF[seq_along(IF_correction)] +
    n * IF_correction

  IF
}
```


## 4. The three causal estimators

This section applies the IF framework from Section 3 to g-computation, IPW, and
matching. The variance formula is always the same (@eq-var-if); what differs is
the estimating equations and whether Channel 1 is zero.

### 4.1 G-computation

**Estimating equations.** G-computation fits an outcome model
$E[Y_i \mid A_i, L_i] = g^{-1}(X_i^T \beta)$ and computes
$\hat\mu_a = \frac{1}{n}\sum_i g^{-1}(X_i^{*T}\hat\beta)$ where $X_i^*$ sets
treatment to $a$. The complete estimating equations are:

$$
\begin{pmatrix}
\psi_\beta(O_i; \beta) \\[4pt]
\psi_\mu(O_i; \beta, \mu_a)
\end{pmatrix}
=
\begin{pmatrix}
\text{GLM score for outcome model} \\[4pt]
g^{-1}(X_i^{*T}\beta) - \mu_a
\end{pmatrix}
$$

**IF.** From Section 3.2 (@eq-if-gcomp):

$$
\text{IF}_i(\hat\mu_a) =
\underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_
{\text{Channel 1 (nonzero)}}
\;+\;
\underbrace{
\color{#31A354}{J_a}
\color{#3182BD}{A^{-1}}
\psi_i
}_{\text{Channel 2}}
$$

**Both channels are needed** because predictions
$g^{-1}(\hat\beta_0 + \hat\beta_1 a + \hat\beta_2 L_i)$ vary across
individuals through $L_i$. The delta method shortcut
$\color{#31A354}{J} V_\beta \color{#31A354}{J^T}$ underestimates.

**Variance**: $\frac{1}{n^2}\sum_i \text{IF}_i^2$ using the full two-channel IF.

### 4.2 Inverse probability weighting (IPW)

IPW is more complex because two models are involved, and the uncertainty from
the first (propensity scores) must propagate into the second (MSM).

#### The two-model setup

1. **Propensity score model**: estimate $\hat\alpha$ ($q$-vector) from
   $\text{logit}[P(A_i = 1 \mid L_i)] = L_i^T \alpha$.
2. **Weighted outcome model (MSM)**: using the estimated weights $w_i =
   1/\hat{P}(A_i \mid L_i; \hat\alpha)$, fit $E[Y_i \mid A_i] =
   g^{-1}(\beta_0 + \beta_1 A_i)$ by weighted regression to get $\hat\beta$
   ($p$-vector).

The weights $w_i$ depend on $\hat\alpha$, so $\hat\beta$ depends on
$\hat\alpha$.

#### Why the naive approach fails

A naive sandwich on the outcome model alone — treating the weights as *known
constants* — ignores the variability in $\hat\alpha$. Each new sample produces
different $\hat\alpha$, hence different weights, hence different $\hat\beta$.
The true variability of $\hat\beta$ has two sources:

- **Outcome variability**: randomness in $Y_i$ given fixed weights.
- **Weight variability**: randomness in $\hat\alpha$ (and hence $w_i$).

The naive sandwich captures only the first. The direction of the bias depends
on the estimand: for the **ATE**, using estimated (rather than known) propensity
scores actually *reduces* the asymptotic variance, so the naive sandwich is
conservative — it **overestimates** the SE (Hirano et al., 2003; Lunceford &
Davidian, 2004). For the **ATT**, the effect can go either direction.

#### The correct approach: joint M-estimation

Treat both models together as a single M-estimation problem. Define the joint
parameter $\theta = (\alpha, \beta)$ — a $(q+p)$-vector — and the stacked
estimating function:

$$
\underset{(q+p) \times 1}{\Psi(O_i; \theta)} =
\begin{pmatrix}
\color{#3182BD}{\psi_\alpha(O_i; \alpha)} \\[4pt]
\color{#E6550D}{\psi_\beta(O_i; \alpha, \beta)}
\end{pmatrix}
$$

where $\color{#3182BD}{\psi_\alpha}$ ($q \times 1$) is the score of the
propensity model, and $\color{#E6550D}{\psi_\beta}$ ($p \times 1$) is the
*weighted* score of the outcome model (which depends on $\alpha$ through $w_i$).

The bread matrix $\hat A = -\frac{1}{n}\sum_i
\frac{\partial\Psi_i}{\partial\theta^T}$ has a block structure:

$$
\underset{(q+p) \times (q+p)}{\hat{A}} =
\begin{pmatrix}
\underset{q \times q}{\color{#3182BD}{A_{\alpha\alpha}}} &
\underset{q \times p}{0} \\[6pt]
\underset{p \times q}{\color{#E6550D}{A_{\beta\alpha}}} &
\underset{p \times p}{\color{#E6550D}{A_{\beta\beta}}}
\end{pmatrix}
$$

Each block has a concrete meaning:

- $\color{#3182BD}{A_{\alpha\alpha}}$ ($q \times q$): the Hessian of the
  propensity model — how $\psi_\alpha$ changes with $\alpha$.
- $\color{#E6550D}{A_{\beta\beta}}$ ($p \times p$): the weighted Hessian of
  the outcome model — how $\psi_\beta$ changes with $\beta$.
- $\color{#E6550D}{A_{\beta\alpha}}$ ($p \times q$): the **cross-term** — how
  the weighted outcome score changes with $\alpha$. This involves
  $\frac{\partial w_i}{\partial\alpha^T}$ (how the weights shift when the
  propensity parameters change). This is the block that captures the dependency.
- Upper-right = $0$ ($q \times p$): the propensity score does not depend on
  $\beta$.

The full sandwich $\frac{1}{n}\hat A^{-1}\hat B(\hat A^{-1})^T$ applied to
the stacked system correctly propagates weight uncertainty into
$\text{Var}(\hat\beta)$ through the $A_{\beta\alpha}$ block.

::: {.callout-tip}
### In practice
`WeightIt::glm_weightit()` fits the stacked system and its `vcov()` method
returns the M-estimation sandwich that accounts for weight uncertainty.
:::

#### The full IF for $\hat\mu_a$

The stacked sandwich above gives $V_\beta$ — the variance of the MSM
parameters $\hat\beta$, accounting for weight uncertainty. But we need
$\text{Var}(\hat\mu_a)$, not $\text{Var}(\hat\beta)$. Following the same
approach as Section 3.5, we add the **mean equation** to the stack:

$$
\underset{(q+p+1) \times 1}{\Psi(O_i;\,\alpha,\,\beta,\,\mu_a)}
= \begin{pmatrix}
\color{#3182BD}{\psi_\alpha(O_i;\,\alpha)} \\[4pt]
\color{#E6550D}{\psi_\beta(O_i;\,\alpha,\,\beta)} \\[4pt]
\psi_\mu(O_i;\,\beta,\,\mu_a)
\end{pmatrix}
$$

The bread is now $3 \times 3$ block-lower-triangular:

$$
A = \begin{pmatrix}
\color{#3182BD}{A_{\alpha\alpha}} & 0 & 0 \\[4pt]
\color{#E6550D}{A_{\beta\alpha}} &
\color{#E6550D}{A_{\beta\beta}} & 0 \\[4pt]
0 & A_{\mu\beta} & A_{\mu\mu}
\end{pmatrix}
$$

where $A_{\mu\beta} = -\color{#31A354}{J}$ (negative Jacobian of predictions
w.r.t. $\beta$) and $A_{\mu\mu} = 1$ (same as Section 3.5). The zero in the
$(\mu, \alpha)$ position arises because the mean equation
$g^{-1}(X_i^{*T}\beta) - \mu_a$ does not depend on $\alpha$ directly — it
depends on $\alpha$ only through $\beta$.

Inverting by back-substitution and extracting the $\mu_a$ row gives:

$$
\text{IF}_i(\hat\mu_a)
= \underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_
  {\text{Channel 1 (direct)}}
+ \underbrace{\color{#31A354}{J}\;
  \color{#E6550D}{A_{\beta\beta}^{-1}}\;
  \psi_{\beta,i}}_{\text{MSM correction}}
+ \underbrace{\color{#31A354}{J}\;
  \color{#E6550D}{A_{\beta\beta}^{-1}}\;
  \color{#E6550D}{A_{\beta\alpha}}\;
  \color{#3182BD}{A_{\alpha\alpha}^{-1}}\;
  \psi_{\alpha,i}}_{\text{propensity correction}}
$$ {#eq-if-ipw}

Reading each term:

- **Channel 1**: how individual $i$'s prediction deviates from the mean (same
  as g-comp). Zero when predictions are constant (saturated MSM + static
  intervention); nonzero for non-static interventions or non-saturated MSMs.
- **MSM correction**: how $i$'s outcome residual (captured in $\psi_{\beta,i}$)
  propagates to $\hat\mu_a$ through the MSM parameters. Same structure as the
  g-comp Channel 2.
- **Propensity correction**: how $i$'s propensity score residual (captured in
  $\psi_{\alpha,i}$) propagates to $\hat\mu_a$ through the weights. The chain
  is: $\psi_{\alpha,i} \to \hat\alpha \to w_i \to \hat\beta \to \hat\mu_a$.
  The matrix product
  $\color{#E6550D}{A_{\beta\alpha}}\color{#3182BD}{A_{\alpha\alpha}^{-1}}$
  captures the first two links (propensity parameters → weights → MSM
  parameters); $\color{#31A354}{J}\color{#E6550D}{A_{\beta\beta}^{-1}}$
  captures the last two (MSM parameters → predictions → estimand).

The second and third terms can be combined:

$$
\text{IF}_i(\hat\mu_a)
= \text{Ch.1}_i
\;+\;
\color{#31A354}{J}\;
\underbrace{
\color{#E6550D}{A_{\beta\beta}^{-1}}
\left(
\psi_{\beta,i}
+ \color{#E6550D}{A_{\beta\alpha}}\;
\color{#3182BD}{A_{\alpha\alpha}^{-1}}\;
\psi_{\alpha,i}
\right)}_{\text{IF}_i(\hat\beta) \text{ from the stacked }
(\alpha,\beta) \text{ system}}
$$

The expression in parentheses is the **adjusted score**: the MSM score
$\psi_{\beta,i}$ corrected for propensity estimation. This is exactly what
`WeightIt::glm_weightit()` computes internally when producing its
M-estimation sandwich — we are now exposing it per-individual.

::: {.callout-note}
### Where are the bread and meat?
The bread appears three times:
$\color{#3182BD}{A_{\alpha\alpha}^{-1}}$ (propensity bread inverse),
$\color{#E6550D}{A_{\beta\beta}^{-1}}$ (MSM bread inverse), and
$\color{#31A354}{J}$ (Jacobian from the mean equation). The individual
scores $\psi_{\alpha,i}$ and $\psi_{\beta,i}$ are what would form the meat
if we squared and summed them. The IF avoids forming the
$(q+p+1) \times (q+p+1)$ meat matrix.
:::

### 4.3 Matching

After matching, the outcome model is fitted on the matched sample. Two issues
require special treatment.

#### Cluster-robust variance

Matched individuals within the same subclass (pair or stratum) are correlated —
they were selected together. The standard sandwich (which assumes independence)
would understate the meat.

The **cluster-robust** sandwich accounts for this. The bread
$\color{#3182BD}{A}$ is unchanged, but the meat sums scores *within* each
subclass before squaring:

$$
V_\beta = \color{#3182BD}{A^{-1}}
\;\color{#E6550D}{B_{\text{CL}}}\;
\color{#3182BD}{(A^{-1})^T}
$$

where:

$$
\color{#E6550D}{B_{\text{CL}}} = \frac{1}{n}\sum_{c=1}^C
\bar\psi_c \, \bar\psi_c^T,
\qquad \bar\psi_c = \sum_{i \in \text{subclass } c} \psi_i
\qquad (p \times 1)
$$

Squaring the summed scores (rather than summing the squared scores) allows
within-cluster correlations to inflate the meat.

#### The full IF for matching

The per-individual IF has the same two-channel structure as g-comp:

$$
\text{IF}_i(\hat\mu_a) =
\underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_
{\text{Channel 1}}
\;+\;
\underbrace{
\color{#31A354}{J_a}\;
\color{#3182BD}{A_{\beta\beta}^{-1}}\;
\psi_{\beta,i}}_{\text{Channel 2}}
$$

where $\psi_{\beta,i}$ is the outcome model score for individual $i$ on the
matched data. The difference from g-comp is in the **aggregation step**: instead
of treating each individual's IF as independent, we account for within-subclass
correlation by summing IFs within each matched set before squaring:

$$
\widehat{\text{Var}}(\hat\mu_a) = \frac{1}{n^2}\sum_{c=1}^{C}
\left(\sum_{i \in \text{subclass } c} \text{IF}_i\right)^2
$$

This is the IF analogue of the cluster-robust sandwich: instead of summing
squared IFs (which assumes independence), we sum the squared *cluster-level*
IFs. It is identical to $\color{#3182BD}{A^{-1}}
\color{#E6550D}{B_{\text{CL}}} \color{#3182BD}{(A^{-1})^T}$ from the
cluster-robust meat above, applied to the stacked system.

### 4.4 Summary: one method, three estimators

All three methods use the same variance formula — the IF squared and summed
(@eq-var-if). What differs is the estimating equations and how Channel 2
is computed.

| Method | Stacked equations | Channel 2 models | Aggregation |
|---|---|---|---|
| **G-comp** | Outcome model + mean eq. | 1 (outcome) | Standard |
| **IPW** | Propensity + MSM + mean eq. | 2 (propensity, MSM) | Standard |
| **Matching** | Outcome model + mean eq. | 1 (outcome) | Cluster-robust |
| **ICE** | $K\!+\!1$ outcome models + mean eq. | $K\!+\!1$ (backward chain) | Standard |

The IF formula handles all cases uniformly. The only per-method variation is
**which models contribute** to Channel 2 and **how IFs are aggregated** (summed
within clusters for matching, directly for all others).


## 5. Chained estimators: ICE g-computation

### 5.1 The chain rule for influence functions

Section 3 derived the IF for a two-step estimator (fit $\hat\beta$, then
compute $\hat\mu_a$). What happens with **multiple models in a chain**, where
each model's response is a prediction from the next? The same logic applies
recursively. If $\hat\mu = f(\hat\beta_0, \hat\beta_1, \ldots, \hat\beta_K)$
and each $\hat\beta_k$ is an M-estimator with bread $\color{#3182BD}{A_k}$ and
score $\psi_k$, the IF of $\hat\mu$ is (Hampel et al., 1986; Huber, 1981):

$$
\text{IF}_i(\hat\mu)
\;=\;
\text{IF}_i^{\text{direct}}
\;+\;
\sum_{k=0}^{K}
\frac{\partial \hat\mu}{\partial \hat\beta_k^T}
\;\cdot\;
\color{#3182BD}{A_k^{-1}} \; \psi_k(O_i; \hat\beta_k)
$$ {#eq-if-chain}

Each term in the sum is a "Channel 2" from Section 3.2 applied to model $k$:
the Jacobian $\partial\hat\mu/\partial\hat\beta_k^T$ (how $\hat\mu$ depends on
model $k$) times the IF of model $k$'s parameters
($\color{#3182BD}{A_k^{-1}}\psi_{k,i}$).

::: {.callout-note}
### Connections to other frameworks
This IF-based approach is equivalent to:

1. **Stacked estimating equations** (Stefanski & Boos, 2002; Zivich et al.,
   2024): concatenate all models' estimating equations and apply the generic
   sandwich to the full stack.
2. **Semiparametric efficiency theory** (Tsiatis, 2006): derive the efficient
   influence function (EIF) and project onto the parametric model.

All yield the same variance estimate.
:::

### 5.2 ICE g-computation: the motivating example

Iterated conditional expectation (ICE) g-computation estimates the
counterfactual mean under a treatment strategy in longitudinal data (Robins,
1986; Bang & Robins, 2005; Zivich et al., 2024). It fits $K+1$ outcome models
backward in time:

```{mermaid}
%%| label: fig-ice-chain
%%| fig-cap: "ICE backward iteration. Models are fit from final time to baseline. Each model's predictions become the next model's response."
flowchart RL
  MK["Model K (final time)<br/>Fit: E[Y | A_K, H_K]<br/>Predict → Ỹ_K"] -->|"Ỹ_K becomes<br/>response for"| MK1["Model K-1<br/>Fit: E[Ỹ_K | A_{K-1}, H_{K-1}]<br/>Predict → Ỹ_{K-1}"]
  MK1 -->|"Ỹ_{K-1} becomes<br/>response for"| Mdots["⋯"]
  Mdots -->|"response for"| M0["Model 0 (baseline)<br/>Fit: E[Ỹ_1 | A_0, H_0]<br/>Predict → Ỹ_0"]
  M0 -->|"average"| MU["μ̂ = mean(Ỹ_0)"]

  style MK fill:#f0f4ff,stroke:#3182BD
  style MK1 fill:#f0f4ff,stroke:#3182BD
  style M0 fill:#f0f4ff,stroke:#3182BD
  style MU fill:#e8f5e9,stroke:#31A354
```

Each model's response ($\tilde{Y}_k$) is a *prediction* from the next model in
the chain. This creates a cascade of dependencies.

### 5.3 Why naive approaches fail

**Naive 1: Variance of pseudo-outcomes only.** Uses only Channel 1 — ignores
all model correction terms.

**Naive 2: $\color{#31A354}{J}V_\beta\color{#31A354}{J^T}$ from model 0
only.** Captures Channel 2 for model 0 but ignores models $1, \ldots, K$.

**Correct: Full IF chain.** Account for all $K+1$ models via the chain rule.

### 5.4 Deriving the IF for chained models

#### The IF decomposition

Applying the chain rule (Section 5.1):

$$
\text{IF}_i
= \underbrace{\frac{1}{n_t}(\tilde{Y}_{0,i} - \hat\mu)}_{\text{Channel 1}}
+ \sum_{k=0}^{K}
\underbrace{
\color{#E6550D}{d_{k,i}} \cdot r^{\text{score}}_{k,i}
}_{\text{Channel 2 for model } k}
$$ {#eq-if-ice}

where $\color{#E6550D}{d_{k,i}}$ is the **sensitivity** — how much a
perturbation at model $k$ propagates to $\hat\mu$ — and
$r^{\text{score}}_{k,i}$ is individual $i$'s **score residual** at
model $k$:

$$
r^{\text{score}}_{k,i}
\;=\; (Y_{k,i} - \hat{Y}_{k,i}) \cdot
\frac{d\mu/d\eta\big|_{\eta_{k,i}}}{V(\hat{Y}_{k,i})}
$$

i.e. the response residual $r_{k,i} = Y_{k,i} - \hat{Y}_{k,i}$ weighted
by the link-scale factor from the GLM score (Section 1.2, route 3).
For **canonical** links (logistic-logit, Poisson-log, Gaussian-identity)
the ratio $(d\mu/d\eta)/V(\mu) = 1$ and $r^{\text{score}}_{k,i}$
collapses to the ordinary response residual $r_{k,i}$. For
**non-canonical** links (probit, cloglog, Gamma-log, Poisson-sqrt,
etc.) the factor is not $1$ and must be applied.

::: {.callout-warning}
### Why the score residual, not the response residual
$\text{Ch.2}_i = J_k A_k^{-1} \psi_{k,i}$ contains the full GLM score
$\psi_{k,i} = (Y - \mu)\cdot(d\mu/d\eta)/V(\mu)\cdot X_i$, not the bare
residual $(Y - \mu)\cdot X_i$. Working through the algebra:

$$
\color{#31A354}{J_k}\color{#3182BD}{A_k^{-1}}\psi_{k,i}
\;=\; n\cdot d_{k,i}\cdot (Y_{k,i} - \hat Y_{k,i})\cdot
\frac{d\mu/d\eta\big|_{\eta_{k,i}}}{V(\hat Y_{k,i})}
\;=\; n\cdot d_{k,i}\cdot r^{\text{score}}_{k,i}
$$

The factor of $n$ arises from $A_k^{-1} = n(X_k^T W_k X_k)^{-1}$
(Section 5.5). Omitting the $(d\mu/d\eta)/V$ factor and using the plain
response residual is correct only when it equals $1$ — i.e., only for
canonical links. This mistake is silent on logistic/Poisson-log/Gaussian
models but miscalibrates probit and cloglog.
:::

#### What the sensitivities measure

Each model correction requires a **sensitivity** $\color{#E6550D}{d_{k,i}}$: a
scalar that answers "if individual $i$'s residual at model $k$ increased by one
unit, how much would $\hat\mu$ change?" Computing $\color{#E6550D}{d_{k,i}}$
takes three steps, each with a clear role:

1. **Gradient** $g_k$ ($p_k \times 1$): the total sensitivity of $\hat\mu$ to
   model $k$'s coefficient vector $\beta_k$. This is a Jacobian — the same
   object as $\color{#31A354}{J_a}$ from Section 3.2, but computed recursively
   for $k > 0$.
2. **Solve** $h_k = (X_k^T W_k X_k)^{-1} g_k$ ($p_k \times 1$): convert from
   "how $\hat\mu$ responds to $\beta_k$" to "how $\hat\mu$ responds to
   individual score contributions." This is the bread inverse
   $\color{#3182BD}{A_k^{-1}}$ applied to $g_k$.
3. **Distribute** $\color{#E6550D}{d_{k,i}} = h_k^T X_{k,i}$ (scalar): project
   onto individual $i$'s covariate vector to get their personal sensitivity.

#### Base case: model 0

$\hat\mu$ is a direct average of model 0's predictions. The gradient of
$\hat\mu$ w.r.t. $\beta_0$ is therefore the Jacobian from Section 3.2:

$$
\underset{p_0 \times 1}{g_0}
= \frac{1}{n_t}\sum_{j \in \text{target}}
\frac{d\mu}{d\eta}\bigg|_{\eta_{0,j}^*}
\;\cdot\; X_{0,j}^*
$$

$g_0$ tells us the total sensitivity of $\hat\mu$ to $\beta_0$, but the IF
requires a per-individual quantity. We need to ask: "of the total shift in
$\hat\beta_0$ caused by the data, how much came from individual $i$?" The
answer uses two facts: (a) individual $i$'s influence on $\hat\beta_0$ is
proportional to their score $\psi_{0,i} = (d\mu/d\eta)_i / V(\mu_i) \cdot
r_{0,i} \cdot X_{0,i}$; and (b) the constant of proportionality involves
the bread inverse $(X_0^T W_0 X_0)^{-1}$. Combining:

$$
h_0 = \underbrace{(X_0^T W_0 X_0)^{-1}}_{\text{bread inverse}} g_0,
\qquad
\color{#E6550D}{d_{0,i}} = h_0^T X_{0,i}
$$

Now $\color{#E6550D}{d_{0,i}} \cdot r^{\text{score}}_{0,i}$ gives
individual $i$'s Channel 2 contribution: their **score residual**
$r^{\text{score}}_{0,i} = (Y_{0,i} - \hat Y_{0,i}) \cdot (d\mu/d\eta)_i /
V(\mu_i)$ (how much their data "disagrees" with model 0, on the
link-scaled score metric) scaled by $\color{#E6550D}{d_{0,i}}$ (how much
that disagreement matters for $\hat\mu$). For canonical links
$(d\mu/d\eta)/V = 1$ and the score residual collapses to the response
residual $r_{0,i}$. For non-canonical links the link-scale factor must
be kept. This is equivalent to
$\color{#31A354}{J_0}\color{#3182BD}{A_0^{-1}}\psi_{0,i}$ from
Section 3.2, factored into the three steps above.

#### Recursive case: model $k = 1, \ldots, K$

Model $k$ does not feed into $\hat\mu$ directly. Instead, model $k$'s
predictions become model $k-1$'s pseudo-outcomes (response variable). We already
know how sensitive $\hat\mu$ is to each of those pseudo-outcomes — that is
$\color{#E6550D}{d_{k-1,j}}$ from the previous step.

The gradient of $\hat\mu$ w.r.t. $\beta_k$ therefore weights each individual's
contribution by the previously-computed sensitivity, further scaled by the
prior weight $w_{k-1,j}$ that the $(k-1)$-th score function carries for
row $j$:

$$
\underset{p_k \times 1}{g_k}
= \sum_{j \in \text{fit}_{k-1}}
w_{k-1,j}\;\cdot\;
\color{#E6550D}{d_{k-1,j}}
\;\cdot\;
\frac{d\mu}{d\eta}\bigg|_{\eta_{k,j}^*}
\;\cdot\; X_{k,j}^*
$$

The $w_{k-1,j}$ factor comes from $A_{k-1,k} =
E[\partial s_{k-1,j}/\partial \beta_k]$, and $s_{k-1,j}$ inherits
$w_{k-1,j}$ because the $(k-1)$-th pseudo-outcome model is a weighted
fit. In the unweighted case $w_{k-1,j} \equiv 1$ and the factor drops
out; dropping it under non-uniform external weights (e.g. IPCW,
survey weights) silently underestimates the sandwich variance by a
factor that scales with weight heterogeneity.

Compare this to $g_0$: the uniform weight $\frac{1}{n_t}$ over the target
population is replaced by $w_{k-1,j} \cdot \color{#E6550D}{d_{k-1,j}}$ over
the previous model's fitting set. Everything else — solve and distribute —
is identical:

$$
h_k = (X_k^T W_k X_k)^{-1} g_k,
\qquad
\color{#E6550D}{d_{k,i}} = h_k^T X_{k,i}
$$

### 5.5 Scaling: the factor of $n$

Each model correction involves $\color{#3182BD}{A_k^{-1}}$. Since
$\color{#3182BD}{A_k} = \frac{1}{n_k} X_k^T W_k X_k$, we have
$\color{#3182BD}{A_k^{-1}} = n_k(X_k^T W_k X_k)^{-1}$. So the correction at
step $k$ is:

$$
n_k \cdot \color{#E6550D}{d_{k,i}} \cdot r^{\text{score}}_{k,i}
$$

The factor $n_k$ ensures each IF contribution is $O(1)$ per individual, giving
the correct $\text{Var}(\hat\mu) = O(1/n)$.

::: {.callout-warning}
### A common bug
Omitting $n_k$ makes each correction $O(1/n)$, yielding $\text{Var} = O(1/n^3)$
— confidence intervals that are much too narrow.
:::

### 5.6 Pseudo-code: IF for a chained estimator

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Pseudo-code: chained_influence()"
chained_influence <- function(models, data_iv, target) {
  n <- number_of_individuals
  K <- length(models) - 1
  mu_hat <- mean(pseudo_final[target])
  n_target <- sum(target)

  # Channel 1 with target-subpopulation scaling (see Sec. 3.6).
  IF <- rep(0, n)
  IF[target] <- (n / n_target) * (pseudo_final[target] - mu_hat)

  d <- rep(0, n)  # sensitivities from previous step

  for (step in 0:K) {
    model_k <- models[[step + 1]]
    family_k <- model_k$family
    X_k <- model.matrix(model_k)         # fitting design matrix
    eta_k <- model_k$linear.predictors
    mu_eta_k <- family_k$mu.eta(eta_k)   # dmu/deta at fitted values
    V_k <- family_k$variance(fitted(model_k))
    W_k <- mu_eta_k^2 / V_k              # IWLS working weights
    fit_ids_k <- model_k$fitting_ids
    n_k <- length(fit_ids_k)

    # Score residual = response residual * (dmu/deta) / V(mu). For
    # canonical links this factor is 1; for non-canonical (probit,
    # cloglog, Gamma-log, ...) it is not.
    r_response_k <- residuals(model_k, type = "response")
    r_score_k <- r_response_k * mu_eta_k / V_k

    # Intervention design matrix and link derivatives (dmu/deta)
    X_star_k <- model.matrix(model_k, newdata = intervention_data[[step + 1]])
    eta_star <- X_star_k %*% coef(model_k)
    mu_eta_star <- family_k$mu.eta(eta_star)

    # Step 1: Gradient g_k = d mu_hat / d beta_k. The step > 0 branch
    # carries w_{k-1} from the previous-step fit's prior weights.
    if (step == 0) {
      g_k <- crossprod(X_star_k[target, ], mu_eta_star[target]) / n_target
    } else {
      prev_ids <- models[[step]]$fitting_ids
      w_prev <- models[[step]]$prior.weights  # 1-vector if unweighted
      g_k <- crossprod(
        X_star_k[prev_ids, ],
        w_prev * d[prev_ids] * mu_eta_star[prev_ids]
      )
    }

    # Step 2: Solve h_k = (X'WX)^{-1} g_k
    h_k <- solve(crossprod(X_k, X_k * W_k)) %*% g_k

    # Step 3: Distribute d_{k,i} = h_k' X_{k,i}
    d <- rep(0, n)
    d[fit_ids_k] <- X_k %*% h_k

    # Channel 2 for model k: n_k * d_{k,i} * r_score_{k,i}
    r_vec <- rep(0, n)
    r_vec[fit_ids_k] <- r_score_k
    IF <- IF + n_k * d * r_vec
  }
  IF
}
```

### 5.7 From individual IFs to the variance

Once we have the per-individual $\text{IF}_i$ from the recursion above, variance
estimation follows from the same formula as always (@eq-var-if):

$$
\widehat{\text{Var}}(\hat\mu)
= \frac{1}{n^2}\sum_{i=1}^n \text{IF}_i^2
$$

For two interventions $a$ and $b$ on the same data:

$$
\widehat{\text{Cov}}(\hat\mu_a, \hat\mu_b)
= \frac{1}{n^2}\sum_{i=1}^n \text{IF}_i^{(a)} \cdot \text{IF}_i^{(b)}
$$

The full $k \times k$ covariance matrix is then used for contrasts.


## 6. Summary

1. **One method**: the influence function. For any causal estimand $\hat\mu$,
   write down the complete estimating equations (nuisance models + target
   estimand), compute each individual's $\text{IF}_i$, and take
   $\widehat{\text{Var}}(\hat\mu) = \frac{1}{n^2}\sum_i \text{IF}_i^2$.

2. **The IF is the sandwich**: the IF formula is the sandwich
   $\color{#3182BD}{A^{-1}} \color{#E6550D}{B} \color{#3182BD}{(A^{-1})^T}$
   applied to the stacked estimating equations, with the block-triangular
   bread inverted by back-substitution. The bread, meat, and scores are all
   present — the IF just computes $A^{-1}\psi_i$ per individual instead of
   forming the full matrices.

3. **Two channels**: the IF always decomposes into a **direct term**
   (Channel 1: covariate sampling) and **model corrections** (Channel 2:
   one per nuisance model). What varies across methods is only which models
   contribute to Channel 2.

4. **G-computation**: one model correction (outcome model). Channel 1 nonzero.

5. **IPW**: two model corrections (MSM + propensity). The propensity
   correction captures weight estimation uncertainty via the cross-derivative
   $A_{\beta\alpha}$.

6. **Matching**: one model correction (outcome model), same as g-comp.
   Aggregation is cluster-robust (sum IFs within matched subclasses).

7. **Chained estimators** (ICE): $K\!+\!1$ model corrections computed by
   backward recursion of sensitivities.

8. **The delta method** ($\color{#31A354}{J} V_\beta
   \color{#31A354}{J^T}$) is a shortcut that captures Channel 2 only. It is
   exact when Channel 1 = 0 (constant predictions), but underestimates
   otherwise. The full IF avoids this limitation.


## References

- Bang H, Robins JM (2005). Doubly robust estimation in missing data and
  causal inference models. *Biometrics* 61:962–973.
- Hampel FR, Ronchetti EM, Rousseeuw PJ, Stahel WA (1986). *Robust
  Statistics: The Approach Based on Influence Functions*. Wiley.
- Hirano K, Imbens GW, Ridder G (2003). Efficient estimation of average
  treatment effects using the estimated propensity score.
  *Econometrica* 71:1161–1189.
- Hernán MA, Robins JM (2025). *Causal Inference: What If*.
  Chapman & Hall/CRC.
- Huber PJ (1967). The behavior of maximum likelihood estimates under
  nonstandard conditions. *Proceedings of the Fifth Berkeley Symposium*
  1:221–233.
- Huber PJ (1981). *Robust Statistics*. Wiley.
- Lunceford JK, Davidian M (2004). Stratification and weighting via the
  propensity score in estimation of causal treatment effects: a comparative
  study. *Statistics in Medicine* 23:2937–2960.
- Robins JM (1986). A new approach to causal inference in mortality studies
  with a sustained exposure period. *Mathematical Modelling* 7:1393–1512.
- Stefanski LA, Boos DD (2002). The calculus of M-estimation.
  *The American Statistician* 56:29–38.
- Tsiatis AA (2006). *Semiparametric Theory and Missing Data*. Springer.
- van der Laan MJ, Robins JM (2003). *Unified Methods for Censored
  Longitudinal Data and Causality*. Springer.
- White H (1980). A heteroskedasticity-consistent covariance matrix estimator
  and a direct test for heteroskedasticity. *Econometrica* 48:817–838.
- White H (1982). Maximum likelihood estimation of misspecified models.
  *Econometrica* 50:1–25.
- Zeileis A (2004). Econometric computing with HC and HAC covariance matrix
  estimators. *Journal of Statistical Software* 11(10):1–17.
- Zeileis A (2006). Object-oriented computation of sandwich estimators.
  *Journal of Statistical Software* 16(9):1–16.
- Zivich PN, Klose M, Cole SR, Edwards JK, Shook-Sa BE (2022).
  Delicatessen: a Python package for the generalized computation of the
  sandwich estimator. *arXiv preprint* arXiv:2203.11300.
- Zivich PN, Ross RK, Shook-Sa BE, Cole SR, Edwards JK (2024). Empirical
  sandwich variance estimator for iterated conditional expectation
  g-computation. *Statistics in Medicine* 43:5562–5572.
