---
title: "Variance estimation for causal effects: sandwich estimators and influence functions"
toc-depth: 4
vignette: >
  %\VignetteIndexEntry{Variance estimation for causal effects: sandwich estimators and influence functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
mjx-container[jax="CHTML"][display="true"] {
  overflow-x: auto;
  overflow-y: visible;
  padding: 0.5em 0;
}
</style>
```

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette develops the theory behind variance estimation for causal effect
estimators — the *sandwich estimator*, the *delta method*, and *influence
functions* — with a focus on the ideas rather than any specific software. We
build up from first principles, use concrete examples, and include pseudo-code
so that every step is reproducible.

```{r}
#| label: setup-data
#| code-fold: true
#| code-summary: "Running example: simulated data"
set.seed(42)
n <- 200
L <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * L))
Y <- rbinom(n, 1, plogis(-1 + 0.8 * A + 0.6 * L))
dat <- data.frame(Y = Y, A = A, L = L)

fit <- glm(Y ~ A + L, data = dat, family = binomial)
beta_hat <- coef(fit)
```

## Setting: the causal effect estimation pipeline

Most causal effect estimators follow a two-stage pattern:

1. **Fit a nuisance model** $\hat\beta$ (an outcome regression, a propensity
   score model, or both).
2. **Compute the target quantity** $\hat\mu = g(\hat\beta)$ by plugging
   $\hat\beta$ into a functional of the data (e.g. a standardised mean, an
   inverse-probability-weighted mean, etc.).

To construct confidence intervals for $\hat\mu$, we need $\text{Var}(\hat\mu)$.
The bootstrap resamples the entire pipeline to get this, but it can be slow. The
*sandwich estimator* gives a closed-form alternative.

```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "The variance estimation pipeline. Each box is a mathematical object; each arrow is an operation covered in this vignette. The sandwich gives V_β from the fitted model; the delta method propagates V_β to the causal estimand via the Jacobian J; the influence function provides an equivalent per-individual decomposition."
flowchart LR
  D["Data<br/>(X, Y)"] -->|"fit model<br/>(Sec 1.1–1.3)"| B["β̂, ψ_i<br/>(coefficients,<br/>scores)"]
  B -->|"sandwich<br/>A⁻¹ B (A⁻¹)ᵀ<br/>(Sec 1.4)"| VB["V_β<br/>(param variance)"]
  VB -->|"delta method<br/>J V_β Jᵀ<br/>(Sec 2)"| VM["Var(μ̂)<br/>(estimand variance)"]
  B -->|"IF = A⁻¹ψ_i<br/>(Sec 4)"| IF["IF_i<br/>(per-individual)"]
  IF -->|"Var = Σ IF²/n²<br/>(Sec 4)"| VM

  style D fill:#f0f4ff,stroke:#3182BD
  style B fill:#fff5eb,stroke:#E6550D
  style VB fill:#f0f4ff,stroke:#3182BD
  style VM fill:#e8f5e9,stroke:#31A354
  style IF fill:#fff5eb,stroke:#E6550D
```

## 1. The sandwich estimator for parametric models

### 1.1 What is an estimating equation?

The starting point is a simple idea: many statistical estimators can be defined
as the solution to an equation of the form

$$
\sum_{i=1}^n \psi(O_i; \beta) = 0
$$

where $O_i$ is the observed data for individual $i$, $\beta$ is the parameter
we want to estimate, and $\psi$ is a function — called the **estimating
function** — that maps each observation and a candidate parameter value to a
vector. The estimator $\hat\beta$ is the value of $\beta$ that makes this sum
equal to zero.

But what *is* $\psi$, concretely? And where does it come from?

### 1.2 Deriving $\psi$: three routes to the same place

#### Route 1: Score equations (maximum likelihood)

The most familiar source of $\psi$ is maximum likelihood. If observations have
log-likelihood $\ell_i(\beta) = \log f(O_i; \beta)$, the MLE maximises
$\sum_i \ell_i(\beta)$. Setting the derivative to zero gives:

$$
\sum_{i=1}^n \underbrace{\frac{\partial \ell_i(\beta)}{\partial \beta}}_
{\psi(O_i;\beta)} = 0
$$

So $\psi$ is the **score function** — the gradient of the individual
log-likelihood. For a GLM with outcome $Y_i$, covariates $X_i$, linear
predictor $\eta_i = X_i^T\beta$, and mean function $\mu_i = g^{-1}(\eta_i)$:

$$
\psi(O_i; \beta) =
\underset{\color{#E6550D}{\text{residual}}}{\color{#E6550D}{(Y_i - \mu_i)}}
\;\cdot\;
\underset{\color{#3182BD}{\text{scale factor}}}
{\color{#3182BD}{\frac{d\mu/d\eta}{V(\mu_i)}}}
\;\cdot\;
\underset{\color{#31A354}{\text{direction}}}{\color{#31A354}{X_i}}
$$

where $V(\mu_i)$ is the variance function of the exponential family
($V(\mu) = \mu(1-\mu)$ for binomial, $V(\mu) = \mu$ for Poisson, $V(\mu) =
1$ for Gaussian — note that $\sigma^2$ enters separately as the dispersion
parameter $\phi$, not as $V(\mu)$). Reading the three factors:

- $\color{#E6550D}{(Y_i - \mu_i)}$ — **Residual**: how far the prediction is
  from the observation. This is zero on average at the truth, which is why
  $E[\psi_i] = 0$.
- $\color{#3182BD}{(d\mu/d\eta) / V(\mu_i)}$ — **Scale factor**: combines the
  link derivative (converting from $\eta$-scale to $\mu$-scale) with the
  variance weight (downweighting noisy observations). For canonical links this
  product simplifies (see below).
- $\color{#31A354}{X_i}$ — **Direction**: the covariates tell us *which*
  parameters this observation informs. An observation with large $|X_{ij}|$
  provides more information about $\beta_j$.

::: {.callout-note}
### Example: logistic regression
For logistic regression, $\mu_i = \text{expit}(\eta_i)$, so $d\mu/d\eta =
\mu_i(1-\mu_i)$ and $V(\mu_i) = \mu_i(1 - \mu_i)$. The scale factor cancels
to $1$, giving:

$$\psi(O_i;\beta) = (Y_i - \mu_i) \cdot X_i$$

The estimating equation $\sum_i (Y_i - \mu_i) X_i = 0$ says: at the MLE, the
covariates are uncorrelated with the residuals.
:::

#### Route 2: Moment conditions (method of moments)

Not every estimating equation comes from a likelihood. The **generalised method
of moments** (GMM) starts from a theoretical property of the true parameter
$\beta_0$ — a *moment condition* — and turns it into $\psi$.

For example, if we know that $E[X_i(Y_i - X_i^T\beta_0)] = 0$, we can define:

$$
\psi(O_i; \beta) = X_i(Y_i - X_i^T\beta)
$$

Setting $\sum_i \psi = 0$ gives the OLS normal equations. No distributional
assumption was needed — just the moment condition $E[X_i \varepsilon_i] = 0$.

This is the key insight of M-estimation: the estimating function $\psi$ does not
have to be a score function. It just needs to satisfy $E[\psi(O_i; \beta_0)] =
0$ at the true parameter value (the **unbiasedness** condition). This is what
guarantees consistency of $\hat\beta$.

#### Route 3: Quasi-likelihood

Quasi-likelihood provides a middle ground. We specify only the first two moments
of $Y_i \mid X_i$ — the mean function $\mu_i = g^{-1}(X_i^T\beta)$ and a
variance function $V(\mu_i)$ — without assuming a full distributional form. The
quasi-score is:

$$
\psi(O_i; \beta) = \frac{(Y_i - \mu_i)}{V(\mu_i)} \cdot
\frac{d\mu_i}{d\eta_i} \cdot X_i
$$

This looks identical to the GLM score, but now $V(\mu)$ is a *working*
variance — it might be wrong. The sandwich estimator (Section 1.4) protects
against this misspecification.

### 1.3 The general M-estimation framework

All three routes produce estimators of the same form: find $\hat\beta$ solving
$\sum_i \psi(O_i; \hat\beta) = 0$. The M-estimation framework (Huber, 1964;
Stefanski & Boos, 2002) unifies them under two conditions:

1. **Unbiasedness**: $E[\psi(O_i; \beta_0)] = 0$ at the true $\beta_0$. This
   guarantees that the estimating equation has the right solution in
   expectation.
2. **Regularity**: $\psi$ is smooth enough in $\beta$ for a Taylor expansion.

Given these, $\hat\beta$ is consistent and asymptotically normal, and its
variance has the sandwich form — regardless of whether $\psi$ came from a
likelihood, a moment condition, or a quasi-likelihood.

::: {.callout-tip}
### Why "M-estimation"?
The "M" stands for "maximum" (as in maximum-likelihood), but the framework is
much broader. Any estimator defined as the root of a sum of functions of the
data and parameters is an M-estimator — this includes MLE, weighted least
squares, quantile regression, robust regression (Huber loss), generalised
estimating equations, and the causal estimators in Sections 3–5 (Stefanski &
Boos, 2002).
:::

### 1.4 Asymptotic variance of M-estimators

We want $\text{Var}(\hat\beta)$. The derivation has three steps: linearise
$\hat\beta$ around the truth, identify the variance of the linear
approximation, then estimate it from data.

#### Step 1: Linearise via Taylor expansion

We know that $\hat\beta$ solves $\sum_i \psi(O_i; \hat\beta) = 0$. We also
know that the true $\beta_0$ satisfies $E[\psi(O_i; \beta_0)] = 0$, but the
finite-sample sum $\sum_i \psi(O_i; \beta_0)$ is generally *not* zero — it is
a random quantity that fluctuates around zero. The idea is to relate these two
facts via a Taylor expansion.

Expand $\sum_i \psi(O_i; \hat\beta)$ around $\beta_0$, keeping only the
first-order term (valid for large $n$ by regularity):

$$
\underbrace{\sum_i \psi(O_i; \hat\beta)}_{= \, 0 \text{ (by definition)}}
\approx \sum_i \psi(O_i; \beta_0) +
\left[\sum_i \frac{\partial \psi(O_i;\beta_0)}
{\partial \beta^T}\right]
(\hat\beta - \beta_0)
$$

Rearranging for the estimation error $\hat\beta - \beta_0$:

$$
\hat\beta - \beta_0 \approx
-\left[\sum_i \frac{\partial \psi(O_i;
\beta_0)}{\partial \beta^T}\right]^{-1}
\sum_i \psi(O_i; \beta_0)
$$

Dividing numerator and denominator by $n$ and labelling the two factors:

$$
\underset{p \times 1}{\hat\beta - \beta_0}
\;\approx\;
\underbrace{\left[-\frac{1}{n}\sum_i \frac{\partial \psi_i}
{\partial \beta^T}\right]^{-1}}_{\to\;
\color{#3182BD}{A^{-1}} \;(p \times p)}
\;\cdot\;
\underbrace{\frac{1}{n}\sum_i \psi(O_i; \beta_0)}_{\text{mean zero,
covariance } \color{#E6550D}{B/n}}
$$

This is the key linearisation: the estimation error is approximately the bread
inverse times the average score. The first factor converges to the constant
matrix $\color{#3182BD}{A^{-1}}$; the second is a sum of iid mean-zero random
$p$-vectors whose covariance we call $\color{#E6550D}{B}$.

#### Step 2: Identify the variance

Name the two matrices that appeared in Step 1:

- **Bread** $\color{#3182BD}{A}$: the expected negative Jacobian of the
  estimating function ($p \times p$):

$$
\color{#3182BD}{A}
\;\equiv\; -E\!\left[\frac{\partial \psi(O_i; \beta_0)}
{\partial \beta^T}\right]
\qquad (p \times p)
$$

For a GLM, $\color{#3182BD}{A} = E[w_i X_i X_i^T]$ where $w_i =
(d\mu/d\eta)^2 / V(\mu)$ are the IWLS working weights — essentially the
curvature of the (quasi-)log-likelihood.

- **Meat** $\color{#E6550D}{B}$: the covariance of the estimating function
  ($p \times p$):

$$
\color{#E6550D}{B}
\;\equiv\; E\!\left[\psi(O_i; \beta_0)\,\psi(O_i; \beta_0)^T\right]
\qquad (p \times p)
$$

This measures how much the individual scores $\psi_i$ vary across observations.
It appeared in Step 1 as the covariance of the second factor: $\text{Var}\!
\left(\frac{1}{n}\sum_i \psi_i\right) = \frac{1}{n}\color{#E6550D}{B}$.

Now propagate through the linearisation. Taking $\text{Var}(\cdot)$ of both
sides of the Step 1 equation:

$$
\underset{p \times p}{\text{Var}(\hat\beta)}
\;\approx\;
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\cdot\;
\underset{p \times p}{\frac{\color{#E6550D}{B}}{n}}
\;\cdot\;
\underset{p \times p}{\color{#3182BD}{(A^{-1})^T}}
\;=\;
\frac{1}{n}\,
\color{#3182BD}{A^{-1}}\,
\color{#E6550D}{B}\,
\color{#3182BD}{(A^{-1})^T}
$$

#### Step 3: Estimate from data

Replace the population expectations $A$ and $B$ with sample averages:

$$
\color{#3182BD}{\hat{A}}
= -\frac{1}{n}\sum_i
\frac{\partial \psi(O_i; \hat\beta)}{\partial \beta^T}
\qquad (p \times p)
$$

$$
\color{#E6550D}{\hat{B}}
= \frac{1}{n}\sum_i
\psi(O_i; \hat\beta)\,\psi(O_i; \hat\beta)^T
\qquad (p \times p)
$$

The **sandwich estimator** is:

$$
\underset{p \times p}{\widehat{\text{Var}}(\hat\beta)}
= \frac{1}{n}\;
\underset{p \times p}{\color{#3182BD}{\hat{A}^{-1}}}
\;
\underset{p \times p}{\color{#E6550D}{\hat{B}}}
\;
\underset{p \times p}{\color{#3182BD}{(\hat{A}^{-1})^T}}
$$ {#eq-sandwich}

so called because the meat $\color{#E6550D}{\hat B}$ is "sandwiched" between
two slices of bread $\color{#3182BD}{\hat A^{-1}}$.

::: {.callout-important}
### Why is this robust?
The bread $\color{#3182BD}{A}$ and meat $\color{#E6550D}{B}$ are estimated
*separately*. Under correct model specification (e.g. the GLM variance function
is right), $A = B$ and the sandwich simplifies to $\frac{1}{n} A^{-1}$ — the
usual model-based variance. But if the variance function is wrong, $A \neq B$:
the bread still captures the curvature correctly (it only depends on the mean
model), while the meat captures the *actual* variability of the scores. The
sandwich remains consistent regardless — this is the Huber–White robustness
property (Huber, 1967; White, 1982).
:::

### 1.5 Pseudo-code: sandwich variance for a GLM

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Pseudo-code: sandwich_variance()"
sandwich_variance <- function(X, Y, beta_hat, family) {
  n <- length(Y)
  eta <- X %*% beta_hat
  mu <- family$linkinv(eta)
  mu_eta <- family$mu.eta(eta)    # dμ/dη
  V <- family$variance(mu)        # Var(Y|X) under the model

  # Bread: A = (1/n) X^T W X where W = diag((dμ/dη)² / V(μ))
  W <- mu_eta^2 / V
  A <- (1 / n) * crossprod(X, X * W)

  # Meat: B = (1/n) Σ ψ_i ψ_i^T
  residuals <- Y - mu
  score_weights <- mu_eta / V
  psi <- X * (residuals * score_weights)  # n × p matrix of score contributions
  B <- (1 / n) * crossprod(psi)

  # Sandwich: (1/n) A^{-1} B (A^{-1})^T
  A_inv <- solve(A)
  V_beta <- (1 / n) * A_inv %*% B %*% t(A_inv)
  V_beta
}
```

::: {.callout-tip}
### In practice
You almost never code this yourself. In R, `sandwich::sandwich(model)` computes
$\hat A^{-1} \hat B (\hat A^{-1})^T$ from any `glm` object. The `sandwich`
package (Zeileis, 2004, 2006) provides variants for clustered data
(`vcovCL()`), panel data (`vcovPL()`), and heteroscedasticity-autocorrelation
(`vcovHAC()`).
:::

### 1.6 Concrete example: linear regression

Let us work through the sandwich for OLS step by step.

**Model**: $Y_i = X_i^T\beta + \varepsilon_i$ where $E[\varepsilon_i \mid X_i]
= 0$.

**Step 1: Write down $\psi$.**
The OLS estimator minimises $\sum_i (Y_i - X_i^T\beta)^2$. Setting the
gradient to zero gives $\sum_i X_i(Y_i - X_i^T\beta) = 0$, so:

$$
\psi(O_i; \beta) = X_i(Y_i - X_i^T\beta)
\qquad (p \times 1)
$$

At $\beta_0$: $E[\psi_i] = E[X_i \varepsilon_i] = 0$ by exogeneity.

**Step 2: Compute the bread $\color{#3182BD}{A}$.**

$$
\frac{\partial \psi(O_i; \beta)}{\partial \beta^T}
= \frac{\partial}{\partial \beta^T}\left[X_i(Y_i - X_i^T\beta)\right]
= -X_i X_i^T
\qquad (p \times p)
$$

So $\color{#3182BD}{A} = E[X_i X_i^T]$, estimated by
$\color{#3182BD}{\hat{A}} = \frac{1}{n} X^T X$.

**Step 3: Compute the meat $\color{#E6550D}{B}$.**
Since $\psi_i = X_i \hat e_i$, the outer product is $\psi_i\psi_i^T =
\hat e_i^2 X_iX_i^T$:

$$
\color{#E6550D}{\hat{B}} = \frac{1}{n}\sum_i \hat{e}_i^2 \, X_i X_i^T
\qquad (p \times p)
$$

The meat depends on the *squared residuals*, not on any assumed variance. This
is the source of robustness.

**Step 4: Assemble the sandwich.**

$$
\underset{p \times p}{\widehat{\text{Var}}(\hat\beta)}
= \frac{1}{n}\,
\color{#3182BD}{\hat{A}^{-1}}\;
\color{#E6550D}{\hat{B}}\;
\color{#3182BD}{(\hat{A}^{-1})^T}
$$

Substituting:

$$
= \underset{p \times p}{\color{#3182BD}{(X^TX)^{-1}}}
\;\;
\underset{p \times p}{\color{#E6550D}{\left[\sum_i
\hat{e}_i^2\,X_iX_i^T\right]}}
\;\;
\underset{p \times p}{\color{#3182BD}{(X^TX)^{-1}}}
$$

This is the *HC0* heteroscedasticity-consistent estimator (White, 1980).

**What if homoscedasticity holds?** If $\text{Var}(\varepsilon_i \mid X_i) =
\sigma^2$ for all $i$, then $E[\hat{e}_i^2 X_i X_i^T] = \sigma^2 E[X_i
X_i^T]$, so $\color{#E6550D}{B} = \sigma^2 \color{#3182BD}{A}$ and the
sandwich simplifies to $\frac{\sigma^2}{n} \color{#3182BD}{A^{-1}}
= \hat\sigma^2(X^TX)^{-1}$, the usual OLS variance.

## 2. The delta method: from model parameters to causal effects

### 2.1 The problem

The sandwich gives us $\widehat{\text{Var}}(\hat\beta)$ — the variance of the
*model coefficients*. But the causal estimand is typically a nonlinear function
of $\hat\beta$. For instance, the g-computation marginal mean is:

$$
\hat\mu_a = \frac{1}{n_t}\sum_{i \in \text{target}}
g^{-1}(X_i^{*T} \hat\beta)
$$

where $X_i^*$ is individual $i$'s covariate vector with treatment set to $a$,
$g^{-1}(\cdot)$ is the inverse link function, and $n_t$ is the size of the
target population. We need $\text{Var}(\hat\mu_a)$, not
$\text{Var}(\hat\beta)$.

### 2.2 The delta method: statement and derivation

The **delta method** propagates variance through a function. The idea: if the
input is approximately normal, the output is approximately normal with a
variance determined by the function's slope.

**Scalar version (intuition).** If $\hat\beta \sim \mathcal{N}(\beta,
\sigma^2_\beta)$ and $\hat\mu = f(\hat\beta)$, then by a first-order Taylor
expansion: $\hat\mu \approx f(\beta) + f'(\beta)(\hat\beta - \beta)$. Taking
variance: $\text{Var}(\hat\mu) \approx [f'(\beta)]^2 \sigma^2_\beta$. The
derivative $f'$ converts uncertainty in $\beta$ to uncertainty in $\mu$.

**Multivariate version.** If $\hat\beta$ is a $p$-vector with variance
$V_\beta$ (the $p \times p$ sandwich from Section 1.4) and
$\hat\mu = f(\hat\beta)$ is a $k$-vector ($k$ = number of interventions):

$$
\underset{k \times k}{\text{Var}(\hat\mu)}
\;=\;
\underset{k \times p}{\color{#31A354}{J}}
\;\;
\underset{p \times p}{V_\beta}
\;\;
\underset{p \times k}{\color{#31A354}{J^T}}
$$ {#eq-delta}

where $\color{#31A354}{J} = \frac{\partial f}{\partial \beta^T}$ is the
$k \times p$ Jacobian matrix — the multivariate analogue of $f'$.
Each row $j$ of $\color{#31A354}{J}$ tells us how the $j$-th marginal mean
$\hat\mu_j$ changes as each of the $p$ model coefficients shifts.

### 2.3 Deriving the Jacobian for the standardised mean

For $\hat\mu_a = \frac{1}{n_t}\sum_{i} g^{-1}(X_i^{*T}\beta)$, differentiate
with respect to $\beta^T$. By the chain rule, $\frac{d}{d\beta^T}
g^{-1}(X_i^{*T}\beta) = \frac{d\mu}{d\eta}\big|_{\eta_i^*} \cdot X_i^{*T}$
(scalar $\times$ $1 \times p$ row). Summing:

$$
\underset{1 \times p}{\color{#31A354}{J_a}}
= \frac{\partial \hat\mu_a}{\partial \beta^T}
= \frac{1}{n_t} \sum_{i \in \text{target}}
\underbrace{\frac{d\mu}{d\eta}\bigg|_{\eta_i^*}}_{\text{scalar}}
\;\cdot\;
\underbrace{X_i^{*T}}_{1 \times p}
$$

The link derivatives for common GLMs:

| Family | Link | $d\mu/d\eta$ | Effect |
|---|---|---|---|
| Gaussian | Identity | $1$ | All observations contribute equally |
| Binomial | Logit | $\mu(1 - \mu)$ | Observations near $\mu = 0.5$ contribute most |
| Poisson | Log | $\mu$ | Observations with larger means contribute more |

### 2.4 Putting it together: sandwich + delta method

For a **causal risk difference** $\hat\Delta = \hat\mu_1 - \hat\mu_0$:

1. Fit the outcome model → $\hat\beta$ and
   $V_\beta = \frac{1}{n}\color{#3182BD}{A^{-1}}\color{#E6550D}{B}
   \color{#3182BD}{(A^{-1})^T}$ (Section 1.4).
2. Compute $\color{#31A354}{J_1}$ and $\color{#31A354}{J_0}$ (Section 2.3).
3. Stack into $\color{#31A354}{J} = \binom{J_1}{J_0}$ ($2 \times p$) and
   compute $\text{Var}(\hat\mu_1, \hat\mu_0) = \color{#31A354}{J} V_\beta
   \color{#31A354}{J^T}$ ($2 \times 2$).
4. The variance of the difference is $\text{Var}(\hat\Delta) =
   \text{Var}(\hat\mu_1) + \text{Var}(\hat\mu_0) -
   2\text{Cov}(\hat\mu_1, \hat\mu_0)$, all read from the $2 \times 2$ matrix.

::: {.callout-note}
### Why the covariance matters
Both $\hat\mu_1$ and $\hat\mu_0$ are functions of the *same* $\hat\beta$, so
they are correlated. This covariance typically *reduces* the variance of the
difference. Ignoring it (computing variances separately and adding them) would
overestimate the SE.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Pseudo-code: delta_method_variance()"
delta_method_variance <- function(model, data_list, target_idx, V_beta) {
  beta_hat <- coef(model)
  pred_terms <- delete.response(terms(model))

  # Compute k × p Jacobian (k = number of interventions)
  J <- t(vapply(data_list, function(df) {
    df_t <- df[target_idx, ]
    X_j <- model.matrix(pred_terms, data = df_t)
    eta_j <- as.numeric(X_j %*% beta_hat)
    mu_eta_j <- model$family$mu.eta(eta_j)  # dμ/dη at intervention
    colMeans(X_j * mu_eta_j)                # (1/n_t) Σ (dμ/dη) X_i*
  }, numeric(length(beta_hat))))

  # Propagate: Var(μ̂) = J V_β J^T
  J %*% V_beta %*% t(J)
}
```

## 3. Method-specific sandwich estimators

The plug-and-play structure above ("get $V_\beta$, then propagate via
$\color{#31A354}{J} V_\beta \color{#31A354}{J^T}$") works for any parametric
model, but the choice of $V_\beta$ depends on the estimation method.

### 3.1 G-computation

The outcome model is a standard GLM fitted on the observed data. The
Huber–White sandwich from Section 1.4 is used directly:

$$
V_\beta = \frac{1}{n}\color{#3182BD}{A^{-1}} \color{#E6550D}{B}
\color{#3182BD}{(A^{-1})^T}
$$

### 3.2 Inverse probability weighting (IPW)

IPW is more subtle because the estimation pipeline involves *two* models, and
the uncertainty from the first must propagate into the second.

#### The two-model setup

1. **Propensity score model**: estimate $\hat\alpha$ ($q$-vector) from
   $\text{logit}[P(A_i = 1 \mid L_i)] = L_i^T \alpha$.
2. **Weighted outcome model (MSM)**: using the estimated weights $w_i =
   1/\hat{P}(A_i \mid L_i; \hat\alpha)$, fit $E[Y_i \mid A_i] =
   g^{-1}(\beta_0 + \beta_1 A_i)$ by weighted regression to get $\hat\beta$
   ($p$-vector).

The weights $w_i$ depend on $\hat\alpha$, so $\hat\beta$ depends on
$\hat\alpha$.

#### Why the naive approach fails

A naive sandwich on the outcome model alone — treating the weights as *known
constants* — ignores the variability in $\hat\alpha$. Each new sample produces
different $\hat\alpha$, hence different weights, hence different $\hat\beta$.
The true variability of $\hat\beta$ has two sources:

- **Outcome variability**: randomness in $Y_i$ given fixed weights.
- **Weight variability**: randomness in $\hat\alpha$ (and hence $w_i$).

The naive sandwich captures only the first. The direction of the bias depends
on the estimand: for the **ATE**, using estimated (rather than known) propensity
scores actually *reduces* the asymptotic variance, so the naive sandwich is
conservative — it **overestimates** the SE (Hirano et al., 2003; Lunceford &
Davidian, 2004). For the **ATT**, the effect can go either direction.

#### The correct approach: joint M-estimation

Treat both models together as a single M-estimation problem. Define the joint
parameter $\theta = (\alpha, \beta)$ — a $(q+p)$-vector — and the stacked
estimating function:

$$
\underset{(q+p) \times 1}{\Psi(O_i; \theta)} =
\begin{pmatrix}
\color{#3182BD}{\psi_\alpha(O_i; \alpha)} \\[4pt]
\color{#E6550D}{\psi_\beta(O_i; \alpha, \beta)}
\end{pmatrix}
$$

where $\color{#3182BD}{\psi_\alpha}$ ($q \times 1$) is the score of the
propensity model, and $\color{#E6550D}{\psi_\beta}$ ($p \times 1$) is the
*weighted* score of the outcome model (which depends on $\alpha$ through $w_i$).

The bread matrix $\hat A = -\frac{1}{n}\sum_i
\frac{\partial\Psi_i}{\partial\theta^T}$ has a block structure:

$$
\underset{(q+p) \times (q+p)}{\hat{A}} =
\begin{pmatrix}
\underset{q \times q}{\color{#3182BD}{A_{\alpha\alpha}}} &
\underset{q \times p}{0} \\[6pt]
\underset{p \times q}{\color{#E6550D}{A_{\beta\alpha}}} &
\underset{p \times p}{\color{#E6550D}{A_{\beta\beta}}}
\end{pmatrix}
$$

Each block has a concrete meaning:

- $\color{#3182BD}{A_{\alpha\alpha}}$ ($q \times q$): the Hessian of the
  propensity model — how $\psi_\alpha$ changes with $\alpha$.
- $\color{#E6550D}{A_{\beta\beta}}$ ($p \times p$): the weighted Hessian of
  the outcome model — how $\psi_\beta$ changes with $\beta$.
- $\color{#E6550D}{A_{\beta\alpha}}$ ($p \times q$): the **cross-term** — how
  the weighted outcome score changes with $\alpha$. This involves
  $\frac{\partial w_i}{\partial\alpha^T}$ (how the weights shift when the
  propensity parameters change). This is the block that captures the dependency.
- Upper-right = $0$ ($q \times p$): the propensity score does not depend on
  $\beta$.

The full sandwich $\frac{1}{n}\hat A^{-1}\hat B(\hat A^{-1})^T$ applied to
the stacked system correctly propagates weight uncertainty into
$\text{Var}(\hat\beta)$ through the $A_{\beta\alpha}$ block.

::: {.callout-tip}
### In practice
`WeightIt::glm_weightit()` fits the stacked system and its `vcov()` method
returns the M-estimation sandwich that accounts for weight uncertainty.
:::

### 3.3 Matching

After matching, the outcome model is fitted on the matched sample. Matched
individuals within the same subclass (pair or stratum) are correlated — they
were selected together. The standard sandwich (which assumes independence)
would understate the meat.

The **cluster-robust** sandwich accounts for this. The bread
$\color{#3182BD}{A}$ is unchanged, but the meat sums scores *within* each
subclass before squaring:

$$
V_\beta = \color{#3182BD}{A^{-1}}
\;\color{#E6550D}{B_{\text{CL}}}\;
\color{#3182BD}{(A^{-1})^T}
$$

where:

$$
\color{#E6550D}{B_{\text{CL}}} = \frac{1}{n}\sum_{c=1}^C
\bar\psi_c \, \bar\psi_c^T,
\qquad \bar\psi_c = \sum_{i \in \text{subclass } c} \psi_i
\qquad (p \times 1)
$$

Squaring the summed scores (rather than summing the squared scores) allows
within-cluster correlations to inflate the meat.

## 4. Influence functions

### 4.1 The idea: how much does each observation matter?

Imagine removing observation $i$ from the dataset and re-estimating
$\hat\theta$. The change $\hat\theta_{(-i)} - \hat\theta$ tells you how much
observation $i$ "influenced" the estimate. The **influence function** (IF) is
the first-order approximation to this quantity, scaled by $n$:

$$
\text{IF}_i \approx -n(\hat\theta_{(-i)} - \hat\theta)
$$

But we do not need to actually refit the model $n$ times. For M-estimators,
the IF can be computed in closed form.

### 4.2 Deriving the IF from the linearisation

In Section 1.4, Step 1, we derived the linearisation:

$$
\underset{p \times 1}{\hat\beta - \beta_0}
\;\approx\;
\frac{1}{n}\sum_{i=1}^n
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\;
\underset{p \times 1}{\psi(O_i; \beta_0)}
$$

Each summand $\color{#3182BD}{A^{-1}}\psi_i$ is the contribution of
observation $i$ to the total estimation error. This is exactly the **influence
function**:

$$
\underset{p \times 1}{\text{IF}_i}
\;=\;
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\cdot\;
\underset{p \times 1}{\psi(O_i; \hat\beta)}
$$ {#eq-if}

Reading each factor:

- $\psi_i$ ($p \times 1$): the score from Section 1.2 — how much observation
  $i$'s data "disagrees" with the current $\hat\beta$.
- $\color{#3182BD}{A^{-1}}$ ($p \times p$): the bread inverse from Section
  1.4 — converts from score-space to parameter-space, accounting for curvature.
- $\text{IF}_i$ ($p \times 1$): the resulting shift in $\hat\beta$ attributable
  to observation $i$ (times $n$).

No new derivation is needed. The IF *is* the linearisation, read term-by-term.

### 4.3 Two properties that make IFs useful

**Property 1: Linearisation.** The IF lets us write a complex estimator as a
simple average:

$$
\hat\beta - \beta_0 \approx \frac{1}{n}\sum_i \text{IF}_i
$$

**Property 2: Variance recovery.** Since $E[\text{IF}_i] = 0$:

$$
\text{Var}(\hat\beta) \approx
\frac{1}{n^2}\sum_i \text{IF}_i \, \text{IF}_i^T
$$

Expanding $\text{IF}_i \text{IF}_i^T = \color{#3182BD}{A^{-1}} \psi_i
\psi_i^T \color{#3182BD}{(A^{-1})^T}$ and factoring:

$$
\frac{1}{n^2}\sum_i \color{#3182BD}{A^{-1}} \psi_i\psi_i^T
\color{#3182BD}{(A^{-1})^T}
= \frac{1}{n} \color{#3182BD}{A^{-1}}
\underbrace{\frac{1}{n}\sum_i \psi_i\psi_i^T}_
{\color{#E6550D}{B}\text{ (the meat)}}
\color{#3182BD}{(A^{-1})^T}
$$

This is the sandwich from Section 1.4.

::: {.callout-note}
### The key equivalence
$$
\underbrace{\frac{1}{n^2}\sum_i \text{IF}_i\,\text{IF}_i^T}_
{\text{IF-based variance}}
\;=\;
\underbrace{\frac{1}{n}\,\color{#3182BD}{A^{-1}}\,\color{#E6550D}{B}\,
\color{#3182BD}{(A^{-1})^T}}_{\text{sandwich variance}}
$$ {#eq-if-sandwich}

The sandwich operates on matrices. The IF operates on per-individual vectors.
Same answer, different representation.
:::

### 4.4 Example: IF for a sample mean

$\hat\mu = \frac{1}{n}\sum_i Y_i$. Estimating equation: $\psi(O_i; \mu)
= Y_i - \mu$. Bread: $A = 1$. IF: $\text{IF}_i = Y_i - \hat\mu$. Variance:
$\frac{1}{n^2}\sum_i(Y_i - \hat\mu)^2 = s^2/n$.

### 4.5 Example: IF for g-computation (the two-channel derivation)

The g-computation marginal mean is a **two-step** estimator: first estimate
$\hat\beta$, then compute $\hat\mu_a = \frac{1}{n}\sum_i g^{-1}(X_i^{*T}
\hat\beta)$. Individual $i$ affects $\hat\mu_a$ through two channels.

#### Channel 1: Direct contribution

Even if $\hat\beta$ were fixed, observation $i$'s prediction
$g^{-1}(X_i^{*T}\hat\beta)$ is one of the $n$ terms being averaged:

$$
\text{IF}_i^{\text{direct}} = g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a
\qquad (\text{scalar})
$$

#### Channel 2: Indirect contribution (via $\hat\beta$)

Observation $i$ shifts $\hat\beta$ by approximately $\frac{1}{n}
\color{#3182BD}{A^{-1}}\psi_i$ (from Section 4.2). This shift changes
$\hat\mu_a$ by:

$$
\delta\hat\mu_a \;\approx\;
\underset{1 \times p}{\color{#31A354}{J_a}}
\;\cdot\;
\underset{p \times 1}{\frac{1}{n}\color{#3182BD}{A^{-1}}\psi_i}
$$

where $\color{#31A354}{J_a}$ ($1 \times p$) is the Jacobian from Section 2.3.

#### Combining both channels

$$
\text{IF}_i(\hat\mu_a) =
\underbrace{g^{-1}(X_i^{*T}\hat\beta) - \hat\mu_a}_
{\text{Channel 1 (scalar)}}
\;+\;
\underbrace{
\underset{1 \times p}{\color{#31A354}{J_a}}
\;\;
\underset{p \times p}{\color{#3182BD}{A^{-1}}}
\;\;
\underset{p \times 1}{\psi_i}
}_{\text{Channel 2 (scalar)}}
$$ {#eq-if-gcomp}

**Why two channels matter**: Channel 1 alone would treat $\hat\beta$ as fixed
— ignoring estimation uncertainty. Channel 2 adds the uncertainty propagated
from the model fit. This is the $\color{#31A354}{J} V_\beta
\color{#31A354}{J^T}$ term from Section 2, now expressed per-individual.

### 4.6 Why influence functions matter for multi-model estimators

For a **single-model** estimator, the IF is equivalent to
$\color{#31A354}{J} V_\beta \color{#31A354}{J^T}$ — useful but not essential.
IFs become necessary for **multi-model** estimators:

1. **Compositional**: IFs from different models can be *summed* to get the
   total IF. No large joint covariance matrix needed.
2. **Covariance for free**: $\text{Cov}(\hat\mu_a, \hat\mu_b)
   = \frac{1}{n^2}\sum_i \text{IF}_i^{(a)} \cdot \text{IF}_i^{(b)}$.
3. **Diagnostics**: large $|\text{IF}_i|$ flags influential observations.

## 5. Influence functions for chained estimators

### 5.1 The chain rule for influence functions

Section 4.5 derived the IF for a two-step estimator (fit $\hat\beta$, then
compute $\hat\mu_a$). The result had two channels: a direct term and a model
correction via $\color{#31A354}{J_a}\color{#3182BD}{A^{-1}}\psi_i$.

What happens with **multiple models in a chain**, where each model's response
is a prediction from the next? The same logic applies recursively. If
$\hat\mu = f(\hat\beta_0, \hat\beta_1, \ldots, \hat\beta_K)$ and each
$\hat\beta_k$ is an M-estimator with bread $\color{#3182BD}{A_k}$ and score
$\psi_k$, the IF of $\hat\mu$ is (Hampel et al., 1986; Huber, 1981):

$$
\text{IF}_i(\hat\mu)
\;=\;
\text{IF}_i^{\text{direct}}
\;+\;
\sum_{k=0}^{K}
\frac{\partial \hat\mu}{\partial \hat\beta_k^T}
\;\cdot\;
\color{#3182BD}{A_k^{-1}} \; \psi_k(O_i; \hat\beta_k)
$$ {#eq-if-chain}

Each term in the sum is a "Channel 2" from Section 4.5 applied to model $k$:
the Jacobian $\partial\hat\mu/\partial\hat\beta_k^T$ (how $\hat\mu$ depends on
model $k$) times the IF of model $k$'s parameters
($\color{#3182BD}{A_k^{-1}}\psi_{k,i}$).

::: {.callout-note}
### Connections to other frameworks
This IF-based approach is equivalent to:

1. **Stacked estimating equations** (Stefanski & Boos, 2002; Zivich et al.,
   2024): concatenate all models' estimating equations and apply the generic
   sandwich to the full stack.
2. **Semiparametric efficiency theory** (Tsiatis, 2006): derive the efficient
   influence function (EIF) and project onto the parametric model.

All yield the same variance estimate.
:::

### 5.2 ICE g-computation: the motivating example

Iterated conditional expectation (ICE) g-computation estimates the
counterfactual mean under a treatment strategy in longitudinal data (Robins,
1986; Bang & Robins, 2005; Zivich et al., 2024). It fits $K+1$ outcome models
backward in time:

```{mermaid}
%%| label: fig-ice-chain
%%| fig-cap: "ICE backward iteration. Models are fit from final time to baseline. Each model's predictions become the next model's response."
flowchart RL
  MK["Model K (final time)<br/>Fit: E[Y | A_K, H_K]<br/>Predict → Ỹ_K"] -->|"Ỹ_K becomes<br/>response for"| MK1["Model K-1<br/>Fit: E[Ỹ_K | A_{K-1}, H_{K-1}]<br/>Predict → Ỹ_{K-1}"]
  MK1 -->|"Ỹ_{K-1} becomes<br/>response for"| Mdots["⋯"]
  Mdots -->|"response for"| M0["Model 0 (baseline)<br/>Fit: E[Ỹ_1 | A_0, H_0]<br/>Predict → Ỹ_0"]
  M0 -->|"average"| MU["μ̂ = mean(Ỹ_0)"]

  style MK fill:#f0f4ff,stroke:#3182BD
  style MK1 fill:#f0f4ff,stroke:#3182BD
  style M0 fill:#f0f4ff,stroke:#3182BD
  style MU fill:#e8f5e9,stroke:#31A354
```

Each model's response ($\tilde{Y}_k$) is a *prediction* from the next model in
the chain. This creates a cascade of dependencies.

### 5.3 Why naive approaches fail

**Naive 1: Variance of pseudo-outcomes only.** Uses only Channel 1 — ignores
all model correction terms.

**Naive 2: $\color{#31A354}{J}V_\beta\color{#31A354}{J^T}$ from model 0
only.** Captures Channel 2 for model 0 but ignores models $1, \ldots, K$.

**Correct: Full IF chain.** Account for all $K+1$ models via the chain rule.

### 5.4 Deriving the IF for chained models

#### The IF decomposition

Applying the chain rule (Section 5.1):

$$
\text{IF}_i
= \underbrace{\frac{1}{n_t}(\tilde{Y}_{0,i} - \hat\mu)}_{\text{Channel 1}}
+ \sum_{k=0}^{K}
\underbrace{
\color{#E6550D}{d_{k,i}} \cdot r_{k,i}
}_{\text{Channel 2 for model } k}
$$ {#eq-if-ice}

where $r_{k,i} = Y_{k,i} - \hat{Y}_{k,i}$ is individual $i$'s residual at
model $k$, and $\color{#E6550D}{d_{k,i}}$ is the **sensitivity** — how much a
perturbation at model $k$ propagates to $\hat\mu$.

#### What the sensitivities measure

Each model correction requires a **sensitivity** $\color{#E6550D}{d_{k,i}}$: a
scalar that answers "if individual $i$'s residual at model $k$ increased by one
unit, how much would $\hat\mu$ change?" Computing $\color{#E6550D}{d_{k,i}}$
takes three steps, each with a clear role:

1. **Gradient** $g_k$ ($p_k \times 1$): the total sensitivity of $\hat\mu$ to
   model $k$'s coefficient vector $\beta_k$. This is a Jacobian — the same
   object as $\color{#31A354}{J_a}$ from Section 2.3, but computed recursively
   for $k > 0$.
2. **Solve** $h_k = (X_k^T W_k X_k)^{-1} g_k$ ($p_k \times 1$): convert from
   "how $\hat\mu$ responds to $\beta_k$" to "how $\hat\mu$ responds to
   individual score contributions." This is the bread inverse
   $\color{#3182BD}{A_k^{-1}}$ applied to $g_k$.
3. **Distribute** $\color{#E6550D}{d_{k,i}} = h_k^T X_{k,i}$ (scalar): project
   onto individual $i$'s covariate vector to get their personal sensitivity.

#### Base case: model 0

$\hat\mu$ is a direct average of model 0's predictions. The gradient of
$\hat\mu$ w.r.t. $\beta_0$ is therefore the Jacobian from Section 2.3:

$$
\underset{p_0 \times 1}{g_0}
= \frac{1}{n_t}\sum_{j \in \text{target}}
\frac{d\mu}{d\eta}\bigg|_{\eta_{0,j}^*}
\;\cdot\; X_{0,j}^*
$$

$g_0$ tells us the total sensitivity of $\hat\mu$ to $\beta_0$, but the IF
requires a per-individual quantity. We need to ask: "of the total shift in
$\hat\beta_0$ caused by the data, how much came from individual $i$?" The
answer uses two facts: (a) individual $i$'s influence on $\hat\beta_0$ is
proportional to their score $\psi_{0,i}$, which is proportional to $r_{0,i}
\cdot X_{0,i}$; and (b) the constant of proportionality involves the bread
inverse $(X_0^T W_0 X_0)^{-1}$. Combining:

$$
h_0 = \underbrace{(X_0^T W_0 X_0)^{-1}}_{\text{bread inverse}} g_0,
\qquad
\color{#E6550D}{d_{0,i}} = h_0^T X_{0,i}
$$

Now $\color{#E6550D}{d_{0,i}} \cdot r_{0,i}$ gives individual $i$'s Channel 2
contribution: their residual $r_{0,i}$ (how much their data "disagrees" with
model 0) scaled by $\color{#E6550D}{d_{0,i}}$ (how much that disagreement
matters for $\hat\mu$). This is equivalent to
$\color{#31A354}{J_0}\color{#3182BD}{A_0^{-1}}\psi_{0,i}$ from Section 4.5,
factored into the three steps above.

#### Recursive case: model $k = 1, \ldots, K$

Model $k$ does not feed into $\hat\mu$ directly. Instead, model $k$'s
predictions become model $k-1$'s pseudo-outcomes (response variable). We already
know how sensitive $\hat\mu$ is to each of those pseudo-outcomes — that is
$\color{#E6550D}{d_{k-1,j}}$ from the previous step.

The gradient of $\hat\mu$ w.r.t. $\beta_k$ therefore weights each individual's
contribution by the previously-computed sensitivity:

$$
\underset{p_k \times 1}{g_k}
= \sum_{j \in \text{fit}_{k-1}}
\color{#E6550D}{d_{k-1,j}}
\;\cdot\;
\frac{d\mu}{d\eta}\bigg|_{\eta_{k,j}^*}
\;\cdot\; X_{k,j}^*
$$

Compare this to $g_0$: the uniform weight $\frac{1}{n_t}$ over the target
population is replaced by $\color{#E6550D}{d_{k-1,j}}$ over the previous
model's fitting set. Everything else — solve and distribute — is identical:

$$
h_k = (X_k^T W_k X_k)^{-1} g_k,
\qquad
\color{#E6550D}{d_{k,i}} = h_k^T X_{k,i}
$$

### 5.5 Scaling: the factor of $n$

Each model correction involves $\color{#3182BD}{A_k^{-1}}$. Since
$\color{#3182BD}{A_k} = \frac{1}{n_k} X_k^T W_k X_k$, we have
$\color{#3182BD}{A_k^{-1}} = n_k(X_k^T W_k X_k)^{-1}$. So the correction at
step $k$ is:

$$
n_k \cdot \color{#E6550D}{d_{k,i}} \cdot r_{k,i}
$$

The factor $n_k$ ensures each IF contribution is $O(1)$ per individual, giving
the correct $\text{Var}(\hat\mu) = O(1/n)$.

::: {.callout-warning}
### A common bug
Omitting $n_k$ makes each correction $O(1/n)$, yielding $\text{Var} = O(1/n^3)$
— confidence intervals that are much too narrow.
:::

### 5.6 Pseudo-code: IF for a chained estimator

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Pseudo-code: chained_influence()"
chained_influence <- function(models, data_iv, target) {
  n <- number_of_individuals
  K <- length(models) - 1
  mu_hat <- mean(pseudo_final[target])
  n_target <- sum(target)

  IF <- rep(0, n)
  IF[target] <- (pseudo_final[target] - mu_hat) / n_target  # Channel 1

  d <- rep(0, n)  # sensitivities from previous step

  for (step in 0:K) {
    model_k <- models[[step + 1]]
    X_k <- model.matrix(model_k)         # fitting design matrix
    r_k <- residuals(model_k)            # r_{k,i}
    W_k <- working_weights(model_k)
    fit_ids_k <- model_k$fitting_ids
    n_k <- length(fit_ids_k)

    # Intervention design matrix and link derivatives (dμ/dη)
    X_star_k <- model.matrix(model_k, newdata = intervention_data[[step + 1]])
    eta_star <- X_star_k %*% coef(model_k)
    mu_eta_star <- model_k$family$mu.eta(eta_star)

    # Step 1 — Gradient: g_k = ∂μ̂/∂β_k
    if (step == 0) {
      g_k <- crossprod(X_star_k[target, ], mu_eta_star[target]) / n_target
    } else {
      prev_ids <- models[[step]]$fitting_ids
      g_k <- crossprod(X_star_k[prev_ids, ], d[prev_ids] * mu_eta_star[prev_ids])
    }

    # Step 2 — Solve: h_k = (X'WX)^{-1} g_k
    h_k <- solve(crossprod(X_k, X_k * W_k)) %*% g_k

    # Step 3 — Distribute: d_{k,i} = h_k' X_{k,i}
    d <- rep(0, n)
    d[fit_ids_k] <- X_k %*% h_k

    # Channel 2 for model k: n_k * d_{k,i} * r_{k,i} (Section 5.5 for n_k)
    r_vec <- rep(0, n)
    r_vec[fit_ids_k] <- r_k
    IF <- IF + n_k * d * r_vec
  }
  IF
}
```

### 5.7 From individual IFs to the variance

Once we have the per-individual $\text{IF}_i$ from the recursion above, variance
estimation follows from the same identity as Section 4.3 (Property 2):

$$
\widehat{\text{Var}}(\hat\mu)
= \frac{1}{n^2}\sum_{i=1}^n \text{IF}_i^2
$$

For two interventions $a$ and $b$ on the same data:

$$
\widehat{\text{Cov}}(\hat\mu_a, \hat\mu_b)
= \frac{1}{n^2}\sum_{i=1}^n \text{IF}_i^{(a)} \cdot \text{IF}_i^{(b)}
$$

The full $k \times k$ covariance matrix is then used for contrasts as in
Section 2.

## 6. Summary

1. **M-estimation**: any root-of-sum estimator has a sandwich variance.
2. **The sandwich** $\color{#3182BD}{A^{-1}}\color{#E6550D}{B}
   \color{#3182BD}{(A^{-1})^T}$: robust to variance misspecification.
3. **Delta method**: propagates parameter variance to functionals via
   $\color{#31A354}{J}V_\beta\color{#31A354}{J^T}$.
4. **Influence functions**: per-individual decomposition; $\frac{1}{n^2}\sum
   \text{IF}_i^2$ = sandwich variance.
5. **Chained estimators**: the IF chain rule propagates uncertainty through
   multi-model pipelines. Ignoring upstream models underestimates variance.
6. **Equivalent frameworks**: direct IF, stacked estimating equations
   (Stefanski & Boos, 2002; Zivich et al., 2024), semiparametric EIF
   (Tsiatis, 2006).

## References

- Bang H, Robins JM (2005). Doubly robust estimation in missing data and
  causal inference models. *Biometrics* 61:962–973.
- Hampel FR, Ronchetti EM, Rousseeuw PJ, Stahel WA (1986). *Robust
  Statistics: The Approach Based on Influence Functions*. Wiley.
- Hirano K, Imbens GW, Ridder G (2003). Efficient estimation of average
  treatment effects using the estimated propensity score.
  *Econometrica* 71:1161–1189.
- Hernán MA, Robins JM (2025). *Causal Inference: What If*.
  Chapman & Hall/CRC.
- Huber PJ (1967). The behavior of maximum likelihood estimates under
  nonstandard conditions. *Proceedings of the Fifth Berkeley Symposium*
  1:221–233.
- Huber PJ (1981). *Robust Statistics*. Wiley.
- Lunceford JK, Davidian M (2004). Stratification and weighting via the
  propensity score in estimation of causal treatment effects: a comparative
  study. *Statistics in Medicine* 23:2937–2960.
- Robins JM (1986). A new approach to causal inference in mortality studies
  with a sustained exposure period. *Mathematical Modelling* 7:1393–1512.
- Stefanski LA, Boos DD (2002). The calculus of M-estimation.
  *The American Statistician* 56:29–38.
- Tsiatis AA (2006). *Semiparametric Theory and Missing Data*. Springer.
- van der Laan MJ, Robins JM (2003). *Unified Methods for Censored
  Longitudinal Data and Causality*. Springer.
- White H (1980). A heteroskedasticity-consistent covariance matrix estimator
  and a direct test for heteroskedasticity. *Econometrica* 48:817–838.
- White H (1982). Maximum likelihood estimation of misspecified models.
  *Econometrica* 50:1–25.
- Zeileis A (2004). Econometric computing with HC and HAC covariance matrix
  estimators. *Journal of Statistical Software* 11(10):1–17.
- Zeileis A (2006). Object-oriented computation of sandwich estimators.
  *Journal of Statistical Software* 16(9):1–16.
- Zivich PN, Klose M, Cole SR, Edwards JK, Shook-Sa BE (2022).
  Delicatessen: a Python package for the generalized computation of the
  sandwich estimator. *arXiv preprint* arXiv:2203.11300.
- Zivich PN, Ross RK, Shook-Sa BE, Cole SR, Edwards JK (2024). Empirical
  sandwich variance estimator for iterated conditional expectation
  g-computation. *Statistics in Medicine* 43:5562–5572.
